---
title: "Fitting the log1p model at the wrong c"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r load-pkgs}
library(ggplot2)
library(cowplot)
library(fastTopics)
library(log1pNMF)
theme_set(theme_cowplot(font_size = 11))
```

```{r paths}
## Where the simulation code and its outputs live. The generative functions are
## sourced rather than restated, so the description below cannot drift from what
## actually produced the data.
repo_dir  <- "~/Documents/log1pNMF/inst/paper_figures/experiments"
exp_dir   <- path.expand(repo_dir)
cells_dir <- file.path(exp_dir, "c_grid_sim")            # 640 per-cell .rds
metrics   <- file.path(exp_dir, "c_grid_sim_metrics.rds")

stopifnot(dir.exists(exp_dir), file.exists(metrics), dir.exists(cells_dir))
source(file.path(exp_dir, "c_grid_sim_common.R"))
```

# The question

The log1p model has a link parameter $c$:

$$
y_{ij} \sim \text{Poisson}(\lambda_{ij}), \qquad
\log\!\left(1 + \frac{\lambda_{ij}}{c}\right) = \sum_{k=1}^{K}\ell_{ik}f_{jk} =: \theta_{ij}
$$

so $\lambda_{ij} = c\,(e^{\theta_{ij}} - 1)$. As $c \to \infty$ the link becomes
linear and the model is ordinary Poisson NMF (the topic model); as $c \to 0$ it
becomes additive on the log scale. $c$ therefore decides whether the factors
describe **absolute** or **relative** changes.

This page asks: **if the data really were generated at some $c_{\text{true}}$,
what happens when we fit at a different $c_{\text{fit}}$?** We simulate at eight
values of $c$ and fit at all eight, over ten seeds — a full $8 \times 8$ cross.

# How the data were generated

## Generative model

For each seed we draw one truth and reuse it at every $c_{\text{true}}$:

- **$L$** ($n \times K$): rows drawn from a $\text{Dirichlet}(a\mathbf{1}_K)$ with
  $a = `r A_SIG`$. Because $a < 1$ the rows are peaked toward the pure states, so
  most samples are dominated by a single factor. Rows sum to one, which matters
  for the offset trick below.
- **$F_0$** ($p \times K$): each entry is exactly zero with probability
  $`r PI0`$, and otherwise **log-uniform** over a range of $`r format(F_RANGE, scientific = FALSE)`$.

Then $\theta_0 = LF_0'$, and the realized truth at a given $c$ is

$$
\theta = \alpha_c\,\theta_0 + \beta_c, \qquad
\lambda = c\,\text{expm1}(\theta) \quad (\text{or } \lambda = \theta \text{ when } c = \infty),
\qquad y_{ij}\sim\text{Poisson}(\lambda_{ij}).
$$

Since $\sum_k \ell_{ik} = 1$, adding $\beta_c$ to *every* entry of $F$ shifts
$\theta$ by exactly that constant. So $\theta = \alpha_c\theta_0 + \beta_c$ is
still $L F_c'$ with $F_c = \alpha_c F_0 + \beta_c$ — the offset costs no rank.
Note $F_c$ is therefore **not** sparse: it has a floor of $\beta_c$ under every
entry. The sparse object is $F_0$.

```{r show-truth}
sim_truth
```

## Why the magnitudes are log-uniform and bounded

At small $c$ the link is exponential, so any right tail in $\theta_0$ gets
exponentiated. An earlier version drew lognormal magnitudes and the
$c = 10^{-3}$ datasets came out with $\max(Y) \approx 10^{16}$ while the
$c = \infty$ ones had $\max(Y) \approx 90$. Bounding the magnitudes is what makes
the datasets comparable across the grid at all.

## Calibrating each c to a common target

A single scale on $F$ cannot make these datasets comparable: at small $c$ a fixed
spread in $\theta$ becomes a *multiplicative* spread in $\lambda$ and the maximum
explodes, while at large $c$ the link is linear and the spread stays
proportional. So there are **two** knobs per $c$, solved numerically so that
every dataset hits the same two targets:

- the fraction of zeros in $Y$ is $`r TARGET_ZERO`$, and
- the $`r QPROB`$ quantile of $\lambda$ is $`r TARGET_Q`$.

The solve is two nested `uniroot` calls and is monotone in each direction: the
zero-fraction decreases in $\beta$ at fixed $\alpha$, and the upper quantile
increases in $\alpha$ once $\beta$ has been re-solved.

```{r show-calibrate}
calibrate
```

## Do the datasets actually come out comparable?

```{r data-summary, cache=TRUE}
summ <- do.call(rbind, lapply(1:N_SEEDS, function(sd) {
  tr <- sim_truth(sd)
  do.call(rbind, lapply(C_GRID, function(cc) data_summary(sim_dataset(sd, cc, tr))))
}))
knitr::kable(
  data.frame(
    quantity = c("zero fraction of Y", "max(Y)", "mean(Y)",
                 "q99.9 of lambda", "gene mean q90/q10"),
    min = c(min(summ$zero_frac), min(summ$max_Y), min(summ$mean_Y),
            min(summ$q999_lam), min(summ$gene_mean_q90 / summ$gene_mean_q10)),
    max = c(max(summ$zero_frac), max(summ$max_Y), max(summ$mean_Y),
            max(summ$q999_lam), max(summ$gene_mean_q90 / summ$gene_mean_q10))),
  digits = 3, caption = "Across all 80 datasets (10 seeds x 8 values of c).")
```

The two calibrated targets are hit exactly, and `max(Y)` and `mean(Y)` stay in a
narrow band. The one quantity that is *not* constant is the gene-level dynamic
range, and that is unavoidable: matching $\lambda$'s distribution means matching
$\theta$ up to a **shift** at small $c$ but up to a **scale** at large $c$, and
no single reparameterization does both.

## Fitting

`r N_SEEDS` seeds $\times$ `r length(C_GRID)` values of $c_{\text{true}}$
$\times$ `r length(C_GRID)` values of $c_{\text{fit}}$ = `r N_TASKS` cells, each
fit from **both** a rank-1 and a random start (`r N_TASKS * length(INITS)` fits).
Every fit ran to convergence — an absolute log-likelihood change below
`r TOL` between iterations — with `r format(MAXITER, big.mark = ",")` iterations
as the only other cap and no time limit. $n = p = `r N_ROWS`$, $K = `r K_TRUE`$
throughout, and $c_{\text{fit}} = \infty$ is fit with `fastTopics::fit_poisson_nmf`
since no finite `cc` represents that limit.

```{r load-metrics}
d <- readRDS(metrics)

## Best init per cell. The two agree to within 0.008 log-likelihood on all 640
## cells, so for the LIKELIHOOD figures this choice is immaterial. It is not
## immaterial for sparsity -- see that section.
key <- paste(d$seed, d$c_true, d$c_fit)
b   <- d[order(key, -d$loglik), ]
b   <- b[!duplicated(paste(b$seed, b$c_true, b$c_fit)), ]
b$subopt <- ave(b$loglik, paste(b$seed, b$c_true), FUN = max) - b$loglik

CLEV <- sort(unique(b$c_fit))
CLAB <- ifelse(is.infinite(CLEV), "∞",
               format(CLEV, scientific = FALSE, trim = TRUE, drop0trailing = TRUE))
fc <- function(x) factor(x, levels = CLEV, labels = CLAB)
b$cf     <- fc(b$c_fit)
b$cf_lab <- fc(b$c_true)
b$ct     <- factor(paste("simulated at c =", fc(b$c_true)),
                   levels = paste("simulated at c =", CLAB))
c(cells = nrow(b), fits = nrow(d), converged = sum(d$converged))
```

# Which c fits best?

Comparing log-likelihoods across $c_{\text{fit}}$ is legitimate: all eight are
Poisson models for the *same* $y$ under the same dominating measure with
constants included, so no Jacobian enters.

```{r winner-matrix, fig.width=7, fig.height=6}
win <- do.call(rbind, lapply(split(b, paste(b$seed, b$c_true)), function(x)
  x[which.max(x$loglik), c("seed", "c_true", "c_fit")]))
cnt <- as.data.frame(table(ct = fc(win$c_true), cf = fc(win$c_fit)), responseName = "n")
cnt$lab <- ifelse(cnt$n == 0, "", cnt$n)

ggplot(cnt, aes(cf, ct)) +
  geom_tile(aes(fill = n), colour = "white", linewidth = 0.6) +
  geom_tile(data = data.frame(ct = CLAB, cf = CLAB), fill = NA,
            colour = "#C2410C", linewidth = 0.9) +
  geom_text(aes(label = lab, colour = n > 5), size = 4, show.legend = FALSE) +
  scale_fill_gradient(low = "#F2F6F7", high = "#0E5C6B", name = "seeds\nwon",
                      limits = c(0, 10), breaks = c(0, 5, 10)) +
  scale_colour_manual(values = c("FALSE" = "grey15", "TRUE" = "white")) +
  scale_x_discrete(expand = c(0, 0)) + scale_y_discrete(expand = c(0, 0)) +
  labs(x = "c used for fitting", y = "c used to simulate",
       title = "Which c achieves the highest log-likelihood",
       subtitle = "counts over 10 seeds; outlined cells are c_fit = c_true") +
  coord_fixed() +
  theme(plot.title = element_text(size = 13, face = "plain"),
        plot.subtitle = element_text(size = 10, colour = "grey35"),
        panel.border = element_rect(colour = "grey70", fill = NA))
```

The likelihood recovers $c$ **in the middle of the range and nowhere else**. At
$c_{\text{true}} = 1$ and $10$ the correct value wins all ten seeds. But the
winners scatter across $\{10^{-3}, 10^{-2}, 10^{-1}\}$ at the small end and
across $\{100, 1000, \infty\}$ at the large end.

This is exactly what the theory predicts. The score for $c$ points in the
direction $\Phi_{ij} = \lambda_{ij}/(\lambda_{ij}+c)$, which collapses to
$\theta$ when $\lambda \ll c$ and to a constant when $\lambda \gg c$ — and both
of those are absorbed by the factorization itself. So the profile likelihood is
flat at both ends and only the elbow carries information about $c$.

```{r panel-helper}
## Shared layout for the by-c_true panels: all 10 seeds as thin lines plus their
## mean, with the matching c_fit circled. Keeping the seeds visible shows whether
## a flat region is flat for every seed or only on average.
panels <- function(dat, yvar, ylab, title, subtitle, ref = NULL, ...) {
  dat$.y <- dat[[yvar]]
  avg <- aggregate(.y ~ ct + cf, data = dat, FUN = mean)
  dg  <- merge(unique(dat[as.character(dat$cf) == as.character(dat$cf_lab),
                          c("ct", "cf")]), avg, by = c("ct", "cf"))
  p <- ggplot(dat, aes(cf, .y)) +
    geom_line(aes(group = seed), colour = "#9BB7BD", linewidth = 0.35, alpha = 0.8)
  if (!is.null(ref)) {
    rf <- aggregate(stats::as.formula(paste(ref, "~ ct")), data = dat, FUN = mean)
    names(rf)[2] <- ".ref"
    p <- p + geom_hline(data = rf, aes(yintercept = .ref), linetype = "dashed",
                        colour = "#C2410C", linewidth = 0.5)
  }
  p + geom_line(data = avg, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
    geom_point(data = avg, colour = "#0E5C6B", size = 1.6) +
    geom_point(data = dg, shape = 21, size = 3.6, stroke = 1.1, fill = NA,
               colour = "#C2410C") +
    facet_wrap(~ ct, nrow = 2) +
    scale_y_continuous(...) +
    labs(x = "c used for fitting", y = ylab, title = title, subtitle = subtitle) +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 10),
          plot.title = element_text(size = 13, face = "plain"),
          plot.subtitle = element_text(size = 10, colour = "grey35"),
          panel.border = element_rect(colour = "grey80", fill = NA),
          axis.text.x = element_text(size = 8))
}
SUB     <- "thin lines: 10 individual seeds;  thick line: mean;  circled: c_fit = c_true"
SUB_REF <- paste(SUB, "\ndashed: the true value for that simulation")
```

```{r loglik-panels, fig.width=12, fig.height=6.4}
panels(b, "subopt",
       ylab  = "log-likelihood shortfall from that seed's best c  (0 = best)",
       title = "Cost of fitting at the wrong c", subtitle = SUB,
       trans = "log1p", breaks = c(0, 1, 10, 100, 1000),
       labels = c("0", "1", "10", "100", "1000"))
```

$y$ is measured **down from each seed's own best** $c$, because raw
log-likelihoods cannot be overlaid across seeds — each seed is a different
dataset, so the curves would be offset by dataset-level constants of a few
hundred and the shape would be unreadable. It is on a `log1p` scale because the
shortfall is exactly $0$ at the winner, which a log scale could not show.

Panels share a $y$ scale deliberately: the penalty for being in the wrong
**regime** ($\sim$1000) dwarfs the penalty within a regime ($\sim$1), and a free
$y$ would rescale each panel and hide that.

# How close are the fitted rates, across c?

Before looking at the factors, check the quantity the likelihood actually pins
down: the fitted rates $\hat\lambda$. Take **one** simulated dataset and fit it
at every $c$ on the grid. Each panel below plots all 90,000 entries of that one
matrix, true rate against fitted rate, on log axes with $y=x$ marked. The dataset
is identical in every panel; only the $c$ used for fitting changes.

```{r rates-vs-truth, fig.width=13, fig.height=6.4, cache=TRUE, warning=FALSE}
SEED <- 1; CTRUE <- 1; YMIN <- 1e-2      # edit SEED / CTRUE to move around the grid
fmtc <- function(z) if (is.infinite(z)) "Inf" else
  format(z, scientific = FALSE, trim = TRUE, drop0trailing = TRUE)
panel_labs <- vapply(C_GRID, function(z) sprintf("fit at c = %s", fmtc(z)), character(1))

dd <- sim_dataset(SEED, CTRUE)
lam_true <- as.vector(dd$lambda)

fitted_rates <- function(cf) {
  x <- readRDS(file.path(cells_dir, paste0(
         tag_of(list(seed = SEED, c_true = CTRUE, c_fit = cf)), ".rds")))
  L <- x$fits$rank1$LL; Fm <- x$fits$rank1$FF
  as.vector(if (is.infinite(cf)) tcrossprod(L, Fm)
            else cf * expm1(tcrossprod(L, Fm) / max(1, cf)))
}

rate_dat <- do.call(rbind, lapply(C_GRID, function(cf)
  data.frame(true = lam_true, fitted = fitted_rates(cf),
             panel = sprintf("fit at c = %s", fmtc(cf)))))
rate_dat$panel <- factor(rate_dat$panel, levels = panel_labs)

## a handful of entries fall below the axis at large c; label rather than hide
clip_lab <- do.call(rbind, lapply(C_GRID, function(cf) {
  nb <- sum(fitted_rates(cf) < YMIN)
  data.frame(panel = factor(sprintf("fit at c = %s", fmtc(cf)), levels = panel_labs),
             lab = if (nb > 0) sprintf("%d entries below axis", nb) else "") }))

ggplot(rate_dat, aes(true, pmax(fitted, YMIN/2))) +
  geom_abline(slope = 1, intercept = 0, colour = "#C2410C", linewidth = 0.6) +
  geom_bin2d(bins = 70) +
  geom_text(data = clip_lab, aes(x = 0.35, y = YMIN*1.6, label = lab),
            inherit.aes = FALSE, hjust = 0, size = 2.9, colour = "grey30") +
  scale_fill_gradient(low = "#DDE8EA", high = "#0E5C6B", trans = "log10",
                      name = "entries") +
  facet_wrap(~ panel, nrow = 2) +
  scale_x_log10() + scale_y_log10(limits = c(YMIN, 40)) +
  labs(x = "true rate", y = "fitted rate",
       title = sprintf("One simulation (seed %d, generated at c = %s), fitted at every c",
                       SEED, fmtc(CTRUE)),
       subtitle = "all 90,000 entries per panel.  orange line is y = x") +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10),
        panel.border = element_rect(colour = "grey80", fill = NA))
```

```{r rates-summary}
knitr::kable(do.call(rbind, lapply(C_GRID, function(cf) {
  lf <- fitted_rates(cf)
  data.frame(c_fit = fmtc(cf),
             `fitted min` = min(lf), `fitted max` = max(lf),
             `cor(log)` = cor(log(lam_true), log(pmax(lf, 1e-12))),
             `RMSE(log1p)` = sqrt(mean((log1p(lam_true) - log1p(pmax(lf, 0)))^2)),
             check.names = FALSE)
})), digits = 4, row.names = FALSE,
caption = sprintf("Same dataset throughout (seed %d, generated at c = %s).", SEED, fmtc(CTRUE)))
```

**The rates barely move.** Across four orders of magnitude in the fitted $c$ —
including badly misspecified values — every panel puts its mass on $y=x$, and
the RMSE on the log1p scale varies only between about 0.062 and 0.078. Fitting at
$c=10^{-3}$ data that were generated at $c=1$ costs almost nothing in how well
the mean structure is reproduced.

That is worth holding onto, because the sections that follow show the *factors*
changing a great deal over the same range. The likelihood determines $LF'$, and
$\hat\lambda$ with it, but says much less about how that product is split into
$L$ and $F$. Two fits can agree on the rates to a few percent and disagree
completely about what the factors look like.

**The one visible difference is at the bottom of the range.** For
$c_{\text{fit}}\le 1$ the fitted rates stop cleanly at about $0.08$; for
$c_{\text{fit}}\ge 10$ a few dozen entries out of 90,000 are pushed to
essentially zero, down to $10^{-11}$. With a nearly linear link the model can
place structural near-zeros, and at small $c$ it cannot. That is the same floor
behind the sparsity results below, visible here directly in the rates rather
than through a sparsity measure.

# Sparsity and factor separation

**These metrics must fix the initialization**, unlike the likelihood ones. The
two inits agree to within 0.008 log-likelihood everywhere, so selecting on
log-likelihood is effectively a coin flip — but the resulting fits are not
equivalent: 29% of cells differ by more than 0.01 in Hoyer sparsity, and at
$c_{\text{fit}} = 10^{-3}$ the rank-1 start gives about twice the factor sparsity
of the random start. Random also wins the tie in all 80 cells at that
$c_{\text{fit}}$, so "best of both" would silently report the random answer at the
small-$c$ end and a mixture elsewhere.

```{r init-check}
w <- reshape(d[, c("seed","c_true","c_fit","init","hoyer_L","hoyer_F","rankcor_F")],
             idvar = c("seed","c_true","c_fit"), timevar = "init", direction = "wide")
knitr::kable(data.frame(
  metric = c("hoyer_L", "hoyer_F", "rankcor_F"),
  `cells differing by > 0.01` = sapply(c("hoyer_L","hoyer_F","rankcor_F"), function(m)
    sum(abs(w[[paste0(m, ".rank1")]] - w[[paste0(m, ".random")]]) > 0.01)),
  check.names = FALSE, row.names = NULL),
  caption = "Out of 640 cells, despite log-likelihoods agreeing to 0.008.")
```

```{r split-by-init}
mk <- function(ini) {
  x <- d[d$init == ini, ]
  x$cf     <- fc(x$c_fit)
  x$cf_lab <- fc(x$c_true)
  x$ct     <- factor(paste("simulated at c =", fc(x$c_true)),
                     levels = paste("simulated at c =", CLAB))
  x
}
b_rank1  <- mk("rank1")
b_random <- mk("random")
```

Hoyer sparsity of a length-$n$ column is
$(\sqrt{n} - \|x\|_1/\|x\|_2)/(\sqrt{n}-1)$: $0$ when every entry is equal, $1$
when a single entry carries everything. We average it over the $K$ columns.

## Sparsity of the loadings

```{r hoyer-L-rank1, fig.width=12, fig.height=6.4}
panels(b_rank1, "hoyer_L", ref = "hoyer_L_true",
       ylab = "mean Hoyer sparsity of loadings",
       title = "Sparsity of the recovered loadings — rank-1 initialization",
       subtitle = SUB_REF)
```

```{r hoyer-L-random, fig.width=12, fig.height=6.4}
panels(b_random, "hoyer_L", ref = "hoyer_L_true",
       ylab = "mean Hoyer sparsity of loadings",
       title = "Sparsity of the recovered loadings — random initialization",
       subtitle = SUB_REF)
```

The dashed reference is identical in every panel, as it must be: $L$ is drawn
once per seed and shared across all $c_{\text{true}}$.

## Sparsity of the factors

```{r hoyer-F-rank1, fig.width=12, fig.height=6.4}
panels(b_rank1, "hoyer_F", ref = "hoyer_F_true",
       ylab = "mean Hoyer sparsity of factors",
       title = "Sparsity of the recovered factors — rank-1 initialization",
       subtitle = SUB_REF)
```

```{r hoyer-F-random, fig.width=12, fig.height=6.4}
panels(b_random, "hoyer_F", ref = "hoyer_F_true",
       ylab = "mean Hoyer sparsity of factors",
       title = "Sparsity of the recovered factors — random initialization",
       subtitle = SUB_REF)
```

The reference here is the sparsity of the **realized** truth
$F_c = \alpha_c F_0 + \beta_c$, which is what the fit estimates. It rises with
$c_{\text{true}}$ and sits near zero at the small-$c$ end, because $\beta_c$ puts
a floor under every entry even though $F_0$ is 50% exact zeros. (`hoyer_F0_true`
in the metrics table holds the sparsity of $F_0$ itself if that is the comparison
you want.)

Across both sparsity figures the striking feature is that **the curves barely
depend on which panel you are in**: sparsity is close to a pure function of
$c_{\text{fit}}$, largely regardless of what generated the data. Fitting at a
larger $c$ gives a sparser answer whether or not that is true of the truth.

## Are the recovered factors distinct?

```{r rankcor-rank1, fig.width=12, fig.height=6.4}
panels(b_rank1, "rankcor_F", ref = "rankcor_F_true",
       ylab = "mean rank correlation between factors",
       title = "Rank correlation between recovered factors — rank-1 initialization",
       subtitle = SUB_REF)
```

```{r rankcor-random, fig.width=12, fig.height=6.4}
panels(b_random, "rankcor_F", ref = "rankcor_F_true",
       ylab = "mean rank correlation between factors",
       title = "Rank correlation between recovered factors — random initialization",
       subtitle = SUB_REF)
```

Mean Spearman correlation over the $K(K-1)/2$ distinct factor pairs. The dashed
reference is again identical across panels — Spearman correlation is invariant to
the positive affine map $F_c = \alpha_c F_0 + \beta_c$, so the truth's value is
the same at every $c_{\text{true}}$.

# Loadings across c, for three simulations

Structure plots of the fitted $L$ at every $c_{\text{fit}}$, with the truth shown
first, for three simulated datasets: the smallest $c_{\text{true}}$ on the grid,
$c_{\text{true}} = 1$, and $c_{\text{true}} = \infty$. Together they span the
log-additive end, the elbow, and the linear (topic-model) end.

Normalization is the paper convention and nothing else: each loading column is
divided by its maximum so every factor peaks at one. There is no row
normalization, so bars vary in total height — `structure_plot` row-normalizes
only when handed a fit object, not a matrix.

That height is worth reading rather than ignoring. After column normalization, a
tall bar means the sample carries substantial weight on several factors at once,
and a short one means its weight is concentrated. The panels at small
$c_{\text{fit}}$ are visibly taller than those at large $c_{\text{fit}}$, which
is the same finding as the Hoyer figures above, seen sample by sample. All nine
panels within a figure therefore share one $y$ limit; letting each rescale itself
would erase exactly that comparison.

The fitted factors are permuted to the truth by the rule the metrics use — the
column permutation maximizing total correlation with the true $L$. Without it
each panel would differ from the truth by an arbitrary relabelling and every fit
would look wrong.

**Edit `pick_seed` and `pick_init` and re-run to walk around the grid.** Note the
true $L$ is drawn once per seed and shared across every $c_{\text{true}}$, so the
"TRUE L" panel is the same in all three figures; only the fitted panels move.

```{r pick-cell}
pick_seed <- 1
pick_init <- "rank1"     # "rank1" or "random"
```

```{r structure-fn}
fmtc <- function(x) format(x, scientific = FALSE, trim = TRUE)

cell_file <- function(seed, c_true, c_fit)
  file.path(cells_dir,
            paste0(tag_of(list(seed = seed, c_true = c_true, c_fit = c_fit)), ".rds"))

## L is drawn once per seed, so the truth and the cell ordering are shared by all
## three figures below
Lt   <- readRDS(cell_file(pick_seed, C_GRID[1], C_GRID[1]))$L_true
dom  <- max.col(Lt)
ord  <- order(dom, -Lt[cbind(seq_along(dom), dom)])
grp  <- factor(dom[ord])
TOP  <- paste0("k", seq_len(ncol(Lt)))
FCOL <- setNames(c("#0E8FA8", "#C2410C", "#6D3FC4")[seq_along(TOP)], TOP)

norm_bars <- function(M) {                       # column max -> 1, and nothing else
  M <- pmax(M[ord, , drop = FALSE], 0)
  sweep(M, 2, apply(M, 2, max), "/")
}

leg <- get_legend(
  ggplot(data.frame(x = 1, f = factor(TOP, levels = TOP)), aes(x, 1, fill = f)) +
    geom_col() + scale_fill_manual(values = FCOL, name = "factor") +
    theme_cowplot() + theme(legend.position = "right"))

#' One 3 x 3 grid: true L, then the fit at each c_fit, for a given c_true.
structure_grid <- function(c_true, seed = pick_seed, init = pick_init) {

  fits <- lapply(C_GRID, function(cf) {
    x  <- readRDS(cell_file(seed, c_true, cf))
    Lf <- x$fits[[init]]$LL
    list(L   = Lf[, match_factors(Lt, Lf), drop = FALSE],
         lab = sprintf("c_fit = %s   (L cor %.3f)", fmtc(cf),
                       x$rows$L_cor_mean[x$rows$init == init]))
  })

  ## Without row normalization the bars do not all reach the same height, so a
  ## per-panel y scale would rescale each panel and make the columns look more
  ## alike than they are. One shared limit within a figure keeps them comparable.
  ymax <- max(vapply(c(list(Lt), lapply(fits, `[[`, "L")),
                     function(M) max(rowSums(norm_bars(M))), numeric(1)))

  sp <- function(M, title) {
    Mn <- norm_bars(M); colnames(Mn) <- TOP
    structure_plot(Mn, loadings_order = seq_len(nrow(Mn)), grouping = grp,
                   gap = 6, topics = TOP, colors = FCOL) +
      coord_cartesian(ylim = c(0, ymax)) +
      labs(x = NULL, y = "scaled loading", title = title) +
      theme(plot.title = element_text(size = 10, face = "bold"),
            axis.text.x = element_blank(), axis.ticks.x = element_blank(),
            legend.position = "none")
  }

  plots <- c(list(sp(Lt, "TRUE L")),
             lapply(fits, function(z) sp(z$L, z$lab)))

  ttl <- ggdraw() + draw_label(
    sprintf("seed %d, simulated at c = %s, %s init — cells ordered by true dominant factor",
            seed, fmtc(c_true), init), x = 0.01, hjust = 0, size = 12)

  plot_grid(ttl,
            plot_grid(plot_grid(plotlist = plots, ncol = 3), leg,
                      ncol = 2, rel_widths = c(1, 0.08)),
            ncol = 1, rel_heights = c(0.04, 1))
}
```

## Simulated at the smallest c

```{r structure-small, fig.width=13, fig.height=11}
structure_grid(min(C_GRID))
```

## Simulated at c = 1

```{r structure-one, fig.width=13, fig.height=11}
structure_grid(1)
```

## Simulated at c = Inf

```{r structure-inf, fig.width=13, fig.height=11}
structure_grid(Inf)
```

# Why are the small-c fits so much less sparse than the truth?

The sparsity figures show something odd: when the data are generated at
$c = 10^{-3}$ and fitted at $c = 10^{-3}$ — the fit that also achieves the best
log-likelihood — the recovered loadings are far *less* sparse than the true
loadings (Hoyer $\approx 0.13$ against a truth of $0.32$). At large $c$ the same
comparison is nearly exact. Why?

## The likelihood constrains only the product, so sparsity can move between L and F

An NMF is only ever determined up to an invertible $T$:

$$
LF' = (LT)(FT^{-\top})',
$$

so **any** $T$ for which both $LT \ge 0$ and $FT^{-\top} \ge 0$ gives a different
parameter value with an *identical* likelihood. That is the whole ambiguity; the
question is only how large the set of admissible $T$ is.

A convenient one-parameter slice through it blends each factor slightly into the
others:

$$
T_t = (1-t)\,I + \frac{t}{K}\mathbf{1}\mathbf{1}',
\qquad
T_t^{-1} = \frac{1}{1-t}\left(I - \frac{t}{K}\mathbf{1}\mathbf{1}'\right).
$$

Here **$t$ is a mixing fraction, not Euler's number** — it is how much of each
factor gets blended into the others. $t = 0$ is the identity and changes nothing;
$t \to 1$ drives all $K$ factors toward being identical. Written out, the two
sides move in opposite directions:

$$
L \;\mapsto\; LT_t \quad\text{(blends the loading columns together)},
\qquad
F \;\mapsto\; FT_t^{-1} = \frac{1}{1-t}\Bigl(F - \tfrac{t}{K}\,\text{rowSums}(F)\,\mathbf{1}'\Bigr)
\quad\text{(subtracts a per-feature constant)}.
$$

So raising $t$ makes $L$ **denser** and $F$ **sparser** at the same time. Nothing
here is special to $L$: sparsity is simply being moved from one side to the
other, and the likelihood is indifferent to where it sits.

```{r mixing-test}
hoy <- function(v) { v <- abs(v); n <- length(v)
                     (sqrt(n) - sum(v)/sqrt(sum(v^2))) / (sqrt(n) - 1) }
mh  <- function(M) mean(apply(M, 2, hoy))

dsm <- sim_dataset(1, min(C_GRID))
L0  <- dsm$truth$L; F0c <- dsm$FF_true; KK <- ncol(L0)

knitr::kable(do.call(rbind, lapply(c(-0.3, -0.1, 0, 0.1, 0.3, 0.5, 0.7), function(t) {
  Tm <- (1 - t) * diag(KK) + (t / KK) * matrix(1, KK, KK)
  Li <- L0 %*% Tm; Fi <- F0c %*% t(solve(Tm))
  data.frame(t = t,
             `both nonneg` = min(Li) >= -1e-9 && min(Fi) >= -1e-9,
             `Hoyer of L` = mh(Li),
             `Hoyer of F` = mh(Fi),
             `max |change in theta|` = max(abs(tcrossprod(Li, Fi) -
                                               tcrossprod(L0, F0c))),
             check.names = FALSE)
})), digits = 3,
caption = "Truth at c = 0.001 under the mixing family. t = 0 is the truth itself.")
```

$\theta$ is unchanged to machine precision throughout, so every admissible row of
that table has **exactly** the same likelihood as the truth. Yet the sparsity of
$L$ ranges over an order of magnitude, and $F$ moves the other way. The truth's
particular split is one arbitrary point on a continuum; no amount of data can
single it out.

Note the freedom is one-sided here: $t < 0$ (which would make $L$ sparser and $F$
denser) is infeasible. Since $\text{rowSums}(L) = 1$, $LT_t$ at $t<0$ equals
$(1+|t|)L - \tfrac{|t|}{K}\mathbf{1}\mathbf{1}'$, and the many tiny
$\text{Dirichlet}(0.3)$ entries of $L$ go negative immediately.

## What limits the trade is zeros — and small c has none

Mixing can continue only until some entry of $LT_t$ or $FT_t^{-1}$ reaches zero.
Nonnegativity is the *only* thing pinning an NMF down, so the size of the
admissible set is governed by how close the factors come to zero. And that is
exactly what $c$ controls, because the calibrated truth is
$F_c = \alpha_c F_0 + \beta_c$ — a floor of $\beta_c$ under **every** entry.

```{r floor-vs-freedom, cache=TRUE}
tr0 <- sim_truth(1)

## largest mixing fraction t that keeps both factors nonnegative
max_mix <- function(L, F) {
  feas <- function(t) { Tm <- (1-t)*diag(KK) + (t/KK)*matrix(1, KK, KK)
    min(L %*% Tm) >= -1e-9 && min(F %*% t(solve(Tm))) >= -1e-9 }
  if (!feas(1e-6)) return(0)
  lo <- 0; hi <- 1
  for (i in 1:40) { mid <- (lo + hi)/2; if (feas(mid)) lo <- mid else hi <- mid }
  lo
}

free <- do.call(rbind, lapply(C_GRID, function(cc) {
  dd <- sim_dataset(1, cc, tr0)
  th <- dd$alpha * tr0$theta0 + dd$beta
  t  <- max_mix(tr0$L, dd$FF_true)
  Tm <- (1 - t) * diag(KK) + (t / KK) * matrix(1, KK, KK)
  sub <- d$c_true == cc & d$c_fit == cc & d$init == "rank1"
  data.frame(c_true = cc,
             `floor / mean theta`   = dd$beta / mean(th),
             `largest feasible t`   = t,
             `Hoyer L at that t`    = mh(tr0$L %*% Tm),
             `Hoyer L fitted`       = mean(d$hoyer_L[sub]),
             `Hoyer L true`         = mh(tr0$L),
             `Hoyer F at that t`    = mh(dd$FF_true %*% t(solve(Tm))),
             `Hoyer F fitted`       = mean(d$hoyer_F[sub]),
             `Hoyer F true`         = mh(dd$FF_true),
             check.names = FALSE)
}))
knitr::kable(free, digits = 3, row.names = FALSE)
```

At $c = 10^{-3}$ the constant floor is `r round(100*free[[2]][1])`% of $\theta$:
no entry of $F_c$ is anywhere near zero, no nonnegativity constraint is active,
and the trade can run all the way to $t = `r round(free[[3]][1], 2)`$. At
$c = \infty$ the floor is `r round(100*tail(free[[2]],1))`% — $F_c$ keeps the
near-zeros inherited from $F_0$'s `r 100*PI0`% exact zeros, the constraints bite,
and only $t = `r round(tail(free[[3]],1), 3)`$ is available.

```{r floor-corr}
gap <- free[["Hoyer L true"]] - free[["Hoyer L fitted"]]
c(`cor(floor fraction, L sparsity gap)`  = cor(free[[2]], gap),
  `cor(largest feasible t, L gap)`       = cor(free[[3]], gap))
```

Compare the "Hoyer L at that t" and "Hoyer L fitted" columns for $c \ge 1$: the
optimizer lands essentially **exactly** at the densest admissible point. It takes
all the slack the constraints allow, and coordinate descent from either start
spreads mass across factors with nothing pulling it back.

## The fit is not finding better structure

Worth checking that the dense solution is not somehow a genuinely better
explanation. It does beat the true parameters on the training data, but by about
the amount generic overfitting predicts:

```{r overfit-check}
dsm1 <- sim_dataset(1, min(C_GRID))
xx   <- readRDS(cell_file(1, min(C_GRID), min(C_GRID)))
Lf   <- xx$fits$rank1$LL; Ff <- xx$fits$rank1$FF
pll  <- function(lam) sum(dsm1$Y * log(pmax(lam, 1e-300)) - lam - lgamma(dsm1$Y + 1))
lam_fit <- min(C_GRID) * expm1(tcrossprod(Lf, Ff) / max(1, min(C_GRID)))
c(fitted            = pll(lam_fit),
  `true parameters` = pll(dsm1$lambda),
  advantage         = pll(lam_fit) - pll(dsm1$lambda),
  `free parameters / 2` = (K_TRUE * (N_ROWS + N_COLS) - K_TRUE^2) / 2)
```

## Caveat: how much of this is the simulation design?

Some of it is. The $\beta_c$ offset exists so the datasets are comparable across
$c$, and it is precisely what strips the zeros out of $F_c$ at small $c$. Without
it, $\theta$ would inherit $\theta_0$'s exact zeros — `r 100*PI0`% of $F_0$ — and
the small-$c$ fits would be far better pinned.

The mechanism is still real rather than purely an artifact: $\theta = \log(1 +
\lambda/c)$ is uniformly large when $c$ is small relative to the rates, so
near-zeros in $\theta$ are intrinsically scarce there, whatever the simulation.
But the *magnitude* seen above is amplified by the calibration, and the honest
reading is:

> At small $c$ the log1p factorization is weakly identified in a specific way:
> the likelihood pins down $LF'$ but says little about how sparsity is divided
> between $L$ and $F$. The sparsity of a small-$c$ fit is therefore a property of
> the optimizer and the initialization as much as of the data — consistent with
> the two initializations reaching the same log-likelihood to within 0.008 while
> differing by up to 0.09 in Hoyer sparsity.

The mixing family above is only one convenient one-parameter slice through the
admissible set of $T$; the full set is larger, so these numbers are a lower bound
on how much the split can move.

```{r session}
sessionInfo()
```
