---
title: "Factor Sim"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2026-06-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

```{r load-pkgs}
library(ggplot2)
library(fastTopics)
library(log1pNMF)
library(cowplot)
theme_set(theme_cowplot())
set.seed(1)
```

## Model

Here, we generate a $1000 \times 400$ data matrix $\mathbf{Y}$ from a rank-3 Poisson NMF model. Specifically, we first define a "baseline" factor $\mathbf{f}_0$ and two sparse change factors $\mathbf{f}_A$ and $\mathbf{f}_B$. Ordering the $p = 400$ features in four blocks of $100$, we define these factors as

\begin{align}
\mathbf{f}_0 &= \big(\underbrace{2, \ldots, 2}_{100},\ \underbrace{1000, \ldots, 1000}_{100},\ \underbrace{2, \ldots, 2}_{100},\ \underbrace{1000, \ldots, 1000}_{100}\big), \\
\mathbf{f}_A &= \big(\underbrace{18, \ldots, 18}_{100},\ \underbrace{100, \ldots, 100}_{100},\ \underbrace{0, \ldots, 0}_{100},\ \underbrace{0, \ldots, 0}_{100}\big), \\
\mathbf{f}_B &= \big(\underbrace{0, \ldots, 0}_{100},\ \underbrace{0, \ldots, 0}_{100},\ \underbrace{18, \ldots, 18}_{100},\ \underbrace{100, \ldots, 100}_{100}\big).
\end{align}

Then, we define four groups of samples based on their loadings on these factors. Specifically, we define loadings as

\begin{equation}
\mathbf{l}_i =
\begin{cases}
(1, 0, 0) & \text{if } 1 \leq i \leq 250 \\
(1, 1, 0)  & \text{if } 250 < i \leq 500 \\
(1, 0, 1)  & \text{if } 500 < i \leq 750 \\
(1, 1, 1)  & \text{if } 750 < i \leq 1000
\end{cases}
\end{equation}

where each loading is ordered $(l_{i0}, l_{iA}, l_{iB})$. For each sample $i$ we define $\boldsymbol{\lambda}_i = l_{i0} \mathbf{f}_0 + l_{iA} \mathbf{f}_A + l_{iB} \mathbf{f}_B$, and draw $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$ independently across $i$ and $j$.

We designed the change factors so that each one acts on two kinds of features, distinguished by their baseline rate. The two kinds respond very differently when their change factor turns on:

| Feature type  | Baseline rate ($\mathbf{f}_0$) | Rate when active | Fold change | Absolute change |
|:--------------|:------------------------------:|:----------------:|:-----------:|:---------------:|
| High-baseline | $1000$                         | $1100$           | $1.1\times$ | $+100$          |
| Low-baseline  | $2$                            | $20$             | $10\times$  | $+18$           |

That is, high-baseline features undergo a large *absolute* change but a small *relative* (fold) change, whereas low-baseline features undergo a large *relative* change but a small *absolute* change.

We note that while we have written this model using an additive link, we could equivalently describe it as a log1p NMF model with any value of $c$. To do this for a given value of $c$, we can keep the loadings $\mathbf{l}_i$ unchanged and replace the factors as follows:

\begin{align}
\mathbf{f}_0^{(c)} &= \alpha_c \log\!\left(1 + \tfrac{1}{c}\,\mathbf{f}_0\right), \\
\mathbf{f}_A^{(c)} &= \alpha_c \left[\log\!\left(1 + \tfrac{1}{c}(\mathbf{f}_0 + \mathbf{f}_A)\right) - \log\!\left(1 + \tfrac{1}{c}\,\mathbf{f}_0\right)\right], \\
\mathbf{f}_B^{(c)} &= \alpha_c \left[\log\!\left(1 + \tfrac{1}{c}(\mathbf{f}_0 + \mathbf{f}_B)\right) - \log\!\left(1 + \tfrac{1}{c}\,\mathbf{f}_0\right)\right],
\end{align}

where $\log(\cdot)$ is applied elementwise. With these factors, $\alpha_c \log(1 + \lambda_{ij} / c) = l_{i0} f^{(c)}_{0j} + l_{iA} f^{(c)}_{Aj} + l_{iB} f^{(c)}_{Bj}$ holds exactly for every sample $i$ and feature $j$. Such a construction exists for any $c$ only because factors $A$ and $B$ have disjoint support, in the sense that no feature is non-zero in both change factors, so the nonlinear link never has to represent an interaction between them.

```{r design}
nP <- 100                         # genes per program panel
p  <- 4 * nP

off_rel <- 2;    on_rel <- 20     # relative-change genes (10x fold, +18)
off_abs <- 1000; on_abs <- 1100   # absolute-change genes (1.1x fold, +100)

panel <- factor(rep(c("A-rel", "A-abs", "B-rel", "B-abs"),
                    c(nP, nP, nP, nP)),
                levels = c("A-rel", "A-abs", "B-rel", "B-abs"))

# the three true factors: baseline + two additive, up-regulating programs
baseP <- c(rep(off_rel, nP), rep(off_abs, nP), rep(off_rel, nP), rep(off_abs, nP))
devA  <- c(rep(on_rel - off_rel, nP), rep(on_abs - off_abs, nP),
           rep(0, nP), rep(0, nP))
devB  <- c(rep(0, nP), rep(0, nP),
           rep(on_rel - off_rel, nP), rep(on_abs - off_abs, nP))
```

## Results

We simulate the four groups of samples, then fit two models: a topic model (the
log1p model in the limit $c \to \infty$) and the log1p model at $c = 1$. As we did in the paper, both models are initialized by first fitting a rank-1 model, and then other factors are initialized randomly. 

Below are the results. It's clear that both the log1pNMF model with $c = 1$ and the $c = \infty$ recover almost exactly the same loading structure. However, they differ greatly in the relative ordering of the features in the factors. In the $c = \infty$ fit, the most highly ranked features are those with greatest absolute changes due to that factor, whereas in the $c = 1$ fit the most highly ranked features are those with the greatest relative changes. 

```{r simulate}
npop    <- 250                     # samples per group
La      <- c(0, 1, 0, 1)           # program-A indicator per group
Lb      <- c(0, 0, 1, 1)           # program-B indicator per group
groupname <- c("Group 1 (Baseline)", "Group 2 (Baseline + A)",
               "Group 3 (Baseline + B)", "Group 4 (Baseline + A + B)")

prof   <- sapply(1:4, function(i) baseP + La[i] * devA + Lb[i] * devB)  # p x 4
Lambda <- t(prof[, rep(1:4, each = npop)])                             # n x p
n      <- nrow(Lambda)

Y <- matrix(rpois(n * p, as.vector(t(Lambda))), nrow = n, ncol = p, byrow = TRUE)
rownames(Y) <- paste0("cell", 1:n)
colnames(Y) <- paste0("gene", 1:p)

# true loadings: baseline on in all samples, programs on per group
Ltrue <- cbind(Baseline = 1, A = La[rep(1:4, each = npop)],
                             B = Lb[rep(1:4, each = npop)])
```

```{r fit}
# topic model (= log1p as c -> Inf): rank-1 init has to be built manually
r1 <- fastTopics:::fit_pnmf_rank1(Y)
init_LL <- cbind(r1$L, matrix(1e-3, n, 2)); rownames(init_LL) <- rownames(Y)
init_FF <- cbind(r1$F, matrix(1e-3, p, 2)); rownames(init_FF) <- colnames(Y)
tm <- fit_poisson_nmf(Y, fit0 = init_poisson_nmf(Y, F = init_FF, L = init_LL),
                      verbose = "none")

# log1p model at c = 1 (rank-1 is the default init_method)
set.seed(1)
lp <- fit_poisson_log1p_nmf(Y, K = 3, loglik = "exact", init_method = "rank1",
                            control = list(maxiter = 200, verbose = FALSE))

# --- helpers: put fitted factors in (Baseline, A, B) order so all plots agree ---
ord_of <- function(L) {
  cc <- function(t) suppressWarnings(apply(L, 2, function(x) cor(x, t)))
  jA <- which.max(cc(Ltrue[, "A"])); jB <- which.max(cc(Ltrue[, "B"]))
  c(setdiff(seq_len(ncol(L)), c(jA, jB)), jA, jB)
}
relabel <- function(L) {
  M <- L[, ord_of(L), drop = FALSE]; colnames(M) <- c("Baseline", "A", "B"); M
}

# shared factor colors, and a role palette kept clear of the A/B (red/teal) hues:
# amber = high-baseline features, purple = low-baseline; darker = up, lighter = flat.
# The role palette is validated colorblind-safe (adjacent CVD dE >= 8, all four in
# the L 0.43-0.77 band with real chroma); the legend supplies the labels the
# lighter fills need for contrast relief.
FACCOL  <- c("Baseline" = "grey60", "A" = "#AE2012", "B" = "#005F73")
ROLECOL <- c("High Rate, Up"   = "#B36A00", "High Rate, Flat" = "#E0A81E",
             "Low Rate, Up"     = "#574494", "Low Rate, Flat"  = "#9B7FD4")
```


```{r big-figure, fig.width=10, fig.height=10.5}
## ---------------- Panel A: ground-truth rates by group ----------------
# split each group's rate into a grey baseline plus an additive, colored cap for
# each active program; geom_rect with data-space ymin/ymax keeps the segment
# boundaries at the true rates under the log1p axis
seg1 <- function(k) {
  out <- data.frame(gene = 1:p, ymin = 0, ymax = baseP, factor = "Baseline")
  if (La[k] == 1) out <- rbind(out, data.frame(gene = 1:p, ymin = baseP,
      ymax = baseP + devA, factor = "A")[devA > 0, ])
  if (Lb[k] == 1) out <- rbind(out, data.frame(gene = 1:p, ymin = baseP,
      ymax = baseP + devB, factor = "B")[devB > 0, ])
  out$factor <- factor(out$factor, levels = c("Baseline", "A", "B")); out
}
# a single faceted plot so the four groups have EQUAL panel widths and share one
# scale; facet_wrap then shows the y-axis only on the leftmost panel
segA <- do.call(rbind, lapply(1:4, function(k)
  transform(seg1(k), group = factor(groupname[k], levels = groupname))))

# one shared "Factor" legend, placed left of Panel A and right of Panel B
leg_factor <- get_legend(
  ggplot(data.frame(x = 1, f = factor(c("Baseline", "A", "B"),
                                      levels = c("Baseline", "A", "B"))),
         aes(x, 1, fill = f)) + geom_col() +
    scale_fill_manual(values = FACCOL, name = "Factor") +
    theme(legend.position = "right"))

panelA <- ggplot(segA, aes(xmin = gene - 0.5, xmax = gene + 0.5,
                           ymin = ymin, ymax = ymax, fill = factor)) +
  geom_rect() +
  facet_wrap(~ group, nrow = 1) +
  scale_fill_manual(values = FACCOL, name = "Factor", drop = FALSE) +
  scale_y_continuous(trans = "log1p", breaks = c(0, 2, 20, 1000), limits = c(0, 1100)) +
  labs(x = "Feature", y = "True rate") +
  theme(legend.position = "none", strip.background = element_blank(),
        strip.text = element_text(size = 11, face = "bold"))

## ---------------- Panel B: structure plots (baseline on the bottom) ----------------
grp <- factor(rep(1:4, each = npop))            # numeric group per sample
SP <- function(L, title, ylab = FALSE) {
  g <- structure_plot(log1pNMF:::normalize_bars(pmax(L, 0)),
                      loadings_order = seq_len(nrow(L)), grouping = grp, gap = 24,
                      topics = c("B", "A", "Baseline"), colors = FACCOL) +
    ggtitle(title) + labs(x = "Group", y = if (ylab) "Membership" else NULL) +
    theme(plot.title = element_text(size = 11, hjust = 0.5), legend.position = "none",
          axis.title.y = element_text(size = 14),
          axis.text.x = element_text(angle = 0, hjust = 0.5))
  if (!ylab) g <- g + theme(axis.text.y = element_blank(),
                            axis.ticks.y = element_blank())
  g
}
panelB <- plot_grid(SP(Ltrue, "True Sample Scores", ylab = TRUE),
                    SP(relabel(tm$L),  "c = ∞ Sample Scores"),
                    SP(relabel(lp$LL), "c = 1 Sample Scores"), nrow = 1)

## ---------------- Panel C: factor-value histograms by gene role ----------------
role_for <- function(prog) {
  other <- setdiff(c("A", "B"), prog)
  factor(ifelse(panel == paste0(prog,  "-abs"), "High Rate, Up",
         ifelse(panel == paste0(other, "-abs"), "High Rate, Flat",
         ifelse(panel == paste0(prog,  "-rel"), "Low Rate, Up",
                                                "Low Rate, Flat"))),
         levels = c("High Rate, Up", "High Rate, Flat",
                    "Low Rate, Up",  "Low Rate, Flat"))
}
fac_df <- function(F, L, model) {
  Fo <- F[, ord_of(L)]                              # (Baseline, A, B)
  rbind(data.frame(value = Fo[, 2] / max(Fo[, 2]), role = role_for("A"),
                   factor = "Factor A", model = model),
        data.frame(value = Fo[, 3] / max(Fo[, 3]), role = role_for("B"),
                   factor = "Factor B", model = model))
}
facC <- rbind(fac_df(tm$F, tm$L, "c = ∞"), fac_df(lp$FF, lp$LL, "c = 1"))
facC$model <- factor(facC$model, levels = c("c = ∞", "c = 1"))

panelC <- ggplot(facC, aes(value, fill = role)) +
  geom_histogram(position = "identity", alpha = 0.85, bins = 60) +
  facet_grid(model ~ factor) +
  scale_fill_manual(values = ROLECOL, name = "Feature Type") +
  labs(x = "Factor Value (Normalized to Max 1)", y = "Count") +
  theme(strip.background = element_blank(), strip.text = element_text(face = "bold"),
        strip.text.y = element_text(angle = 0),
        panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.7))

## ---------------- assemble: A over B share one legend (right, centered); C centered below ----------------
AB  <- plot_grid(panelA, panelB, ncol = 1, labels = c("A", "B"), rel_heights = c(1, 1))
top <- plot_grid(AB, leg_factor, ncol = 2, rel_widths = c(1, 0.10))
Crow <- plot_grid(NULL, panelC, NULL, ncol = 3, rel_widths = c(0.12, 1, 0.12),
                  labels = c("", "C", ""))
plot_grid(top, Crow, ncol = 1, rel_heights = c(2, 1.05))
```
