---
title: "Analysis of Greedy Initialization for PBMC Purified Analysis"
output: html_document
date: "2024-04-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, comment = FALSE, message = FALSE)
options(warn=-1)
```

## Introduction

Here, I analyzed the PBMC purified dataset using three different methods: (1) Poisson NMF (as implemented in fastTopics), (2) log1p transformed frobenius NMF, and (3) Poisson NMF with a log1p link function.

```{r}
library(dplyr)
library(tidyr)
library(purrr)
library(stringr)
```

```{r}
load("~/Documents/data/fastglmpca/raw_data/pbmc_purified.RData")
set.seed(1)
sz <- Matrix::rowSums(counts) / mean(Matrix::rowSums(counts))
counts <- counts[, Matrix::colSums(counts) >= 100]
genes_map <- genes$ensembl
names(genes_map) <- genes$symbol

genes_map_inv <- genes$symbol
names(genes_map_inv) <- genes$ensembl
```


Below is the code to run the log1p link Poisson NMF factorization method greedily.

```{r, eval=FALSE}
library(Matrix)
set.seed(1)
log1p_greedy_fit <- fit_factor_model_log1p_quad_approx_sparse_greedy(
  Y = counts,
  K = 25,
  iter_per_factor = 8,
  approx_range = c(0, 1.25),
  s = as.vector(sz)
)
```

Below is the code to run the log1p link Poisson NMF factorization method with a greedy initialization.

```{r, eval=FALSE}
set.seed(1)
log1p_fit_init_greedy <- fit_factor_model_log1p_quad_approx_sparse(
  Y = counts,
  K = 25,
  maxiter = 50,
  approx_range = c(0, 1.25),
  s = as.vector(sz),
  init_U = log1p_fit$U,
  init_V = log1p_fit$V
)
```

Finally, below below is the code to run the log1p link Poisson NMF factorization method with a Frobenius NMF initialization.

```{r, eval=FALSE}
log1p_fit_init_nmf <- fit_factor_model_log1p_quad_approx_sparse(
  Y = counts,
  K = 25,
  maxiter = 25,
  approx_range = c(0, 1.25),
  s = as.vector(sz),
  init_method = "frob_nmf"
)
```

```{r, echo=FALSE}
log1p_greedy_fit <- readr::read_rds(
  "~/Documents/log1p_experiments/output/pbmc_purified_25K_greedy_log1p_fit.rds"
)
log1p_fit_init_greedy <- readr::read_rds(
  "~/Documents/log1p_experiments/output/pbmc_purified_25K_greedy_init_log1p_fit.rds"
  )
log1p_fit_init_nmf <- readr::read_rds(
  "~/Documents/log1p_experiments/output/pbmc_purified_25K_frob_nmf_init_log1p_fit.rds"
  )
```

## Analysis

# Heatmaps

Below are the heatmaps of the different models:

```{r, fig.width=12, fig.height=16}
FF_log1p_greedy <- log1p_greedy_fit$U
FF_log1p_greedy <- scale(
  FF_log1p_greedy,
  center = FALSE, 
  scale = apply(FF_log1p_greedy, 2, max)
)
colnames(FF_log1p_greedy) <- paste0("k", 1:ncol(FF_log1p_greedy))

FF_log1p_greedy_init <- log1p_fit_init_greedy$U
FF_log1p_greedy_init <- scale(
  FF_log1p_greedy_init, 
  center = FALSE,
  scale = apply(FF_log1p_greedy_init, 2, max)
)
colnames(FF_log1p_greedy_init) <- paste0("k", 1:ncol(FF_log1p_greedy_init))

FF_log1p_nmf_init <- log1p_fit_init_nmf$U
FF_log1p_nmf_init <- scale(
  FF_log1p_nmf_init, 
  center = FALSE,
  scale = apply(FF_log1p_nmf_init, 2, max)
)
colnames(FF_log1p_nmf_init) <- paste0("k", 1:ncol(FF_log1p_nmf_init))

cell.type <- samples$celltype

# Downsample the number of cells and sort them using tSNE.
set.seed(8675309)
cell.idx <- numeric(0)
cell.types <- levels(cell.type)
for (i in 1:length(cell.types)) {
  which.idx <- which(cell.type == cell.types[i])
  # Downsample common cell types.
  if (length(which.idx) > 1250) {
    which.idx <- sample(which.idx, 1250)
  }
  # Don't include rare cell types.
  if (length(which.idx) > 20) {
    # Sort using tsne.
    tsne.res <- Rtsne::Rtsne(
      FF_log1p_greedy[which.idx, ],
      dims = 1,
      pca = FALSE,
      normalize = FALSE,
      perplexity = min(100, floor((length(which.idx) - 1) / 3) - 1),
      theta = 0.1,
      max_iter = 1000,
      eta = 200,
      check_duplicates = FALSE
    )$Y[, 1]
    which.idx <- which.idx[order(tsne.res)]
    cell.idx <- c(cell.idx, which.idx)
  }
}

cell.type <- cell.type[cell.idx]
cell.type <- droplevels(cell.type)

FF_log1p_greedy <- FF_log1p_greedy[cell.idx, ]
FF_log1p_greedy_init <- FF_log1p_greedy_init[cell.idx, ]
FF_log1p_nmf_init <- FF_log1p_nmf_init[cell.idx, ]

make.heatmap.tib <- function(FF) {
  tib <- as_tibble(scale(FF, center = FALSE, scale = apply(FF, 2, max))) %>%
    mutate(Cell.type = cell.type) %>%
    arrange(Cell.type) %>%
    mutate(Cell.idx = row_number())
  
  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(str_extract(Factor, "[0-9]+")))
  
  return(tib)
}

log1p_greedy_tib <- make.heatmap.tib(FF_log1p_greedy)
log1p_greedy_init_tib <- make.heatmap.tib(FF_log1p_greedy_init)
log1p_nmf_init_tib <- make.heatmap.tib(FF_log1p_nmf_init)

heatmap.tib <- log1p_greedy_tib %>% mutate(Method = "log1p Poisson NMF Greedy Fit") %>%
  bind_rows(log1p_greedy_init_tib %>% mutate(Method = "log1p Poisson NMF Greedy Init.")) %>%
  bind_rows(log1p_nmf_init_tib %>% mutate(Method = "log1p Poisson NMF Frob NMF Init.")) %>%
  mutate(Method = factor(Method, levels = c("log1p Poisson NMF Greedy Fit", "log1p Poisson NMF Greedy Init.", "log1p Poisson NMF Frob NMF Init.")))

tib <- heatmap.tib %>%
  group_by(Cell.type, Cell.idx) %>%
  summarize()

cell_type_breaks <- c(1, which(tib$Cell.type[-1] != tib$Cell.type[-nrow(tib)]))
label_pos <- cell_type_breaks / 2 + c(cell_type_breaks[-1], nrow(tib)) / 2

library(ggplot2)

plt <- ggplot(heatmap.tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(y = "") +
  scale_y_continuous(breaks = -label_pos,
                     minor_breaks = NULL,
                     labels = levels(cell.type)) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  theme_minimal() +
  geom_hline(yintercept = -cell_type_breaks, size = 0.1) +
  facet_wrap(~Method, ncol = 1, axes = "all") +
  theme(legend.position = "none", 
        strip.text = element_text(size = 16)) 

plt
```

The greedy fit looks generally as expected, except that the 24th factor also looks like an intercept, which is strange. With greedy initialization, the fit looks somewhere in between the NMF init and the greedy init. The first factor stays mostly intact, but you can see that some if it has been replaced by other factors. 

# GO Analysis

First, we exclude genes that are very highly expressed across many factors.


```{r}

rownames(log1p_greedy_fit$V) <- genes_map_inv[rownames(log1p_greedy_fit$V)]
rownames(log1p_fit_init_greedy$V) <- genes_map_inv[rownames(log1p_fit_init_greedy$V)]
rownames(log1p_fit_init_nmf$V) <- genes_map_inv[rownames(log1p_fit_init_nmf$V)]

log1p_greedy_fit$V <- log1p_greedy_fit$V[
  rownames(log1p_greedy_fit$V) != "MALAT1" &
    !(rownames(log1p_greedy_fit$V) %in% rownames(log1p_greedy_fit$V)[grep("^RPS", rownames(log1p_greedy_fit$V))]) &
    !(rownames(log1p_greedy_fit$V) %in% rownames(log1p_greedy_fit$V)[grep("^RPL", rownames(log1p_greedy_fit$V))]) &
    !(rownames(log1p_greedy_fit$V) %in% rownames(log1p_greedy_fit$V)[grep("\\.", rownames(log1p_greedy_fit$V))]), 
]

log1p_fit_init_greedy$V <- log1p_fit_init_greedy$V[
  rownames(log1p_fit_init_greedy$V) != "MALAT1" &
    !(rownames(log1p_fit_init_greedy$V) %in% rownames(log1p_fit_init_greedy$V)[grep("^RPS", rownames(log1p_fit_init_greedy$V))]) &
    !(rownames(log1p_fit_init_greedy$V) %in% rownames(log1p_fit_init_greedy$V)[grep("^RPL", rownames(log1p_fit_init_greedy$V))]) &
    !(rownames(log1p_fit_init_greedy$V) %in% rownames(log1p_fit_init_greedy$V)[grep("\\.", rownames(log1p_fit_init_greedy$V))]), 
]

log1p_fit_init_nmf$V <- log1p_fit_init_nmf$V[
  rownames(log1p_fit_init_nmf$V) != "MALAT1" &
    !(rownames(log1p_fit_init_nmf$V) %in% rownames(log1p_fit_init_nmf$V)[grep("^RPS", rownames(log1p_fit_init_nmf$V))]) &
    !(rownames(log1p_fit_init_nmf$V) %in% rownames(log1p_fit_init_nmf$V)[grep("^RPL", rownames(log1p_fit_init_nmf$V))]) &
    !(rownames(log1p_fit_init_nmf$V) %in% rownames(log1p_fit_init_nmf$V)[grep("\\.", rownames(log1p_fit_init_nmf$V))]), 
]
```


```{r}
library(clusterProfiler)
library(fgsea)
library(AnnotationDbi)
library(org.Hs.eg.db)

get_go_terms <- function(V) {
  
  K <- ncol(V)
  
  genes_vec <- c()
  go_terms_vec <- c()
  
  for (k in 1:K) {
    
    driving_genes <- names(sort(V[,k], decreasing = TRUE))[1:20]
    driving_genes_ensembl <- genes_map[driving_genes]
    
    
    go_result <- enrichGO(gene = driving_genes_ensembl,
                          OrgDb = org.Hs.eg.db,
                          keyType = "ENSEMBL",
                          ont = "BP", 
                          pAdjustMethod = "bonferroni",
                          pvalueCutoff = 0.01,
                          qvalueCutoff = 0.1)@result
    
    go_result <- go_result %>%
      dplyr::filter(p.adjust < .01)
    
    go_terms <- go_result$Description
    
    genes_vec <- c(genes_vec, paste(driving_genes, collapse = ", "))
    go_terms_vec <- c(go_terms_vec, paste(go_terms, collapse = ", "))
    
  }
  
  go_df <- data.frame(
    driving_genes = genes_vec,
    go_terms = go_terms_vec,
    factor = 1:K
  )
  
  return(go_df)
  
}

gdf <- get_go_terms(log1p_greedy_fit$V)
gdf2 <- get_go_terms(log1p_fit_init_greedy$V)
gdf3 <- get_go_terms(log1p_fit_init_nmf$V)
```

Go Terms for log1p link Poisson NMF greedy fit

```{r}
knitr::kable(gdf)
```

Go Terms for log1p link Poisson NMF greedy initialization

```{r}
knitr::kable(gdf2)
```

Go Terms for log1p link Poisson NMF Frob. NMF initialization

```{r}
knitr::kable(gdf3)
```
