---
title: "GTEX Sex Differences"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2024-10-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Introduction

Here, I wanted to again examine the thesis of multiplicative vs. additive effects by analysis differences in expression between the sexes in GTEX data. Below, I look at GTEX data from the coronary artery. I fit a negative binomial regression model for each gene where the only regression variable is sex. That is, for each gene $j$ I fit the model:

\begin{align*}
y_{ij} &\sim NB(s_{i} \mu_{ij}, \phi_{j}) \\
\mu_{ij} &= \log(\alpha_{j} + \beta_{j}M_{i}),
\end{align*}

where $s_{i}$ is the total library size and $M_{i}$ is $1$ if the sample is from a male and $0$ otherwise. I also perform shrinkage on the values of $\beta_{j}$ using `ashr`. 

```{r, eval=FALSE}
df <- read.delim(file="~/Downloads/gene_reads_2017-06-05_v8_artery_coronary.gct", skip = 2)

counts <- df %>%
  dplyr::select(-c(id, Name, Description)) %>%
  as.matrix()

rownames(counts) <- df$Description

# filter out very low expression genes
counts <- counts[Matrix::rowSums(counts) >= 100, ]

md <- read.delim("~/Downloads/GTEx_Analysis_v8_Annotations_SubjectPhenotypesDS.txt")

col_obs_ids <- sub("\\..*", "", sub("\\.", "-", colnames(counts)))

md <- md %>% dplyr::filter(SUBJID %in% col_obs_ids)
colnames(counts) <- col_obs_ids

md$gender <- as.factor(ifelse(md$SEX == 1, "M", "F"))

counts <- counts[,md$SUBJID]
tot <- Matrix::colSums(counts)
md$tot <- tot
b0_vec <- c()
b1_vec <- c()
b1_se_vec <- c()
gene_vec <- c()

it <- 1

for (gene in rownames(counts)) {

  print(round(it / nrow(counts), 2))

  md$expr <- counts[gene, ]

  # Wrap the model fitting and extraction in tryCatch
  tryCatch({
    mod <- MASS::glm.nb(
      expr ~ gender + offset(log(tot)),
      data = md
    )

    b0_vec <- c(b0_vec, coef(summary(mod))["(Intercept)", "Estimate"])
    b1_vec <- c(b1_vec, coef(summary(mod))["genderM", "Estimate"])
    b1_se_vec <- c(b1_se_vec, coef(summary(mod))["genderM", "Std. Error"])
    gene_vec <- c(gene_vec, gene)

  }, error = function(e) {
    message(paste("Error with gene:", gene, "- Skipping this gene"))
  })

  it <- it + 1
}

out_df <- data.frame(
  gene = gene_vec,
  b1 = b1_vec,
  b0 = b0_vec,
  b1_se = b1_se_vec
)

library(ashr)
a_out <- ash(betahat = out_df$b1, sebetahat = out_df$b1_se)

out_df$b1_pm <- ashr::get_pm(a_out)

off_m <- mean(tot)

out_df <- out_df %>%
  dplyr::mutate(
    pred_diff = off_m * exp(b0) - off_m * exp(b0 + b1_pm)
  )

out_df <- out_df %>%
  dplyr::mutate(
    baseline = if_else(
      pred_diff > 0, off_m * exp(b0), off_m * exp(b0 + b1)
    )
  )
```

```{r}
library(ggplot2)
library(ggpubr)
out_df <- readr::read_rds("~/Documents/data/passPCA/experiment_results/gtex_artery_sex_diff.rds")
```

The plot below is for all genes:

```{r}
g1 <- ggplot(data = out_df, aes(x = baseline, y = abs(pred_diff))) +
  geom_point() +
  geom_smooth() +
  ggtitle(glue::glue("Rank correlation = {round(cor(out_df$baseline, abs(out_df$pred_diff), method = 'spearman'), 2)}")) +
  xlab("Baseline Expression") +
  ylab("|Difference in Expression|")

g2 <- ggplot(data = out_df, aes(x = log(baseline), y = abs(b1_pm))) +
  geom_point() +
  geom_smooth() +
  ggtitle(glue::glue("Rank correlation = {round(cor(log(out_df$baseline), abs(out_df$b1_pm), method = 'spearman'), 2)}"))  +
  xlab("log(Baseline Expression)") +
  ylab("|LFC|")

ggarrange(g1, g2, nrow = 1, ncol = 2)
```
The plot below is for genes $j$ where $lfsr(\beta_{j}) < 0.05$:

```{r}
out_sig_df <- out_df %>% dplyr::filter(lfsr < 0.05)

g1 <- ggplot(data = out_sig_df, aes(x = baseline, y = abs(pred_diff))) +
  geom_point() +
  geom_smooth() +
  ggtitle(glue::glue("Rank correlation = {round(cor(out_sig_df$baseline, abs(out_sig_df$pred_diff), method = 'spearman'), 2)}")) +
  xlab("Baseline Expression") +
  ylab("|Difference in Expression|")

g2 <- ggplot(data = out_sig_df, aes(x = log(baseline), y = abs(b1_pm))) +
  geom_point() +
  geom_smooth() +
  ggtitle(glue::glue("Rank correlation = {round(cor(log(out_sig_df$baseline), abs(out_sig_df$b1_pm), method = 'spearman'), 2)}")) +
  xlab("log(Baseline Expression)") +
  ylab("|LFC|")

ggarrange(g1, g2, nrow = 1, ncol = 2)
```


