---
title: "Why the sparsity of a log1p fit is decided by Y and c"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r pkgs}
library(ggplot2); library(cowplot); library(log1pNMF)
theme_set(theme_cowplot(font_size = 11))
exp_dir   <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
cells_dir <- file.path(exp_dir, "c_grid_sim")
source(file.path(exp_dir, "c_grid_sim_common.R"))

## Hoyer sparsity of a nonnegative vector: 0 if all entries are equal, 1 if one
## entry carries everything.
hoyer <- function(x) { x <- abs(x); n <- length(x)
  (sqrt(n) - sum(x)/sqrt(sum(x^2))) / (sqrt(n) - 1) }

## pm(): peak-to-mean. The ONLY summary used on this page. It takes a vector or
## a matrix and returns one number, so pm(L[, k]) is a column's peak-to-mean and
## pm(theta) is the whole matrix's.
pm <- function(x) max(x) / mean(x)

## Axis labels for values of c, with Inf spelled as the symbol.
fmtc <- function(v) ifelse(is.infinite(v), "∞",
  format(v, scientific = FALSE, trim = TRUE, drop0trailing = TRUE))
```

# The question

You have a count matrix $Y$. You fit the log1p model at several values of $c$.
The factors come out visibly more or less sparse depending on which $c$ you used,
even though nothing about the data changed. What decides that?

This page answers that question. It is about **Hoyer** sparsity — a continuous
measure that requires no exact zeros — expressed purely in terms of $Y$ and $c$.
A companion section at the end shows that the same budget also decides how
mutually correlated the fitted factors are.

**The answer in one sentence.** The Hoyer sparsity that $L$ and $F$ can achieve
is jointly capped by the *peak-to-mean ratio* of $\theta = \log(1+\lambda/c)$;
the rates $\lambda$ are fixed by $Y$; and $c$ alone slides that ratio between
$1$ and $\lambda$'s own peak-to-mean.

The rest of the page is that sentence in four steps.

# Notation

Only one summary statistic appears on this page, applied to different things.
For any nonnegative vector or matrix $v$ with at least one positive entry,

$$
\operatorname{pm}(v) \;=\; \frac{\max(v)}{\operatorname{mean}(v)} \;\ge\; 1 ,
$$

the **peak-to-mean ratio**: how large the biggest entry is relative to a typical
one. $\operatorname{pm}(v) = 1$ exactly when $v$ is constant.

The objects it gets applied to:

| symbol | shape | meaning |
|---|---|---|
| $Y$ | $n \times p$ | the observed counts |
| $\lambda$ | $n \times p$ | the rates, $\mathbb{E}[Y]$ |
| $\theta = \log(1+\lambda/c)$ | $n \times p$ | the linear predictor |
| $L$, $F$ | $n\times K$, $p\times K$ | the factorization, $\theta = LF'$ |
| $\ell_{\cdot k}$ | length $n$ | column $k$ of $L$ — every sample's loading on factor $k$ |
| $f_{\cdot k}$ | length $p$ | column $k$ of $F$ — every feature's weight on factor $k$ |
| $\ell_{i\cdot}$ | length $K$ | row $i$ of $L$ — sample $i$'s loadings across factors |

So $\max(\ell_{\cdot k}) = \max_i \ell_{ik}$ is the largest loading any sample
places on factor $k$, and $\operatorname{pm}(\ell_{\cdot k})$ is that divided by
the average loading on factor $k$. Hoyer sparsity is always computed per column
and then averaged over the $K$ columns, matching the earlier pages.

# Step A: Hoyer sparsity is peak-to-mean in disguise

For a nonnegative vector $x$ of length $n$, Hoyer sparsity is

$$
H(x) = \frac{\sqrt{n} - \|x\|_1/\|x\|_2}{\sqrt{n}-1} .
$$

Use $\|x\|_2^2 = \sum_i x_i^2 \le \max(x)\sum_i x_i = \|x\|_\infty\|x\|_1$, and
note $\|x\|_1/\|x\|_\infty = n\,\operatorname{mean}(x)/\max(x) = n/\operatorname{pm}(x)$:

$$
\frac{\|x\|_1}{\|x\|_2} \;\ge\; \frac{\|x\|_1}{\sqrt{\|x\|_\infty\|x\|_1}}
\;=\; \sqrt{\frac{\|x\|_1}{\|x\|_\infty}} \;=\; \sqrt{\frac{n}{\operatorname{pm}(x)}},
$$

and therefore

$$
\boxed{\;H(x) \;\le\; \frac{\sqrt{n}-\sqrt{n/\operatorname{pm}(x)}}{\sqrt{n}-1}
\;\approx\; 1 - \frac{1}{\sqrt{\operatorname{pm}(x)}}\;}
$$

In words: **a vector can only be Hoyer-sparse if some of its entries stick far
above its mean.** If $\operatorname{pm}(x) = 1$ every entry equals the mean, the
vector is constant, and $H(x) = 0$ — however many entries it has, and whether or
not any of them are zero. Sparsity needs *range*, not zeros.

```{r stepA, fig.width=7, fig.height=3.4}
set.seed(1); n <- 500
sim <- do.call(rbind, lapply(seq(0, 1, length.out = 60), function(a) {
  do.call(rbind, lapply(1:8, function(r) {
    x <- (1 - a) + a * rexp(n)^2          # dial from constant to heavy-tailed
    data.frame(pm = pm(x), H = hoyer(x)) }) ) }))
curve <- data.frame(pm = seq(1, max(sim$pm), length.out = 300))
curve$bound <- 1 - 1/sqrt(curve$pm)

ggplot(sim, aes(pm, H)) +
  geom_point(colour = "#0E8FA8", alpha = 0.5, size = 1.2) +
  geom_line(data = curve, aes(pm, bound), colour = "#C2410C", linewidth = 1) +
  scale_x_log10() +
  labs(x = "pm(x)  =  max / mean", y = "Hoyer sparsity  H(x)",
       title = "Sparsity is bounded by peak-to-mean",
       subtitle = "points: random nonnegative vectors.   line: 1 - 1/sqrt(pm)")
```

# Step B: L and F share a single budget

The fit gives $\theta = LF'$ with $L, F \ge 0$. Fix one factor $k$. Let
$i^\star$ be the sample achieving $\max(\ell_{\cdot k})$ and $j^\star$ the
feature achieving $\max(f_{\cdot k})$. Because **every term of the sum is
nonnegative**, we may discard all but the $k$th:

$$
\max(\theta) \;\ge\; \theta_{i^\star j^\star}
\;=\; \sum_{m=1}^{K}\ell_{i^\star m}\,f_{j^\star m}
\;\ge\; \ell_{i^\star k}\,f_{j^\star k}
\;=\; \max(\ell_{\cdot k})\cdot\max(f_{\cdot k}) .
$$

For the mean, the double average factorizes across the sum over factors:

$$
\operatorname{mean}(\theta)
= \frac{1}{np}\sum_{i,j}\sum_{m}\ell_{im}f_{jm}
= \sum_{m}\Bigl(\frac{1}{n}\sum_i \ell_{im}\Bigr)\Bigl(\frac{1}{p}\sum_j f_{jm}\Bigr)
= \sum_{m}\operatorname{mean}(\ell_{\cdot m})\operatorname{mean}(f_{\cdot m}).
$$

Let

$$
w_k \;=\; \frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}{\operatorname{mean}(\theta)} \;\in\;(0,1)
$$

be the share of $\operatorname{mean}(\theta)$ that factor $k$ contributes; by the
display above the $w_k$ sum to one.

Now take the inequality $\max(\theta)\ge\max(\ell_{\cdot k})\max(f_{\cdot k})$,
divide both sides by $\operatorname{mean}(\theta)$, and then multiply the
right-hand side by
$\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})$ over
itself — which is just multiplying by $1$ — and regroup the four factors:

$$
\begin{aligned}
\operatorname{pm}(\theta)
\;=\;\frac{\max(\theta)}{\operatorname{mean}(\theta)}
&\;\ge\; \frac{\max(\ell_{\cdot k})\,\max(f_{\cdot k})}{\operatorname{mean}(\theta)}
&&\text{(divide by }\operatorname{mean}(\theta)\text{)}\\[4pt]
&\;=\; \frac{\max(\ell_{\cdot k})\,\max(f_{\cdot k})}{\operatorname{mean}(\theta)}
\cdot\underbrace{\frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}
{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}}_{=\,1}
&&\text{(insert }1\text{)}\\[4pt]
&\;=\; \underbrace{\frac{\max(\ell_{\cdot k})}{\operatorname{mean}(\ell_{\cdot k})}}_{\operatorname{pm}(\ell_{\cdot k})}
\cdot\underbrace{\frac{\max(f_{\cdot k})}{\operatorname{mean}(f_{\cdot k})}}_{\operatorname{pm}(f_{\cdot k})}
\cdot\underbrace{\frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}{\operatorname{mean}(\theta)}}_{w_k}
&&\text{(regroup)} .
\end{aligned}
$$

The middle step is the only trick, and it is not much of one: the two
$\operatorname{mean}$ terms are introduced solely so that each $\max$ can be
paired with the matching $\operatorname{mean}$ to form a $\operatorname{pm}$,
with whatever is left over collecting into $w_k$. Reading the chain end to end,

$$
\boxed{\;\operatorname{pm}(\ell_{\cdot k})\;\cdot\;\operatorname{pm}(f_{\cdot k})\;\cdot\;w_k
\;\;\le\;\; \operatorname{pm}(\theta)\;}
\qquad \text{for every factor } k .
$$

Read $\operatorname{pm}(\theta)$ as a **budget**: how much peak-to-mean the
fitted $\theta$ has available. The two factors' peak-to-mean ratios enter
*multiplicatively* against it, so

- they **trade**: at fixed $\operatorname{pm}(\theta)$, making $\ell_{\cdot k}$
  more peaked forces $f_{\cdot k}$ flatter;
- they are **jointly capped**: if $\operatorname{pm}(\theta)$ is near $1$ then
  neither can be peaked, since their product cannot exceed it.

Combined with Step A: if $\theta$ has no range, neither factor can be
Hoyer-sparse, and no optimizer or initialization can change that.

This also reconciles the two behaviours you may have noticed. Holding the data
and $c$ fixed and moving among equally-good factorizations slides *along* the
budget constraint — a trade. Changing $c$ moves the constraint itself — so both
can rise together.

## The same budget, written in Hoyer units

Step B constrains peak-to-mean; what you actually measure is Hoyer sparsity. The
two are interchangeable, because Step A's inequality can be read backwards. Define
the **density** of a length-$m$ nonnegative vector,

$$
d(x) \;=\; 1-\gamma_m H(x) \;=\; \frac{\|x\|_1}{\sqrt m\,\|x\|_2}
\;\in\;\Bigl[\tfrac{1}{\sqrt m},\,1\Bigr],
\qquad \gamma_m = 1-\frac{1}{\sqrt m},
$$

so $d(x)\approx 1-H(x)$ for large $m$. The second equality is worth pausing on: it
is exact, and it says the density is nothing but the **normalized
$\ell_1/\ell_2$ ratio**, of which Hoyer sparsity is the affine rescaling onto
$[0,1]$. That is why the results below are natively multiplicative in $d$ — the
budget is a product of norm ratios — and why restating them in Hoyer units costs
a constant $\gamma_m$ per column.

Solving the boxed inequality of Step A for $\operatorname{pm}$ — it is an
equivalence, not a weakening — gives

$$
H(x)\le\frac{\sqrt m-\sqrt{m/\operatorname{pm}(x)}}{\sqrt m-1}
\iff
\operatorname{pm}(x)\;\ge\;\frac{1}{d(x)^2} .
$$

Apply that to $\ell_{\cdot k}$ and to $f_{\cdot k}$, multiply, and read Step B
from the other end:

$$
\frac{1}{d(\ell_{\cdot k})^{2}\,d(f_{\cdot k})^{2}}
\;\le\;\operatorname{pm}(\ell_{\cdot k})\operatorname{pm}(f_{\cdot k})
\;\le\;\frac{\operatorname{pm}(\theta)}{w_k} .
$$

Taking square roots:

$$
d(\ell_{\cdot k})\cdot d(f_{\cdot k})\;\ge\;r_k,
\qquad r_k \;=\; \sqrt{\frac{w_k}{\operatorname{pm}(\theta)}},
\qquad\text{for every factor } k .
$$

This is Step B with nothing thrown away: a **hyperbola in the
$(H(\ell_{\cdot k}), H(f_{\cdot k}))$ plane**, with the feasible region on the
origin side of it. Both columns of a factor cannot be sparse at once, and
lowering $c$ — which lowers $\operatorname{pm}(\theta)$, Step C — pulls the whole
frontier toward the origin.

## The budget as a sum of sparsities

The hyperbola is exact but reads awkwardly, because expanding it leaves a cross
term,

$$
\gamma_n H(\ell_{\cdot k}) + \gamma_p H(f_{\cdot k})
- \gamma_n\gamma_p\,H(\ell_{\cdot k})H(f_{\cdot k}) \;\le\; 1-r_k .
$$

Ask instead the question the picture suggests: **how large can the two
sparsities be *added together*?** Substitute $u=d(\ell_{\cdot k})$,
$v=d(f_{\cdot k})$, so that $H(\ell_{\cdot k})=(1-u)/\gamma_n$ and likewise for
$v$, and the constraint is just $uv\ge r_k$ on $u,v\in(0,1]$. Then by the
weighted AM–GM inequality

$$
\frac{u}{\gamma_n}+\frac{v}{\gamma_p}\;\ge\;2\sqrt{\frac{uv}{\gamma_n\gamma_p}}\;\ge\;2\sqrt{\frac{r_k}{\gamma_n\gamma_p}},
$$

and since $H(\ell_{\cdot k})+H(f_{\cdot k}) = 1/\gamma_n+1/\gamma_p-(u/\gamma_n+v/\gamma_p)$,

$$
\boxed{\;H(\ell_{\cdot k}) + H(f_{\cdot k})
\;\le\; \frac{1}{\gamma_n}+\frac{1}{\gamma_p}-\frac{2\sqrt{r_k}}{\sqrt{\gamma_n\gamma_p}}
\;\xrightarrow[\;n,p\to\infty\;]{}\;
2\left(1-\Bigl(\frac{w_k}{\operatorname{pm}(\theta)}\Bigr)^{1/4}\right)\;}
$$

Here $n=p=300$, so $\gamma_n=\gamma_p=\gamma=`r round(1-1/sqrt(300), 4)`$ and the
bound is simply $\tfrac{2}{\gamma}\bigl(1-\sqrt{r_k}\bigr)=`r round(2/(1-1/sqrt(300)), 3)`\,(1-\sqrt{r_k})$.
This is the statement to remember: **the two columns of a factor split a single
sparsity allowance, and $c$ sets how large that allowance is.**

Two things make it worth preferring to the hyperbola in practice.

**It gives up nothing.** AM–GM is an equality at $u=v$, and that point is
feasible, so the right-hand side is the *exact maximum* of
$H(\ell_{\cdot k})+H(f_{\cdot k})$ over the constraint set — the sharpest
statement that exists about the sum, not a relaxation of the hyperbola. The only
information discarded is how the total gets split between the two columns, and
the bound is attained whenever they are equally sparse.

**It is much stronger than the one-column corollary below.** Halving it gives an
average per column, which the table shows carries less than half the slack of the
cap obtained by discarding $d(f_{\cdot k})$ outright.

## Corollary: a bound on one column at a time

Since $d(f_{\cdot k})\le 1$, drop it:

$$
H(\ell_{\cdot k})\;\le\;\frac{1-\sqrt{w_k/\operatorname{pm}(\theta)}}{1-1/\sqrt n}
\;\approx\;1-\sqrt{\frac{w_k}{\operatorname{pm}(\theta)}},
$$

and symmetrically for $f_{\cdot k}$. This is the upper half of the
[appendix](#the-sandwich) sandwich. It is a genuine theorem about a single
column, but it is much weaker than the hyperbola, and the $w_k$ cannot be removed:
take $K=2$ with a flat dominant factor plus a vanishing spike,
$\ell_{\cdot 2}=\varepsilon e_1$, $f_{\cdot 2}=\varepsilon e_1$. Then
$\operatorname{pm}(\theta)\to 1$ while $\operatorname{pm}(\ell_{\cdot 2})=n$; the
factor stays admissible only because $w_2\to 0$ with it.

Averaging the corollary over $k$ and using $\sqrt{w_k}\ge w_k$, so that
$\sum_k\sqrt{w_k}\ge\sum_k w_k=1$, removes $w$ entirely and bounds the quantity
the simulation figures actually plot:

$$
\frac1K\sum_{k}H(\ell_{\cdot k})\;\;\le\;\;1-\frac{1}{K\sqrt{\operatorname{pm}(\theta)}} .
$$

The observed $\sum_k\sqrt{w_k}$ is $1.73=\sqrt3=\sqrt K$ almost exactly — the
equal-shares value — so in practice the $K$ in the denominator behaves like
$\sqrt K$. That is an observation about these fits, not a theorem.

## How much is lost going from the pair to one column

All 640 cells, both initializations, $K=3$: 3,840 factor-columns.

```{r stepB-columns, cache=TRUE}
dens <- function(x) 1 - hoyer(x) * (1 - 1/sqrt(length(x)))

col <- do.call(rbind, lapply(list.files(cells_dir, "\\.rds$", full.names = TRUE),
  function(f) {
    x <- readRDS(f)
    do.call(rbind, lapply(names(x$fits), function(init) {
      L <- x$fits[[init]]$LL; Fm <- x$fits[[init]]$FF
      th <- tcrossprod(L, Fm)              # pm and H are both scale invariant,
      w  <- colMeans(L) * colMeans(Fm)     # so the alpha rescaling is irrelevant
      data.frame(c_fit = x$rows$c_fit[1], init = init, k = seq_len(ncol(L)),
                 pm_theta = pm(th), w = w / sum(w),
                 HL = apply(L, 2, hoyer), HF = apply(Fm, 2, hoyer),
                 dL = apply(L, 2, dens),  dF = apply(Fm, 2, dens),
                 pmL = apply(L, 2, pm),   pmF = apply(Fm, 2, pm))
    }))
  }))
gam      <- 1 - 1/sqrt(300)                            # n = p = 300 throughout
col$rhs  <- sqrt(col$w / col$pm_theta)                 # r_k, the hyperbola bound
col$capL <- (1 - col$rhs) / gam                        # the one-column cap
col$capS <- pmin(2, (2/gam) * (1 - sqrt(col$rhs)))     # the sum-form cap

qs <- function(v) sprintf("%.2f   (%.2f – %.2f)", median(v), min(v), max(v))
knitr::kable(data.frame(
  quantity = c("Step B itself:  pm(theta) / [pm(l) pm(f) w]",
               "w_k, the per-factor price   (1/K = 0.33)",
               "pm(f_.k), discarded to reach one column",
               "hyperbola slack:  d(l) d(f) / r_k",
               "sum-form slack:  cap - [H(l) + H(f)]   (Hoyer units)",
               "|H(l_.k) - H(f_.k)|, where the sum form is exact at 0",
               "one-column slack:  cap - H(l_.k)   (Hoyer units)"),
  `median (min – max)` = c(
    qs(col$pm_theta / (col$pmL * col$pmF * col$w)), qs(col$w), qs(col$pmF),
    qs(col$dL * col$dF / col$rhs), qs(col$capS - col$HL - col$HF),
    qs(abs(col$HL - col$HF)), qs(col$capL - col$HL)),
  check.names = FALSE), row.names = FALSE)

cat("violations — hyperbola:", sum(col$dL * col$dF < col$rhs - 1e-9),
    " sum form:", sum(col$HL + col$HF > col$capS + 1e-9),
    " one-column L:", sum(col$HL > col$capL + 1e-9),
    " one-column F:", sum(col$HF > col$capL + 1e-9), "of", nrow(col), "\n")
```

Read the table top to bottom and the accounting is complete. **Step B is nearly
an equality** on real fits — median slack 1.09, minimum 1.00 — so the budget
argument is not merely true, it is spent. **$w_k$ behaves**: it sits at $1/K$, so
going per-factor costs a clean factor of $K$ and nothing worse. Everything else
is the third row: the only route from a statement about the pair to a statement
about one column is to discard $\operatorname{pm}(f_{\cdot k})\ge 1$, and on these
fits that throws away a factor of about eight. That is why the one-column bound
carries 0.56 of slack in Hoyer units while the hyperbola carries a factor of two
— and the factor of two that remains is Step A's slack, not the budget argument's.

The sum form sits where you would want it: 0.47 of slack on the *pair*, so 0.23
per column, against 0.56 for the one-column cap. And the columns stay reasonably
near the diagonal $H(\ell_{\cdot k})=H(f_{\cdot k})$ where the sum bound is
attained — median separation 0.16 — so little is lost by declining to say how the
allowance is split.

```{r stepB-hyperbola, fig.width=8, fig.height=7, cache=TRUE}
pA <- ggplot(col, aes(rhs, dL * dF)) +
  geom_abline(slope = 1, intercept = 0, colour = "#C2410C", linewidth = 0.9) +
  geom_point(colour = "#0E8FA8", alpha = 0.25, size = 0.9) +
  scale_x_log10() + scale_y_log10() +
  labs(x = "sqrt(w_k / pm(theta))   — the bound",
       y = "d(l) x d(f)   — the fit",
       title = "Every column obeys the hyperbola",
       subtitle = "3,840 factor-columns; points must lie above the line")

SHOW <- c(1e-3, 1, 100, Inf)
sub  <- col[col$c_fit %in% SHOW, ]
sub$panel <- factor(paste("c_fit =", fmtc(sub$c_fit)),
                    levels = paste("c_fit =", fmtc(SHOW)))
stopifnot(!any(is.na(sub$panel)))
frontier <- do.call(rbind, lapply(levels(sub$panel), function(p) {
  r  <- median(sub$rhs[sub$panel == p])
  hl <- seq(0, 0.995, length.out = 400); d <- 1 - hl * gam
  data.frame(panel = factor(p, levels = levels(sub$panel)), HL = hl,
             HF = (1 - r/d) / gam, cap_sum = (2/gam) * (1 - sqrt(r)))
}))
touch <- unique(frontier[, c("panel", "cap_sum")])       # tangency, on H(l)=H(f)
touch$HL <- touch$HF <- touch$cap_sum / 2
pB <- ggplot(sub, aes(HL, HF)) +
  geom_abline(data = frontier, aes(slope = -1, intercept = cap_sum),
              colour = "#0E5C6B", linetype = "dashed", linewidth = 0.7) +
  geom_line(data = frontier, colour = "#C2410C", linewidth = 0.9) +
  geom_point(data = touch, shape = 21, size = 2.6, stroke = 0.9,
             colour = "#0E5C6B", fill = "white") +
  geom_point(colour = "#0E8FA8", alpha = 0.4, size = 0.9) +
  facet_wrap(~ panel, nrow = 1) +
  coord_cartesian(xlim = c(0, 0.8), ylim = c(0, 0.9)) +
  labs(x = "H(l_.k)", y = "H(f_.k)",
       title = "The frontier moves out as c grows",
       subtitle = "solid: the hyperbola.  dashed: the sum form, touching it at the open circle")
plot_grid(pA, pB, nrow = 2)
```

The lower row is the picture worth keeping. The feasible region is the area
below each solid curve; only $c$ moves it. At $c=10^{-3}$ the frontier is pinched
into the corner and the fits sit right against it — there is no sparser
factorization to find. By $c=100$ the frontier has swung out and the fits, now
much sparser in absolute terms, are no longer anywhere near it. The dashed line
is the sum form: a line of slope $-1$ touching the hyperbola exactly on the
diagonal, which is the geometric content of "it gives up nothing about the sum."

# Step C: c is what sets the budget

Here is the crux. $\theta$ is not free; it is the rates seen through the link,
$\theta = \log(1+\lambda/c)$. Take five rates and never change them, then look at
what $c$ does to their peak-to-mean:

```{r stepC-table}
lam <- c(0.1, 0.5, 1, 5, 20)
cs  <- c(1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3, Inf)
knitr::kable(do.call(rbind, lapply(cs, function(cc) {
  th <- if (is.infinite(cc)) lam else log1p(lam/cc)
  data.frame(c = format(cc),
             `theta = log1p(lambda/c)` = paste(sprintf("%.3g", th), collapse = "  "),
             `pm(theta)` = pm(th),
             `sparsity cap 1-1/sqrt(pm)` = 1 - 1/sqrt(pm(th)),
             check.names = FALSE) })),
  digits = 3, row.names = FALSE,
  caption = "The same five rates in every row. Only c changes.")
```

The rates never move; their own peak-to-mean is
$\operatorname{pm}(\lambda) = `r round(pm(lam), 2)`$. But the budget the fit
actually receives runs from about `r round(pm(log1p(lam/1e-3)), 2)` at
$c=10^{-3}$ up to `r round(pm(lam), 2)` at $c=\infty$, and the sparsity cap
roughly triples with it.

**Why.** For $c$ well below the rates,
$\log(1+\lambda/c)\approx \log(1/c) + \log\lambda$: every entry picks up the
*same* large additive constant $\log(1/c)$, and a shared additive constant is
exactly what destroys peak-to-mean. For $c$ well above the rates,
$\log(1+\lambda/c)\approx\lambda/c$, a pure rescaling, which leaves peak-to-mean
untouched. Formally, writing $\theta_c = \log(1+\lambda/c)$,

$$
\operatorname{pm}(\theta_c)\;\xrightarrow[c\to 0]{}\;1,
\qquad
\operatorname{pm}(\theta_c) - 1 \;\approx\; \frac{\log\max(\lambda)-\operatorname{mean}(\log\lambda)}{\log(1/c)},
\qquad
\operatorname{pm}(\theta_c)\;\xrightarrow[c\to\infty]{}\;\operatorname{pm}(\lambda).
$$

```{r stepC-plot, fig.width=6.5, fig.height=3.6}
grid <- 10^seq(-4, 4, length.out = 200)
d <- data.frame(c = grid, pm = sapply(grid, function(cc) pm(log1p(lam/cc))))
ggplot(d, aes(c, pm)) +
  geom_hline(yintercept = pm(lam), linetype = "dashed", colour = "#C2410C") +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey50") +
  geom_line(colour = "#0E5C6B", linewidth = 1) +
  scale_x_log10() +
  annotate("text", x = 1e3, y = pm(lam), vjust = -0.7, size = 3.2,
           colour = "#C2410C", label = "pm(lambda)") +
  annotate("text", x = 1e-4, y = 1, vjust = -0.7, hjust = 0, size = 3.2,
           colour = "grey35", label = "no range left") +
  labs(x = "c", y = "pm(theta)",
       title = "The sparsity budget, as a function of c",
       subtitle = "same rates throughout; only the link changes")
```

# Step D: the rates come from Y, not from c

The last link is what lets us treat $\lambda$ as given. The Poisson
log-likelihood is

$$
\ell(\lambda) = \sum_{ij}\bigl[y_{ij}\log\lambda_{ij}-\lambda_{ij}\bigr] - \sum_{ij}\log y_{ij}!,
$$

which contains **no $c$ anywhere**. $c$ appears only in the feasible set — which
rate matrices are reachable, namely those for which $\log(1+\lambda/c)$ has
nonnegative rank $K$. So $c$ does not change what a good set of rates is; the
fitted rates are governed by $Y$.

That means $\lambda$ can be estimated from $Y$ before fitting anything, and for
this purpose the crudest estimate is enough — the rank-1 independence model

$$
m_{ij} = \frac{y_{i\cdot}\,y_{\cdot j}}{y_{\cdot\cdot}} .
$$

# The recipe

Given $Y$ and a candidate $c$, with no model fitted:

1. $m_{ij} = y_{i\cdot}y_{\cdot j}/y_{\cdot\cdot}$
2. compute $\operatorname{pm}\bigl(\log(1+m/c)\bigr)$
3. the joint sparsity cap is $\approx 1 - 1/\sqrt{\operatorname{pm}}$, shared
   between $L$ and $F$

If $\operatorname{pm}$ comes out near 1, sparse factors are unavailable at that
$c$, and whatever sparsity a fit reports is an artifact of the optimizer's path.

```{r recipe}
sparsity_budget <- function(Y, c) {
  m  <- outer(rowSums(Y), colSums(Y)) / sum(Y)   # rank-1 estimate of the rates
  th <- if (is.infinite(c)) m else log1p(m / c)
  c(pm_theta = pm(th), cap = 1 - 1/sqrt(pm(th)))
}
Y <- sim_dataset(1, 1)$Y                          # any count matrix
round(t(sapply(C_GRID, function(cc) c(c = cc, sparsity_budget(Y, cc)))), 3)
```

# Does it hold on real fits?

Against the 640 fits from the [c-grid simulation](c_grid_sim.html). The Step B
inequality should never be violated, and if the argument is explanatory rather
than merely true it should be roughly *tight* — the optimizer should spend what
it is given.

```{r verify, cache=TRUE}
ver <- do.call(rbind, lapply(C_GRID, function(cf) {
  r <- do.call(rbind, lapply(1:10, function(sd) {
    Yd <- sim_dataset(sd, 1)$Y
    x  <- readRDS(file.path(cells_dir, paste0(
            tag_of(list(seed = sd, c_true = 1, c_fit = cf)), ".rds")))
    L <- x$fits$rank1$LL; Fm <- x$fits$rank1$FF
    th <- tcrossprod(L, Fm) / max(1, if (is.infinite(cf)) 1 else cf)
    w  <- colMeans(L) * colMeans(Fm); w <- w / sum(w)
    pmL <- apply(L,  2, pm)          # pm(l_.k) for each factor k
    pmF <- apply(Fm, 2, pm)          # pm(f_.k) for each factor k
    c(`pm predicted from Y,c`  = unname(sparsity_budget(Yd, cf)["pm_theta"]),
      `pm(theta) of the fit`   = pm(th),
      `pm(l)pm(f)w, worst k`   = max(pmL * pmF * w),
      `mean H(l_.k)`           = mean(apply(L,  2, hoyer)),
      `mean H(f_.k)`           = mean(apply(Fm, 2, hoyer)),
      violations               = sum(pmL * pmF * w > pm(th) * (1 + 1e-8)))
  }))
  as.data.frame(t(c(c_fit = cf, colMeans(r))))
}))
knitr::kable(ver, digits = c(3, 2, 2, 2, 3, 3, 0), row.names = FALSE)
```

Three things to check in that table.

**The inequality is never violated**, as it must not be, being an identity plus
nonnegativity.

**It is tight.** The `pm(l)pm(f)w` column sits just under `pm(theta) of the fit`
in every row. The fits use their whole budget, which is what makes the bound
explanatory: at small $c$ the optimizer is not failing to find sparse factors,
there are none to find.

**The budget predicted from $Y$ and $c$ alone tracks the realized one**, so the
whole chain can be evaluated before fitting.

```{r verify-plot, fig.width=8, fig.height=3.6}
lab <- ifelse(is.infinite(ver$c_fit), "∞",
              format(ver$c_fit, scientific = FALSE, trim = TRUE, drop0trailing = TRUE))
ver$cf <- factor(lab, levels = lab)
long <- rbind(
  data.frame(cf = ver$cf, what = "budget:  pm(theta)",   v = ver$`pm(theta) of the fit`),
  data.frame(cf = ver$cf, what = "spent:   pm(l)pm(f)w", v = ver$`pm(l)pm(f)w, worst k`))
p1 <- ggplot(long, aes(cf, v, colour = what, group = what)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = c("#0E5C6B", "#C2410C"), name = NULL) +
  labs(x = "c used for fitting", y = "peak-to-mean",
       title = "The budget, and what the fit spends") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.text = element_text(size = 8))
p2 <- ggplot(ver, aes(`pm(theta) of the fit`)) +
  geom_line(aes(y = `mean H(l_.k)`, colour = "H(l)"), linewidth = 1) +
  geom_line(aes(y = `mean H(f_.k)`, colour = "H(f)"), linewidth = 1) +
  geom_line(aes(y = 1 - 1/sqrt(`pm(theta) of the fit`), colour = "cap"),
            linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = c(`H(l)` = "#0E8FA8", `H(f)` = "#C2410C",
                                 cap = "grey40"), name = NULL) +
  scale_x_log10() +
  labs(x = "pm(theta)", y = "mean Hoyer sparsity",
       title = "Both factors rise with the budget")
plot_grid(p1, p2, nrow = 1)
```

# A companion: small c also forces the factors apart

Sparsity is not the only thing that moves with $c$. In the
[c-grid simulation](c_grid_sim.html) the mean pairwise correlation *among* the
$K$ fitted factors — a measure of how separated they are, not of how accurate
they are — runs from about $-0.35$ at $c=10^{-3}$ to about $-0.03$ at
$c=\infty$, even though the factors that generated the data are mutually
uncorrelated. The fits at small $c$ are strongly anti-correlated for a reason
that has nothing to do with the truth, and it is the same budget argument.

**The mechanism.** Row $i$ of $\theta$ is a *nonnegative* combination of the
factors, $\theta_{i\cdot}=\sum_k \ell_{ik}f_{\cdot k}$. Center it over $j$; the
row means drop out and

$$
\sum_k \ell_{ik}\,\tilde f_{\cdot k} \;=\; \tilde\theta_{i\cdot},
\qquad \tilde f_{\cdot k}=f_{\cdot k}-\operatorname{mean}(f_{\cdot k})\mathbf{1}.
$$

When $c$ is small $\theta$ is nearly constant, so the right-hand side is nearly
zero — and $K$ vectors whose *positive* combination is near zero cannot be
mutually uncorrelated. They are pushed apart.

**The identity.** Write $s_k=\|\tilde f_{\cdot k}\|$, $v_k=\tilde f_{\cdot k}/s_k$
and $a_{ik}=\ell_{ik}s_k\ge 0$, so that $\tilde\theta_{i\cdot}=\sum_k a_{ik}v_k$
and $\langle v_k,v_{k'}\rangle=\rho_{kk'}$ is exactly the Pearson correlation
between factors $k$ and $k'$. Expanding the squared norm and summing over rows,

$$
\underbrace{\sum_i\|\tilde\theta_{i\cdot}\|^2}_{E}
=\underbrace{\sum_{i,k}a_{ik}^2}_{A}
+\sum_{k\neq k'}\Bigl(\sum_i a_{ik}a_{ik'}\Bigr)\rho_{kk'} .
$$

The weights $\sum_i a_{ik}a_{ik'}$ are nonnegative and sum to $B-A$, where
$B=\sum_i(\sum_k a_{ik})^2$. So the weighted mean pairwise correlation of the
factors is not merely bounded — it is *determined*:

$$
\boxed{\;\bar\rho \;=\; \frac{E-A}{B-A}\;}
$$

**The floor.** By Cauchy–Schwarz within each row, $(\sum_k a_{ik})^2\le K\sum_k a_{ik}^2$,
so $B\le KA$; since $A/(B-A)$ is increasing in $A$ this gives
$A/(B-A)\ge 1/(K-1)$ and therefore

$$
\bar\rho \;\le\; -\frac{1}{K-1} \;+\; \frac{E}{B-A}
\qquad\xrightarrow[\;E\to 0\;]{}\qquad -\frac{1}{K-1} .
$$

$E$ is the row-centered variation of $\theta$ — the same spread that $c$ controls
in Step C, here in squared-Frobenius rather than peak-to-mean form. As the budget
vanishes the factors are driven to the algebraic floor $-1/(K-1)$, which is
$-0.5$ at $K=3$. (The same algebra is why Dirichlet loadings, whose rows sum
exactly to one, have mean pairwise correlation $\approx-1/(K-1)$ — and the true
$L$ here measures $-0.44$.)

Two caveats before the numbers. $B>A$ requires that some row load on more than
one factor, which fails only for an exactly block-structured factorization. And
$\bar\rho$ constrains the factors *collectively*: it is a weighted average, so it
says nothing about any named pair.

```{r corr-verify, cache=TRUE}
cors <- do.call(rbind, lapply(list.files(cells_dir, "\\.rds$", full.names = TRUE),
  function(f) {
    x <- readRDS(f)
    if (is.data.frame(x)) return(NULL)        # the metrics file lives here too
    rt <- cor(x$FF_true)                      # the truth, for reference
    do.call(rbind, lapply(names(x$fits), function(init) {
      L <- x$fits[[init]]$LL; Fm <- x$fits[[init]]$FF; K <- ncol(L)
      th <- tcrossprod(L, Fm)
      Fc <- sweep(Fm, 2, colMeans(Fm))                # centered factors
      a  <- sweep(L, 2, sqrt(colSums(Fc^2)), "*")     # a_ik = l_ik ||f~_.k||
      A  <- sum(a^2); B <- sum(rowSums(a)^2)
      E  <- sum((th - rowMeans(th))^2)                # row-centered variation
      rp <- cor(Fm); rs <- cor(Fm, method = "spearman")
      data.frame(
        c_fit = as.numeric(sub(".*_cfit", "", sub("\\.rds$", "", basename(f)))),
        init = init, K = K, floor = -1/(K - 1),
        rho_bar = (E - A)/(B - A), slack = E/(B - A),
        pearson = mean(rp[upper.tri(rp)]), spearman = mean(rs[upper.tri(rs)]),
        truth = mean(rt[upper.tri(rt)]))
    }))
  }))

r1  <- cors[cors$init == "rank1", ]
agg <- do.call(rbind, lapply(sort(unique(r1$c_fit)), function(cc) {
  g <- r1[r1$c_fit == cc, ]
  data.frame(c_fit = cc, `rho-bar (identity)` = mean(g$rho_bar),
             `E/(B-A)` = mean(g$slack), `bound` = mean(g$floor + g$slack),
             `floor -1/(K-1)` = g$floor[1], `measured Pearson` = mean(g$pearson),
             `measured Spearman` = mean(g$spearman),
             `true factors` = mean(g$truth), check.names = FALSE)
}))
knitr::kable(agg, digits = 3, row.names = FALSE)

cat("identity below the bound:", sum(cors$rho_bar <= cors$floor + cors$slack + 1e-9),
    "  above the floor:", sum(cors$rho_bar >= cors$floor - 1e-9),
    "of", nrow(cors), "fits\n")
cat("max |identity - measured Pearson|:", round(max(abs(cors$rho_bar - cors$pearson)), 3), "\n")
```

The identity tracks the plainly-measured pairwise Pearson correlation to about
two decimals across the whole grid, so the quantity in the theory is essentially
the quantity in the figures — even though one is weighted by $a_{ik}a_{ik'}$ and
the other is not. The true factors sit at $\approx 0$, confirming that the
anti-correlation at small $c$ is manufactured by the budget rather than inherited
from the data.

```{r corr-plot, fig.width=9.5, fig.height=3.8, cache=TRUE}
lv   <- fmtc(sort(unique(cors$c_fit)))
long <- rbind(
  data.frame(cf = fmtc(agg$c_fit), what = "rho-bar (identity)", v = agg$`rho-bar (identity)`),
  data.frame(cf = fmtc(agg$c_fit), what = "measured Pearson",   v = agg$`measured Pearson`),
  data.frame(cf = fmtc(agg$c_fit), what = "measured Spearman",  v = agg$`measured Spearman`))
long$cf <- factor(long$cf, levels = lv)
stopifnot(!any(is.na(long$cf)))
qA <- ggplot(long, aes(cf, v, colour = what, group = what)) +
  geom_hline(yintercept = cors$floor[1], linetype = "dashed", colour = "#C2410C") +
  geom_line(linewidth = 0.9) + geom_point(size = 2) +
  annotate("text", x = 1, y = cors$floor[1], vjust = -0.6, hjust = 0, size = 3.1,
           colour = "#C2410C", label = "floor  -1/(K-1)") +
  scale_colour_manual(name = NULL,
    values = c(`rho-bar (identity)` = "#0E5C6B", `measured Pearson` = "#0E8FA8",
               `measured Spearman` = "#B45309")) +
  labs(x = "c used for fitting", y = "mean pairwise correlation of the factors",
       title = "Small c forces the factors apart") +
  theme(legend.text = element_text(size = 8))

line <- data.frame(x = 10^seq(-3, log10(max(cors$slack)), length.out = 300))
line$y <- cors$floor[1] + line$x
qB <- ggplot(cors, aes(slack, rho_bar)) +
  geom_line(data = line, aes(x, y), colour = "#C2410C", linewidth = 0.9) +
  geom_hline(yintercept = cors$floor[1], linetype = "dashed", colour = "grey45") +
  geom_point(colour = "#0E8FA8", alpha = 0.3, size = 0.9) +
  scale_x_log10() + coord_cartesian(ylim = c(-0.55, 0.15)) +
  annotate("text", x = 1.4e-3, y = cors$floor[1], vjust = -0.6, hjust = 0,
           size = 3.1, colour = "grey35", label = "floor  -1/(K-1)") +
  labs(x = "E / (B - A)   — the term c controls", y = "mean pairwise correlation",
       title = "Everything sits between the floor and the bound",
       subtitle = "line: -1/(K-1) + E/(B-A).   dashed: the floor it decays to")
plot_grid(qA, qB, nrow = 1)
```

**How much this is worth.** The *identity* is exact and informative everywhere:
it says the mean pairwise correlation of the factors is a deterministic function
of how flat $\theta$ is, which is a function of $c$. The *bound* is only
non-vacuous at $c\le 0.01$ here — the term $E/(B-A)$ exceeds $1/(K-1)$ as soon as
$\theta$ has any real range, and then $-1/(K-1)+E/(B-A)$ is above zero and says
nothing. So the floor explains the small-$c$ end and not the large-$c$ end.

The result is also about **Pearson** correlation. Spearman, which is what the
simulation page plots, agrees closely at small $c$ ($-0.35$ against $-0.34$) but
departs at large $c$ ($-0.20$ against $-0.03$), where ties and skew from the now
very sparse factors dominate the ranks. That gap is a sparsity artifact, not a
budget one.

Finally, this is about how separated the fitted factors are from *each other*,
which is a different question from whether they are *correct*. Accuracy against
the true factors peaks sharply at $c_{\text{fit}}=c_{\text{true}}$ and is a
misspecification story, not a budget story.

# What to do with this

- **Compute `sparsity_budget(Y, c)` before you fit.** It costs nothing and tells
  you whether sparse factors are even available at that $c$.
- **Do not compare sparsity across $c$.** "The $c=0.01$ factors are denser than
  the $c=10$ factors" is a statement about $\log(1+\cdot/c)$, not about the data:
  the budgets differ, so the comparison is meaningless.
- **Do compare quantities that depend on $LF'$** — fitted rates, held-out
  log-likelihood, which samples resemble which. Those do not depend on how the
  budget was split.
- **If sparsity matters scientifically, impose it.** The likelihood will not
  choose it for you; only a penalty, a prior, or an anchor assumption will.

A companion page,
[Non-identifiability of the log1p factorization](log1p_identifiability.html),
covers the other half: for a *fixed* budget, exactly which factorizations are
equally likely, and how sparsity moves between $L$ and $F$ along that constraint.

# Appendix: a two-sided bound on sparsity

*Nothing above depends on this section.* The upper bound is the result of the
page; this appendix states the full two-sided version and proves the half that
has not been proved yet. The lower half turns out to be valid but weak, so it is
best read as a statement about what is *possible*, not about what happens.

## The sandwich

Throughout, $\theta = LF'$ with $L, F\ge 0$ of rank $K$, and

$$
w_k \;=\; \frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}{\operatorname{mean}(\theta)},
\qquad \sum_k w_k = 1 ,
$$

as in Step B. Define, for a target peak-to-mean $P$ and a length $m$,

$$
H_{\min}(P,m) \;=\; \frac{\sqrt{m}-m\big/\sqrt{P^{2}+(m-1)q^{2}}}{\sqrt{m}-1},
\qquad q=\frac{m-P}{m-1},
\qquad\text{so}\qquad
H_{\min}(P,m)\approx 1-\frac{1}{\sqrt{1+P^{2}/m}} .
$$

> **Sandwich.** For any nonnegative factorization $\theta = LF'$,
>
> **(upper — proved in Steps A and B)** for *every* factor $k$ and *both* sides,
> $$
> H(\ell_{\cdot k}) \;\le\; 1-\sqrt{\frac{w_k}{\operatorname{pm}(\theta)}},
> \qquad
> H(f_{\cdot k}) \;\le\; 1-\sqrt{\frac{w_k}{\operatorname{pm}(\theta)}} ;
> $$
>
> **(lower — proved below)** for *at least one* factor $k$,
> $$
> \max\bigl\{H(\ell_{\cdot k}),\,H(f_{\cdot k})\bigr\}
> \;\ge\; H_{\min}\!\bigl(\sqrt{\operatorname{pm}(\theta)},\;\max(n,p)\bigr) .
> $$
>
> Both bounds are increasing in $\operatorname{pm}(\theta)$.

The two halves say different things, and the quantifiers matter. The upper half
constrains **every** column: nothing can be sparser than this. The lower half
constrains **some** column: at least one side of one factor must be at least
this sparse. Raising $c$ raises $\operatorname{pm}(\theta)$ and so lifts both
ends of the band, which is the formal counterpart of the rising lower envelope
seen in the simulations.

The upper half needs no new work: it is the one-column corollary proved at the
end of Step B, where the stronger hyperbola form
$d(\ell_{\cdot k})d(f_{\cdot k})\ge\sqrt{w_k/\operatorname{pm}(\theta)}$ and the
cost of weakening it to a single column are also given.

The rest of this appendix proves the lower half, in two lemmas.

## Lemma 1: the least sparse a vector can be at a given peak-to-mean

> For nonnegative $x$ of length $m$ with $\operatorname{pm}(x)=P$,
> $$H(x)\;\ge\;H_{\min}(P,m),$$
> with equality for $x^\star=(P,q,\dots,q)$, $q=(m-P)/(m-1)$.

**Proof.** $H$ is a decreasing function of $\|x\|_1/\|x\|_2$, so minimising $H$
means minimising $\|x\|_2$. Both $H$ and $\operatorname{pm}$ are scale
invariant, so normalise $\operatorname{mean}(x)=1$; then $\|x\|_1=m$ is fixed and
the constraint $\operatorname{pm}(x)=P$ says $\max(x)=P$.

One coordinate is therefore pinned at $P$, contributing $P^2$ to $\|x\|_2^2$
regardless. The remaining $m-1$ coordinates are nonnegative and must sum to
$m-P$, and among vectors with a fixed sum the squared norm
$\sum_{i\ge 2}x_i^2$ is minimised when all entries are equal — by Cauchy–Schwarz,
$\sum_{i\ge2}x_i^2\ge\bigl(\sum_{i\ge2}x_i\bigr)^2/(m-1)$, with equality exactly
when they are. That common value is $q=(m-P)/(m-1)$, which satisfies
$0\le q\le P$ for every $1\le P\le m$, so $x^\star$ is feasible and $P$ really is
its maximum. Substituting $\|x\|_1=m$ and
$\|x\|_2^2=P^2+(m-1)q^2$ into the definition of $H$ gives $H_{\min}(P,m)$. $\square$

Note the shape. Step A's upper bound involves $P$; this lower bound involves
$P^{2}/m$. A lone spike among many coordinates barely moves $\ell_1/\ell_2$, so
the lower bound only becomes interesting once $P$ approaches $\sqrt{m}$.

```{r A1-check}
Hmin <- function(P, m) { q <- (m - P)/(m - 1)
  (sqrt(m) - m/sqrt(P^2 + (m-1)*q^2)) / (sqrt(m) - 1) }

set.seed(1); m <- 400
knitr::kable(do.call(rbind, lapply(c(2, 10, 50, 200), function(P) {
  best <- Inf                       # random x with pm exactly P: mean 1, one entry P
  for (rep in 1:8000) {
    y <- rexp(m - 1)^runif(1, 0.2, 4); y <- y/sum(y)*(m - P)
    if (max(y) > P) next
    best <- min(best, hoyer(c(P, y)))
  }
  data.frame(P = P, `H_min(P, m)` = Hmin(P, m),
             `H of x-star` = hoyer(c(P, rep((m - P)/(m - 1), m - 1))),
             `best of 8000 random` = best, check.names = FALSE)
})), digits = 5, row.names = FALSE,
caption = "x-star attains the bound exactly; random vectors never beat it.")
```

## Lemma 2: the reverse of the budget inequality

> $\displaystyle \operatorname{pm}(\theta)\;\le\;\max_k\ \operatorname{pm}(\ell_{\cdot k})\cdot\operatorname{pm}(f_{\cdot k}).$

**Proof.** Step B bounded $\max(\theta)$ from *below* by discarding all but one
term. Bound it from *above* instead by taking the largest possible value of each
term: for every $i,j$,

$$
\theta_{ij}=\sum_m \ell_{im}f_{jm}\;\le\;\sum_m \max(\ell_{\cdot m})\max(f_{\cdot m}),
\qquad\text{so}\qquad
\max(\theta)\;\le\;\sum_m \max(\ell_{\cdot m})\max(f_{\cdot m}).
$$

The mean factorizes as before,
$\operatorname{mean}(\theta)=\sum_m\operatorname{mean}(\ell_{\cdot m})\operatorname{mean}(f_{\cdot m})$.
Put $a_m=\max(\ell_{\cdot m})\max(f_{\cdot m})$ and
$b_m=\operatorname{mean}(\ell_{\cdot m})\operatorname{mean}(f_{\cdot m})>0$, and
apply the mediant inequality $\bigl(\sum_m a_m\bigr)/\bigl(\sum_m b_m\bigr)\le\max_m (a_m/b_m)$:

$$
\operatorname{pm}(\theta)=\frac{\max(\theta)}{\operatorname{mean}(\theta)}
\;\le\;\frac{\sum_m a_m}{\sum_m b_m}
\;\le\;\max_m \frac{a_m}{b_m}
\;=\;\max_m\ \operatorname{pm}(\ell_{\cdot m})\operatorname{pm}(f_{\cdot m}). \qquad\square
$$

Together with Step B the product is now pinned on both sides:

$$
\operatorname{pm}(\theta)\;\le\;\max_k\ \operatorname{pm}(\ell_{\cdot k})\operatorname{pm}(f_{\cdot k})
\;\le\;\frac{\operatorname{pm}(\theta)}{\min_k w_k}.
$$

## Proof of the lower half

By Lemma 2 there is a factor $k$ with
$\operatorname{pm}(\ell_{\cdot k})\operatorname{pm}(f_{\cdot k})\ge\operatorname{pm}(\theta)$.
A product of two numbers cannot have both factors below the square root of the
product, so

$$
\max\bigl\{\operatorname{pm}(\ell_{\cdot k}),\ \operatorname{pm}(f_{\cdot k})\bigr\}
\;\ge\;\sqrt{\operatorname{pm}(\theta)} .
$$

Apply Lemma 1 to whichever column achieves that maximum. Its length is $n$ or
$p$, and $H_{\min}(P,m)$ is decreasing in $m$, so using $\max(n,p)$ is valid
either way:

$$
\max\bigl\{H(\ell_{\cdot k}),\,H(f_{\cdot k})\bigr\}\;\ge\;H_{\min}\!\bigl(\sqrt{\operatorname{pm}(\theta)},\ \max(n,p)\bigr).
\qquad\square
$$

## Both halves, checked on the fits

```{r A3-check, cache=TRUE}
knitr::kable(do.call(rbind, lapply(C_GRID, function(cf) {
  r <- do.call(rbind, lapply(1:10, function(sd) {
    x <- readRDS(file.path(cells_dir, paste0(
           tag_of(list(seed = sd, c_true = 1, c_fit = cf)), ".rds")))
    L <- x$fits$rank1$LL; Fm <- x$fits$rank1$FF
    th <- tcrossprod(L, Fm) / max(1, if (is.infinite(cf)) 1 else cf)
    w  <- colMeans(L) * colMeans(Fm); w <- w/sum(w)
    P  <- pm(th); mm <- max(nrow(L), nrow(Fm))
    HL <- apply(L, 2, hoyer); HF <- apply(Fm, 2, hoyer)
    ub <- 1 - sqrt(w/P)                                  # upper, every column
    k  <- which.max(apply(L, 2, pm) * apply(Fm, 2, pm))  # the factor Lemma 2 gives
    c(`pm(theta)`       = P,
      `upper bound`     = max(ub),
      `largest H seen`  = max(c(HL, HF)),
      `max H at that k` = max(HL[k], HF[k]),
      `lower bound`     = Hmin(sqrt(P), mm),
      `upper violations`= sum(HL > ub + 1e-9) + sum(HF > ub + 1e-9),
      `lower violations`= as.numeric(max(HL[k], HF[k]) < Hmin(sqrt(P), mm) - 1e-9))
  }))
  as.data.frame(t(c(c_fit = cf, colMeans(r))))
})), digits = c(3, 2, 3, 3, 3, 4, 0, 0), row.names = FALSE,
caption = "Columns are ordered as they appear in the sandwich: upper bound, what is observed, lower bound.")
```

## How useful is the lower half?

Not very, quantitatively. The guarantee moves from about $10^{-4}$ to $0.02$
across the grid while the fits deliver $0.18$ to $0.59$ — the right *shape*, off
by more than an order of magnitude. Lemma 1 explains why: the bound depends on
$P^{2}/m$, and here $\sqrt{\operatorname{pm}(\theta)}\le 4.4$ against
$m=300$, so it never comes near the regime where it bites.

The gap is structural, not loose algebra. Lemmas 1 and 2 hold for *every*
nonnegative factorization of $\theta$, and a maximum-likelihood fit is not
adversarial. Closing it would need an argument about what maximum likelihood
actually does — that the fit concentrates because concentrating fits the data,
not merely because the geometry permits it. That is a different theorem and this
page does not have it.

So the pairing worth remembering:

- the **upper** bound is tight and explanatory — the fits spend their entire
  budget, which is why it predicts their behaviour;
- the **lower** bound is valid and correctly monotone but loose — it establishes
  that the trend cannot be reversed, not that it must be as strong as observed.

```{r session}
sessionInfo()
```
