---
title: "Why the sparsity of a log1p fit is decided by Y and c"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r pkgs}
library(ggplot2); library(cowplot); library(log1pNMF)
theme_set(theme_cowplot(font_size = 11))
exp_dir   <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
cells_dir <- file.path(exp_dir, "c_grid_sim")
source(file.path(exp_dir, "c_grid_sim_common.R"))

hoyer <- function(x) { x <- abs(x); n <- length(x)
  (sqrt(n) - sum(x)/sqrt(sum(x^2))) / (sqrt(n) - 1) }
p2m <- function(x) max(x) / mean(x)          # peak-to-mean ratio
```

# The question

You have a count matrix $Y$. You fit the log1p model at several values of $c$.
The factors come out visibly more or less sparse depending on which $c$ you used,
even though nothing about the data changed. What decides that?

This page answers exactly that, and nothing else. It is about **Hoyer** sparsity
— a continuous measure that does not require any exact zeros — expressed purely
in terms of $Y$ and $c$.

**The answer in one sentence.** The Hoyer sparsity that $L$ and $F$ can achieve
is jointly capped by the *peak-to-mean ratio* of $\theta = \log(1+\lambda/c)$;
the rates $\lambda$ are fixed by $Y$; and $c$ alone slides that ratio between
$1$ and $\lambda$'s own peak-to-mean.

The rest of the page is that sentence in four steps.

# Step A: Hoyer sparsity is peak-to-mean in disguise

For a nonnegative vector $x$ of length $n$,

$$
H(x) = \frac{\sqrt{n} - \|x\|_1/\|x\|_2}{\sqrt{n}-1},
\qquad\text{and write}\qquad
P(x) = \frac{x_{\max}}{\bar{x}} .
$$

$P$ asks a simple question: how big is the largest entry compared with a typical
one? Now use $\|x\|_2^2 = \sum x_i^2 \le x_{\max}\sum x_i = \|x\|_\infty\|x\|_1$:

$$
\frac{\|x\|_1}{\|x\|_2} \;\ge\; \frac{\|x\|_1}{\sqrt{\|x\|_\infty\|x\|_1}}
\;=\; \sqrt{\frac{\|x\|_1}{\|x\|_\infty}} \;=\; \sqrt{\frac{n}{P}},
$$

and therefore

$$
\boxed{\;H(x) \;\le\; \frac{\sqrt{n}-\sqrt{n/P}}{\sqrt{n}-1} \;\approx\; 1 - \frac{1}{\sqrt{P}}\;}
$$

This is the whole of Step A, and it is worth reading in words: **a vector can only
be Hoyer-sparse if some of its entries stick far above its mean.** If $P = 1$
every entry equals the mean, the vector is constant, and $H = 0$ — no matter how
many entries there are and regardless of whether any are zero. Sparsity requires
*range*, not zeros.

```{r stepA, fig.width=7, fig.height=3.4}
set.seed(1); n <- 500
sim <- do.call(rbind, lapply(seq(0, 1, length.out = 60), function(a) {
  do.call(rbind, lapply(1:8, function(r) {
    x <- (1 - a) + a * rexp(n)^2          # dial from constant to heavy-tailed
    data.frame(P = p2m(x), H = hoyer(x)) }) ) }))
curve <- data.frame(P = seq(1, max(sim$P), length.out = 300))
curve$bound <- 1 - 1/sqrt(curve$P)

ggplot(sim, aes(P, H)) +
  geom_point(colour = "#0E8FA8", alpha = 0.5, size = 1.2) +
  geom_line(data = curve, aes(P, bound), colour = "#C2410C", linewidth = 1) +
  scale_x_log10() +
  labs(x = "peak-to-mean  P = max / mean", y = "Hoyer sparsity",
       title = "Sparsity is bounded by peak-to-mean",
       subtitle = "points: random nonnegative vectors.  line: 1 - 1/sqrt(P)")
```

# Step B: L and F share a single budget

The fit gives $\theta = LF'$ with $L, F \ge 0$. Fix a factor $k$, and let
$i^\star$ be the sample with the largest loading on it and $j^\star$ the feature
with the largest weight on it. Because **every term in the sum is nonnegative**
we may throw all the others away:

$$
\theta_{\max} \;\ge\; \theta_{i^\star j^\star}
= \sum_{m}\ell_{i^\star m}f_{j^\star m}
\;\ge\; \ell^{\max}_k\, f^{\max}_k .
$$

The mean is $\bar\theta = \sum_m \bar\ell_m\bar f_m$. Let
$w_k = \bar\ell_k\bar f_k/\bar\theta \in (0,1)$ be the share of $\bar\theta$
contributed by factor $k$. Dividing the display by $\bar\theta$ and inserting
$\bar\ell_k\bar f_k$ top and bottom,

$$
\boxed{\;P_L(k)\;\cdot\;P_F(k)\;\cdot\;w_k \;\;\le\;\; P_\theta\;}
\qquad
P_\theta=\frac{\theta_{\max}}{\bar\theta},\quad
P_L(k)=\frac{\ell^{\max}_k}{\bar\ell_k},\quad
P_F(k)=\frac{f^{\max}_k}{\bar f_k}.
$$

Read it as a **budget**. $P_\theta$ is how much peak-to-mean the fitted $\theta$
has to spend. The two factors' peakedness multiply against it, so:

- they **trade**: at fixed $P_\theta$, making $L$ more peaked forces $F$ flatter;
- they are **jointly capped**: neither can be peaked at all if $P_\theta$ is near
  $1$, because the product cannot exceed it.

Combining with Step A, if $\theta$ has no range then neither factor can be
Hoyer-sparse, and no choice of optimizer or initialization can change that.

This is also why the two things you may have noticed are not in conflict. Holding
the data and $c$ fixed and moving between equally-good factorizations slides
along the budget line — a trade. Changing $c$ moves the budget line itself — so
both can rise together.

# Step C: c is what sets the budget

Here is the crux. $\theta$ is not free: it is the rates seen through the link,

$$
\theta = \log\!\left(1+\frac{\lambda}{c}\right).
$$

Take five rates and never change them. Look at what $c$ does to their
peak-to-mean:

```{r stepC-table}
lam <- c(0.1, 0.5, 1, 5, 20)
cs  <- c(1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3, Inf)
knitr::kable(do.call(rbind, lapply(cs, function(cc) {
  th <- if (is.infinite(cc)) lam else log1p(lam/cc)
  data.frame(c = format(cc),
             theta = paste(sprintf("%.3g", th), collapse = "  "),
             P_theta = p2m(th), `sparsity cap` = 1 - 1/sqrt(p2m(th)),
             check.names = FALSE) })),
  digits = 3, row.names = FALSE,
  caption = "The same five rates throughout. Only c changes.")
```

The rates never move. Their own peak-to-mean is `r round(p2m(lam), 2)`. But the
budget the fit actually gets runs from about `r round(p2m(log1p(lam/1e-3)), 2)`
at $c=10^{-3}$ up to `r round(p2m(lam), 2)` at $c=\infty$, and with it the
sparsity cap roughly triples.

**Why.** For $c$ much smaller than the rates,
$\log(1+\lambda/c)\approx \log(1/c) + \log\lambda$: every entry acquires the same
large constant $\log(1/c)$, and a shared additive constant is exactly what
destroys peak-to-mean. For $c$ much larger than the rates,
$\log(1+\lambda/c)\approx\lambda/c$, a pure rescaling, which preserves
peak-to-mean exactly. Formally

$$
P_\theta(c)\;\xrightarrow[c\to 0]{}\;1,
\qquad
P_\theta(c) - 1 \;\approx\; \frac{\log\lambda_{\max}-\overline{\log\lambda}}{\log(1/c)},
\qquad
P_\theta(c)\;\xrightarrow[c\to\infty]{}\;\frac{\lambda_{\max}}{\bar\lambda}.
$$

```{r stepC-plot, fig.width=6.5, fig.height=3.6}
grid <- 10^seq(-4, 4, length.out = 200)
d <- data.frame(c = grid,
                P = sapply(grid, function(cc) p2m(log1p(lam/cc))))
ggplot(d, aes(c, P)) +
  geom_hline(yintercept = p2m(lam), linetype = "dashed", colour = "#C2410C") +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey50") +
  geom_line(colour = "#0E5C6B", linewidth = 1) +
  scale_x_log10() +
  annotate("text", x = 1e3, y = p2m(lam), vjust = -0.7, size = 3.2,
           colour = "#C2410C", label = "max/mean of the rates") +
  annotate("text", x = 1e-4, y = 1, vjust = -0.7, hjust = 0, size = 3.2,
           colour = "grey35", label = "no range left") +
  labs(x = "c", y = expression(P[theta]),
       title = "The sparsity budget, as a function of c",
       subtitle = "same rates throughout; only the link changes")
```

# Step D: the rates come from Y, not from c

The last link is why we may treat $\lambda$ as given. The Poisson
log-likelihood is

$$
\ell(\lambda) = \sum_{ij}\bigl[y_{ij}\log\lambda_{ij}-\lambda_{ij}\bigr] - \sum_{ij}\log y_{ij}!,
$$

which contains **no $c$ anywhere**. $c$ appears only in the feasible set — which
rate matrices are reachable, namely those with $\log(1+\lambda/c)$ of nonnegative
rank $K$. So $c$ does not change what a good set of rates is; the fitted rates
are governed by $Y$.

That means $\lambda$ can be estimated from $Y$ before fitting anything. The
crudest estimate suffices for this purpose — the rank-1 independence model

$$
m_{ij} = \frac{y_{i\cdot}\,y_{\cdot j}}{y_{\cdot\cdot}} .
$$

# The recipe

Given $Y$ and a candidate $c$, with no model fitted:

1. $m_{ij} = y_{i\cdot}y_{\cdot j}/y_{\cdot\cdot}$
2. $P_\theta(c) = \max_{ij}\log(1+m_{ij}/c) \big/ \overline{\log(1+m/c)}$
3. sparsity cap $\approx 1 - 1/\sqrt{P_\theta(c)}$, shared between $L$ and $F$

If $P_\theta(c)$ is near 1, sparse factors are unobtainable at that $c$ and any
sparsity you do see is an artifact of the optimizer's path.

```{r recipe}
sparsity_budget <- function(Y, c) {
  m  <- outer(rowSums(Y), colSums(Y)) / sum(Y)
  th <- if (is.infinite(c)) m else log1p(m / c)
  P  <- max(th) / mean(th)
  c(P_theta = P, cap = 1 - 1/sqrt(P))
}
Y <- sim_dataset(1, 1)$Y                       # any count matrix
round(t(sapply(C_GRID, function(cc) c(c = cc, sparsity_budget(Y, cc)))), 3)
```

# Does it hold on real fits?

Against the 640 fits from the [c-grid simulation](c_grid_sim.html): the budget
inequality of Step B should never be violated, and if the argument is right it
should be roughly *tight* — the optimizer should spend what it is given.

```{r verify, cache=TRUE}
ver <- do.call(rbind, lapply(C_GRID, function(cf) {
  r <- do.call(rbind, lapply(1:10, function(sd) {
    Yd <- sim_dataset(sd, 1)$Y
    x  <- readRDS(file.path(cells_dir, paste0(
            tag_of(list(seed = sd, c_true = 1, c_fit = cf)), ".rds")))
    L <- x$fits$rank1$LL; Fm <- x$fits$rank1$FF
    th <- tcrossprod(L, Fm) / max(1, if (is.infinite(cf)) 1 else cf)
    w  <- colMeans(L) * colMeans(Fm); w <- w / sum(w)
    PL <- apply(L, 2, p2m); PF <- apply(Fm, 2, p2m)
    c(`predicted P (from Y,c)` = unname(sparsity_budget(Yd, cf)["P_theta"]),
      `actual P of theta-hat`  = p2m(th),
      `P_L P_F w`              = max(PL * PF * w),
      `Hoyer L`                = mean(apply(L,  2, hoyer)),
      `Hoyer F`                = mean(apply(Fm, 2, hoyer)),
      violations               = sum(PL * PF * w > p2m(th) * (1 + 1e-8)))
  }))
  as.data.frame(t(c(c_fit = cf, colMeans(r))))
}))
knitr::kable(ver, digits = c(3, 2, 2, 2, 3, 3, 0), row.names = FALSE)
```

Three things to check in that table.

**The inequality is never violated** — as it must not be, being an identity plus
nonnegativity.

**It is tight.** `P_L P_F w` sits just under `actual P of theta-hat` in every
row. The fits use their whole budget, which is what makes the bound explanatory
rather than merely true: at small $c$ the optimizer is not failing to find sparse
factors, there are none to find.

**The budget predicted from $Y$ and $c$ alone tracks the realized one**, so the
whole chain can be evaluated before fitting.

```{r verify-plot, fig.width=8, fig.height=3.6}
lab <- ifelse(is.infinite(ver$c_fit), "∞",
              format(ver$c_fit, scientific = FALSE, trim = TRUE, drop0trailing = TRUE))
ver$cf <- factor(lab, levels = lab)
long <- rbind(
  data.frame(cf = ver$cf, what = "budget P (theta)", v = ver$`actual P of theta-hat`),
  data.frame(cf = ver$cf, what = "spent  P_L P_F w", v = ver$`P_L P_F w`))
p1 <- ggplot(long, aes(cf, v, colour = what, group = what)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = c("#0E5C6B", "#C2410C"), name = NULL) +
  labs(x = "c used for fitting", y = "peak-to-mean",
       title = "The budget, and what the fit spends") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8))
p2 <- ggplot(ver, aes(`actual P of theta-hat`)) +
  geom_line(aes(y = `Hoyer L`, colour = "L"), linewidth = 1) +
  geom_line(aes(y = `Hoyer F`, colour = "F"), linewidth = 1) +
  geom_line(aes(y = 1 - 1/sqrt(`actual P of theta-hat`), colour = "cap"),
            linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = c(L = "#0E8FA8", F = "#C2410C", cap = "grey40"),
                      name = NULL) +
  scale_x_log10() +
  labs(x = expression(P[theta]), y = "mean Hoyer sparsity",
       title = "Both factors rise with the budget")
plot_grid(p1, p2, nrow = 1)
```

# What to do with this

- **Compute `sparsity_budget(Y, c)` before you fit.** It costs nothing and tells
  you whether sparse factors are even available at that $c$.
- **Do not compare sparsity across $c$.** "The $c=0.01$ factors are denser than
  the $c=10$ factors" is a statement about $\log(1+\cdot/c)$, not about the data.
  The budgets differ, so the comparison is meaningless.
- **Do compare things that depend on $LF'$** — fitted rates, held-out
  log-likelihood, which samples resemble which. Those do not depend on how the
  budget was split.
- **If sparsity is scientifically important, impose it.** The likelihood will not
  choose it for you; only a penalty, a prior, or an anchor assumption will.

A companion page,
[Non-identifiability of the log1p factorization](log1p_identifiability.html),
covers the other half of the story: for a *fixed* budget, exactly which
factorizations are equally likely and how sparsity moves between $L$ and $F$
along that constraint.

```{r session}
sessionInfo()
```
