---
title: "Why the sparsity of a log1p fit is decided by Y and c"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r pkgs}
library(ggplot2); library(cowplot); library(log1pNMF)
theme_set(theme_cowplot(font_size = 11))
exp_dir   <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
cells_dir <- file.path(exp_dir, "c_grid_sim")
source(file.path(exp_dir, "c_grid_sim_common.R"))

## Hoyer sparsity of a nonnegative vector: 0 if all entries are equal, 1 if one
## entry carries everything.
hoyer <- function(x) { x <- abs(x); n <- length(x)
  (sqrt(n) - sum(x)/sqrt(sum(x^2))) / (sqrt(n) - 1) }

## pm(): peak-to-mean. The ONLY summary used on this page. It takes a vector or
## a matrix and returns one number, so pm(L[, k]) is a column's peak-to-mean and
## pm(theta) is the whole matrix's.
pm <- function(x) max(x) / mean(x)
```

# The question

You have a count matrix $Y$. You fit the log1p model at several values of $c$.
The factors come out visibly more or less sparse depending on which $c$ you used,
even though nothing about the data changed. What decides that?

This page answers exactly that, and nothing else. It is about **Hoyer** sparsity
— a continuous measure that requires no exact zeros — expressed purely in terms
of $Y$ and $c$.

**The answer in one sentence.** The Hoyer sparsity that $L$ and $F$ can achieve
is jointly capped by the *peak-to-mean ratio* of $\theta = \log(1+\lambda/c)$;
the rates $\lambda$ are fixed by $Y$; and $c$ alone slides that ratio between
$1$ and $\lambda$'s own peak-to-mean.

The rest of the page is that sentence in four steps.

# Notation

Only one summary statistic appears on this page, applied to different things.
For any nonnegative vector or matrix $v$ with at least one positive entry,

$$
\operatorname{pm}(v) \;=\; \frac{\max(v)}{\operatorname{mean}(v)} \;\ge\; 1 ,
$$

the **peak-to-mean ratio**: how large the biggest entry is relative to a typical
one. $\operatorname{pm}(v) = 1$ exactly when $v$ is constant.

The objects it gets applied to:

| symbol | shape | meaning |
|---|---|---|
| $Y$ | $n \times p$ | the observed counts |
| $\lambda$ | $n \times p$ | the rates, $\mathbb{E}[Y]$ |
| $\theta = \log(1+\lambda/c)$ | $n \times p$ | the linear predictor |
| $L$, $F$ | $n\times K$, $p\times K$ | the factorization, $\theta = LF'$ |
| $\ell_{\cdot k}$ | length $n$ | column $k$ of $L$ — every sample's loading on factor $k$ |
| $f_{\cdot k}$ | length $p$ | column $k$ of $F$ — every feature's weight on factor $k$ |
| $\ell_{i\cdot}$ | length $K$ | row $i$ of $L$ — sample $i$'s loadings across factors |

So $\max(\ell_{\cdot k}) = \max_i \ell_{ik}$ is the largest loading any sample
places on factor $k$, and $\operatorname{pm}(\ell_{\cdot k})$ is that divided by
the average loading on factor $k$. Hoyer sparsity is always computed per column
and then averaged over the $K$ columns, matching the earlier pages.

# Step A: Hoyer sparsity is peak-to-mean in disguise

For a nonnegative vector $x$ of length $n$, Hoyer sparsity is

$$
H(x) = \frac{\sqrt{n} - \|x\|_1/\|x\|_2}{\sqrt{n}-1} .
$$

Use $\|x\|_2^2 = \sum_i x_i^2 \le \max(x)\sum_i x_i = \|x\|_\infty\|x\|_1$, and
note $\|x\|_1/\|x\|_\infty = n\,\operatorname{mean}(x)/\max(x) = n/\operatorname{pm}(x)$:

$$
\frac{\|x\|_1}{\|x\|_2} \;\ge\; \frac{\|x\|_1}{\sqrt{\|x\|_\infty\|x\|_1}}
\;=\; \sqrt{\frac{\|x\|_1}{\|x\|_\infty}} \;=\; \sqrt{\frac{n}{\operatorname{pm}(x)}},
$$

and therefore

$$
\boxed{\;H(x) \;\le\; \frac{\sqrt{n}-\sqrt{n/\operatorname{pm}(x)}}{\sqrt{n}-1}
\;\approx\; 1 - \frac{1}{\sqrt{\operatorname{pm}(x)}}\;}
$$

In words: **a vector can only be Hoyer-sparse if some of its entries stick far
above its mean.** If $\operatorname{pm}(x) = 1$ every entry equals the mean, the
vector is constant, and $H(x) = 0$ — however many entries it has, and whether or
not any of them are zero. Sparsity needs *range*, not zeros.

```{r stepA, fig.width=7, fig.height=3.4}
set.seed(1); n <- 500
sim <- do.call(rbind, lapply(seq(0, 1, length.out = 60), function(a) {
  do.call(rbind, lapply(1:8, function(r) {
    x <- (1 - a) + a * rexp(n)^2          # dial from constant to heavy-tailed
    data.frame(pm = pm(x), H = hoyer(x)) }) ) }))
curve <- data.frame(pm = seq(1, max(sim$pm), length.out = 300))
curve$bound <- 1 - 1/sqrt(curve$pm)

ggplot(sim, aes(pm, H)) +
  geom_point(colour = "#0E8FA8", alpha = 0.5, size = 1.2) +
  geom_line(data = curve, aes(pm, bound), colour = "#C2410C", linewidth = 1) +
  scale_x_log10() +
  labs(x = "pm(x)  =  max / mean", y = "Hoyer sparsity  H(x)",
       title = "Sparsity is bounded by peak-to-mean",
       subtitle = "points: random nonnegative vectors.   line: 1 - 1/sqrt(pm)")
```

# Step B: L and F share a single budget

The fit gives $\theta = LF'$ with $L, F \ge 0$. Fix one factor $k$. Let
$i^\star$ be the sample achieving $\max(\ell_{\cdot k})$ and $j^\star$ the
feature achieving $\max(f_{\cdot k})$. Because **every term of the sum is
nonnegative**, we may discard all but the $k$th:

$$
\max(\theta) \;\ge\; \theta_{i^\star j^\star}
\;=\; \sum_{m=1}^{K}\ell_{i^\star m}\,f_{j^\star m}
\;\ge\; \ell_{i^\star k}\,f_{j^\star k}
\;=\; \max(\ell_{\cdot k})\cdot\max(f_{\cdot k}) .
$$

For the mean, the double average factorizes across the sum over factors:

$$
\operatorname{mean}(\theta)
= \frac{1}{np}\sum_{i,j}\sum_{m}\ell_{im}f_{jm}
= \sum_{m}\Bigl(\frac{1}{n}\sum_i \ell_{im}\Bigr)\Bigl(\frac{1}{p}\sum_j f_{jm}\Bigr)
= \sum_{m}\operatorname{mean}(\ell_{\cdot m})\operatorname{mean}(f_{\cdot m}).
$$

Let

$$
w_k \;=\; \frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}{\operatorname{mean}(\theta)} \;\in\;(0,1)
$$

be the share of $\operatorname{mean}(\theta)$ that factor $k$ contributes; by the
display above the $w_k$ sum to one.

Now take the inequality $\max(\theta)\ge\max(\ell_{\cdot k})\max(f_{\cdot k})$,
divide both sides by $\operatorname{mean}(\theta)$, and then multiply the
right-hand side by
$\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})$ over
itself — which is just multiplying by $1$ — and regroup the four factors:

$$
\begin{aligned}
\operatorname{pm}(\theta)
\;=\;\frac{\max(\theta)}{\operatorname{mean}(\theta)}
&\;\ge\; \frac{\max(\ell_{\cdot k})\,\max(f_{\cdot k})}{\operatorname{mean}(\theta)}
&&\text{(divide by }\operatorname{mean}(\theta)\text{)}\\[4pt]
&\;=\; \frac{\max(\ell_{\cdot k})\,\max(f_{\cdot k})}{\operatorname{mean}(\theta)}
\cdot\underbrace{\frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}
{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}}_{=\,1}
&&\text{(insert }1\text{)}\\[4pt]
&\;=\; \underbrace{\frac{\max(\ell_{\cdot k})}{\operatorname{mean}(\ell_{\cdot k})}}_{\operatorname{pm}(\ell_{\cdot k})}
\cdot\underbrace{\frac{\max(f_{\cdot k})}{\operatorname{mean}(f_{\cdot k})}}_{\operatorname{pm}(f_{\cdot k})}
\cdot\underbrace{\frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}{\operatorname{mean}(\theta)}}_{w_k}
&&\text{(regroup)} .
\end{aligned}
$$

The middle step is the only trick, and it is not much of one: the two
$\operatorname{mean}$ terms are introduced solely so that each $\max$ can be
paired with the matching $\operatorname{mean}$ to form a $\operatorname{pm}$,
with whatever is left over collecting into $w_k$. Reading the chain end to end,

$$
\boxed{\;\operatorname{pm}(\ell_{\cdot k})\;\cdot\;\operatorname{pm}(f_{\cdot k})\;\cdot\;w_k
\;\;\le\;\; \operatorname{pm}(\theta)\;}
\qquad \text{for every factor } k .
$$

Read $\operatorname{pm}(\theta)$ as a **budget**: how much peak-to-mean the
fitted $\theta$ has available. The two factors' peak-to-mean ratios enter
*multiplicatively* against it, so

- they **trade**: at fixed $\operatorname{pm}(\theta)$, making $\ell_{\cdot k}$
  more peaked forces $f_{\cdot k}$ flatter;
- they are **jointly capped**: if $\operatorname{pm}(\theta)$ is near $1$ then
  neither can be peaked, since their product cannot exceed it.

Combined with Step A: if $\theta$ has no range, neither factor can be
Hoyer-sparse, and no optimizer or initialization can change that.

This also reconciles the two behaviours you may have noticed. Holding the data
and $c$ fixed and moving among equally-good factorizations slides *along* the
budget constraint — a trade. Changing $c$ moves the constraint itself — so both
can rise together.

# Step C: c is what sets the budget

Here is the crux. $\theta$ is not free; it is the rates seen through the link,
$\theta = \log(1+\lambda/c)$. Take five rates and never change them, then look at
what $c$ does to their peak-to-mean:

```{r stepC-table}
lam <- c(0.1, 0.5, 1, 5, 20)
cs  <- c(1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3, Inf)
knitr::kable(do.call(rbind, lapply(cs, function(cc) {
  th <- if (is.infinite(cc)) lam else log1p(lam/cc)
  data.frame(c = format(cc),
             `theta = log1p(lambda/c)` = paste(sprintf("%.3g", th), collapse = "  "),
             `pm(theta)` = pm(th),
             `sparsity cap 1-1/sqrt(pm)` = 1 - 1/sqrt(pm(th)),
             check.names = FALSE) })),
  digits = 3, row.names = FALSE,
  caption = "The same five rates in every row. Only c changes.")
```

The rates never move; their own peak-to-mean is
$\operatorname{pm}(\lambda) = `r round(pm(lam), 2)`$. But the budget the fit
actually receives runs from about `r round(pm(log1p(lam/1e-3)), 2)` at
$c=10^{-3}$ up to `r round(pm(lam), 2)` at $c=\infty$, and the sparsity cap
roughly triples with it.

**Why.** For $c$ well below the rates,
$\log(1+\lambda/c)\approx \log(1/c) + \log\lambda$: every entry picks up the
*same* large additive constant $\log(1/c)$, and a shared additive constant is
exactly what destroys peak-to-mean. For $c$ well above the rates,
$\log(1+\lambda/c)\approx\lambda/c$, a pure rescaling, which leaves peak-to-mean
untouched. Formally, writing $\theta_c = \log(1+\lambda/c)$,

$$
\operatorname{pm}(\theta_c)\;\xrightarrow[c\to 0]{}\;1,
\qquad
\operatorname{pm}(\theta_c) - 1 \;\approx\; \frac{\log\max(\lambda)-\operatorname{mean}(\log\lambda)}{\log(1/c)},
\qquad
\operatorname{pm}(\theta_c)\;\xrightarrow[c\to\infty]{}\;\operatorname{pm}(\lambda).
$$

```{r stepC-plot, fig.width=6.5, fig.height=3.6}
grid <- 10^seq(-4, 4, length.out = 200)
d <- data.frame(c = grid, pm = sapply(grid, function(cc) pm(log1p(lam/cc))))
ggplot(d, aes(c, pm)) +
  geom_hline(yintercept = pm(lam), linetype = "dashed", colour = "#C2410C") +
  geom_hline(yintercept = 1, linetype = "dotted", colour = "grey50") +
  geom_line(colour = "#0E5C6B", linewidth = 1) +
  scale_x_log10() +
  annotate("text", x = 1e3, y = pm(lam), vjust = -0.7, size = 3.2,
           colour = "#C2410C", label = "pm(lambda)") +
  annotate("text", x = 1e-4, y = 1, vjust = -0.7, hjust = 0, size = 3.2,
           colour = "grey35", label = "no range left") +
  labs(x = "c", y = "pm(theta)",
       title = "The sparsity budget, as a function of c",
       subtitle = "same rates throughout; only the link changes")
```

# Step D: the rates come from Y, not from c

The last link is what lets us treat $\lambda$ as given. The Poisson
log-likelihood is

$$
\ell(\lambda) = \sum_{ij}\bigl[y_{ij}\log\lambda_{ij}-\lambda_{ij}\bigr] - \sum_{ij}\log y_{ij}!,
$$

which contains **no $c$ anywhere**. $c$ appears only in the feasible set — which
rate matrices are reachable, namely those for which $\log(1+\lambda/c)$ has
nonnegative rank $K$. So $c$ does not change what a good set of rates is; the
fitted rates are governed by $Y$.

That means $\lambda$ can be estimated from $Y$ before fitting anything, and for
this purpose the crudest estimate is enough — the rank-1 independence model

$$
m_{ij} = \frac{y_{i\cdot}\,y_{\cdot j}}{y_{\cdot\cdot}} .
$$

# The recipe

Given $Y$ and a candidate $c$, with no model fitted:

1. $m_{ij} = y_{i\cdot}y_{\cdot j}/y_{\cdot\cdot}$
2. compute $\operatorname{pm}\bigl(\log(1+m/c)\bigr)$
3. the joint sparsity cap is $\approx 1 - 1/\sqrt{\operatorname{pm}}$, shared
   between $L$ and $F$

If $\operatorname{pm}$ comes out near 1, sparse factors are unavailable at that
$c$, and whatever sparsity a fit reports is an artifact of the optimizer's path.

```{r recipe}
sparsity_budget <- function(Y, c) {
  m  <- outer(rowSums(Y), colSums(Y)) / sum(Y)   # rank-1 estimate of the rates
  th <- if (is.infinite(c)) m else log1p(m / c)
  c(pm_theta = pm(th), cap = 1 - 1/sqrt(pm(th)))
}
Y <- sim_dataset(1, 1)$Y                          # any count matrix
round(t(sapply(C_GRID, function(cc) c(c = cc, sparsity_budget(Y, cc)))), 3)
```

# Does it hold on real fits?

Against the 640 fits from the [c-grid simulation](c_grid_sim.html). The Step B
inequality should never be violated, and if the argument is explanatory rather
than merely true it should be roughly *tight* — the optimizer should spend what
it is given.

```{r verify, cache=TRUE}
ver <- do.call(rbind, lapply(C_GRID, function(cf) {
  r <- do.call(rbind, lapply(1:10, function(sd) {
    Yd <- sim_dataset(sd, 1)$Y
    x  <- readRDS(file.path(cells_dir, paste0(
            tag_of(list(seed = sd, c_true = 1, c_fit = cf)), ".rds")))
    L <- x$fits$rank1$LL; Fm <- x$fits$rank1$FF
    th <- tcrossprod(L, Fm) / max(1, if (is.infinite(cf)) 1 else cf)
    w  <- colMeans(L) * colMeans(Fm); w <- w / sum(w)
    pmL <- apply(L,  2, pm)          # pm(l_.k) for each factor k
    pmF <- apply(Fm, 2, pm)          # pm(f_.k) for each factor k
    c(`pm predicted from Y,c`  = unname(sparsity_budget(Yd, cf)["pm_theta"]),
      `pm(theta) of the fit`   = pm(th),
      `pm(l)pm(f)w, worst k`   = max(pmL * pmF * w),
      `mean H(l_.k)`           = mean(apply(L,  2, hoyer)),
      `mean H(f_.k)`           = mean(apply(Fm, 2, hoyer)),
      violations               = sum(pmL * pmF * w > pm(th) * (1 + 1e-8)))
  }))
  as.data.frame(t(c(c_fit = cf, colMeans(r))))
}))
knitr::kable(ver, digits = c(3, 2, 2, 2, 3, 3, 0), row.names = FALSE)
```

Three things to check in that table.

**The inequality is never violated**, as it must not be, being an identity plus
nonnegativity.

**It is tight.** The `pm(l)pm(f)w` column sits just under `pm(theta) of the fit`
in every row. The fits use their whole budget, which is what makes the bound
explanatory: at small $c$ the optimizer is not failing to find sparse factors,
there are none to find.

**The budget predicted from $Y$ and $c$ alone tracks the realized one**, so the
whole chain can be evaluated before fitting.

```{r verify-plot, fig.width=8, fig.height=3.6}
lab <- ifelse(is.infinite(ver$c_fit), "∞",
              format(ver$c_fit, scientific = FALSE, trim = TRUE, drop0trailing = TRUE))
ver$cf <- factor(lab, levels = lab)
long <- rbind(
  data.frame(cf = ver$cf, what = "budget:  pm(theta)",   v = ver$`pm(theta) of the fit`),
  data.frame(cf = ver$cf, what = "spent:   pm(l)pm(f)w", v = ver$`pm(l)pm(f)w, worst k`))
p1 <- ggplot(long, aes(cf, v, colour = what, group = what)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_colour_manual(values = c("#0E5C6B", "#C2410C"), name = NULL) +
  labs(x = "c used for fitting", y = "peak-to-mean",
       title = "The budget, and what the fit spends") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
        legend.text = element_text(size = 8))
p2 <- ggplot(ver, aes(`pm(theta) of the fit`)) +
  geom_line(aes(y = `mean H(l_.k)`, colour = "H(l)"), linewidth = 1) +
  geom_line(aes(y = `mean H(f_.k)`, colour = "H(f)"), linewidth = 1) +
  geom_line(aes(y = 1 - 1/sqrt(`pm(theta) of the fit`), colour = "cap"),
            linetype = "dashed", linewidth = 0.8) +
  scale_colour_manual(values = c(`H(l)` = "#0E8FA8", `H(f)` = "#C2410C",
                                 cap = "grey40"), name = NULL) +
  scale_x_log10() +
  labs(x = "pm(theta)", y = "mean Hoyer sparsity",
       title = "Both factors rise with the budget")
plot_grid(p1, p2, nrow = 1)
```

# What to do with this

- **Compute `sparsity_budget(Y, c)` before you fit.** It costs nothing and tells
  you whether sparse factors are even available at that $c$.
- **Do not compare sparsity across $c$.** "The $c=0.01$ factors are denser than
  the $c=10$ factors" is a statement about $\log(1+\cdot/c)$, not about the data:
  the budgets differ, so the comparison is meaningless.
- **Do compare quantities that depend on $LF'$** — fitted rates, held-out
  log-likelihood, which samples resemble which. Those do not depend on how the
  budget was split.
- **If sparsity matters scientifically, impose it.** The likelihood will not
  choose it for you; only a penalty, a prior, or an anchor assumption will.

A companion page,
[Non-identifiability of the log1p factorization](log1p_identifiability.html),
covers the other half: for a *fixed* budget, exactly which factorizations are
equally likely, and how sparsity moves between $L$ and $F$ along that constraint.

# Appendix: a two-sided bound on sparsity

*Nothing above depends on this section.* The upper bound is the result of the
page; this appendix states the full two-sided version and proves the half that
has not been proved yet. The lower half turns out to be valid but weak, so it is
best read as a statement about what is *possible*, not about what happens.

## The sandwich

Throughout, $\theta = LF'$ with $L, F\ge 0$ of rank $K$, and

$$
w_k \;=\; \frac{\operatorname{mean}(\ell_{\cdot k})\operatorname{mean}(f_{\cdot k})}{\operatorname{mean}(\theta)},
\qquad \sum_k w_k = 1 ,
$$

as in Step B. Define, for a target peak-to-mean $P$ and a length $m$,

$$
H_{\min}(P,m) \;=\; \frac{\sqrt{m}-m\big/\sqrt{P^{2}+(m-1)q^{2}}}{\sqrt{m}-1},
\qquad q=\frac{m-P}{m-1},
\qquad\text{so}\qquad
H_{\min}(P,m)\approx 1-\frac{1}{\sqrt{1+P^{2}/m}} .
$$

> **Sandwich.** For any nonnegative factorization $\theta = LF'$,
>
> **(upper — proved in Steps A and B)** for *every* factor $k$ and *both* sides,
> $$
> H(\ell_{\cdot k}) \;\le\; 1-\sqrt{\frac{w_k}{\operatorname{pm}(\theta)}},
> \qquad
> H(f_{\cdot k}) \;\le\; 1-\sqrt{\frac{w_k}{\operatorname{pm}(\theta)}} ;
> $$
>
> **(lower — proved below)** for *at least one* factor $k$,
> $$
> \max\bigl\{H(\ell_{\cdot k}),\,H(f_{\cdot k})\bigr\}
> \;\ge\; H_{\min}\!\bigl(\sqrt{\operatorname{pm}(\theta)},\;\max(n,p)\bigr) .
> $$
>
> Both bounds are increasing in $\operatorname{pm}(\theta)$.

The two halves say different things, and the quantifiers matter. The upper half
constrains **every** column: nothing can be sparser than this. The lower half
constrains **some** column: at least one side of one factor must be at least
this sparse. Raising $c$ raises $\operatorname{pm}(\theta)$ and so lifts both
ends of the band, which is the formal counterpart of the rising lower envelope
seen in the simulations.

The upper half needs no new work. Step B gives
$\operatorname{pm}(\ell_{\cdot k})\operatorname{pm}(f_{\cdot k})w_k\le\operatorname{pm}(\theta)$,
and $\operatorname{pm}(f_{\cdot k})\ge 1$ always, so
$\operatorname{pm}(\ell_{\cdot k})\le \operatorname{pm}(\theta)/w_k$; substituting
into Step A's $H \le 1-1/\sqrt{\operatorname{pm}}$ gives the stated form, and the
argument is symmetric in $L$ and $F$.

The rest of this appendix proves the lower half, in two lemmas.

## Lemma 1: the least sparse a vector can be at a given peak-to-mean

> For nonnegative $x$ of length $m$ with $\operatorname{pm}(x)=P$,
> $$H(x)\;\ge\;H_{\min}(P,m),$$
> with equality for $x^\star=(P,q,\dots,q)$, $q=(m-P)/(m-1)$.

**Proof.** $H$ is a decreasing function of $\|x\|_1/\|x\|_2$, so minimising $H$
means minimising $\|x\|_2$. Both $H$ and $\operatorname{pm}$ are scale
invariant, so normalise $\operatorname{mean}(x)=1$; then $\|x\|_1=m$ is fixed and
the constraint $\operatorname{pm}(x)=P$ says $\max(x)=P$.

One coordinate is therefore pinned at $P$, contributing $P^2$ to $\|x\|_2^2$
regardless. The remaining $m-1$ coordinates are nonnegative and must sum to
$m-P$, and among vectors with a fixed sum the squared norm
$\sum_{i\ge 2}x_i^2$ is minimised when all entries are equal — by Cauchy–Schwarz,
$\sum_{i\ge2}x_i^2\ge\bigl(\sum_{i\ge2}x_i\bigr)^2/(m-1)$, with equality exactly
when they are. That common value is $q=(m-P)/(m-1)$, which satisfies
$0\le q\le P$ for every $1\le P\le m$, so $x^\star$ is feasible and $P$ really is
its maximum. Substituting $\|x\|_1=m$ and
$\|x\|_2^2=P^2+(m-1)q^2$ into the definition of $H$ gives $H_{\min}(P,m)$. $\square$

Note the shape. Step A's upper bound involves $P$; this lower bound involves
$P^{2}/m$. A lone spike among many coordinates barely moves $\ell_1/\ell_2$, so
the lower bound only becomes interesting once $P$ approaches $\sqrt{m}$.

```{r A1-check}
Hmin <- function(P, m) { q <- (m - P)/(m - 1)
  (sqrt(m) - m/sqrt(P^2 + (m-1)*q^2)) / (sqrt(m) - 1) }

set.seed(1); m <- 400
knitr::kable(do.call(rbind, lapply(c(2, 10, 50, 200), function(P) {
  best <- Inf                       # random x with pm exactly P: mean 1, one entry P
  for (rep in 1:8000) {
    y <- rexp(m - 1)^runif(1, 0.2, 4); y <- y/sum(y)*(m - P)
    if (max(y) > P) next
    best <- min(best, hoyer(c(P, y)))
  }
  data.frame(P = P, `H_min(P, m)` = Hmin(P, m),
             `H of x-star` = hoyer(c(P, rep((m - P)/(m - 1), m - 1))),
             `best of 8000 random` = best, check.names = FALSE)
})), digits = 5, row.names = FALSE,
caption = "x-star attains the bound exactly; random vectors never beat it.")
```

## Lemma 2: the reverse of the budget inequality

> $\displaystyle \operatorname{pm}(\theta)\;\le\;\max_k\ \operatorname{pm}(\ell_{\cdot k})\cdot\operatorname{pm}(f_{\cdot k}).$

**Proof.** Step B bounded $\max(\theta)$ from *below* by discarding all but one
term. Bound it from *above* instead by taking the largest possible value of each
term: for every $i,j$,

$$
\theta_{ij}=\sum_m \ell_{im}f_{jm}\;\le\;\sum_m \max(\ell_{\cdot m})\max(f_{\cdot m}),
\qquad\text{so}\qquad
\max(\theta)\;\le\;\sum_m \max(\ell_{\cdot m})\max(f_{\cdot m}).
$$

The mean factorizes as before,
$\operatorname{mean}(\theta)=\sum_m\operatorname{mean}(\ell_{\cdot m})\operatorname{mean}(f_{\cdot m})$.
Put $a_m=\max(\ell_{\cdot m})\max(f_{\cdot m})$ and
$b_m=\operatorname{mean}(\ell_{\cdot m})\operatorname{mean}(f_{\cdot m})>0$, and
apply the mediant inequality $\bigl(\sum_m a_m\bigr)/\bigl(\sum_m b_m\bigr)\le\max_m (a_m/b_m)$:

$$
\operatorname{pm}(\theta)=\frac{\max(\theta)}{\operatorname{mean}(\theta)}
\;\le\;\frac{\sum_m a_m}{\sum_m b_m}
\;\le\;\max_m \frac{a_m}{b_m}
\;=\;\max_m\ \operatorname{pm}(\ell_{\cdot m})\operatorname{pm}(f_{\cdot m}). \qquad\square
$$

Together with Step B the product is now pinned on both sides:

$$
\operatorname{pm}(\theta)\;\le\;\max_k\ \operatorname{pm}(\ell_{\cdot k})\operatorname{pm}(f_{\cdot k})
\;\le\;\frac{\operatorname{pm}(\theta)}{\min_k w_k}.
$$

## Proof of the lower half

By Lemma 2 there is a factor $k$ with
$\operatorname{pm}(\ell_{\cdot k})\operatorname{pm}(f_{\cdot k})\ge\operatorname{pm}(\theta)$.
A product of two numbers cannot have both factors below the square root of the
product, so

$$
\max\bigl\{\operatorname{pm}(\ell_{\cdot k}),\ \operatorname{pm}(f_{\cdot k})\bigr\}
\;\ge\;\sqrt{\operatorname{pm}(\theta)} .
$$

Apply Lemma 1 to whichever column achieves that maximum. Its length is $n$ or
$p$, and $H_{\min}(P,m)$ is decreasing in $m$, so using $\max(n,p)$ is valid
either way:

$$
\max\bigl\{H(\ell_{\cdot k}),\,H(f_{\cdot k})\bigr\}\;\ge\;H_{\min}\!\bigl(\sqrt{\operatorname{pm}(\theta)},\ \max(n,p)\bigr).
\qquad\square
$$

## Both halves, checked on the fits

```{r A3-check, cache=TRUE}
knitr::kable(do.call(rbind, lapply(C_GRID, function(cf) {
  r <- do.call(rbind, lapply(1:10, function(sd) {
    x <- readRDS(file.path(cells_dir, paste0(
           tag_of(list(seed = sd, c_true = 1, c_fit = cf)), ".rds")))
    L <- x$fits$rank1$LL; Fm <- x$fits$rank1$FF
    th <- tcrossprod(L, Fm) / max(1, if (is.infinite(cf)) 1 else cf)
    w  <- colMeans(L) * colMeans(Fm); w <- w/sum(w)
    P  <- pm(th); mm <- max(nrow(L), nrow(Fm))
    HL <- apply(L, 2, hoyer); HF <- apply(Fm, 2, hoyer)
    ub <- 1 - sqrt(w/P)                                  # upper, every column
    k  <- which.max(apply(L, 2, pm) * apply(Fm, 2, pm))  # the factor Lemma 2 gives
    c(`pm(theta)`       = P,
      `upper bound`     = max(ub),
      `largest H seen`  = max(c(HL, HF)),
      `max H at that k` = max(HL[k], HF[k]),
      `lower bound`     = Hmin(sqrt(P), mm),
      `upper violations`= sum(HL > ub + 1e-9) + sum(HF > ub + 1e-9),
      `lower violations`= as.numeric(max(HL[k], HF[k]) < Hmin(sqrt(P), mm) - 1e-9))
  }))
  as.data.frame(t(c(c_fit = cf, colMeans(r))))
})), digits = c(3, 2, 3, 3, 3, 4, 0, 0), row.names = FALSE,
caption = "Columns are ordered as they appear in the sandwich: upper bound, what is observed, lower bound.")
```

## How useful is the lower half?

Not very, quantitatively. The guarantee moves from about $10^{-4}$ to $0.02$
across the grid while the fits deliver $0.18$ to $0.59$ — the right *shape*, off
by more than an order of magnitude. Lemma 1 explains why: the bound depends on
$P^{2}/m$, and here $\sqrt{\operatorname{pm}(\theta)}\le 4.4$ against
$m=300$, so it never comes near the regime where it bites.

The gap is structural, not loose algebra. Lemmas 1 and 2 hold for *every*
nonnegative factorization of $\theta$, and a maximum-likelihood fit is not
adversarial. Closing it would need an argument about what maximum likelihood
actually does — that the fit concentrates because concentrating fits the data,
not merely because the geometry permits it. That is a different theorem and this
page does not have it.

So the pairing worth remembering:

- the **upper** bound is tight and explanatory — the fits spend their entire
  budget, which is why it predicts their behaviour;
- the **lower** bound is valid and correctly monotone but loose — it establishes
  that the trend cannot be reversed, not that it must be as strong as observed.

```{r session}
sessionInfo()
```
