---
title: "Approximation Comparison"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2024-12-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Here, I applied both the exact and approximate `log1p` matrix factorization methods to get a sense of how accurate the approximation is for different values of $c$. Below, column A was optimized using the exact algorithm, column B was optimized using the sparse approximation to the exact algorithm, and column C is frobenius NMF (using `NNLM`) applied to the transformed data.

```{r}
library(Matrix)
library(dplyr)
library(passPCA)

#load("/home/ericw456/pbmc/liu_data.Rdata")
load("~/Documents/data/passPCA/liu_data.Rdata")


counts <- counts[,Matrix::colSums(counts) > 0]
# require that all used genes appear in at least 5 cells
s <- Matrix::rowSums(counts)
s <- s / mean(s)
genes_to_use <- which(Matrix::colSums(counts>0)>4)
counts <- counts[,genes_to_use]
K <- 25
cc_vec <- c(1e-3, 1e-2, 1e-1, 1, 1e1, 1e2, 1e3)
```


```{r, eval=FALSE}
n <- nrow(counts)
p <- ncol(counts)

for (cc in cc_vec) {

  print(cc)

  set.seed(1)
  log1p_k1 <- fit_factor_model_log1p_exact(
    Y = counts,
    K = 1,
    maxiter = 10,
    s = cc * s,
    init_method = "frob_nmf"
  )

  init_LL <- log1p_k1$U %>%
    cbind(
      matrix(
        data = rexp(
          n = n * (K - 1), rate = 15
        ),
        nrow = n,
        ncol = K - 1
      )
    )

  init_FF <- log1p_k1$V %>%
    cbind(
      matrix(
        data = rexp(
          n = p * (K - 1), rate = 15
        ),
        nrow = p,
        ncol = K - 1
      )
    )

  tictoc::tic()
  set.seed(1)
  fit <- fit_factor_model_log1p_quad_approx_sparse(
    Y = counts,
    K = K,
    init_U = init_LL,
    init_V = init_FF,
    maxiter = 100,
    s = cc * s,
    approx_method = "taylor"
  )
  total_time <- tictoc::toc()

  fit[["total_time"]] <- total_time$toc
  rownames(fit$U) <- rownames(counts)
  rownames(fit$V) <- colnames(counts)

  readr::write_rds(
    fit, glue::glue("~/Documents/data/passPCA/liu_pbmc_log1p_c{cc}_k25_approx_taylor_100_iter.rds")
  )

}
```

```{r}
fit_list_exact <- list()
fit_list_approx_taylor <- list()
fit_list_approx_frob <- list()

for (cc in cc_vec) {
  
  fit_list_exact[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/paper_figures_res/liu_pbmc/liu_pbmc_log1p_c{cc}_k25_exact_100_iter.rds"
      )
    )
  
  fit_list_approx_taylor[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/paper_figures_res/liu_pbmc/liu_pbmc_log1p_c{cc}_k25_approx_taylor_100_iter.rds"
      )
    )
  
  fit_list_approx_frob[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/paper_figures_res/liu_pbmc/liu_pbmc_log1p_c{cc}_k25_approx_frob_100_iter.rds"
      )
    )
  
}
```

```{r}
normalize_bars <- function(LL) {

  max_col <- apply(LL, 2, max)
  sweep(LL, 2, max_col, FUN = "/")

}
```

```{r, eval=FALSE}
Y <- as.matrix(counts)

ll_vec_exact <- c()

for (cc in cc_vec) {

  fit <- fit_list_exact[[as.character(cc)]]
  B <- fit$U %*% t(fit$V)
  Lambda <- cc * (exp(B) - 1)
  Lambda <- as.matrix(Matrix::Diagonal(x = s) %*% Lambda)

  ll <- sum(
    dpois(
      x = as.vector(Y),
      lambda = as.vector(Lambda),
      log = TRUE
    )
  )

  ll_vec_exact <- c(ll_vec_exact, ll)

}

ll_vec_approx <- c()

for (cc in cc_vec) {

  fit <- fit_list_approx_taylor[[as.character(cc)]]
  B <- fit$U %*% t(fit$V)
  Lambda <- cc * (exp(B) - 1)
  Lambda <- as.matrix(Matrix::Diagonal(x = s) %*% Lambda)

  ll <- sum(
    dpois(
      x = as.vector(Y),
      lambda = as.vector(Lambda),
      log = TRUE
    )
  )

  ll_vec_approx <- c(ll_vec_approx, ll)

}

ll_df <- data.frame(
  ll_diff = pmax(1, ll_vec_exact - ll_vec_approx),
  c = cc_vec
)
```

```{r, eval=FALSE}
library(ggplot2)
ggplot(data = ll_df, aes(x = c, y = ll_diff)) +
  geom_point() +
  geom_line() +
  ylab("Approximation Error (Log-Likelihood)") +
  xlab("c") +
  scale_y_continuous(transform = "log10") +
  scale_x_continuous(transform = "log10") +
  cowplot::theme_cowplot()
```


```{r, fig.width=20, fig.height=40}
library(tidyr)
library(stringi)
library(stringr)
library(ggpubr)
cell.type <- as.factor(ct$celltype)

LL_c1_exact <- normalize_bars(fit_list_exact$`1`$U)

# Downsample the number of cells and sort them using tSNE.
set.seed(8675309)
cell.idx <- numeric(0)
cell.types <- levels(cell.type)
for (i in 1:length(cell.types)) {
  which.idx <- which(cell.type == cell.types[i])
  # Downsample common cell types.
  if (length(which.idx) > 2000) {
    which.idx <- sample(which.idx, 2000)
  }
  # Don't include rare cell types.
  if (length(which.idx) > 10) {
    # Sort using tsne.
    tsne.res <- Rtsne::Rtsne(
      LL_c1_exact[which.idx, ],
      dims = 1,
      pca = FALSE,
      normalize = FALSE,
      perplexity = min(100, floor((length(which.idx) - 1) / 3) - 1),
      theta = 0.1,
      max_iter = 1000,
      eta = 200,
      check_duplicates = FALSE
    )$Y[, 1]
    which.idx <- which.idx[order(tsne.res)]
    cell.idx <- c(cell.idx, which.idx)
  }
}

cell.type <- cell.type[cell.idx]
cell.type <- droplevels(cell.type)

LL_c1_exact <- LL_c1_exact[cell.idx, ]
LL_c001_exact <- normalize_bars(fit_list_exact$`0.001`$U)
LL_c001_exact <- LL_c001_exact[cell.idx, ]
LL_c01_exact <- normalize_bars(fit_list_exact$`0.01`$U)
LL_c01_exact <- LL_c01_exact[cell.idx, ]
LL_cp1_exact <- normalize_bars(fit_list_exact$`0.1`$U)
LL_cp1_exact <- LL_cp1_exact[cell.idx, ]
LL_c10_exact <- normalize_bars(fit_list_exact$`10`$U)
LL_c10_exact <- LL_c10_exact[cell.idx, ]
LL_c100_exact <- normalize_bars(fit_list_exact$`100`$U)
LL_c100_exact <- LL_c100_exact[cell.idx, ]
LL_c1000_exact <- normalize_bars(fit_list_exact$`1000`$U)
LL_c1000_exact <- LL_c1000_exact[cell.idx, ]

LL_c1_approx_taylor <- normalize_bars(fit_list_approx_taylor$`1`$U)
LL_c1_approx_taylor <- LL_c1_approx_taylor[cell.idx, ]
LL_c001_approx_taylor <- normalize_bars(fit_list_approx_taylor$`0.001`$U)
LL_c001_approx_taylor <- LL_c001_approx_taylor[cell.idx, ]
LL_c01_approx_taylor <- normalize_bars(fit_list_approx_taylor$`0.01`$U)
LL_c01_approx_taylor <- LL_c01_approx_taylor[cell.idx, ]
LL_cp1_approx_taylor <- normalize_bars(fit_list_approx_taylor$`0.1`$U)
LL_cp1_approx_taylor <- LL_cp1_approx_taylor[cell.idx, ]
LL_c10_approx_taylor <- normalize_bars(fit_list_approx_taylor$`10`$U)
LL_c10_approx_taylor <- LL_c10_approx_taylor[cell.idx, ]
LL_c100_approx_taylor <- normalize_bars(fit_list_approx_taylor$`100`$U)
LL_c100_approx_taylor <- LL_c100_approx_taylor[cell.idx, ]
LL_c1000_approx_taylor <- normalize_bars(fit_list_approx_taylor$`1000`$U)
LL_c1000_approx_taylor <- LL_c1000_approx_taylor[cell.idx, ]

LL_c1_approx_frob <- normalize_bars(fit_list_approx_frob$`1`$W)
LL_c1_approx_frob <- LL_c1_approx_frob[cell.idx, ]
LL_c001_approx_frob <- normalize_bars(fit_list_approx_frob$`0.001`$W)
LL_c001_approx_frob <- LL_c001_approx_frob[cell.idx, ]
LL_c01_approx_frob <- normalize_bars(fit_list_approx_frob$`0.01`$W)
LL_c01_approx_frob <- LL_c01_approx_frob[cell.idx, ]
LL_cp1_approx_frob <- normalize_bars(fit_list_approx_frob$`0.1`$W)
LL_cp1_approx_frob <- LL_cp1_approx_frob[cell.idx, ]
LL_c10_approx_frob <- normalize_bars(fit_list_approx_frob$`10`$W)
LL_c10_approx_frob <- LL_c10_approx_frob[cell.idx, ]
LL_c100_approx_frob <- normalize_bars(fit_list_approx_frob$`100`$W)
LL_c100_approx_frob <- LL_c100_approx_frob[cell.idx, ]
LL_c1000_approx_frob <- normalize_bars(fit_list_approx_frob$`1000`$W)
LL_c1000_approx_frob <- LL_c1000_approx_frob[cell.idx, ]

make.heatmap.tib <- function(FF) {
  tib <- as_tibble(scale(FF, center = FALSE, scale = apply(FF, 2, max))) %>%
    mutate(Cell.type = cell.type) %>%
    arrange(Cell.type) %>%
    mutate(Cell.idx = row_number())
  
  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(str_extract(Factor, "[0-9]+")))
  
  return(tib)
}

LL_c001_tib_exact <- make.heatmap.tib(LL_c001_exact)
LL_c01_tib_exact <- make.heatmap.tib(LL_c01_exact)
LL_cp1_tib_exact <- make.heatmap.tib(LL_cp1_exact)
LL_c1_tib_exact <- make.heatmap.tib(LL_c1_exact)
LL_c10_tib_exact <- make.heatmap.tib(LL_c10_exact)
LL_c100_tib_exact <- make.heatmap.tib(LL_c100_exact)
LL_c1000_tib_exact <- make.heatmap.tib(LL_c1000_exact)

LL_c001_tib_approx_taylor <- make.heatmap.tib(LL_c001_approx_taylor)
LL_c01_tib_approx_taylor <- make.heatmap.tib(LL_c01_approx_taylor)
LL_cp1_tib_approx_taylor <- make.heatmap.tib(LL_cp1_approx_taylor)
LL_c1_tib_approx_taylor <- make.heatmap.tib(LL_c1_approx_taylor)
LL_c10_tib_approx_taylor <- make.heatmap.tib(LL_c10_approx_taylor)
LL_c100_tib_approx_taylor <- make.heatmap.tib(LL_c100_approx_taylor)
LL_c1000_tib_approx_taylor <- make.heatmap.tib(LL_c1000_approx_taylor)

LL_c001_tib_approx_frob <- make.heatmap.tib(LL_c001_approx_frob)
LL_c01_tib_approx_frob <- make.heatmap.tib(LL_c01_approx_frob)
LL_cp1_tib_approx_frob <- make.heatmap.tib(LL_cp1_approx_frob)
LL_c1_tib_approx_frob <- make.heatmap.tib(LL_c1_approx_frob)
LL_c10_tib_approx_frob <- make.heatmap.tib(LL_c10_approx_frob)
LL_c100_tib_approx_frob <- make.heatmap.tib(LL_c100_approx_frob)
LL_c1000_tib_approx_frob <- make.heatmap.tib(LL_c1000_approx_frob)

heatmap.tib_exact <- LL_c001_tib_exact %>% mutate(Method = "c = 0.001") %>%
  bind_rows(LL_c01_tib_exact %>% mutate(Method = "c = 0.01")) %>%
  bind_rows(LL_cp1_tib_exact %>% mutate(Method = "c = 0.1")) %>%
  bind_rows(LL_c1_tib_exact %>% mutate(Method = "c = 1")) %>%
  bind_rows(LL_c10_tib_exact %>% mutate(Method = "c = 10")) %>%
  bind_rows(LL_c100_tib_exact %>% mutate(Method = "c = 100")) %>%
  bind_rows(LL_c1000_tib_exact %>% mutate(Method = "c = 1000")) %>%
  mutate(Method = factor(
    Method, 
    levels = c(
      "c = 0.001", "c = 0.01", "c = 0.1", "c = 1", "c = 10", "c = 100", "c = 1000"
      )))

heatmap.tib_approx_taylor <- LL_c001_tib_approx_taylor %>% mutate(Method = "c = 0.001") %>%
  bind_rows(LL_c01_tib_approx_taylor %>% mutate(Method = "c = 0.01")) %>%
  bind_rows(LL_cp1_tib_approx_taylor %>% mutate(Method = "c = 0.1")) %>%
  bind_rows(LL_c1_tib_approx_taylor %>% mutate(Method = "c = 1")) %>%
  bind_rows(LL_c10_tib_approx_taylor %>% mutate(Method = "c = 10")) %>%
  bind_rows(LL_c100_tib_approx_taylor %>% mutate(Method = "c = 100")) %>%
  bind_rows(LL_c1000_tib_approx_taylor %>% mutate(Method = "c = 1000")) %>%
  mutate(Method = factor(
    Method, 
    levels = c(
      "c = 0.001", "c = 0.01", "c = 0.1", "c = 1", "c = 10", "c = 100", "c = 1000"
      )))

heatmap.tib_approx_frob <- LL_c001_tib_approx_frob %>% mutate(Method = "c = 0.001") %>%
  bind_rows(LL_c01_tib_approx_frob %>% mutate(Method = "c = 0.01")) %>%
  bind_rows(LL_cp1_tib_approx_frob %>% mutate(Method = "c = 0.1")) %>%
  bind_rows(LL_c1_tib_approx_frob %>% mutate(Method = "c = 1")) %>%
  bind_rows(LL_c10_tib_approx_frob %>% mutate(Method = "c = 10")) %>%
  bind_rows(LL_c100_tib_approx_frob %>% mutate(Method = "c = 100")) %>%
  bind_rows(LL_c1000_tib_approx_frob %>% mutate(Method = "c = 1000")) %>%
  mutate(Method = factor(
    Method, 
    levels = c(
      "c = 0.001", "c = 0.01", "c = 0.1", "c = 1", "c = 10", "c = 100", "c = 1000"
      )))

tib_exact <- heatmap.tib_exact %>%
  group_by(Cell.type, Cell.idx) %>%
  summarize()

tib_approx_taylor <- heatmap.tib_approx_taylor %>%
  group_by(Cell.type, Cell.idx) %>%
  summarize()

tib_approx_frob <- heatmap.tib_approx_frob %>%
  group_by(Cell.type, Cell.idx) %>%
  summarize()

cell_type_breaks <- c(1, which(tib_exact$Cell.type[-1] != tib_exact$Cell.type[-nrow(tib_exact)]))
label_pos <- cell_type_breaks / 2 + c(cell_type_breaks[-1], nrow(tib_exact)) / 2

library(ggplot2)

plt_exact <- ggplot(heatmap.tib_exact, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(y = "") +
  scale_y_continuous(breaks = -label_pos,
                     minor_breaks = NULL,
                     labels = levels(cell.type)) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  theme_minimal() +
  geom_hline(yintercept = -cell_type_breaks, size = 0.1) +
  facet_wrap(~Method, ncol = 1, axes = "all") +
  theme(legend.position = "none", 
        strip.text = element_text(size = 16)) 

plt_approx_taylor <- ggplot(heatmap.tib_approx_taylor, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(y = "") +
  scale_y_continuous(breaks = -label_pos,
                     minor_breaks = NULL,
                     labels = levels(cell.type)) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  theme_minimal() +
  geom_hline(yintercept = -cell_type_breaks, size = 0.1) +
  facet_wrap(~Method, ncol = 1, axes = "all") +
  theme(legend.position = "none", 
        strip.text = element_text(size = 16)) 

plt_approx_frob <- ggplot(heatmap.tib_approx_frob, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(y = "") +
  scale_y_continuous(breaks = -label_pos,
                     minor_breaks = NULL,
                     labels = levels(cell.type)) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  theme_minimal() +
  geom_hline(yintercept = -cell_type_breaks, size = 0.1) +
  facet_wrap(~Method, ncol = 1, axes = "all") +
  theme(legend.position = "none", 
        strip.text = element_text(size = 16)) 

ggarrange(plt_exact, plt_approx_taylor, plt_approx_frob, nrow = 1, labels = "AUTO")
```
