---
title: "Non-identifiability of the log1p factorization, and how c controls it"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r pkgs}
library(ggplot2)
library(cowplot)
library(log1pNMF)
theme_set(theme_cowplot(font_size = 11))
```

```{r paths}
exp_dir   <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
cells_dir <- file.path(exp_dir, "c_grid_sim")       # the 640 per-cell fits
source(file.path(exp_dir, "c_grid_sim_common.R"))   # generator, for the examples
stopifnot(dir.exists(cells_dir))

hoyer <- function(v) { v <- abs(v); n <- length(v)
                       (sqrt(n) - sum(v)/sqrt(sum(v^2))) / (sqrt(n) - 1) }
mean_hoyer <- function(M) mean(apply(M, 2, hoyer))
```

Fitting the log1p model at different $c$ gives factorizations whose sparsity
differs a lot, even when the log-likelihoods are indistinguishable. This page
works out why, in general rather than for one simulation: what exactly is
unidentified, how much freedom there is, and how $c$ controls that freedom.

The short version, stated up front:

> The likelihood sees only $\theta = LF'$. For a **fixed** $\theta$, $c$ has no
> effect on identifiability whatsoever. What $c$ changes is *which* $\theta$ is
> needed to represent data with given rates — and as $c \to 0$ that $\theta$
> becomes a large constant plus a bounded perturbation, which is precisely the
> shape with the largest ambiguity.

# What the likelihood can and cannot see

The model is

$$
y_{ij}\sim\text{Poisson}(\lambda_{ij}), \qquad
\lambda_{ij} = c\left(e^{\theta_{ij}}-1\right), \qquad
\theta = LF', \quad L\in\mathbb{R}^{n\times K}_{\ge 0},\ F\in\mathbb{R}^{p\times K}_{\ge 0}.
$$

The map $\theta \mapsto \lambda$ is a fixed, strictly monotone, elementwise
transform once $c$ is chosen. So the log-likelihood is a function of $\theta$
alone: **any two parameter values with the same product $LF'$ have exactly the
same likelihood, for every $c$.** Whatever is not determined by $\theta$ is not
determined by the data, no matter how much data there is.

# The ambiguity group

Suppose $\theta$ has rank $K$ and admits a nonnegative rank-$K$ factorization
$\theta = LF'$. If $\tilde L\tilde F' = LF'$ is another one, the two share a
column space, so $\tilde L = LA$ for some invertible $A$, and then
$\tilde F' = A^{-1}F'$. Hence **every** alternative factorization has the form

$$
(L, F) \;\longmapsto\; \bigl(LA,\; FA^{-\top}\bigr),
\qquad A \in \mathcal{A}(L,F) := \bigl\{A \ \text{invertible} \ K\times K : LA \ge 0,\; FA^{-\top} \ge 0\bigr\}.
$$

Invertibility is all that is asked of $A$ itself, and it is forced by the
construction, since $A^{-1}$ has to exist. In particular $A$ is **not** required
to be nonnegative — only the two *products* $LA$ and $FA^{-\top}$ are. That is
where the whole ambiguity lives: if $A$ and $A^{-1}$ were both forced
nonnegative, $A$ would have to be monomial and there would be nothing to
discuss.

$\mathcal{A}$ always contains the **monomial** matrices — a permutation times a
positive diagonal — which merely relabel and rescale the factors. That part is
harmless. Non-identifiability in the sense that matters is $\mathcal{A}$
containing anything else.

## A geometric reading

Write $a_k$ for the $k$th column of $A$ and $\tilde a_k$ for the $k$th *row* of
$A^{-1}$. Then

$$
(LA)_{ik} = \langle \ell_i, a_k\rangle, \qquad
(FA^{-\top})_{jk} = \langle f_j, \tilde a_k\rangle,
$$

so the constraints are

$$
a_k \in \mathcal{C}_L^{*} \ \ \forall k,
\qquad
\tilde a_k \in \mathcal{C}_F^{*} \ \ \forall k,
$$

where $\mathcal{C}_L$ is the cone generated by the rows of $L$ and
$\mathcal{C}_L^{*} = \{y : \langle \ell_i, y\rangle \ge 0\ \forall i\}$ is its
dual, and likewise for $F$.

This makes the picture concrete. Since $L \ge 0$, the cone $\mathcal{C}_L$ sits
inside the nonnegative orthant, so its dual *contains* the orthant. The more
$\mathcal{C}_L$ fills the orthant, the tighter its dual, and the less room $A$
has.

**Uniqueness by separability.** If $L$ has a *pure* row for each factor — a
sample loading on one factor only — then $\mathcal{C}_L$ is the whole orthant, so
$\mathcal{C}_L^{*}$ is the orthant and $A \ge 0$. If $F$ likewise has a pure row
per factor, then $A^{-1} \ge 0$. A nonnegative matrix with a nonnegative inverse
is monomial, so the factorization is unique up to relabelling. Pure rows —
equivalently, *zeros in the right places* — are what pin an NMF down.

# Where c enters

Nothing above mentions $c$. For a fixed $\theta$ the ambiguity set
$\mathcal{A}$ is fixed, so **$c$ cannot affect identifiability directly**. It
enters one step earlier: to model data whose rates are $\lambda$, the required
linear predictor is

$$
\theta_c = \log\!\left(1 + \frac{\lambda}{c}\right),
$$

and the *shape* of $\theta_c$ — hence the geometry above — depends strongly on
$c$. Two limits make this sharp. Let $\lambda_{\min}$ and $\lambda_{\max}$ be the
smallest and largest rates.

**Large $c$.** $\theta_c \approx \lambda/c$. The shape of $\theta$ is the shape of
$\lambda$, exactly. Any zeros or near-zeros in $\lambda$ survive into $\theta$,
and with them the pure rows that pin the factorization.

**Small $c$, with $\lambda_{\min} > 0$.**

$$
\theta_{\min} = \log\!\left(1+\frac{\lambda_{\min}}{c}\right) \xrightarrow[c\to 0]{} \infty,
\qquad
\theta_{\max}-\theta_{\min}
= \log\frac{c+\lambda_{\max}}{c+\lambda_{\min}}
\xrightarrow[c\to 0]{} \log\frac{\lambda_{\max}}{\lambda_{\min}},
$$

which is **finite**. The level of $\theta$ diverges while its spread stays
bounded, so $\theta / \theta_{\min} \to \mathbf{1}\mathbf{1}'$. As $c\to 0$ the
linear predictor becomes a huge constant plus a bounded perturbation.

A constant matrix is the *least* identifiable thing a nonnegative factorization
can be asked to represent: with $\theta \approx \kappa\,\mathbf{1}\mathbf{1}'$
every row of $L$ points at the centre of the simplex, $\mathcal{C}_L$ collapses
to a ray far from the orthant boundary, and $\mathcal{C}_L^{*}$ becomes enormous.

The one escape is $\lambda_{\min} = 0$: exact zeros in the rates give
$\theta_{ij}=0$ for **every** $c$, since $\log(1+0/c) = 0$. Exact zeros are
preserved by the link; it is *near*-zeros that get washed out.

```{r two-limits}
## the two limits, on the rates from one simulated dataset
lam <- sim_dataset(1, 1)$lambda
lam <- lam[lam > 0]
shape <- function(cc) { th <- log1p(lam / cc)
  c(level = min(th), spread = max(th) - min(th),
    `spread / level` = (max(th) - min(th)) / min(th)) }
knitr::kable(t(sapply(c(1e-4, 1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3), shape)) |>
             `rownames<-`(format(c(1e-4, 1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3))),
             digits = 3,
             caption = "theta = log1p(lambda/c) for one fixed set of rates. The spread saturates at log(lambda_max/lambda_min) while the level grows without bound.")
```

# How much freedom, concretely

The full set $\mathcal{A}$ is awkward — it constrains $A$ and $A^{-1}$
simultaneously, so it is not convex. A one-parameter slice through it is enough
to get closed forms and the right intuition. Let $J = \mathbf{1}\mathbf{1}'$ and

$$
B_t \;=\; (1-t)\,I \;+\; \frac{t}{K}J,
\qquad
B_t^{-1} \;=\; \frac{1}{1-t}\left(I - \frac{t}{K}J\right).
$$

$t$ is a **mixing fraction**: $t=0$ is the identity, and $t\to 1$ blends all $K$
factors into one. Writing the two sides out, with $r_i = \sum_k \ell_{ik}$ and
$s_j = \sum_k f_{jk}$,

$$
(LB_t)_{ik} = (1-t)\,\ell_{ik} + \frac{t}{K}\,r_i,
\qquad
(FB_t^{-1})_{jk} = \frac{1}{1-t}\left(f_{jk} - \frac{t}{K}\,s_j\right).
$$

So increasing $t$ pulls each row of $L$ toward its own mean — **denser** — while
subtracting a per-feature constant from $F$ — **sparser**. This is the essential
point: sparsity is not lost, it is *moved between the two factors*. Nothing here
is specific to $L$.

## The feasible interval, in closed form

Both expressions are linear in $t$, so the constraints resolve exactly. Define
the smallest normalized share on each side,

$$
q_L = K\min_{i,k}\frac{\ell_{ik}}{r_i}\in[0,1],
\qquad
q_F = K\min_{j,k}\frac{f_{jk}}{s_j}\in[0,1].
$$

For $t \ge 0$ the $L$ constraint is automatic (both terms are nonnegative), and
the $F$ constraint is $f_{jk}\ge \tfrac{t}{K}s_j$ for all $j,k$. For $t<0$ the
roles swap. Hence

$$
\boxed{\;-\frac{q_L}{1-q_L} \;\le\; t \;\le\; q_F\;}
$$

$q_F$ is exactly *how far $F$ is from having a zero*: it is $0$ as soon as any
single $f_{jk}=0$, and $1$ only if every feature loads equally on all factors.
The same reading applies to $q_L$ on the other side. **A single exact zero in $F$
removes the entire positive branch of this family.**

```{r verify-interval}
d0 <- sim_dataset(1, 1e-3)
L0 <- d0$truth$L; F0c <- d0$FF_true; KK <- ncol(L0)

qL <- KK * min(L0  / rowSums(L0))
qF <- KK * min(F0c / rowSums(F0c))

feasible <- function(t) {
  B <- (1 - t) * diag(KK) + (t / KK) * matrix(1, KK, KK)
  min(L0 %*% B) >= -1e-12 && min(F0c %*% t(solve(B))) >= -1e-12
}
bisect <- function(sgn) { lo <- 0; hi <- 1
  for (i in 1:60) { m <- (lo + hi)/2; if (feasible(sgn*m)) lo <- m else hi <- m }; sgn*lo }

knitr::kable(data.frame(
  bound   = c("t_min", "t_max"),
  formula = c(-qL/(1 - qL), qF),
  numeric = c(bisect(-1), bisect(+1))),
  digits = 8, caption = "Closed form against bisection, at c = 0.001.")
```

## Asymptotics in c

Under the small-$c$ limit above, $\theta \approx \kappa\mathbf{1}\mathbf{1}' +
\theta^{*}$ with $\kappa = \theta_{\min}\to\infty$. Normalizing $L$ so that
$r_i = 1$, the constant is carried by $F$: $f_{jk} = f^{*}_{jk} + \kappa$. Then

$$
q_F = K\min_j \frac{\min_k f^{*}_{jk} + \kappa}{S^{*}_j + K\kappa}
\;\xrightarrow[\kappa\to\infty]{}\; 1,
\qquad
1 - q_F \;\approx\; \frac{\max_j\bigl(S^{*}_j - K\min_k f^{*}_{jk}\bigr)}{K\kappa},
$$

with $S^{*}_j = \sum_k f^{*}_{jk}$. So the freedom approaches *total* as $c\to0$,
and it does so at rate $1/\kappa = 1/\log(1/c)$ — slowly, but without bound.
Conversely $\kappa \to 0$ recovers $q_F = K\min_{j,k}f^{*}_{jk}/S^{*}_j$, which is
zero the moment $F^{*}$ has a zero.

The offset is not a modelling artifact: it is pinned by the smallest rate,

$$
\kappa = \theta_{\min} = \log\!\left(1 + \frac{\lambda_{\min}}{c}\right).
$$

# Numerical illustration

Using the calibrated generator (see the [c-grid page](c_grid_sim.html)), which
builds a truth at each $c$ with matched marginals, we can read off the feasible
interval and the sparsity it buys.

```{r sweep, cache=TRUE}
tr <- sim_truth(1)
sweep_c <- do.call(rbind, lapply(C_GRID, function(cc) {
  dd <- sim_dataset(1, cc, tr)
  L <- dd$truth$L; Fm <- dd$FF_true
  th <- dd$alpha * tr$theta0 + dd$beta
  qF <- KK * min(Fm / rowSums(Fm))
  B  <- (1 - qF) * diag(KK) + (qF / KK) * matrix(1, KK, KK)
  data.frame(c = cc,
             `theta level / spread` = min(th) / (max(th) - min(th)),
             q_F = qF,
             `Hoyer L: truth`   = mean_hoyer(L),
             `Hoyer L: at t_max`= mean_hoyer(L %*% B),
             `Hoyer F: truth`   = mean_hoyer(Fm),
             `Hoyer F: at t_max`= mean_hoyer(Fm %*% t(solve(B))),
             check.names = FALSE)
}))
knitr::kable(sweep_c, digits = 3, row.names = FALSE)
```

```{r sweep-plot, fig.width=9, fig.height=4}
lab <- ifelse(is.infinite(sweep_c$c), "∞",
              format(sweep_c$c, scientific = FALSE, trim = TRUE, drop0trailing = TRUE))
sweep_c$cf <- factor(lab, levels = lab)

p1 <- ggplot(sweep_c, aes(cf, q_F, group = 1)) +
  geom_line(colour = "#0E5C6B", linewidth = 1) + geom_point(colour = "#0E5C6B", size = 2) +
  ylim(0, 1) +
  labs(x = "c", y = expression(q[F]),
       title = "Freedom in the mixing family",
       subtitle = "0 = pinned, 1 = the factors can be blended completely")

rng <- rbind(
  data.frame(cf = sweep_c$cf, side = "L", lo = sweep_c$`Hoyer L: at t_max`,
             hi = sweep_c$`Hoyer L: truth`),
  data.frame(cf = sweep_c$cf, side = "F", lo = pmin(sweep_c$`Hoyer F: truth`, sweep_c$`Hoyer F: at t_max`),
             hi = pmax(sweep_c$`Hoyer F: truth`, sweep_c$`Hoyer F: at t_max`)))

p2 <- ggplot(rng, aes(cf, ymin = lo, ymax = hi, colour = side)) +
  geom_linerange(linewidth = 2.2, alpha = 0.85,
                 position = position_dodge(width = 0.35)) +
  scale_colour_manual(values = c(L = "#0E8FA8", F = "#C2410C"), name = NULL) +
  labs(x = "c", y = "achievable mean Hoyer sparsity",
       title = "Sparsity is not determined by the data",
       subtitle = "range spanned without changing the likelihood at all")

plot_grid(p1, p2, nrow = 1, rel_widths = c(1, 1.15))
```

At $c=\infty$ the interval is a point: the factorization is effectively pinned.
At $c=10^{-3}$ the sparsity of $L$ can be driven from the truth's value down by
roughly an order of magnitude, with $F$ moving the other way, and the likelihood
does not change by so much as a rounding error.

## The slice is a lower bound

$B_t$ is one line through $\mathcal{A}$. Searching over all of $A\in\mathbb{R}^{K
\times K}$ gives a slightly wider range, so the closed forms above understate the
true freedom.

```{r full-search, cache=TRUE}
explore <- function(L, Fm, sgn, nstart = 25, seed = 1) {
  K <- ncol(L); set.seed(seed)
  obj <- function(v) {
    A <- matrix(v, K, K); dt <- det(A)
    if (!is.finite(dt) || abs(dt) < 1e-8) return(1e6)
    LA <- L %*% A; FA <- Fm %*% t(solve(A))
    pen <- sum(pmin(LA, 0)^2) + sum(pmin(FA, 0)^2)
    if (pen > 0) return(1e3 + 1e6 * pen)
    -sgn * mean_hoyer(LA)
  }
  best <- NULL; bv <- Inf
  for (s in 1:nstart) {
    o <- try(optim(as.vector(diag(K) + matrix(rnorm(K*K, 0, .15), K, K)), obj,
                   control = list(maxit = 3000, reltol = 1e-10)), silent = TRUE)
    if (!inherits(o, "try-error") && o$value < bv) { bv <- o$value; best <- o$par }
  }
  mean_hoyer(L %*% matrix(best, K, K))
}

knitr::kable(do.call(rbind, lapply(c(1e-3, 1, Inf), function(cc) {
  dd <- sim_dataset(1, cc, tr); L <- dd$truth$L; Fm <- dd$FF_true
  qF <- KK * min(Fm / rowSums(Fm))
  B  <- (1 - qF) * diag(KK) + (qF / KK) * matrix(1, KK, KK)
  data.frame(c = cc, `truth` = mean_hoyer(L),
             `mixing slice, lowest` = mean_hoyer(L %*% B),
             `full search, lowest`  = explore(L, Fm, -1),
             check.names = FALSE)
})), digits = 3, row.names = FALSE, caption = "Mean Hoyer sparsity of L.")
```

# What sets the joint level of sparsity?

Everything above is about *where sparsity sits*: the mixing family moves it from
$L$ to $F$ and back, leaving the likelihood untouched. That cannot explain a
common observation in practice — that fitting at larger $c$ makes **both** $L$
and $F$ sparser. A trade cannot raise both. Something else fixes the level.

## Support intersection

The constraint is combinatorial. Write $S_i = \operatorname{supp}(\ell_i)$ for the
set of factors sample $i$ uses, and $T_j = \operatorname{supp}(f_j)$ for feature
$j$. Since $L, F \ge 0$ there is no cancellation, so

$$
\theta_{ij} = \langle \ell_i, f_j\rangle > 0
\iff
S_i \cap T_j \neq \emptyset .
$$

If $\theta$ is **strictly positive everywhere**, then *every* $S_i$ must meet
*every* $T_j$ — the two families must be cross-intersecting. That is a strong
joint requirement, and it has an immediate consequence:

> If any sample uses a single factor, $S_i = \{k\}$, then every $T_j$ must
> contain $k$: **column $k$ of $F$ is completely dense.**

and symmetrically with the roles swapped. Sparsity on one side is paid for by
density on the other, *and it is the strictly positive $\theta$ that sends the
bill*. A crude but useful sufficient condition is $|S_i| + |T_j| > K$, which
guarantees intersection by pigeonhole; at $K = 3$ two samples and features each
using two of the three factors always overlap, whereas two singletons need not.

So a $\theta$ bounded away from zero cannot be represented by two sparse
nonnegative factors at all — regardless of $c$, regardless of the optimizer.
Conversely, **each zero of $\theta$ relaxes exactly one of these
cross-intersection requirements**, buying sparsity on both sides at once.

This argument is about *exact* zeros, and so it speaks directly to exact
sparsity. Hoyer sparsity is a continuous measure of concentration, and a vector
with no zeros at all can still be very sparse in the Hoyer sense. The next
subsection gives the continuous analogue, which turns out to be about the
**range of values in $\theta$** rather than its zeros.

## The continuous version: a peak-to-mean budget

Hoyer sparsity does not require zeros, so the exact-support argument is not the
whole story. The continuous statement is about range, and it comes in three
pieces.

**Piece 1: Hoyer is controlled by peak-to-mean.** For a nonnegative vector $x$ of
length $n$, write $P(x) = x_{\max}/\bar{x}$. Since
$\|x\|_2^2 \le \|x\|_\infty\|x\|_1$,

$$
\frac{\|x\|_1}{\|x\|_2} \;\ge\; \frac{\|x\|_1}{\sqrt{\|x\|_\infty \|x\|_1}}
= \sqrt{\frac{\|x\|_1}{\|x\|_\infty}} = \sqrt{\frac{n}{P(x)}},
\qquad\text{so}\qquad
H(x) \;\le\; \frac{\sqrt{n}-\sqrt{n/P}}{\sqrt{n}-1} \;\approx\; 1 - \frac{1}{\sqrt{P}} .
$$

A vector can only be Hoyer-sparse if some entries stick far above the mean. If
$P = 1$ the vector is constant and $H = 0$.

Relatedly, if the entries are confined to a band $[b, M]$ with $\rho = M/b$, the
most concentrated arrangement puts a fraction $u^\star = 1/(\rho+1)$ of entries at
$M$ and the rest at $b$, giving

$$
H_{\max}(\rho) = 1 - \frac{2\sqrt{\rho}}{\rho+1} = \frac{(\sqrt{\rho}-1)^2}{\rho+1}.
$$

So $\rho = 1 \Rightarrow H_{\max} = 0$: no range, no sparsity, zeros or not.

**Piece 2: $\theta$'s range is a budget that $L$ and $F$ must share.** Fix a
factor $k$ and let $i^\star = \arg\max_i \ell_{ik}$, $j^\star = \arg\max_j f_{jk}$.
Because every term is nonnegative we may drop all but the $k$th,

$$
\theta_{\max} \;\ge\; \theta_{i^\star j^\star} = \sum_m \ell_{i^\star m} f_{j^\star m}
\;\ge\; \ell^{\max}_k f^{\max}_k .
$$

Meanwhile $\bar\theta = \sum_m \bar\ell_m \bar f_m$. Writing
$w_k = \bar\ell_k \bar f_k / \bar\theta$ for factor $k$'s share of the mean and
dividing,

$$
\boxed{\;P_L(k)\cdot P_F(k)\cdot w_k \;\le\; P_\theta \;}
\qquad
P_\theta = \frac{\theta_{\max}}{\bar\theta},\quad
P_L(k) = \frac{\ell^{\max}_k}{\bar\ell_k},\quad
P_F(k) = \frac{f^{\max}_k}{\bar f_k}.
$$

This is exact, needs no zeros, and is the continuous counterpart of the support
argument. The peakedness of $L$ and of $F$ enter **multiplicatively** against a
single budget $P_\theta$. Combined with Piece 1 it caps both Hoyer values at once:
neither factor can concentrate unless $\theta$ has range to spend.

It also reconciles the two observations that motivated all of this. At fixed
$\theta$ the budget is fixed, so pushing $P_L$ up must push $P_F$ down — that is
the mixing family, a movement *along* the constraint. Changing $c$ changes
$P_\theta$ itself, moving the constraint surface, which lets both rise together.

**Piece 3: $c$ sets $P_\theta$.** With $\theta = \log(1+\lambda/c)$,

$$
P_\theta(c) = \frac{\log(1+\lambda_{\max}/c)}{\overline{\log(1+\lambda/c)}} .
$$

As $c \to \infty$ the logs linearize and $P_\theta \to \lambda_{\max}/\bar\lambda$,
the rates' own peak-to-mean. As $c \to 0$ every term behaves like
$\log(1/c) + \log\lambda$, the shared divergent part dominates, and

$$
P_\theta(c) \;\longrightarrow\; 1,
\qquad
P_\theta(c) - 1 \;\approx\; \frac{\log\lambda_{\max} - \overline{\log\lambda}}{\log(1/c)} .
$$

**Small $c$ compresses $\theta$ toward a constant, which drives the budget to
1, which forces both factors flat.** The rates never changed; only how far apart
the link places them did.

```{r p2m-check, cache=TRUE}
p2m <- function(v) max(v) / mean(v)
knitr::kable(do.call(rbind, lapply(C_GRID, function(cf) {
  r <- do.call(rbind, lapply(1:10, function(sd) {
    x  <- readRDS(file.path(cells_dir, paste0(
            tag_of(list(seed = sd, c_true = 1, c_fit = cf)), ".rds")))
    L <- x$fits$rank1$LL; Fm <- x$fits$rank1$FF
    th <- tcrossprod(L, Fm) / max(1, if (is.infinite(cf)) 1 else cf)
    w  <- colMeans(L) * colMeans(Fm); w <- w / sum(w)
    PL <- apply(L, 2, p2m); PF <- apply(Fm, 2, p2m)
    c(P_theta = p2m(th), P_L = mean(PL), P_F = mean(PF),
      `P_L P_F w` = max(PL * PF * w),
      `Hoyer L` = mean(apply(L, 2, hoyer)),
      `Hoyer F` = mean(apply(Fm, 2, hoyer)),
      violations = sum(PL * PF * w > p2m(th) * (1 + 1e-8)))
  }))
  as.data.frame(t(c(c_fit = cf, colMeans(r))))
})), digits = c(3, 2, 2, 2, 2, 3, 3, 0), row.names = FALSE,
caption = "The bound is exact (no violations in 80 fits x 3 factors) and essentially tight: the fits spend the whole budget.")
```

The budget $P_\theta$ climbs from 1.6 to 19.4 across the grid, $P_L$ and $P_F$
climb with it, and the product sits right against the bound in every row. That
tightness is the point: the optimizer is not leaving sparsity on the table at
small $c$ — there is none to take.

## Where c comes in, again

$\theta_{ij} = 0$ if and only if $\lambda_{ij} = 0$, at *every* $c$ — the link
maps zero to zero. What $c$ controls is the **floor** under the nonzero entries,

$$
\theta_{\min} = \log\!\left(1 + \frac{\lambda_{\min}}{c}\right),
$$

so the useful comparison is not $c$ in the abstract but $c$ against the smallest
rate in the data:

| regime | $\theta_{\min}$ | consequence |
|---|---|---|
| $c \gg \lambda_{\min}$ | $\approx \lambda_{\min}/c \approx 0$ | no floor; both factors free to be sparse |
| $c \ll \lambda_{\min}$ | $\approx \log(\lambda_{\min}/c)$, large | floor; both factors forced dense |

The mechanism is a property of the model. Whether it bites is a property of the
data.

## The floor and the joint sparsity move together

First, in the fits from the [c-grid page](c_grid_sim.html): reconstruct
$\hat\theta = \hat L\hat F'$ at each $c_{\text{fit}}$ and measure how much of it
is an irreducible constant.

```{r fitted-floor}
knitr::kable(do.call(rbind, lapply(C_GRID, function(cf) {
  x  <- readRDS(file.path(cells_dir, paste0(
          tag_of(list(seed = 1, c_true = 1, c_fit = cf)), ".rds")))
  Lh <- x$fits$rank1$LL; Fh <- x$fits$rank1$FF
  th <- tcrossprod(Lh, Fh) / max(1, if (is.infinite(cf)) 1 else cf)
  data.frame(c_fit = cf, `min theta-hat` = min(th),
             `floor share` = min(th)/mean(th),
             `Hoyer L` = mean_hoyer(Lh), `Hoyer F` = mean_hoyer(Fh),
             check.names = FALSE)
})), digits = 3, row.names = FALSE,
caption = "One dataset (seed 1, simulated at c = 1), fitted at each c. 'floor share' is min(theta)/mean(theta).")
```

The floor share falls from 0.81 to 0 and both sparsities rise with it. At
$c_{\text{fit}} \ge 10$ the fit puts *exact* zeros into $\hat\theta$ (minimum
around $10^{-13}$) and sparsity reaches its maximum.

## The decisive test: give the rates exact zeros

Correlation is not mechanism. If the floor is really the cause, then data whose
rates contain exact zeros should show **no** systematic sparsity trend in $c$,
because those zeros survive every $c$.

Build two datasets from one truth with sparse supports on both sides, calibrated
to the *same* 60% zero fraction in $Y$, differing only in whether an offset is
allowed:

```{r zeros-setup, cache=TRUE}
suppressMessages(library(fastTopics))
nn <- 300; pp <- 300; Kz <- 3
set.seed(11)
Lz <- matrix(0, nn, Kz)
for (i in 1:nn) { S <- sample(Kz, sample(1:2, 1)); w <- rgamma(length(S), .5)
                  Lz[i, S] <- w / sum(w) }
Fz <- matrix(0, pp, Kz)
for (j in 1:pp) { S <- sample(Kz, sample(1:2, 1))
                  Fz[j, S] <- exp(runif(length(S), -3.5, 3.5)) }
th0z <- Lz %*% t(Fz)

zfrac  <- function(lam) mean(exp(-lam))
lam_of <- function(a, b, cc) { th <- a*th0z + b
  if (is.infinite(cc)) th else cc*expm1(pmin(th, 700)) }

## zeros arm: offset forced to 0, one knob for the zero fraction
cal_zero <- function(cc, tz = 0.6)
  exp(uniroot(function(la) zfrac(lam_of(exp(la), 0, cc)) - tz,
              c(-14, 10), extendInt = "downX")$root)

## floored arm: two knobs, same zero fraction plus an upper-quantile target
cal_floor <- function(cc, tz = 0.6, tq = 20, qp = 0.999) {
  bmax <- if (is.infinite(cc)) 1e6 else log1p(1e6/cc)
  bfor <- function(a) { f <- function(b) zfrac(lam_of(a, b, cc)) - tz
    if (f(0) <= 0) return(0); uniroot(f, c(0, bmax), tol = 1e-12)$root }
  g  <- function(la) { a <- exp(la)
    as.numeric(quantile(lam_of(a, bfor(a), cc), qp)) - tq }
  hi <- log(200 / max(th0z)); a <- exp(uniroot(g, c(hi - 30, hi), tol = 1e-10)$root)
  c(a = a, b = bfor(a))
}
c(`theta0 exact zeros` = mean(th0z == 0),
  `true Hoyer L` = mean_hoyer(Lz), `true Hoyer F` = mean_hoyer(Fz))
```

```{r zeros-fit, cache=TRUE, dependson="zeros-setup"}
fit_one <- function(Y, cf) {
  set.seed(7)
  if (is.infinite(cf)) {
    f <- fit_poisson_nmf(as(Y, "CsparseMatrix"), k = Kz, numiter = 3000,
           method = "scd", control = list(nc = 2, min.delta.loglik = 1e-6),
           verbose = "none")
    list(L = f$L, F = f$F)
  } else {
    f <- fit_poisson_log1p_nmf(Y, K = Kz, cc = cf, s = FALSE, loglik = "exact",
           init_method = "rank1",
           control = list(maxiter = 4000, tol = 1e-6, verbose = FALSE, threads = 2))
    list(L = f$LL, F = f$FF)
  }
}

CZ <- c(1e-3, 1e-1, 1, 10, Inf)
zres <- do.call(rbind, lapply(c("zeros", "floored"), function(arm) {
  do.call(rbind, lapply(CZ, function(cf) {
    ab  <- if (arm == "zeros") c(a = cal_zero(cf), b = 0) else cal_floor(cf)
    lam <- lam_of(ab[["a"]], ab[["b"]], cf)
    set.seed(5); Y <- matrix(rpois(length(lam), lam), nn, pp)
    Y <- Y[rowSums(Y) > 0, colSums(Y) > 0, drop = FALSE]
    z <- fit_one(Y, cf)
    data.frame(arm = arm, c_fit = cf,
               `min lambda` = min(lam), `zeros in Y` = mean(Y == 0),
               `Hoyer L` = mean_hoyer(z$L), `Hoyer F` = mean_hoyer(z$F),
               check.names = FALSE)
  }))
}))
knitr::kable(zres, digits = 3, row.names = FALSE,
  caption = paste("Same truth, matched zero fraction. True Hoyer:",
                  sprintf("L %.3f, F %.3f.", mean_hoyer(Lz), mean_hoyer(Fz))))
```

```{r zeros-plot, fig.width=8, fig.height=3.6}
## long form: the two Hoyer columns have different names, so build it explicitly
zl <- rbind(
  data.frame(arm = zres$arm, c_fit = zres$c_fit, side = "L", h = zres[["Hoyer L"]]),
  data.frame(arm = zres$arm, c_fit = zres$c_fit, side = "F", h = zres[["Hoyer F"]]))
zl$side <- factor(zl$side, levels = c("L", "F"))
zl$cf <- factor(ifelse(is.infinite(zl$c_fit), "∞",
                       format(zl$c_fit, scientific = FALSE, trim = TRUE,
                              drop0trailing = TRUE)),
                levels = c("0.001", "0.1", "1", "10", "∞"))
truth <- data.frame(side = c("L", "F"),
                    h = c(mean_hoyer(Lz), mean_hoyer(Fz)))

ggplot(zl, aes(cf, h, colour = arm, group = arm)) +
  geom_hline(data = truth, aes(yintercept = h), linetype = "dashed",
             colour = "grey50") +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  facet_wrap(~ side) +
  scale_colour_manual(values = c(zeros = "#0E8FA8", floored = "#C2410C"),
                      name = NULL) +
  labs(x = "c used for fitting", y = "mean Hoyer sparsity",
       title = "The sparsity trend in c needs a floor to exist",
       subtitle = "dashed: the truth. With exact zeros in lambda the trend disappears.")
```

With exact zeros the trend is gone: at $c_{\text{fit}} = 10$ and $\infty$ both
factors are recovered essentially exactly, and small $c$ no longer collapses
them. With a floor the familiar monotone climb returns.

A residual effect does survive at the smallest $c$ in the zeros arm — $L$ comes
out somewhat denser than the truth while $F$ does not. Exact zeros are preserved
at every $c$, but the *nonzero* entries of $\theta$ are still compressed toward a
common large value as $c \to 0$, so near-purity degrades even when exact purity
does not. The effect is far weaker than with a floor, and it does not hit both
factors at once.

## How realistic is the floor?

In the calibrated generator used on the [c-grid page](c_grid_sim.html), the
offset makes $\lambda_{\min} = c\,\text{expm1}(\beta_c)$, which lands at about
$0.47$ against a median rate near $0.52$ — a floor at roughly 90% of the median.
That is far higher than real count data, where the least-expressed feature in the
shallowest sample sits orders of magnitude below the median. With a realistic
$\lambda_{\min}$, $c = 1$ is comfortably in the $c \gg \lambda_{\min}$ regime and
sparsity is not suppressed at all; the effect only appears when fitting with a
$c$ below the smallest rate present.

So the mechanism generalizes, but **the magnitude in that simulation is inflated
by its own calibration**, which is the price paid for making the datasets
comparable across $c$.

# In practice: you never see the rates

Everything so far conditions on the true $\lambda$, which you never have. In
practice you hold $Y$, you fit at several $c$, and most of those $c$ are
misspecified. Saying "the floor of $\hat\theta$ controls the sparsity" is then
circular, because $\hat\theta$ *is* the fit. This section removes the circularity
by tracing the floor back to quantities fixed by $Y$ and $K$ alone.

## Step 1: the objective does not contain c

The Poisson log-likelihood is

$$
\ell(\lambda) = \sum_{ij}\bigl[y_{ij}\log\lambda_{ij} - \lambda_{ij}\bigr] - \sum_{ij}\log y_{ij}!,
$$

a function of the **rates** and the data only. $c$ appears nowhere in it. Where
$c$ does appear is the feasible set,

$$
\mathcal{M}_c(K) = \Bigl\{\lambda \ge 0 \;:\; \log\!\left(1+\tfrac{\lambda}{c}\right) = LF',\ \ L,F\ge 0,\ \text{rank } K\Bigr\},
$$

and the fit is $\hat\lambda = \arg\max_{\lambda\in\mathcal{M}_c(K)}\ell(\lambda)$.
So $c$ never changes *what a good set of rates is*. It changes only *which rate
matrices are reachable*. That is the wedge that breaks the circle: if we can pin
down properties of $\hat\lambda$ without reference to $c$, the rest follows
mechanically.

## Step 2: the zero pattern of any fit is a union of K rectangles

Let $R_k = \{i : \ell_{ik} > 0\}$ and $C_k = \{j : f_{jk} > 0\}$ be the sets of
samples and features that use factor $k$. Because $L, F \ge 0$ there is no
cancellation, so $\theta_{ij} = \sum_k \ell_{ik}f_{jk} > 0$ exactly when some $k$
has $i \in R_k$ and $j \in C_k$. Therefore

$$
\operatorname{supp}(\theta) \;=\; \bigcup_{k=1}^{K} R_k \times C_k ,
$$

a union of $K$ combinatorial rectangles — and since $\lambda_{ij} = 0$ iff
$\theta_{ij}=0$, **the zero set of any fitted rate matrix is the complement of a
union of $K$ rectangles, at every $c$.**

Now add the data. If $y_{ij} > 0$ and $\hat\lambda_{ij}=0$ the log-likelihood is
$-\infty$, so a feasible fit must satisfy

$$
\{(i,j) : y_{ij} > 0\} \;\subseteq\; \bigcup_{k=1}^{K} R_k \times C_k .
$$

**The nonzero pattern of $Y$ must be covered by $K$ rectangles.** For count data
whose nonzeros are scattered — which is what Poisson sampling produces, since
$y_{ij}=0$ occurs independently with probability $e^{-\lambda_{ij}}$ — no small
collection of rectangles can cover the nonzeros while leaving much uncovered.
The rectangles are forced to be nearly the full matrix, so the fit's zero set is
nearly empty and $\hat\lambda > 0$ essentially everywhere.

This is the crucial step and it is **completely free of $c$**: it depends only on
$Y$'s nonzero pattern and on $K$. A rank-$K$ model cannot represent scattered
sampling zeros as structural zeros, so it smooths across them and returns a
strictly positive low-rank rate surface with some positive low end
$\lambda_{\text{low}}$ — a property of the data and the rank, not of the link.

## Step 3: c maps that floor into theta, deterministically

Given a low end $\lambda_{\text{low}}$ fixed by Step 2, the low end of $\theta$
follows with no further modelling:

$$
\theta_{\text{low}} \;=\; \log\!\left(1+\frac{\lambda_{\text{low}}}{c}\right),
\qquad
\frac{\theta_{\text{low}}}{\overline{\theta}}
\;=\;
\frac{\log(1+\lambda_{\text{low}}/c)}{\overline{\log(1+\hat\lambda/c)}} .
$$

The two limits are immediate. For $c \gg \lambda_{\text{low}}$ both numerator and
denominator linearize and the floor share tends to
$\lambda_{\text{low}}/\overline{\lambda}$ — small whenever the rates are spread
out. For $c \ll \lambda_{\text{low}}$ both behave like $\log(1/c)$ and the ratio
tends to **1**: θ becomes all floor.

So the same data, with the same fitted rates, produces a θ whose floor share runs
from $\lambda_{\text{low}}/\overline{\lambda}$ up to $1$ purely as a function of
$c$. Nothing about the fit changed; only the transform did.

## Step 4: a floor in theta forbids joint sparsity

This is the support-intersection argument from the previous section. A θ bounded
away from zero requires every $S_i$ to meet every $T_j$, and a single-factor
sample forces a fully dense factor. Combining the four steps:

$$
\underbrace{Y,\,K}_{\text{fix }\lambda_{\text{low}}}
\;\longrightarrow\;
\underbrace{\theta_{\text{low}} = \log(1+\lambda_{\text{low}}/c)}_{\text{deterministic in } c}
\;\longrightarrow\;
\underbrace{\text{support intersection}}_{\text{combinatorial}}
\;\longrightarrow\;
\text{joint density of } L \text{ and } F .
$$

Only the middle arrow involves $c$, and it involves nothing else.

## Testing it: predict identifiability before fitting

If the chain is right, the floor is predictable from $Y$ and $c$ with no model
fitted at all. Estimate the rates by the rank-1 independence model —
row totals times column totals over the grand total — and push them through the
link:

$$
m_{ij} = \frac{y_{i\cdot}\,y_{\cdot j}}{y_{\cdot\cdot}},
\qquad
\widehat{\text{floor share}} = \frac{q_\alpha\bigl(\log(1+m/c)\bigr)}{\overline{\log(1+m/c)}} .
$$

```{r fitfree, cache=TRUE}
QA <- 0.01     # a low quantile, not the minimum -- see the caveat below
pred_obs <- do.call(rbind, lapply(C_GRID, function(cf) {
  r <- do.call(rbind, lapply(1:10, function(sd) {
    Y <- sim_dataset(sd, 1)$Y
    m  <- outer(rowSums(Y), colSums(Y)) / sum(Y)        # fit-free rate estimate
    tg <- if (is.infinite(cf)) m else log1p(m / cf)

    x <- readRDS(file.path(cells_dir, paste0(
           tag_of(list(seed = sd, c_true = 1, c_fit = cf)), ".rds")))
    g <- lapply(c("rank1", "random"), function(i) {
      L <- x$fits[[i]]$LL; Fm <- x$fits[[i]]$FF
      th <- tcrossprod(L, Fm) / max(1, if (is.infinite(cf)) 1 else cf)
      c(fl = as.numeric(quantile(th, QA)) / mean(th), h = mean_hoyer(L)) })

    c(min_m = min(m),
      predicted = as.numeric(quantile(tg, QA)) / mean(tg),
      observed  = g[[1]]["fl"],
      init_disagreement = abs(g[[1]]["h"] - g[[2]]["h"]))
  }))
  as.data.frame(t(c(c_fit = cf, colMeans(r))))
}))
knitr::kable(pred_obs, digits = c(3, 4, 3, 3, 6), row.names = FALSE,
  caption = "min_m and 'predicted' use only Y and c. Nothing is fitted to produce them.")
```

```{r fitfree-cor}
lg <- function(x) log10(pmax(x, 1e-12))
c(`cor(predicted, observed)`             = cor(pred_obs$predicted, pred_obs$observed),
  `cor(predicted, log10 disagreement)`   = cor(pred_obs$predicted, lg(pred_obs$init_disagreement)),
  `cor(observed,  log10 disagreement)`   = cor(pred_obs$observed,  lg(pred_obs$init_disagreement)))
```

Two things to read here. The prediction tracks the realized floor essentially
exactly, and it also tracks **initialization disagreement** — how far apart two
starts land in Hoyer sparsity at the same $c$. That disagreement is a purely
empirical fingerprint of non-identifiability: same data, log-likelihoods equal to
within 0.008, different answer. Being able to forecast it from $Y$ and $c$ before
fitting is the sense in which the explanation is not circular.

Note also that `min_m` is *identical at every $c$* — it is a property of $Y$
alone. All of the variation in the predicted floor comes from the link.

## Two things that do not work

**Minimum-based statistics are fragile.** Using $\min\hat\theta$ rather than a low
quantile, the floor reads as $0$ for $c \ge 10$ because coordinate descent drives
a single coordinate to around $10^{-13}$, while the bulk of the low end sits near
$0.28$. One entry out of $90{,}000$ flips the diagnostic. The $1\%$ quantile is
stable and tracks the prediction with correlation $1.000$.

**$q_F$ computed on a fit is useless for the same reason.** The closed form
$q_F = K\min_{j,k} f_{jk}/s_j$ is exactly right for the *truth*, but evaluated on
fitted factors it returns $\sim 10^{-12}$ at every $c$ — one numerically-zero
coordinate destroys the minimum, reporting "no freedom" for factorizations that
are visibly loose. Replacing the minimum with a $5\%$ quantile restores it
(correlation $0.96$ with initialization disagreement); at a $25\%$ quantile it is
dead again, so the information really does live near the boundary.

## What this means for comparing fits across c

The fitted rates are pinned by the data; the split of sparsity between $L$ and
$F$ is not; and the *joint level* of sparsity is set by where $c$ places the
data's rate floor on the log1p curve. So:

- **Sparsity is not comparable across $c$.** "The $c = 0.01$ factors are denser
  than the $c = 10$ factors" is a statement about the link, not the data.
- **Quantities that depend on $LF'$ are comparable** — fitted rates, held-out
  log-likelihood, which samples resemble which. Use those for model comparison.
- **Report the floor share alongside any sparsity claim**, and fit from at least
  two initializations. If the floor is high and the two starts disagree, the
  sparsity being reported is a property of $c$.
- If the split genuinely matters for the science, it has to be pinned by
  something outside the likelihood: a sparsity penalty, a prior, or an anchor
  assumption.

# What to take from this

1. **The likelihood determines $LF'$, not $L$ and $F$ separately.** Any reported
   sparsity, or any statement that one factor is "cleaner" than another, is a
   statement about the point in $\mathcal{A}$ your optimizer happened to reach.

2. **Sparsity trades between the two factors.** The mixing family makes $L$
   denser and $F$ sparser simultaneously. Reporting sparsity for one without the
   other can be actively misleading.

   But the *joint level* is a separate question with its own mechanism. For
   exact sparsity it is support intersection: $\theta_{ij} > 0$ requires
   $\operatorname{supp}(\ell_i)$ and $\operatorname{supp}(f_j)$ to meet, so a
   $\theta$ bounded away from zero cannot be built from two sparse factors.
   For **Hoyer** sparsity, which needs no zeros, the continuous analogue is a
   multiplicative budget on peak-to-mean ratios,
   $P_L(k)P_F(k)w_k \le P_\theta$, together with $H(x)\lesssim 1-1/\sqrt{P(x)}$.
   Both factors can only be peaked if $\theta$ has range to spend, and that
   bound is essentially tight in practice.

3. **Zeros are what buy identifiability**, and they must be *exact* or nearly so.
   $q_F$ is zero as soon as one $f_{jk}$ is, and grows smoothly as $F$ retreats
   from the boundary.

4. **$c$ controls both effects through the level of $\theta$.** Small $c$ forces
   $\theta_{\min} = \log(1+\lambda_{\min}/c)$ up while the spread saturates at
   $\log(\lambda_{\max}/\lambda_{\min})$, so $\theta$ tends to a constant: the
   ambiguity tends to total *and* the floor forbids joint sparsity. The right
   comparison is $c$ against $\lambda_{\min}$, not $c$ on its own.

   The exception is exact zeros in $\lambda$, which survive every $c$ since the
   link maps zero to zero. Given them, the sparsity trend in $c$ disappears
   entirely — which is the test that separates "inherent to the model" from
   "a property of this data".

5. **None of this needs the truth.** The likelihood objective contains no $c$;
   $c$ only restricts the feasible set. A rank-$K$ model cannot represent
   scattered sampling zeros as structural zeros -- its zero set is always the
   complement of a union of $K$ rectangles -- so it returns a strictly positive
   rate surface whose low end is fixed by $Y$ and $K$. $c$ then decides where
   that floor lands on the log1p curve, and the floor is what forbids joint
   sparsity. Every link is checkable on your own data: a rank-1 estimate of the
   rates pushed through the link predicts the realized floor, and predicts
   initialization disagreement, before any model is fitted.

6. **Practical consequence.** At small $c$, prefer conclusions that depend on
   $LF'$ (fitted rates, held-out likelihood, which samples resemble which) over
   conclusions that depend on the split (how sparse the factors are, how many
   features a factor "uses"). Sparsity is not comparable across $c$. If the split
   matters, it has to be pinned by something outside the likelihood -- a sparsity
   penalty, a prior, or an anchor assumption.

```{r session}
sessionInfo()
```
