---
title: "Non-identifiability of the log1p factorization, and how c controls it"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-02"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r pkgs}
library(ggplot2)
library(cowplot)
library(log1pNMF)
theme_set(theme_cowplot(font_size = 11))
```

```{r paths}
exp_dir <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
source(file.path(exp_dir, "c_grid_sim_common.R"))   # generator, for the examples

hoyer <- function(v) { v <- abs(v); n <- length(v)
                       (sqrt(n) - sum(v)/sqrt(sum(v^2))) / (sqrt(n) - 1) }
mean_hoyer <- function(M) mean(apply(M, 2, hoyer))
```

Fitting the log1p model at different $c$ gives factorizations whose sparsity
differs a lot, even when the log-likelihoods are indistinguishable. This page
works out why, in general rather than for one simulation: what exactly is
unidentified, how much freedom there is, and how $c$ controls that freedom.

The short version, stated up front:

> The likelihood sees only $\theta = LF'$. For a **fixed** $\theta$, $c$ has no
> effect on identifiability whatsoever. What $c$ changes is *which* $\theta$ is
> needed to represent data with given rates — and as $c \to 0$ that $\theta$
> becomes a large constant plus a bounded perturbation, which is precisely the
> shape with the largest ambiguity.

# What the likelihood can and cannot see

The model is

$$
y_{ij}\sim\text{Poisson}(\lambda_{ij}), \qquad
\lambda_{ij} = c\left(e^{\theta_{ij}}-1\right), \qquad
\theta = LF', \quad L\in\mathbb{R}^{n\times K}_{\ge 0},\ F\in\mathbb{R}^{p\times K}_{\ge 0}.
$$

The map $\theta \mapsto \lambda$ is a fixed, strictly monotone, elementwise
transform once $c$ is chosen. So the log-likelihood is a function of $\theta$
alone: **any two parameter values with the same product $LF'$ have exactly the
same likelihood, for every $c$.** Whatever is not determined by $\theta$ is not
determined by the data, no matter how much data there is.

# The ambiguity group

Suppose $\theta$ has rank $K$ and admits a nonnegative rank-$K$ factorization
$\theta = LF'$. If $\tilde L\tilde F' = LF'$ is another one, the two share a
column space, so $\tilde L = LA$ for some invertible $A$, and then
$\tilde F' = A^{-1}F'$. Hence **every** alternative factorization has the form

$$
(L, F) \;\longmapsto\; \bigl(LA,\; FA^{-\top}\bigr),
\qquad A \in \mathcal{A}(L,F) := \bigl\{A \ \text{invertible} \ K\times K : LA \ge 0,\; FA^{-\top} \ge 0\bigr\}.
$$

Invertibility is all that is asked of $A$ itself, and it is forced by the
construction, since $A^{-1}$ has to exist. In particular $A$ is **not** required
to be nonnegative — only the two *products* $LA$ and $FA^{-\top}$ are. That is
where the whole ambiguity lives: if $A$ and $A^{-1}$ were both forced
nonnegative, $A$ would have to be monomial and there would be nothing to
discuss.

$\mathcal{A}$ always contains the **monomial** matrices — a permutation times a
positive diagonal — which merely relabel and rescale the factors. That part is
harmless. Non-identifiability in the sense that matters is $\mathcal{A}$
containing anything else.

## A geometric reading

Write $a_k$ for the $k$th column of $A$ and $\tilde a_k$ for the $k$th *row* of
$A^{-1}$. Then

$$
(LA)_{ik} = \langle \ell_i, a_k\rangle, \qquad
(FA^{-\top})_{jk} = \langle f_j, \tilde a_k\rangle,
$$

so the constraints are

$$
a_k \in \mathcal{C}_L^{*} \ \ \forall k,
\qquad
\tilde a_k \in \mathcal{C}_F^{*} \ \ \forall k,
$$

where $\mathcal{C}_L$ is the cone generated by the rows of $L$ and
$\mathcal{C}_L^{*} = \{y : \langle \ell_i, y\rangle \ge 0\ \forall i\}$ is its
dual, and likewise for $F$.

This makes the picture concrete. Since $L \ge 0$, the cone $\mathcal{C}_L$ sits
inside the nonnegative orthant, so its dual *contains* the orthant. The more
$\mathcal{C}_L$ fills the orthant, the tighter its dual, and the less room $A$
has.

**Uniqueness by separability.** If $L$ has a *pure* row for each factor — a
sample loading on one factor only — then $\mathcal{C}_L$ is the whole orthant, so
$\mathcal{C}_L^{*}$ is the orthant and $A \ge 0$. If $F$ likewise has a pure row
per factor, then $A^{-1} \ge 0$. A nonnegative matrix with a nonnegative inverse
is monomial, so the factorization is unique up to relabelling. Pure rows —
equivalently, *zeros in the right places* — are what pin an NMF down.

# Where c enters

Nothing above mentions $c$. For a fixed $\theta$ the ambiguity set
$\mathcal{A}$ is fixed, so **$c$ cannot affect identifiability directly**. It
enters one step earlier: to model data whose rates are $\lambda$, the required
linear predictor is

$$
\theta_c = \log\!\left(1 + \frac{\lambda}{c}\right),
$$

and the *shape* of $\theta_c$ — hence the geometry above — depends strongly on
$c$. Two limits make this sharp. Let $\lambda_{\min}$ and $\lambda_{\max}$ be the
smallest and largest rates.

**Large $c$.** $\theta_c \approx \lambda/c$. The shape of $\theta$ is the shape of
$\lambda$, exactly. Any zeros or near-zeros in $\lambda$ survive into $\theta$,
and with them the pure rows that pin the factorization.

**Small $c$, with $\lambda_{\min} > 0$.**

$$
\theta_{\min} = \log\!\left(1+\frac{\lambda_{\min}}{c}\right) \xrightarrow[c\to 0]{} \infty,
\qquad
\theta_{\max}-\theta_{\min}
= \log\frac{c+\lambda_{\max}}{c+\lambda_{\min}}
\xrightarrow[c\to 0]{} \log\frac{\lambda_{\max}}{\lambda_{\min}},
$$

which is **finite**. The level of $\theta$ diverges while its spread stays
bounded, so $\theta / \theta_{\min} \to \mathbf{1}\mathbf{1}'$. As $c\to 0$ the
linear predictor becomes a huge constant plus a bounded perturbation.

A constant matrix is the *least* identifiable thing a nonnegative factorization
can be asked to represent: with $\theta \approx \kappa\,\mathbf{1}\mathbf{1}'$
every row of $L$ points at the centre of the simplex, $\mathcal{C}_L$ collapses
to a ray far from the orthant boundary, and $\mathcal{C}_L^{*}$ becomes enormous.

The one escape is $\lambda_{\min} = 0$: exact zeros in the rates give
$\theta_{ij}=0$ for **every** $c$, since $\log(1+0/c) = 0$. Exact zeros are
preserved by the link; it is *near*-zeros that get washed out.

```{r two-limits}
## the two limits, on the rates from one simulated dataset
lam <- sim_dataset(1, 1)$lambda
lam <- lam[lam > 0]
shape <- function(cc) { th <- log1p(lam / cc)
  c(level = min(th), spread = max(th) - min(th),
    `spread / level` = (max(th) - min(th)) / min(th)) }
knitr::kable(t(sapply(c(1e-4, 1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3), shape)) |>
             `rownames<-`(format(c(1e-4, 1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3))),
             digits = 3,
             caption = "theta = log1p(lambda/c) for one fixed set of rates. The spread saturates at log(lambda_max/lambda_min) while the level grows without bound.")
```

# How much freedom, concretely

The full set $\mathcal{A}$ is awkward — it constrains $A$ and $A^{-1}$
simultaneously, so it is not convex. A one-parameter slice through it is enough
to get closed forms and the right intuition. Let $J = \mathbf{1}\mathbf{1}'$ and

$$
B_t \;=\; (1-t)\,I \;+\; \frac{t}{K}J,
\qquad
B_t^{-1} \;=\; \frac{1}{1-t}\left(I - \frac{t}{K}J\right).
$$

$t$ is a **mixing fraction**: $t=0$ is the identity, and $t\to 1$ blends all $K$
factors into one. Writing the two sides out, with $r_i = \sum_k \ell_{ik}$ and
$s_j = \sum_k f_{jk}$,

$$
(LB_t)_{ik} = (1-t)\,\ell_{ik} + \frac{t}{K}\,r_i,
\qquad
(FB_t^{-1})_{jk} = \frac{1}{1-t}\left(f_{jk} - \frac{t}{K}\,s_j\right).
$$

So increasing $t$ pulls each row of $L$ toward its own mean — **denser** — while
subtracting a per-feature constant from $F$ — **sparser**. This is the essential
point: sparsity is not lost, it is *moved between the two factors*. Nothing here
is specific to $L$.

## The feasible interval, in closed form

Both expressions are linear in $t$, so the constraints resolve exactly. Define
the smallest normalized share on each side,

$$
q_L = K\min_{i,k}\frac{\ell_{ik}}{r_i}\in[0,1],
\qquad
q_F = K\min_{j,k}\frac{f_{jk}}{s_j}\in[0,1].
$$

For $t \ge 0$ the $L$ constraint is automatic (both terms are nonnegative), and
the $F$ constraint is $f_{jk}\ge \tfrac{t}{K}s_j$ for all $j,k$. For $t<0$ the
roles swap. Hence

$$
\boxed{\;-\frac{q_L}{1-q_L} \;\le\; t \;\le\; q_F\;}
$$

$q_F$ is exactly *how far $F$ is from having a zero*: it is $0$ as soon as any
single $f_{jk}=0$, and $1$ only if every feature loads equally on all factors.
The same reading applies to $q_L$ on the other side. **A single exact zero in $F$
removes the entire positive branch of this family.**

```{r verify-interval}
d0 <- sim_dataset(1, 1e-3)
L0 <- d0$truth$L; F0c <- d0$FF_true; KK <- ncol(L0)

qL <- KK * min(L0  / rowSums(L0))
qF <- KK * min(F0c / rowSums(F0c))

feasible <- function(t) {
  B <- (1 - t) * diag(KK) + (t / KK) * matrix(1, KK, KK)
  min(L0 %*% B) >= -1e-12 && min(F0c %*% t(solve(B))) >= -1e-12
}
bisect <- function(sgn) { lo <- 0; hi <- 1
  for (i in 1:60) { m <- (lo + hi)/2; if (feasible(sgn*m)) lo <- m else hi <- m }; sgn*lo }

knitr::kable(data.frame(
  bound   = c("t_min", "t_max"),
  formula = c(-qL/(1 - qL), qF),
  numeric = c(bisect(-1), bisect(+1))),
  digits = 8, caption = "Closed form against bisection, at c = 0.001.")
```

## Asymptotics in c

Under the small-$c$ limit above, $\theta \approx \kappa\mathbf{1}\mathbf{1}' +
\theta^{*}$ with $\kappa = \theta_{\min}\to\infty$. Normalizing $L$ so that
$r_i = 1$, the constant is carried by $F$: $f_{jk} = f^{*}_{jk} + \kappa$. Then

$$
q_F = K\min_j \frac{\min_k f^{*}_{jk} + \kappa}{S^{*}_j + K\kappa}
\;\xrightarrow[\kappa\to\infty]{}\; 1,
\qquad
1 - q_F \;\approx\; \frac{\max_j\bigl(S^{*}_j - K\min_k f^{*}_{jk}\bigr)}{K\kappa},
$$

with $S^{*}_j = \sum_k f^{*}_{jk}$. So the freedom approaches *total* as $c\to0$,
and it does so at rate $1/\kappa = 1/\log(1/c)$ — slowly, but without bound.
Conversely $\kappa \to 0$ recovers $q_F = K\min_{j,k}f^{*}_{jk}/S^{*}_j$, which is
zero the moment $F^{*}$ has a zero.

The offset is not a modelling artifact: it is pinned by the smallest rate,

$$
\kappa = \theta_{\min} = \log\!\left(1 + \frac{\lambda_{\min}}{c}\right).
$$

# Numerical illustration

Using the calibrated generator (see the [c-grid page](c_grid_sim.html)), which
builds a truth at each $c$ with matched marginals, we can read off the feasible
interval and the sparsity it buys.

```{r sweep, cache=TRUE}
tr <- sim_truth(1)
sweep_c <- do.call(rbind, lapply(C_GRID, function(cc) {
  dd <- sim_dataset(1, cc, tr)
  L <- dd$truth$L; Fm <- dd$FF_true
  th <- dd$alpha * tr$theta0 + dd$beta
  qF <- KK * min(Fm / rowSums(Fm))
  B  <- (1 - qF) * diag(KK) + (qF / KK) * matrix(1, KK, KK)
  data.frame(c = cc,
             `theta level / spread` = min(th) / (max(th) - min(th)),
             q_F = qF,
             `Hoyer L: truth`   = mean_hoyer(L),
             `Hoyer L: at t_max`= mean_hoyer(L %*% B),
             `Hoyer F: truth`   = mean_hoyer(Fm),
             `Hoyer F: at t_max`= mean_hoyer(Fm %*% t(solve(B))),
             check.names = FALSE)
}))
knitr::kable(sweep_c, digits = 3, row.names = FALSE)
```

```{r sweep-plot, fig.width=9, fig.height=4}
lab <- ifelse(is.infinite(sweep_c$c), "∞",
              format(sweep_c$c, scientific = FALSE, trim = TRUE, drop0trailing = TRUE))
sweep_c$cf <- factor(lab, levels = lab)

p1 <- ggplot(sweep_c, aes(cf, q_F, group = 1)) +
  geom_line(colour = "#0E5C6B", linewidth = 1) + geom_point(colour = "#0E5C6B", size = 2) +
  ylim(0, 1) +
  labs(x = "c", y = expression(q[F]),
       title = "Freedom in the mixing family",
       subtitle = "0 = pinned, 1 = the factors can be blended completely")

rng <- rbind(
  data.frame(cf = sweep_c$cf, side = "L", lo = sweep_c$`Hoyer L: at t_max`,
             hi = sweep_c$`Hoyer L: truth`),
  data.frame(cf = sweep_c$cf, side = "F", lo = pmin(sweep_c$`Hoyer F: truth`, sweep_c$`Hoyer F: at t_max`),
             hi = pmax(sweep_c$`Hoyer F: truth`, sweep_c$`Hoyer F: at t_max`)))

p2 <- ggplot(rng, aes(cf, ymin = lo, ymax = hi, colour = side)) +
  geom_linerange(linewidth = 2.2, alpha = 0.85,
                 position = position_dodge(width = 0.35)) +
  scale_colour_manual(values = c(L = "#0E8FA8", F = "#C2410C"), name = NULL) +
  labs(x = "c", y = "achievable mean Hoyer sparsity",
       title = "Sparsity is not determined by the data",
       subtitle = "range spanned without changing the likelihood at all")

plot_grid(p1, p2, nrow = 1, rel_widths = c(1, 1.15))
```

At $c=\infty$ the interval is a point: the factorization is effectively pinned.
At $c=10^{-3}$ the sparsity of $L$ can be driven from the truth's value down by
roughly an order of magnitude, with $F$ moving the other way, and the likelihood
does not change by so much as a rounding error.

## The slice is a lower bound

$B_t$ is one line through $\mathcal{A}$. Searching over all of $A\in\mathbb{R}^{K
\times K}$ gives a slightly wider range, so the closed forms above understate the
true freedom.

```{r full-search, cache=TRUE}
explore <- function(L, Fm, sgn, nstart = 25, seed = 1) {
  K <- ncol(L); set.seed(seed)
  obj <- function(v) {
    A <- matrix(v, K, K); dt <- det(A)
    if (!is.finite(dt) || abs(dt) < 1e-8) return(1e6)
    LA <- L %*% A; FA <- Fm %*% t(solve(A))
    pen <- sum(pmin(LA, 0)^2) + sum(pmin(FA, 0)^2)
    if (pen > 0) return(1e3 + 1e6 * pen)
    -sgn * mean_hoyer(LA)
  }
  best <- NULL; bv <- Inf
  for (s in 1:nstart) {
    o <- try(optim(as.vector(diag(K) + matrix(rnorm(K*K, 0, .15), K, K)), obj,
                   control = list(maxit = 3000, reltol = 1e-10)), silent = TRUE)
    if (!inherits(o, "try-error") && o$value < bv) { bv <- o$value; best <- o$par }
  }
  mean_hoyer(L %*% matrix(best, K, K))
}

knitr::kable(do.call(rbind, lapply(c(1e-3, 1, Inf), function(cc) {
  dd <- sim_dataset(1, cc, tr); L <- dd$truth$L; Fm <- dd$FF_true
  qF <- KK * min(Fm / rowSums(Fm))
  B  <- (1 - qF) * diag(KK) + (qF / KK) * matrix(1, KK, KK)
  data.frame(c = cc, `truth` = mean_hoyer(L),
             `mixing slice, lowest` = mean_hoyer(L %*% B),
             `full search, lowest`  = explore(L, Fm, -1),
             check.names = FALSE)
})), digits = 3, row.names = FALSE, caption = "Mean Hoyer sparsity of L.")
```

# What to take from this

1. **The likelihood determines $LF'$, not $L$ and $F$ separately.** Any reported
   sparsity, or any statement that one factor is "cleaner" than another, is a
   statement about the point in $\mathcal{A}$ your optimizer happened to reach.

2. **Sparsity trades between the two factors.** The mixing family makes $L$
   denser and $F$ sparser simultaneously. Reporting sparsity for one without the
   other can be actively misleading.

3. **Zeros are what buy identifiability**, and they must be *exact* or nearly so.
   $q_F$ is zero as soon as one $f_{jk}$ is, and grows smoothly as $F$ retreats
   from the boundary.

4. **$c$ controls this through the level of $\theta$.** Small $c$ forces
   $\theta_{\min} = \log(1+\lambda_{\min}/c)$ up while the spread saturates at
   $\log(\lambda_{\max}/\lambda_{\min})$, so $\theta$ tends to a constant and the
   ambiguity tends to total. The exception is exact zeros in $\lambda$, which
   survive every $c$.

5. **Practical consequence.** At small $c$, prefer conclusions that depend on
   $LF'$ (fitted rates, held-out likelihood, which samples resemble which) over
   conclusions that depend on the split (how sparse the factors are, how many
   features a factor "uses"). If the split matters, it has to be pinned by
   something outside the likelihood — a sparsity penalty, a prior, or an anchor
   assumption.

```{r session}
sessionInfo()
```
