---
title: "When does max-1 loading scaling coincide with topic proportions?"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
date: "2026-08-09"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# The question

In the BBC structure plots, the $c = \infty$ document scores all total
exactly 1 — they look like the topic model's own proportion normalization,
even though the figures apply the paper's loading-scaling rule: divide the
loadings by the size factors, then scale each column of $\mathbf{L}$ so its
maximum is 1. One might wonder whether the rule was actually applied there.

This page proves that no discrepancy exists: the plotted scores **are** the
paper's rule, and the rule coincides with the proportion normalization
exactly when every factor contains a "pure" document. The BBC fit satisfies
that condition; the pancreas fit does not, and its bars visibly exceed 1
under the very same code.

# Setup

A Poisson NMF fit of an $n \times p$ count matrix $Y$:
$y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$ with
$\Lambda = \mathbf{L}\mathbf{F}^{\top}$, $\mathbf{L}, \mathbf{F} \geq 0$.
Write $y_{i\cdot} = \sum_j y_{ij}$ for the row totals,
$F_k = \sum_j f_{jk}$ for the factor totals, and let the size factors be
$s_i = y_{i\cdot}/\bar{y}$ for any convenient constant $\bar{y}$ (only
ratios within a column will matter).

Two normalizations of the loadings:

- **The paper's rule**, as implemented in the figures:
  $$a_{ik} \;=\; \frac{l_{ik}/s_i}{\max_{i'}\left(l_{i'k}/s_{i'}\right)},$$
  i.e. `normalize_bars(diag(1/s) %*% L)`.
- **Topic proportions** (`fastTopics::poisson2multinom`):
  $$p_{ik} \;=\; \frac{l_{ik} F_k}{\sum_{m} l_{im} F_m},
  \qquad \textstyle\sum_k p_{ik} = 1 .$$

# Lemma: fitted row totals equal observed row totals

At any local maximum of the Poisson NMF log-likelihood (with no row of
$\mathbf{L}$ identically zero),
$$\sum_j \lambda_{ij} \;=\; y_{i\cdot} \qquad \text{for every } i.$$

**Proof.** Scaling row $i$ of $\mathbf{L}$ by $t > 0$ is a feasible move that
scales row $i$ of $\Lambda$ by $t$ and leaves all other rows untouched. The
terms of the log-likelihood involving row $i$ are
$\sum_j \left(y_{ij}\log \lambda_{ij} - \lambda_{ij}\right)$, whose
derivative in $t$ at $t = 1$ is $y_{i\cdot} - \sum_j \lambda_{ij}$. At a
maximum this must vanish. $\square$

# Proposition: the rule is a column rescaling of the proportions

$$a_{ik} \;=\; \frac{p_{ik}}{\max_{i'} p_{i'k}} .$$

**Proof.** By the lemma, the denominator of $p_{ik}$ is
$\sum_m l_{im}F_m = \sum_j \lambda_{ij} = y_{i\cdot} = \bar{y}\, s_i$, so
$$p_{ik} \;=\; \frac{l_{ik} F_k}{\bar{y}\, s_i}
\;=\; \frac{l_{ik}}{s_i}\cdot\frac{F_k}{\bar{y}} .$$
Within a column $k$, the factor $F_k/\bar{y}$ is a constant, so
$p_{ik} \propto l_{ik}/s_i$ across $i$, and dividing by the column maximum
cancels the constant:
$$\frac{p_{ik}}{\max_{i'} p_{i'k}}
= \frac{l_{ik}/s_i}{\max_{i'}\left(l_{i'k}/s_{i'}\right)} = a_{ik}. \square$$

**Corollary 1 (the pure-document condition).**
$a_{ik} = p_{ik}$ for all $i, k$ — i.e. the max-1 rule reproduces the
topic proportions exactly — **if and only if** every factor has a pure
document: $\max_i p_{ik} = 1$ for every $k$.

**Corollary 2 (bars can only exceed 1).** The plotted bar heights satisfy
$$\sum_k a_{ik} \;=\; \sum_k \frac{p_{ik}}{\max_{i'} p_{i'k}}
\;\geq\; \sum_k p_{ik} \;=\; 1 ,$$
with equality for document $i$ exactly when every factor that document
touches has a pure document somewhere. So under this rule, bars totalling
exactly 1 is the *extreme* case, attained when the fit is clustered enough
that each factor has anchor documents — it is a property of the fit, not a
different normalization.

# Verification on the paper's fits

`A` below is computed exactly as in the figure scripts; `P` by
`poisson2multinom`.

```{r verify}
suppressMessages(library(fastTopics))
load(path.expand("~/Documents/log1pNMF/inst/paper_figures/data/experiment_results.Rdata"))
normalize_bars <- function(M) sweep(M, 2, apply(M, 2, max), "/")

out <- lapply(c("bbc", "pancreas"), function(ds) {
  tm <- res_list[[ds]][["Inf"]]
  s  <- res_list[[ds]][["1"]]$s
  A  <- normalize_bars(as.matrix(tm$L) / s)
  P  <- poisson2multinom(tm)$L
  ## the lemma, checked directly: fitted row totals must be proportional to
  ## the size factors (i.e., to the observed row totals); r_i constant <=> lemma
  Fk <- colSums(tm$F)
  r  <- as.vector(as.matrix(tm$L) %*% Fk) / s
  data.frame(
    dataset = ds, K = ncol(A),
    `lemma: max/min of fitted rowsum / s` = round(max(r)/min(r), 6),
    `min over k of max_i p_ik` = round(min(apply(P, 2, max)), 3),
    `max |A - P|` = signif(max(abs(A - P)), 3),
    `bar height median` = round(median(rowSums(A)), 3),
    `bar height max` = round(max(rowSums(A)), 3),
    check.names = FALSE)
})
knitr::kable(do.call(rbind, out), row.names = FALSE)
```

```{r purity}
tmb <- res_list$bbc[["Inf"]]
Pb  <- poisson2multinom(tmb)$L
tmp <- res_list$pancreas[["Inf"]]
Pp  <- poisson2multinom(tmp)$L
knitr::kable(data.frame(
  topic = seq_len(10),
  `bbc: max_i p_ik` = round(apply(Pb, 2, max), 4)),
  caption = "BBC: every one of the 10 topics has a pure document, so
  Corollary 1 applies and A = P (up to convergence tolerance).")
knitr::kable(data.frame(
  factor = seq_len(13),
  `pancreas: max_i p_ik` = round(apply(Pp, 2, max), 3)),
  caption = "Pancreas: several factors have no pure cell, so A != P and the
  bars exceed 1 (up to 1.34) under the very same rule.")
```

# Conclusion

The $c = \infty$ document scores in the BBC figures are normalized exactly
by the paper's max-1 rule. Their coincidence with topic-model proportions is a
theorem, not a discrepancy: the rule equals the proportion normalization if
and only if every factor has a document with proportion one, which the BBC
$c = \infty$ fit satisfies for all ten factors: the plotted matrix equals the
proportions matrix up to the fit's convergence tolerance (max absolute
difference 1.4e-5, the same order as the lemma's residual in the table). The pancreas $c = \infty$ fit
fails the condition and its bars exceed one under the same code — the rule
at work, visibly.
