---
title: The MCF-7 data set as possible model system for the Poisson log1p NMF model
author: Peter Carbonetto
output:
  workflowr::wflow_html:
    toc: no
    theme: readable
    highlight: textmate
    lib_dir: site_libs
    self_contained: no
---

Here I explore the possibility of the data from the
[Sanford et al eLife paper][sanford2020elife] as a possible "model
system" for the Poisson log1p NMF model. This paper analyzed the
transcriptional response of human MCF-7 cells to retinoic acid and
TGF-$\beta$, applied individually and in combination.

The following abbreviations were used in their data:
EtOH = ethanol; RA = retinoic acid; TGFb = TGF-$\beta$.

Note that the data were downloaded from GEO, accession GSE152749.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the fastTopics and flashier packages, and a few other packages
needed to retrieve the data and create the plots.

```{r load-pkgs, message=FALSE}
library(fastTopics)
library(flashier)
library(data.table)
library(GEOquery)
library(ggplot2)
library(cowplot)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

## Prepare the data for analysis with fastTopics and flashier

Load the gene information:

```{r load-genes}
genes <- fread("data/Human.GRCh38.p13.annot.tsv.gz",
               sep = "\t",header = TRUE,stringsAsFactors = FALSE)
class(genes) <- "data.frame"
```

Load the RNA-seq counts:

```{r load-rnaseq}
counts <- fread("data/GSE152749_raw_counts_GRCh38.p13_NCBI.tsv.gz",
                sep = "\t",header = TRUE,stringsAsFactors = FALSE)
class(counts) <- "data.frame"
rownames(counts) <- counts$GeneID
counts <- counts[,-1]
counts <- as.matrix(counts)
storage.mode(counts) <- "double"
counts <- t(counts)
ids <- rownames(counts)
```

Load the sample information:

```{r load-sample-info, message=FALSE}
geo <- getGEO(filename = "data/GSE152749_family.soft.gz")
samples <- data.frame(id = names(GSMList(geo)),
                      treatment = sapply(GSMList(geo),
                                         function (x) Meta(x)$title))
samples <- samples[ids,]
rownames(samples) <- NULL
samples <- transform(samples,
                     EtOH = grepl("EtOH",treatment,fixed = TRUE),
                     RA   = grepl("RA",treatment,fixed = TRUE),
                     TGFb = grepl("TGFb",treatment,fixed = TRUE))
samples$label                 <- "EtOH"
samples[samples$RA,"label"]   <- "RA"
samples[samples$TGFb,"label"] <- "TGFb"
samples[with(samples,RA & TGFb),"label"] <- "RA+TGFb"
samples <- transform(samples,
                     label = factor(label,c("EtOH","RA","TGFb","RA+TGFb")))
```

Remove the genes that are expressed in fewer than 4 samples:

```{r filter-genes}
x <- colSums(counts > 0)
i <- which(x > 3)
genes  <- genes[i,]
counts <- counts[,i]
```

## Topic model

Let's first try fitting a (multinomial) topic model to these
data. After a bit of trial and error, it turns out that 4 topics are
needed to get the right result:

```{r fit-topic-model}
fit0 <- fit_poisson_nmf(counts,k = 4,init.method = "random",numiter = 50,
                        verbose = "none",
                        control = list(nc = 4,extrapolate = FALSE))
fit <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 50,verbose = "none",
                       control = list(nc = 4,extrapolate = TRUE))
```

The topic model identifies one topic for the transcriptional responses
to RA, one topic for the responses TGF-$\beta$, and one topic for the
"control" or "baseline" (ethanol), and the response from exposure to
RA + TGF-$\beta$ is a fairly even mixture of the RA and TGF-$\beta$
topics.

```{r structure-plot-topic-model, fig.height=1.75, fig.width=5}
topic_colors <- c("tomato","darkblue","gold","skyblue")
p1 <- structure_plot(fit,grouping = samples$label,topics = 1:4,
                     colors = topic_colors)
p1
```

## NMF of the shifted log counts

ADD TEXT AND CODE HERE.

[sanford2020elife]: https://doi.org/10.7554/eLife.59388
