---
title: The MCF-7 data set as possible model system for the Poisson log1p NMF model
author: Peter Carbonetto
output:
  workflowr::wflow_html:
    toc: no
    theme: readable
    highlight: textmate
    lib_dir: site_libs
    self_contained: no
---

Here I explore the possibility of the data from the
[Sanford et al eLife paper][sanford2020elife] as a possible "model
system" for the Poisson log1p NMF model. This paper analyzed the
transcriptional response of human MCF-7 cells to retinoic acid and
TGF-$\beta$, applied individually and in combination.

The following abbreviations were used in their data:
EtOH = ethanol; RA = retinoic acid; TGFb = TGF-$\beta$.

Note that the data were downloaded from GEO, accession GSE152749.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load the fastTopics and flashier packages, and a few other packages
needed to retrieve the data and create the plots.

```{r load-pkgs, message=FALSE}
library(rsvd)
library(fastglmpca)
library(fastTopics)
library(flashier)
library(data.table)
library(GEOquery)
library(ggplot2)
library(cowplot)
library(passPCA)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

## Prepare the data for analysis with fastTopics and flashier

Load the gene information:

```{r load-genes}
genes <- fread("data/Human.GRCh38.p13.annot.tsv.gz",
               sep = "\t",header = TRUE,stringsAsFactors = FALSE)
class(genes) <- "data.frame"
```

Load the RNA-seq counts:

```{r load-rnaseq}
counts <- fread("data/GSE152749_raw_counts_GRCh38.p13_NCBI.tsv.gz",
                sep = "\t",header = TRUE,stringsAsFactors = FALSE)
class(counts) <- "data.frame"
rownames(counts) <- counts$GeneID
counts <- counts[,-1]
counts <- as.matrix(counts)
storage.mode(counts) <- "double"
counts <- t(counts)
ids <- rownames(counts)
```

Load the sample information:

```{r load-sample-info, message=FALSE}
geo <- getGEO(filename = "data/GSE152749_family.soft.gz")
samples <- data.frame(id = names(GSMList(geo)),
                      treatment = sapply(GSMList(geo),
                                         function (x) Meta(x)$title))
samples <- samples[ids,]
rownames(samples) <- NULL
samples <- transform(samples,
                     EtOH = grepl("EtOH",treatment,fixed = TRUE),
                     RA   = grepl("RA",treatment,fixed = TRUE),
                     TGFb = grepl("TGFb",treatment,fixed = TRUE))
samples$label                 <- "EtOH"
samples[samples$RA,"label"]   <- "RA"
samples[samples$TGFb,"label"] <- "TGFb"
samples[with(samples,RA & TGFb),"label"] <- "RA+TGFb"
samples <- transform(samples,
                     label = factor(label,c("EtOH","RA","TGFb","RA+TGFb")))
```

Remove the genes that are expressed in fewer than 4 samples:

```{r filter-genes}
x <- colSums(counts > 0)
i <- which(x > 3)
genes  <- genes[i,]
counts <- counts[,i]
```

## PCA

Let's see what happens when we apply PCA to the shifted log counts.

First compute the shifted log counts:

```{r shifted-log-counts}
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
Y <- log1p(counts/(a*s))
```

Now run PCA on the shifted log counts, and plot the first 2 PCs:

```{r pca, fig.height=2, fig.width=3}
pca <- rpca(Y,k = 4,center = TRUE,scale = FALSE)
colnames(pca$x) <- paste0("PC",1:4)
pdat <- data.frame(samples,pca$x)
ggplot(pdat,aes(x = PC1,y = PC2,color = label)) +
  geom_point() +
  scale_color_manual(values = c("dodgerblue","tomato","darkblue",
                                "limegreen")) + 
  theme_cowplot(font_size = 10)
```

## fastglmpca

Just for interest, let's see what happens when we compute a GLM-PCA
reduced representation of the counts:

```{r fastglmpca, fig.height=2, fig.width=3, message=FALSE}
fit_glmpca <- init_glmpca_pois(t(counts),K = 3)
fit_glmpca <- fit_glmpca_pois(t(counts),fit0 = fit_glmpca,verbose = FALSE)
pdat <- data.frame(samples,fit_glmpca$V)
ggplot(pdat,aes(x = k_1,y = k_2,color = label)) +
  geom_point() +
  scale_color_manual(values = c("dodgerblue","tomato","darkblue",
                                "limegreen")) + 
  theme_cowplot(font_size = 10)
```

## Topic model

Let's now try fitting a (multinomial) topic model to these data,
which represents an *additive model* of the transcriptional changes in
response to the different exposures. After a bit of trial and error,
it turns out that 4 topics are needed to get the right result:

```{r fit-topic-model}
fit0 <- fit_poisson_nmf(counts,k = 4,init.method = "random",numiter = 50,
                        verbose = "none",
                        control = list(nc = 4,extrapolate = FALSE))
fit <- fit_poisson_nmf(counts,fit0 = fit0,numiter = 50,verbose = "none",
                       control = list(nc = 4,extrapolate = TRUE))
```

The topic model identifies one topic for the transcriptional responses
to RA, one topic for the responses TGF-$\beta$, and one topic for the
"control" or "baseline" (ethanol), and the response from exposure to
RA + TGF-$\beta$ is a fairly even mixture of the RA and TGF-$\beta$
topics.

```{r structure-plot-topic-model, fig.height=1.75, fig.width=5}
topic_colors <- c("tomato","darkblue","gold","skyblue")
p1 <- structure_plot(fit,grouping = samples$label,topics = 1:4,
                     colors = topic_colors)
p1
```

## NMF of the shifted log counts

Now let's try fitting an NMF to the shifted log counts using flashier.
Implicitly, this model is a *multiplicative model* of the
transcriptional changes in response to the different exposures.

Next fit an NMF to the shifted log counts using flashier:

```{r flashier-nmf, cache=TRUE}
n  <- nrow(counts)
x  <- rpois(1e7,1/n)
s1 <- sd(log(x + 1))
fl_nmf <- flash(Y,ebnm_fn = ebnm_point_exponential,var_type = 2, 
                greedy_Kmax = 4,S = s1,backfit = TRUE,verbose = 0)
```

Note that I chose "greedy_Kmax" after some trial and error and found
that 4 was a good setting.

```{r structure-plot-nmf, fig.height=2.25, fig.width=5}
topic_colors <- c("olivedrab","dodgerblue","tomato","darkblue")
res <- ldf(fl_nmf,type = "i")
p2 <- structure_plot(res$L,grouping = samples$label,topics = c(2,3,4,1),
                     colors = topic_colors)
p2
```

Interestingly, the NMF model applied to the shifted log count
decomposes the samples in nearly the same way as the topic model, with
one key difference: it also explicitly has a *baseline factor* (factor
1).

[sanford2020elife]: https://doi.org/10.7554/eLife.59388

## Log1p Fits

Below I fit the log1p model for different values of c.

```{r, eval=FALSE}
cc_vec <- c(1e-3, 1)

fit_list <- list()

for (cc in cc_vec) {
  
  set.seed(1)
  fit_list[[as.character(cc)]] <- fit_poisson_log1p_nmf(
    Y = counts,
    K = 4,
    cc = cc,
    loglik = "exact",
    control = list(verbose = FALSE)
  )
  
}
```

Here is the c = 1 structure plot, which looks very similar to flashier (though is less sparse).

```{r, eval=FALSE}
topic_colors <- c("olivedrab","dodgerblue","tomato","darkblue")
normalized_structure_plot(
  fit = fit_list$`1`,
  grouping = samples$label,
  colors = topic_colors,
  topics = c(2,3,4,1)
)$plot
```

The c = 1e-3 structure plot is below, which looks a bit crazy.

```{r, eval=FALSE}
topic_colors <- c("olivedrab","dodgerblue","tomato","darkblue")
normalized_structure_plot(
  fit = fit_list$`0.001`,
  grouping = samples$label,
  colors = topic_colors,
  topics = c(2,3,4,1)
)$plot
```

```{r, eval=FALSE}
tm_rank1 <- fastTopics:::fit_pnmf_rank1(counts)

init_LL <- cbind(
  tm_rank1$L,
  matrix(
    data = 1e-10,
    nrow = nrow(counts),
    ncol = 3
  )
)
rownames(init_LL) <- rownames(counts)

init_FF <- cbind(
  tm_rank1$F,
  matrix(
    data = 1e-10,
    nrow = ncol(counts),
    ncol = 3
  )
)

rownames(init_FF) <- colnames(counts)
fit0 <- init_poisson_nmf(X = counts, L = init_LL, F = init_FF)

fit <- fit_poisson_nmf(X = counts, fit0 = fit0)

```

Finally, I also fit fasttopics with a rank-1 initialization just to make sure the differences weren't being driven by initialization.

```{r, eval=FALSE}
topic_colors <- c("tomato","darkblue","gold","skyblue")
p1 <- structure_plot(fit,grouping = samples$label,topics = 1:4,
                     colors = topic_colors)
p1$plot
```

## Topic Model with Alternative Initialization

I also attempted to initialize the topic model where each sample is loaded on the mean of only the greatest L2 norm point in PC space corresponding to it's treatment group.

```{r, eval=FALSE}
pdat$l2_pc <- sqrt(
  pdat$PC1 ^ 2 + pdat$PC2 ^ 2
)

# representative samples:
# RA+TGFb: GSM4625622
# RA: GSM4625634
# EtOH: GSM4625648
# TGFb: GSM4625621

FF <- cbind(
  counts["GSM4625622", ],
  counts["GSM4625634", ],
  counts["GSM4625648", ],
  counts["GSM4625621", ]
)

rownames(FF) <- colnames(counts)

LL <- cbind(
  ifelse(pdat$label == "RA+TGFb", 1, 0),
  ifelse(pdat$label == "RA", 1, 0),
  ifelse(pdat$label == "EtOH", 1, 0),
  ifelse(pdat$label == "TGFb", 1, 0)
)
rownames(LL) <- rownames(counts)

fit0 <- init_poisson_nmf(X = counts, F = FF, L = LL)
fit <- fit_poisson_nmf(X = counts, fit0 = fit0)
```

Here is the structure plot at initialization:

```{r, fig.height=1.75, fig.width=5, eval=FALSE}
topic_colors <- c("tomato","darkblue","gold","skyblue")
structure_plot(fit0, grouping = pdat$label, topics = 1:4,
                     colors = topic_colors)$plot
```

And here it is after optimization:

```{r, fig.height=1.75, fig.width=5, eval=FALSE}
topic_colors <- c("tomato","darkblue","gold","skyblue")

structure_plot(fit, grouping = pdat$label, topics = 1:4,
                     colors = topic_colors)$plot
```

```{r, eval=FALSE}
mn <- t(fit$F %*% diag(apply(fit$L, 2, max)))
newX <- log1p(mn)

# 1) center using the training means
#    (pca$center is a length‐p vector of column means of Y)
newX_cent <- sweep(newX, 2, pca$center, FUN = "-")

# 2) since you had scale = FALSE, you do NOT rescale here

# 3) project onto the PC loadings
#    (pca$v is a p×k matrix of the top k loading vectors)
scores <- newX_cent %*% pca$rotation
```

```{r, eval=FALSE}
pd2 <- data.frame(
  pc1 = pdat$PC1,
  pc2 = pdat$PC2,
  label = pdat$label
)

new_pd <- data.frame(
  pc1 = scores[,1],
  pc2 = scores[,2],
  label = c("topic1", "topic2", "topic3", "topic4")
)

pd2 <- rbind(pd2, new_pd)

```

Here's a PC plot with the fitted topics:

```{r, eval=FALSE}
ggplot(pd2,aes(x = pc1,y = pc2,color = label)) +
  geom_point() +
  scale_color_manual(values = c("dodgerblue","tomato","darkblue",
                                "limegreen", "orange", "black", "gold", "maroon")) + 
  theme_cowplot(font_size = 10)
```

