---
title: "Newsgroups 25 Factors"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2025-01-29"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Here, I will apply the `log1p` matrix factorization model to the Newsgroups dataset using 25 factors.

```{r}
load("~/Downloads/newsgroups.RData")
library(dplyr)
library(Matrix)
library(fastTopics)
library(passPCA)
rm(counts)

# I did the following to the original dataset
# counts <- counts[, !(colnames(counts) %in% stopwords("SMART"))]
# colnames(counts) <- wordStem(colnames(counts))

counts <- readr::read_rds(
  "~/Downloads/newsgroups_stemmed_stopwords.rds"
)

topics <- topics[Matrix::rowSums(counts) > 9]
counts <- counts[Matrix::rowSums(counts) > 9, ]
genes_to_use <- which(Matrix::colSums(counts>0)>4)
counts <- counts[,genes_to_use]
s <- Matrix::rowSums(counts)
s <- s / mean(s)


K <- 25
cc_vec <- c(1e-3, 1e-2, 1e-1, 1, 1e1, 1e2, 1e3)

n <- nrow(counts)
p <- ncol(counts)
```

```{r, eval=FALSE}
for (cc in cc_vec) {

  print(cc)

  set.seed(1)
  log1p_k1 <- fit_factor_model_log1p_exact(
    Y = counts,
    K = 1,
    maxiter = 5,
    s = cc * s,
    init_method = "frob_nmf"
  )

  set.seed(1)
  init_LL <- log1p_k1$U %>%
    cbind(
      matrix(
        data = rexp(
          n = n * (K - 1), rate = 15
        ),
        nrow = n,
        ncol = K - 1
      )
    )

  set.seed(1)
  init_FF <- log1p_k1$V %>%
    cbind(
      matrix(
        data = rexp(
          n = p * (K - 1), rate = 15
        ),
        nrow = p,
        ncol = K - 1
      )
    )

  tictoc::tic()
  set.seed(1)
  fit <- fit_factor_model_log1p_exact(
    Y = counts,
    K = K,
    init_U = init_LL,
    init_V = init_FF,
    maxiter = 100,
    s = cc * s
  )
  total_time <- tictoc::toc()

  fit[["total_time"]] <- total_time$toc
  rownames(fit$U) <- rownames(counts)
  rownames(fit$V) <- colnames(counts)

  readr::write_rds(
    fit, glue::glue("~/Documents/data/passPCA/news/news_log1p_c{cc}_k12_exact_100_iter.rds")
  )

}

fit0_nmf <- fastTopics:::fit_pnmf_rank1(counts)

set.seed(1)
init_LL <- fit0_nmf$L %>%
  cbind(
    matrix(
      data = rexp(
        n = n * (K - 1), rate = 15
      ),
      nrow = n,
      ncol = K - 1
    )
  )

set.seed(1)
init_FF <- fit0_nmf$F %>%
  cbind(
    matrix(
      data = rexp(
        n = p * (K - 1), rate = 15
      ),
      nrow = p,
      ncol = K - 1
    )
  )

rownames(init_LL) <- rownames(counts)
rownames(init_FF) <- colnames(counts)

fit0_K <- init_poisson_nmf(
  X = counts, F = init_FF, L = init_LL
)

fit_nmf <- fit_poisson_nmf(
  X = counts,
  fit0 = fit0_K,
  control = list(list(nc = 7))
)

readr::write_rds(
  fit_nmf, glue::glue("~/Documents/data/passPCA/news/news_pois_nmf_k12_exact_100_iter.rds")
)
```

```{r}
library(ggpubr)
library(fastTopics)
fit_list <- list()

for (cc in cc_vec) {

  fit_list[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/news/news_log1p_c{cc}_k{K}_exact_100_iter.rds"
    )
  )

}

fit_list[["Inf"]] <- readr::read_rds(
  glue::glue("~/Documents/data/passPCA/news/news_pois_nmf_k{K}_exact_100_iter.rds")
)
```

# Analysis of fits

## Loadings

```{r}
normalize_bars <- function(LL) {

  max_col <- apply(LL, 2, max)
  sweep(LL, 2, max_col, FUN = "/")

}
```

```{r, fig.width=12, fig.height=33}
library(tidyr)
library(stringi)
library(stringr)
cell.type <- as.factor(topics)

LL_c001 <- normalize_bars(fit_list$`0.001`$U)

# Downsample the number of cells and sort them using tSNE.
set.seed(8675309)
cell.idx <- numeric(0)
cell.types <- levels(cell.type)
for (i in 1:length(cell.types)) {
  which.idx <- which(cell.type == cell.types[i])
  # Downsample common cell types.
  if (length(which.idx) > 2000) {
    which.idx <- sample(which.idx, 2000)
  }
  # Don't include rare cell types.
  if (length(which.idx) > 10) {
    # Sort using tsne.
    tsne.res <- Rtsne::Rtsne(
      LL_c001[which.idx, ],
      dims = 1,
      pca = FALSE,
      normalize = FALSE,
      perplexity = min(100, floor((length(which.idx) - 1) / 3) - 1),
      theta = 0.1,
      max_iter = 1000,
      eta = 200,
      check_duplicates = FALSE
    )$Y[, 1]
    which.idx <- which.idx[order(tsne.res)]
    cell.idx <- c(cell.idx, which.idx)
  }
}

cell.type <- cell.type[cell.idx]
cell.type <- droplevels(cell.type)

LL_c001 <- LL_c001[cell.idx, ]
LL_c01 <- normalize_bars(fit_list$`0.01`$U)
LL_c01 <- LL_c01[cell.idx, ]
LL_cp1 <- normalize_bars(fit_list$`0.1`$U)
LL_cp1 <- LL_cp1[cell.idx, ]
LL_c1 <- normalize_bars(fit_list$`1`$U)
LL_c1 <- LL_c1[cell.idx, ]
LL_c10 <- normalize_bars(fit_list$`10`$U)
LL_c10 <- LL_c10[cell.idx, ]
LL_c100 <- normalize_bars(fit_list$`100`$U)
LL_c100 <- LL_c100[cell.idx, ]
LL_c1000 <- normalize_bars(fit_list$`1000`$U)
LL_c1000 <- LL_c1000[cell.idx, ]
LL_cinf <- Matrix::Diagonal(x = 1 / s) %*% fit_list$`Inf`$L
LL_cinf <- normalize_bars(as.matrix(LL_cinf))
LL_cinf <- LL_cinf[cell.idx, ]

make.heatmap.tib <- function(FF) {
  tib <- as_tibble(scale(FF, center = FALSE, scale = apply(FF, 2, max))) %>%
    mutate(Cell.type = cell.type) %>%
    arrange(Cell.type) %>%
    mutate(Cell.idx = row_number())
  
  tib <- tib %>%
    pivot_longer(
      -c(Cell.idx, Cell.type),
      names_to = "Factor",
      values_to = "Loading",
      values_drop_na = TRUE
    ) %>%
    mutate(Factor = as.numeric(str_extract(Factor, "[0-9]+")))
  
  return(tib)
}

LL_c001_tib <- make.heatmap.tib(LL_c001)
LL_c01_tib <- make.heatmap.tib(LL_c01)
LL_cp1_tib <- make.heatmap.tib(LL_cp1)
LL_c1_tib <- make.heatmap.tib(LL_c1)
LL_c10_tib <- make.heatmap.tib(LL_c10)
LL_c100_tib <- make.heatmap.tib(LL_c100)
LL_c1000_tib <- make.heatmap.tib(LL_c1000)
LL_cinf_tib <- make.heatmap.tib(LL_cinf)

heatmap.tib <- LL_c001_tib %>% mutate(Method = "c = 0.001") %>%
  bind_rows(LL_c01_tib %>% mutate(Method = "c = 0.01")) %>%
  bind_rows(LL_cp1_tib %>% mutate(Method = "c = 0.1")) %>%
  bind_rows(LL_c1_tib %>% mutate(Method = "c = 1")) %>%
  bind_rows(LL_c10_tib %>% mutate(Method = "c = 10")) %>%
  bind_rows(LL_c100_tib %>% mutate(Method = "c = 100")) %>%
  bind_rows(LL_c1000_tib %>% mutate(Method = "c = 1000")) %>%
  bind_rows(LL_cinf_tib %>% mutate(Method = "Topic Model")) %>% 
  mutate(Method = factor(Method, levels = c(
    "c = 0.001", "c = 0.01", "c = 0.1", "c = 1", "c = 10", "c = 100", "c = 1000", "Topic Model")))

tib <- heatmap.tib %>%
  group_by(Cell.type, Cell.idx) %>%
  summarize()

cell_type_breaks <- c(1, which(tib$Cell.type[-1] != tib$Cell.type[-nrow(tib)]))
label_pos <- cell_type_breaks / 2 + c(cell_type_breaks[-1], nrow(tib)) / 2

library(ggplot2)

plt <- ggplot(heatmap.tib, aes(x = Factor, y = -Cell.idx, fill = Loading)) +
  geom_tile() +
  scale_fill_gradient(low = "white", high = "firebrick") +
  labs(y = "") +
  scale_y_continuous(breaks = -label_pos,
                     minor_breaks = NULL,
                     labels = levels(cell.type)) +
  scale_x_continuous(breaks = seq(0, 30, 5)) +
  theme_minimal() +
  geom_hline(yintercept = -cell_type_breaks, size = 0.1) +
  facet_wrap(~Method, ncol = 1, axes = "all") +
  theme(legend.position = "none", 
        strip.text = element_text(size = 16)) 

plt
```

The plot of the fastTopics fit above looks weird, and I believe it is perhaps because the most loaded document on each topic is often much more loaded than any other topic, and so normalizing the columns to have maximum value $1$ distorts the fit. I plot a structure plot of the fastTopcis fit below:

```{r, results='hide'}
sp <- fastTopics::structure_plot(fit_list$`Inf`, grouping = topics, gap = 40)
```

```{r}
sp
```


## Factors

Below, I print out the top $10$ words for each factor from each model.

```{r}
get_keywords <- function(V) {
  
  kw_list <- list()
  
  for (k in 1:ncol(V)) {
    
    kw_list[[k]] <- names(head(sort(V[,k], decreasing = T), 10))
    
  }
  
  return(kw_list)
  
}
```

```{r}
kw_lists <- list()

for (cc in c(cc_vec)) {
  
  fit <- fit_list[[as.character(cc)]]
  kw_lists[[as.character(cc)]] <- get_keywords(fit$V)

}

kw_lists[["Inf"]] <- get_keywords(fit_list$`Inf`$F)
```

### c = 1e-3

```{r}
print(kw_lists[[as.character(1e-3)]])
```

### c = 1e-2

```{r}
print(kw_lists[[as.character(1e-2)]])
```

### c = 1e-1

```{r}
print(kw_lists[[as.character(1e-1)]])
```

### c = 1

```{r}
print(kw_lists[[as.character(1)]])
```

### c = 10

```{r}
print(kw_lists[[as.character(10)]])
```

### c = 100

```{r}
print(kw_lists[[as.character(100)]])
```

### c = 1000

```{r}
print(kw_lists[[as.character(1000)]])
```

### nmf

```{r}
print(kw_lists[["Inf"]])
```
