---
title: "Analysis of Newsgroups Data"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2025-01-23"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Here, I analyzed the Newsgroups dataset.

```{r}
load("~/Downloads/newsgroups.RData")
library(dplyr)
library(Matrix)
library(fastTopics)
library(passPCA)
rm(counts)

# I did the following to the original dataset
# counts <- counts[, !(colnames(counts) %in% stopwords("SMART"))]
# colnames(counts) <- wordStem(colnames(counts))

counts <- readr::read_rds(
  "~/Downloads/newsgroups_stemmed_stopwords.rds"
)

topics <- topics[Matrix::rowSums(counts) > 9]
counts <- counts[Matrix::rowSums(counts) > 9, ]
genes_to_use <- which(Matrix::colSums(counts>0)>4)
counts <- counts[,genes_to_use]
s <- Matrix::rowSums(counts)
s <- s / mean(s)


K <- 12
cc_vec <- c(1e-3, 1e-2, 1e-1, 1, 1e1, 1e2, 1e3)

n <- nrow(counts)
p <- ncol(counts)
```

Now, I fit the log1p model for many values of $c$. I also fit Poisson NMF using `fastTopics`. To fit each model, I first fit a rank $1$ model. Then, I initialize each model with $K = 12$, where the first factor is taken from the rank 1 model and the other $11$ are randomly iniitalized.

```{r, eval=FALSE}
for (cc in cc_vec) {

  print(cc)

  set.seed(1)
  log1p_k1 <- fit_factor_model_log1p_exact(
    Y = counts,
    K = 1,
    maxiter = 5,
    s = cc * s,
    init_method = "frob_nmf"
  )

  set.seed(1)
  init_LL <- log1p_k1$U %>%
    cbind(
      matrix(
        data = rexp(
          n = n * (K - 1), rate = 15
        ),
        nrow = n,
        ncol = K - 1
      )
    )

  set.seed(1)
  init_FF <- log1p_k1$V %>%
    cbind(
      matrix(
        data = rexp(
          n = p * (K - 1), rate = 15
        ),
        nrow = p,
        ncol = K - 1
      )
    )

  tictoc::tic()
  set.seed(1)
  fit <- fit_factor_model_log1p_exact(
    Y = counts,
    K = K,
    init_U = init_LL,
    init_V = init_FF,
    maxiter = 100,
    s = cc * s
  )
  total_time <- tictoc::toc()

  fit[["total_time"]] <- total_time$toc
  rownames(fit$U) <- rownames(counts)
  rownames(fit$V) <- colnames(counts)

  readr::write_rds(
    fit, glue::glue("~/Documents/data/passPCA/news/news_log1p_c{cc}_k12_exact_100_iter.rds")
  )

}

fit0_nmf <- fastTopics:::fit_pnmf_rank1(counts)

set.seed(1)
init_LL <- fit0_nmf$L %>%
  cbind(
    matrix(
      data = rexp(
        n = n * (K - 1), rate = 15
      ),
      nrow = n,
      ncol = K - 1
    )
  )

set.seed(1)
init_FF <- fit0_nmf$F %>%
  cbind(
    matrix(
      data = rexp(
        n = p * (K - 1), rate = 15
      ),
      nrow = p,
      ncol = K - 1
    )
  )

rownames(init_LL) <- rownames(counts)
rownames(init_FF) <- colnames(counts)

fit0_K <- init_poisson_nmf(
  X = counts, F = init_FF, L = init_LL
)

fit_nmf <- fit_poisson_nmf(
  X = counts,
  fit0 = fit0_K,
  control = list(list(nc = 7))
)

readr::write_rds(
  fit_nmf, glue::glue("~/Documents/data/passPCA/news/news_pois_nmf_k12_exact_100_iter.rds")
)
```

Below are the structure plots for each of the models:

```{r}
library(ggpubr)
library(fastTopics)
fit_list <- list()


for (cc in cc_vec) {

  fit_list[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/news/news_log1p_c{cc}_k12_exact_100_iter.rds"
    )
  )

}

fit_list[["Inf"]] <- readr::read_rds(
  glue::glue("~/Documents/data/passPCA/news/news_pois_nmf_k12_exact_100_iter.rds")
)
```


```{r, results='hide'}
normalize_bars <- function(LL) {

  max_col <- apply(LL, 2, max)
  sweep(LL, 2, max_col, FUN = "/")

}

plot_list <- list()

for (cc in cc_vec) {

  plot_list[[glue::glue("{cc}: Poisson")]] <- structure_plot(
    normalize_bars(fit_list[[as.character(cc)]]$U),
    grouping = topics,gap = 50,perplexity = 70,n = Inf, font.size = 12
    )

}

plot_list[["Inf"]] <- structure_plot(
  fit_list[["Inf"]],
  grouping = topics,gap = 50,perplexity = 70,n = Inf,font.size = 12
  )
```

```{r, fig.width=20, fig.height=70}
ggarrange(
  plotlist = plot_list,
  ncol = 1,
  labels = paste0("c = ", names(plot_list))
)
```

```{r}
get_keywords <- function(V) {
  
  kw_list <- list()
  
  for (k in 1:ncol(V)) {
    
    kw_list[[k]] <- names(head(sort(V[,k], decreasing = T), 10))
    
  }
  
  return(kw_list)
  
}
```

## Keywords

```{r}
kw_lists <- list()

for (cc in c(cc_vec)) {
  
  fit <- fit_list[[as.character(cc)]]
  
  kw_lists[[as.character(cc)]] <- get_keywords(fit$V)

}

kw_lists[["Inf"]] <- get_keywords(fit_list$`Inf`$F)
```

### c = 1e-3

```{r}
print(kw_lists[[as.character(1e-3)]])
```

### c = 1e-2

```{r}
print(kw_lists[[as.character(1e-2)]])
```

### c = 1e-1

```{r}
print(kw_lists[[as.character(1e-1)]])
```

### c = 1

```{r}
print(kw_lists[[as.character(1)]])
```

### c = 10

```{r}
print(kw_lists[[as.character(10)]])
```

### c = 100

```{r}
print(kw_lists[[as.character(100)]])
```

### c = 1000

```{r}
print(kw_lists[[as.character(1000)]])
```

### c = Inf

```{r}
print(kw_lists[["Inf"]])
```

