---
title: "Analysis of Pancreas Cel-Seq Data"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2024-12-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Here, I analyzed the pancreas Cel-Seq dataset that Peter has looked at before.

```{r}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)

set.seed(1)

load("~/Downloads/pancreas.RData")

i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i,]
counts      <- counts[i,]
sample_info <- transform(sample_info,celltype = factor(celltype))

counts <- counts[, Matrix::colSums(counts) > 0]
genes_to_use <- which(Matrix::colSums(counts>0)>4)
counts <- counts[,genes_to_use]
s <- Matrix::rowSums(counts)
s <- s / mean(s)

library(passPCA)

K <- 9
cc_vec <- c(1e-3, 1e-2, 1e-1, 1, 1e1, 1e2, 1e3)

n <- nrow(counts)
p <- ncol(counts)
```

Now, I fit the log1p model for many values of $c$. I also fit Poisson NMF using `fastTopics`. To fit each model, I first fit a rank $1$ model. Then, I initialize each model with $K = 9$, where the first factor is taken from the rank 1 model and the other $8$ are randomly iniitalized.

```{r, eval=FALSE}
for (cc in cc_vec) {

  print(cc)

  set.seed(1)
  log1p_k1 <- fit_factor_model_log1p_exact(
    Y = counts,
    K = 1,
    maxiter = 10,
    s = cc * s,
    init_method = "frob_nmf"
  )

  set.seed(1)
  init_LL <- log1p_k1$U %>%
    cbind(
      matrix(
        data = rexp(
          n = n * (K - 1), rate = 15
        ),
        nrow = n,
        ncol = K - 1
      )
    )

  set.seed(1)
  init_FF <- log1p_k1$V %>%
    cbind(
      matrix(
        data = rexp(
          n = p * (K - 1), rate = 15
        ),
        nrow = p,
        ncol = K - 1
      )
    )

  tictoc::tic()
  set.seed(1)
  fit <- fit_factor_model_log1p_exact(
    Y = counts,
    K = K,
    init_U = init_LL,
    init_V = init_FF,
    maxiter = 100,
    s = cc * s
  )
  total_time <- tictoc::toc()

  fit[["total_time"]] <- total_time$toc
  rownames(fit$U) <- rownames(counts)
  rownames(fit$V) <- colnames(counts)

  readr::write_rds(
    fit, glue::glue("~/Documents/data/passPCA/pancreas/pancreas_log1p_c{cc}_k9_exact_100_iter.rds")
  )

}

fit0_nmf <- fastTopics:::fit_pnmf_rank1(counts)

set.seed(1)
init_LL <- fit0_nmf$L %>%
  cbind(
    matrix(
      data = rexp(
        n = n * (K - 1), rate = 15
      ),
      nrow = n,
      ncol = K - 1
    )
  )

set.seed(1)
init_FF <- fit0_nmf$F %>%
  cbind(
    matrix(
      data = rexp(
        n = p * (K - 1), rate = 15
      ),
      nrow = p,
      ncol = K - 1
    )
  )

rownames(init_LL) <- rownames(counts)
rownames(init_FF) <- colnames(counts)

fit0_K <- init_poisson_nmf(
  X = counts, F = init_FF, L = init_LL
)

fit_nmf <- fit_poisson_nmf(
  X = counts,
  fit0 = fit0_K,
  control = list(list(nc = 7))
)

readr::write_rds(
  fit_nmf, glue::glue("~/Documents/data/passPCA/pancreas/pancreas_pois_nmf_k9_exact_100_iter.rds")
)
```

I also ran Frobenius NMF on the transformed counts $\log(1 + y/(c \cdot s))$, where $s$ is the same size factor used above. To attempt to keep things consistent with the Poisson models, I initialized the NMF models with the models from above after one iteration on the $K = 9$ fits.

Below are the structure plots for each of the models:

```{r}
library(ggpubr)
library(fastTopics)
fit_list <- list()
fit_list_nnlm <- list()

for (cc in cc_vec) {

  fit_list[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/pancreas/pancreas_log1p_c{cc}_k9_exact_100_iter.rds"
    )
  )
  
  fit_list[[as.character(cc)]][["Lambda"]] <- cc * (exp(
    fit_list[[as.character(cc)]]$U %*% t(fit_list[[as.character(cc)]]$V)) - 1)
  
  fit_list_nnlm[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/pancreas/pancreas_log1p_c{cc}_nnlm.rds"
      )
  )
    

}

fit_list[["Inf"]] <- readr::read_rds(
  glue::glue("~/Documents/data/passPCA/pancreas/pancreas_pois_nmf_k9_exact_100_iter.rds")
)

fit_list[["Inf"]][["Lambda"]] <- fit_list[["Inf"]]$L %*% t(fit_list[["Inf"]]$F) 
```

```{r}
# library(ggplot2)
# 
# df <- data.frame(
#   hat = Matrix::colMeans(counts),
#   lr_hat = colMeans(fit_list$`1000`$Lambda)
# )
# 
# ggplot(data = df) +
#   geom_point(aes(x = hat, y = lr_hat), alpha = 0.25) +
#   scale_x_continuous(transform = "log10") +
#   scale_y_continuous(transform = "log10") +
#   geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
#   xlab("Average Expression") +
#   ylab("Low Rank Mean Expression") +
#   ggtitle("K = 1000") +
#   cowplot::theme_cowplot()
```


```{r, results='hide'}
normalize_bars <- function(LL) {

  max_col <- apply(LL, 2, max)
  sweep(LL, 2, max_col, FUN = "/")

}

celltype <- sample_info$celltype
celltype <-
  factor(celltype,
         c("acinar","ductal","activated_stellate","quiescent_stellate",
           "endothelial","macrophage","mast","schwann","alpha","beta",
           "delta","gamma","epsilon"))


plot_list <- list()

for (cc in cc_vec) {

  plot_list[[glue::glue("{cc}: Poisson")]] <- structure_plot(
    normalize_bars(fit_list[[as.character(cc)]]$U),
    grouping = celltype,gap = 20,perplexity = 70,n = Inf, font.size = 12
    )
  
  plot_list[[glue::glue("{cc}: Vanilla NMF")]] <- structure_plot(
    normalize_bars(fit_list_nnlm[[as.character(cc)]]$W),
    grouping = celltype,gap = 20,perplexity = 70,n = Inf, font.size = 12
    )

}

plot_list[["Inf"]] <- structure_plot(
  fit_list[["Inf"]],
  grouping = celltype,gap = 20,perplexity = 70,n = Inf,font.size = 12
  )
```

```{r, fig.width=20, fig.height=70}
ggarrange(
  plotlist = plot_list,
  ncol = 1,
  labels = paste0("c = ", names(plot_list))
)
```

