---
title: "Analysis of Pancreas Cel-Seq Data"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2024-12-30"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Here, I analyzed the pancreas Cel-Seq dataset that Peter has looked at before.

```{r}
library(Matrix)
library(fastTopics)
library(ggplot2)
library(cowplot)

set.seed(1)

load("~/Downloads/pancreas.RData")

i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i,]
counts      <- counts[i,]
sample_info <- transform(sample_info,celltype = factor(celltype))

counts <- counts[, Matrix::colSums(counts) > 0]
genes_to_use <- which(Matrix::colSums(counts>0)>4)
counts <- counts[,genes_to_use]
s <- Matrix::rowSums(counts)
s <- s / mean(s)

library(passPCA)

K <- 9
cc_vec <- c(1e-3, 1e-2, 1e-1, 1, 1e1, 1e2, 1e3)

n <- nrow(counts)
p <- ncol(counts)
```

Now, I fit the log1p model for many values of $c$. I also fit Poisson NMF using `fastTopics`. To fit each model, I first fit a rank $1$ model. Then, I initialize each model with $K = 9$, where the first factor is taken from the rank 1 model and the other $8$ are randomly iniitalized.

```{r, eval=FALSE}
for (cc in cc_vec) {

  print(cc)

  set.seed(1)
  log1p_k1 <- fit_factor_model_log1p_exact(
    Y = counts,
    K = 1,
    maxiter = 10,
    s = cc * s,
    init_method = "frob_nmf"
  )

  set.seed(1)
  init_LL <- log1p_k1$U %>%
    cbind(
      matrix(
        data = rexp(
          n = n * (K - 1), rate = 15
        ),
        nrow = n,
        ncol = K - 1
      )
    )

  set.seed(1)
  init_FF <- log1p_k1$V %>%
    cbind(
      matrix(
        data = rexp(
          n = p * (K - 1), rate = 15
        ),
        nrow = p,
        ncol = K - 1
      )
    )

  tictoc::tic()
  set.seed(1)
  fit <- fit_factor_model_log1p_exact(
    Y = counts,
    K = K,
    init_U = init_LL,
    init_V = init_FF,
    maxiter = 100,
    s = cc * s
  )
  total_time <- tictoc::toc()

  fit[["total_time"]] <- total_time$toc
  rownames(fit$U) <- rownames(counts)
  rownames(fit$V) <- colnames(counts)

  readr::write_rds(
    fit, glue::glue("~/Documents/data/passPCA/pancreas/pancreas_log1p_c{cc}_k9_exact_100_iter.rds")
  )

}

fit0_nmf <- fastTopics:::fit_pnmf_rank1(counts)

set.seed(1)
init_LL <- fit0_nmf$L %>%
  cbind(
    matrix(
      data = rexp(
        n = n * (K - 1), rate = 15
      ),
      nrow = n,
      ncol = K - 1
    )
  )

set.seed(1)
init_FF <- fit0_nmf$F %>%
  cbind(
    matrix(
      data = rexp(
        n = p * (K - 1), rate = 15
      ),
      nrow = p,
      ncol = K - 1
    )
  )

rownames(init_LL) <- rownames(counts)
rownames(init_FF) <- colnames(counts)

fit0_K <- init_poisson_nmf(
  X = counts, F = init_FF, L = init_LL
)

fit_nmf <- fit_poisson_nmf(
  X = counts,
  fit0 = fit0_K,
  control = list(list(nc = 7))
)

readr::write_rds(
  fit_nmf, glue::glue("~/Documents/data/passPCA/pancreas/pancreas_pois_nmf_k9_exact_100_iter.rds")
)
```

I also ran Frobenius NMF on the transformed counts $\log(1 + y/(c \cdot s))$, where $s$ is the same size factor used above. To attempt to keep things consistent with the Poisson models, I initialized the NMF models with the models from above after one iteration on the $K = 9$ fits.

```{r, eval=FALSE}
Y_tilde <- Matrix::Diagonal(x = 1/s) %*% counts

library(passPCA)

K <- 9
cc_vec <- c(1e-3, 1e-2, 1e-1, 1, 1e1, 1e2, 1e3)

n <- nrow(counts)
p <- ncol(counts)

for (cc in cc_vec) {
  
  print(cc)
  
  set.seed(1)
  log1p_k1 <- fit_factor_model_log1p_exact(
    Y = counts,
    K = 1,
    maxiter = 10,
    s = cc * s,
    init_method = "frob_nmf"
  )
  
  set.seed(1)
  init_LL <- log1p_k1$U %>%
    cbind(
      matrix(
        data = rexp(
          n = n * (K - 1), rate = 15
        ),
        nrow = n,
        ncol = K - 1
      )
    )
  
  set.seed(1)
  init_FF <- log1p_k1$V %>%
    cbind(
      matrix(
        data = rexp(
          n = p * (K - 1), rate = 15
        ),
        nrow = p,
        ncol = K - 1
      )
    )
  
  fit <- fit_factor_model_log1p_exact(
    Y = counts,
    K = K,
    init_U = init_LL,
    init_V = init_FF,
    maxiter = 1,
    s = cc * s
  )

  Y_tilde_cc <- MatrixExtra::mapSparse(Y_tilde, function(x){log1p(x/cc)})
  
  nmf_fit <- nnmf(
    A = as.matrix(Y_tilde_cc), 
    k = 9,
    init = list(
      W = fit$U, H = t(fit$V)
    )
  )
  
  readr::write_rds(
    nmf_fit, glue::glue("~/Documents/data/passPCA/pancreas/pancreas_log1p_c{cc}_nnlm.rds")
  )
  
}
```


Below are the structure plots for each of the models:

```{r}
library(ggpubr)
library(fastTopics)
fit_list <- list()
fit_list_nnlm <- list()

for (cc in cc_vec) {

  fit_list[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/pancreas/pancreas_log1p_c{cc}_k9_exact_100_iter.rds"
    )
  )

  fit_list_nnlm[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/pancreas/pancreas_log1p_c{cc}_nnlm.rds"
      )
  )
    

}

fit_list[["Inf"]] <- readr::read_rds(
  glue::glue("~/Documents/data/passPCA/pancreas/pancreas_pois_nmf_k9_exact_100_iter.rds")
)

```

```{r}
# library(ggplot2)
# 
# df <- data.frame(
#   hat = Matrix::colMeans(counts),
#   lr_hat = colMeans(fit_list$`1000`$Lambda)
# )
# 
# ggplot(data = df) +
#   geom_point(aes(x = hat, y = lr_hat), alpha = 0.25) +
#   scale_x_continuous(transform = "log10") +
#   scale_y_continuous(transform = "log10") +
#   geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
#   xlab("Average Expression") +
#   ylab("Low Rank Mean Expression") +
#   ggtitle("K = 1000") +
#   cowplot::theme_cowplot()
```


```{r, results='hide'}
normalize_bars <- function(LL) {

  max_col <- apply(LL, 2, max)
  sweep(LL, 2, max_col, FUN = "/")

}

celltype <- sample_info$celltype
celltype <-
  factor(celltype,
         c("acinar","ductal","activated_stellate","quiescent_stellate",
           "endothelial","macrophage","mast","schwann","alpha","beta",
           "delta","gamma","epsilon"))


plot_list <- list()

for (cc in cc_vec) {

  plot_list[[glue::glue("{cc}: Poisson")]] <- structure_plot(
    normalize_bars(fit_list[[as.character(cc)]]$U),
    grouping = celltype,gap = 20,perplexity = 70,n = Inf, font.size = 12
    )
  
  plot_list[[glue::glue("{cc}: Vanilla NMF")]] <- structure_plot(
    normalize_bars(fit_list_nnlm[[as.character(cc)]]$W),
    grouping = celltype,gap = 20,perplexity = 70,n = Inf, font.size = 12
    )

}

plot_list[["Inf"]] <- structure_plot(
  fit_list[["Inf"]],
  grouping = celltype,gap = 20,perplexity = 70,n = Inf,font.size = 12
  )
```

```{r, fig.width=20, fig.height=70}
ggarrange(
  plotlist = plot_list,
  ncol = 1,
  labels = paste0("c = ", names(plot_list))
)
```


## A closer look at the loadings and the factors

First, I looked at the correlations between the factors:

```{r}
cor_plot_list <- list()

cor_df <- expand.grid(1:9, 1:9)
colnames(cor_df) <- c("row", "col")

for (cc in cc_vec) {
  
  cor_df$cor <- as.vector(cor(fit_list[[as.character(cc)]]$V))
    
  library(ggplot2)
  cor_plot_list[[glue::glue("c = {cc}")]] <- ggplot(
      cor_df, 
      aes(x = col, y = row, fill = cor)
    ) +
    geom_tile() +
    # Optional: adjust the color scale
    scale_fill_gradient2(
      low = "blue",
      mid = "white",
      high = "red",
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    theme_minimal() +
    coord_fixed() +
    scale_x_continuous(breaks = 1:9, labels = 1:9) +
    scale_y_reverse(breaks = 1:9, labels = 1:9) +
    xlab("factor") +
    ylab("factor")

}

cor_df$cor <- as.vector(cor(fit_list$`Inf`$F))

cor_plot_list[["c = Inf"]] <- ggplot(
      cor_df, 
      aes(x = col, y = row, fill = cor)
    ) +
    geom_tile() +
    # Optional: adjust the color scale
    scale_fill_gradient2(
      low = "blue",
      mid = "white",
      high = "red",
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    theme_minimal() +
    coord_fixed() +
    scale_x_continuous(breaks = 1:9, labels = 1:9) +
    scale_y_reverse(breaks = 1:9, labels = 1:9) +
    xlab("factor") +
    ylab("factor")
```

```{r, fig.width=15, fig.height=15}
ggarrange(
  plotlist = cor_plot_list,
  ncol = 3,
  nrow = 3,
  labels = names(cor_plot_list),
  common.legend = TRUE,
  legend = "right"
)
```

There is a noticable increase in correlation as $c$ gets larger. 

I also looked at the overlap in the "driving genes." To do this, I took the top $25$ genes for each factor and computed the Jaccard Index for each pair of factors.

```{r}
compare_overlap <- function(X, top = 10, method = c("intersection", "jaccard")) {
  # Check input validity
  if (!is.matrix(X)) stop("X must be a matrix")
  if (is.null(rownames(X))) stop("Matrix X must have rownames")
  method <- match.arg(method)
  
  # For each column, get the top 'top' rownames based on largest values
  top_names <- lapply(seq_len(ncol(X)), function(j) {
    sorted_indices <- order(X[, j], decreasing = TRUE)
    rownames(X)[sorted_indices][1:min(top, length(sorted_indices))]
  })
  names(top_names) <- colnames(X)
  
  # Create a pairwise matrix to store overlap values
  ncols <- ncol(X)
  overlap_mat <- matrix(0, nrow = ncols, ncol = ncols)
  rownames(overlap_mat) <- colnames(X)
  colnames(overlap_mat) <- colnames(X)
  
  # Calculate pairwise overlap between columns
  for(i in seq_len(ncols)) {
    for(j in seq_len(ncols)) {
      inter <- length(intersect(top_names[[i]], top_names[[j]]))
      if (method == "intersection") {
        overlap_mat[i, j] <- inter
      } else if (method == "jaccard") {
        union_length <- length(union(top_names[[i]], top_names[[j]]))
        overlap_mat[i, j] <- inter / union_length
      }
    }
  }
  return(overlap_mat)
}
```

```{r}
ov_nmf <- compare_overlap(fit_list$`0.001`$V, top = 25, method = "jaccard")

cor_plot_list <- list()

cor_df <- expand.grid(1:9, 1:9)
colnames(cor_df) <- c("row", "col")

for (cc in cc_vec) {
  
  cor_df$cor <- as.vector(compare_overlap(
    fit_list[[as.character(cc)]]$V,
    top = 25, 
    method = "jaccard"
    ))
    
  library(ggplot2)
  cor_plot_list[[glue::glue("c = {cc}")]] <- ggplot(
      cor_df, 
      aes(x = col, y = row, fill = cor)
    ) +
    geom_tile() +
    # Optional: adjust the color scale
    scale_fill_gradient(
      low = "white",
      high = "red",
      limits = c(0, 1)
    ) +
    theme_minimal() +
    coord_fixed() +
    scale_x_continuous(breaks = 1:9, labels = 1:9) +
    scale_y_reverse(breaks = 1:9, labels = 1:9) +
    xlab("factor") +
    ylab("factor")

}

  cor_df$cor <- as.vector(compare_overlap(
    fit_list$`Inf`$F,
    top = 25, 
    method = "jaccard"
    ))

cor_plot_list[["c = Inf"]] <- ggplot(
      cor_df, 
      aes(x = col, y = row, fill = cor)
    ) +
    geom_tile() +
    # Optional: adjust the color scale
    scale_fill_gradient2(
      low = "blue",
      mid = "white",
      high = "red",
      limits = c(-1, 1)
    ) +
    theme_minimal() +
    coord_fixed() +
    scale_x_continuous(breaks = 1:9, labels = 1:9) +
    scale_y_reverse(breaks = 1:9, labels = 1:9) +
    xlab("factor") +
    ylab("factor")
```


```{r, fig.width=15, fig.height=15}
ggarrange(
  plotlist = cor_plot_list,
  ncol = 3,
  nrow = 3,
  labels = names(cor_plot_list),
  common.legend = TRUE,
  legend = "right"
)
```

Again, for the models with larger $c$, the overlap in the driving genes between factors is much greater. 

Finally, I wanted to look at the distribution of the values of the loadings and factors for each value of $c$. Because the scale varies greatly for different values of $c$, the simplest way to visualize this is to normalize each column so that its largest value is $1$.

Below are the loadings:

```{r}
LL_c001 <- normalize_bars(fit_list$`0.001`$U)
LL_c01 <- normalize_bars(fit_list$`0.01`$U)
LL_cp1 <- normalize_bars(fit_list$`0.1`$U)
LL_c1 <- normalize_bars(fit_list$`1`$U)
LL_c10 <- normalize_bars(fit_list$`10`$U)
LL_c100 <- normalize_bars(fit_list$`100`$U)
LL_c1000 <- normalize_bars(fit_list$`1000`$U)
LL_cinf <- normalize_bars(fit_list$`Inf`$L)

cc_out_vec <- c(
  rep("1e-3", nrow(counts) * 9),
  rep("1e-2", nrow(counts) * 9),
  rep("1e-1", nrow(counts) * 9),
  rep("1", nrow(counts) * 9),
  rep("1e1", nrow(counts) * 9),
  rep("1e2", nrow(counts) * 9),
  rep("1e3", nrow(counts) * 9),
  rep("Inf", nrow(counts) * 9)
)

cc_out_vector <- factor(
  x = cc_out_vec,
  levels = c("1e-3", "1e-2", "1e-1", "1", "1e1", "1e2", "1e3", "Inf")
)

l_df <- data.frame(
  l = c(
    as.vector(LL_c001),
    as.vector(LL_c01),
    as.vector(LL_cp1),
    as.vector(LL_c1),
    as.vector(LL_c10),
    as.vector(LL_c100),
    as.vector(LL_c1000),
    as.vector(LL_cinf)
  ),
  cc = cc_out_vector
)
```

```{r}
ggplot(l_df, aes(x = l)) +
  geom_histogram(bins = 40, fill = "blue", color = "black", alpha = 0.7) +
  facet_wrap(~ cc) +
  labs(title = "Histograms of loadings by c",
       x = "l",
       y = "Count") +
  theme_minimal()
```

And here are the factors:

```{r}
# LL_c001 <- normalize_bars(fit_list$`0.001`$V)
# LL_c01 <- normalize_bars(fit_list$`0.01`$V)
# LL_cp1 <- normalize_bars(fit_list$`0.1`$V)
# LL_c1 <- normalize_bars(fit_list$`1`$V)
# LL_c10 <- normalize_bars(fit_list$`10`$V)
# LL_c100 <- normalize_bars(fit_list$`100`$V)
# LL_c1000 <- normalize_bars(fit_list$`1000`$V)
# LL_cinf <- normalize_bars(fit_list$`Inf`$F)

LL_c001 <- fit_list$`0.001`$V
LL_c01 <- fit_list$`0.01`$V
LL_cp1 <- fit_list$`0.1`$V
LL_c1 <- fit_list$`1`$V
LL_c10 <- fit_list$`10`$V
LL_c100 <- fit_list$`100`$V
LL_c1000 <- fit_list$`1000`$V
LL_cinf <- fit_list$`Inf`$F

cc_out_vec <- c(
  rep("1e-3", ncol(counts) * 9),
  rep("1e-2", ncol(counts) * 9),
  rep("1e-1", ncol(counts) * 9),
  rep("1", ncol(counts) * 9),
  rep("1e1", ncol(counts) * 9),
  rep("1e2", ncol(counts) * 9),
  rep("1e3", ncol(counts) * 9),
  rep("Inf", ncol(counts) * 9)
)

cc_out_vector <- factor(
  x = cc_out_vec,
  levels = c("1e-3", "1e-2", "1e-1", "1", "1e1", "1e2", "1e3", "Inf")
)

l_df <- data.frame(
  l = c(
    as.vector(LL_c001),
    as.vector(LL_c01),
    as.vector(LL_cp1),
    as.vector(LL_c1),
    as.vector(LL_c10),
    as.vector(LL_c100),
    as.vector(LL_c1000),
    as.vector(LL_cinf)
  ),
  cc = cc_out_vector
)
```

```{r}
ggplot(l_df, aes(x = log1p(l))) +
  geom_histogram(bins = 20, fill = "blue", color = "black", alpha = 1, size = 0) +
  facet_wrap(~ cc, scales = "free_x") +
  labs(title = "Histograms of log1p(factors) by c",
       x = "l",
       y = "Count") +
  theme_minimal() +
  cowplot::theme_cowplot()
```
Interestingly, both the loadings and the factors appear to get more sparse as $c$ increases.

```{r}
Lambda_c001 <- 0.001 * (exp(fit_list$`0.001`$U %*% t(fit_list$`0.001`$V)) - 1)
Lambda_c01 <- 0.01 * (exp(fit_list$`0.01`$U %*% t(fit_list$`0.01`$V)) - 1)
Lambda_cp1 <- 0.1 * (exp(fit_list$`0.1`$U %*% t(fit_list$`0.1`$V)) - 1)
Lambda_c1 <- (exp(fit_list$`1`$U %*% t(fit_list$`1`$V)) - 1)
Lambda_c10 <- 10 * (exp(fit_list$`10`$U %*% t(fit_list$`10`$V)) - 1)
Lambda_c100 <- 100 * (exp(fit_list$`100`$U %*% t(fit_list$`100`$V)) - 1)
Lambda_c1000 <- 1000 * (exp(fit_list$`1000`$U %*% t(fit_list$`1000`$V)) - 1)
Lambda_cinf <- fit_list$`Inf`$L %*% t(fit_list$`Inf`$F)
```

```{r}
sample_size <- round(prod(dim(counts)) * 0.01)
samp_idx <- sample.int(n = prod(dim(counts)), size = sample_size)

l_samp_df <- data.frame(
  lambda_c001 = as.vector(Lambda_c001)[samp_idx],
  lambda_c1000 = as.vector(Lambda_c1000)[samp_idx]
)
ggplot(data = l_samp_df) +
  geom_point(aes(x = log(lambda_c001), y = log(lambda_c1000)), alpha = 0.1) +
  xlab("log(lambda) for c = 1e-3") +
  ylab("log(lambda) for c = 1e3")
```



```{r}
sample_size <- round(prod(dim(counts)) * 0.2)
samp_idx <- sample.int(n = prod(dim(counts)), size = sample_size)

cc_out_vec <- c(
  rep("1e-3", sample_size),
  rep("1e-2", sample_size),
  rep("1e-1", sample_size),
  rep("1", sample_size),
  rep("1e1", sample_size),
  rep("1e2", sample_size),
  rep("1e3", sample_size),
  rep("Inf", sample_size)
)

cc_out_vector <- factor(
  x = cc_out_vec,
  levels = c("1e-3", "1e-2", "1e-1", "1", "1e1", "1e2", "1e3", "Inf")
)

l_df <- data.frame(
  l = c(
    as.vector(Lambda_c001)[samp_idx],
    as.vector(Lambda_c01)[samp_idx],
    as.vector(Lambda_cp1)[samp_idx],
    as.vector(Lambda_c1)[samp_idx],
    as.vector(Lambda_c10)[samp_idx],
    as.vector(Lambda_c100)[samp_idx],
    as.vector(Lambda_c1000)[samp_idx],
    as.vector(Lambda_cinf)[samp_idx]
  ),
  cc = cc_out_vector
)
```

```{r}
rm(fit_list, fit_list_nnlm, Lambda_c001, Lambda_c01, Lambda_c1, Lambda_c10, Lambda_c100, Lambda_c1000, Lambda_cinf, Lambda_cp1, counts, sample_info, i, genes_to_use, cc_out_vec, s, n, p, K, cc, cc_vec, cc_out_vector, sample_size, samp_idx)
gc()
```

```{r}
ggplot(l_df, aes(x = log(l))) +
  geom_histogram(bins = 40, fill = "blue", color = "black", alpha = 0.7) +
  facet_wrap(~ cc) +
  labs(title = "Histograms of factors by c",
       x = "Lambda",
       y = "Count") +
  theme_minimal()
```

