---
title: "Analysis of Pancreas Gamma and Delta Cells in Cel-Seq Data"
output:
  workflowr::wflow_html:
    code_folding: hide
date: "2025-02-14"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

Here, I analyzed the pancreas Cel-Seq dataset that Peter has looked at before, but this time focusing only on gamma and delta cells.

```{r}
library(passPCA)
library(Matrix)
library(dplyr)
set.seed(1)

load("~/Downloads/pancreas.RData")

i           <- which(sample_info$tech == "celseq2")
sample_info <- sample_info[i,]
counts      <- counts[i,]
sample_info <- transform(sample_info,celltype = factor(celltype))
gd_info <- sample_info %>%
  dplyr::filter(celltype %in% c("gamma", "delta"))

counts <- counts[gd_info$id, ]

counts <- counts[, Matrix::colSums(counts) > 0]
genes_to_use <- which(Matrix::colSums(counts>0)>4)
counts <- counts[,genes_to_use]
s <- Matrix::rowSums(counts)
s <- s / mean(s)

K <- 3
cc_vec <- c(1e-4, 1e-3, 1e-2, 1e-1, 1, 1e1, 1e2, 1e3, 1e4)

n <- nrow(counts)
p <- ncol(counts)
```

Now, I fit the log1p model for many values of $c$. I also fit Poisson NMF using `fastTopics`. To fit each model, I first fit a rank $1$ model. Then, I initialize each model with $K = 3$, where the first factor is taken from the rank 1 model and the other $2$ are initialized to very small values.

```{r, eval=FALSE}
for (cc in cc_vec) {

  print(cc)

  set.seed(1)
  fit <- fit_poisson_log1p_nmf(
    Y = counts,
    K = K,
    s = s,
    cc = cc,
    loglik = "exact",
    init_method = "rank1",
    control = list(
      maxiter = 250,
      threads = 35
    )
  )

  readr::write_rds(
    fit, glue::glue("pancreas_log1p_c{cc}_k3_gamma_delta_exact_250_iter.rds")
  )

}

```

Below are the structure plots for each of the models:

```{r, results='hide'}
library(ggpubr)
library(fastTopics)
fit_list <- list()

for (cc in cc_vec) {
  
  fit_list[[as.character(cc)]] <- readr::read_rds(
    glue::glue(
      "~/Documents/data/passPCA/pancreas_log1p_c{cc}_k3_gamma_delta_exact_250_iter.rds"
    )
  )

}

fit_list[["Inf"]] <- readr::read_rds(
  glue::glue("~/Documents/data/passPCA/pancreas_nmf_k3_gamma_delta_exact_250_iter.rds")
)
```

```{r, results='hide'}
normalize_bars <- function(LL) {

  max_col <- apply(LL, 2, max)
  sweep(LL, 2, max_col, FUN = "/")

}

celltype <- gd_info$celltype
celltype <-
  factor(celltype,
         c("delta","gamma"))


plot_list <- list()

for (cc in cc_vec) {

  plot_list[[glue::glue("{cc}: Poisson")]] <- structure_plot(
    normalize_bars(fit_list[[as.character(cc)]]$LL),
    grouping = celltype,gap = 20,perplexity = 70,n = Inf, font.size = 12
    )

}

plot_list[["Inf"]] <- structure_plot(
  fit_list[["Inf"]],
  grouping = celltype,gap = 20,perplexity = 70,n = Inf,font.size = 12
  )
```

```{r, fig.width=20, fig.height=70}
ggarrange(
  plotlist = plot_list,
  ncol = 1,
  labels = paste0("c = ", names(plot_list))
)
```

Below are the spearman correlations between the factors:

```{r}
cor_plot_list <- list()

cor_df <- expand.grid(1:3, 1:3)
colnames(cor_df) <- c("row", "col")

for (cc in cc_vec) {
  
  cor_df$cor <- as.vector(cor(fit_list[[as.character(cc)]]$FF, method = "spearman"))
    
  library(ggplot2)
  cor_plot_list[[glue::glue("c = {cc}")]] <- ggplot(
      cor_df, 
      aes(x = col, y = row, fill = cor)
    ) +
    geom_tile() +
    # Optional: adjust the color scale
    scale_fill_gradient2(
      low = "blue",
      mid = "white",
      high = "red",
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    theme_minimal() +
    coord_fixed() +
    scale_x_continuous(breaks = 1:3, labels = 1:3) +
    scale_y_reverse(breaks = 1:3, labels = 1:3) +
    xlab("factor") +
    ylab("factor")

}

cor_df$cor <- as.vector(cor(fit_list$`Inf`$F, method = "spearman"))

cor_plot_list[["c = Inf"]] <- ggplot(
      cor_df, 
      aes(x = col, y = row, fill = cor)
    ) +
    geom_tile() +
    # Optional: adjust the color scale
    scale_fill_gradient2(
      low = "blue",
      mid = "white",
      high = "red",
      midpoint = 0,
      limits = c(-1, 1)
    ) +
    theme_minimal() +
    coord_fixed() +
    scale_x_continuous(breaks = 1:3, labels = 1:3) +
    scale_y_reverse(breaks = 1:3, labels = 1:3) +
    xlab("factor") +
    ylab("factor")
```

```{r, fig.width=15, fig.height=15}
ggarrange(
  plotlist = cor_plot_list,
  ncol = 3,
  nrow = 4,
  labels = names(cor_plot_list),
  common.legend = TRUE,
  legend = "right"
)
```

```{r, fig.height=8}
l_df <- data.frame(
  factr = rep(c(
    rep("1", nrow(fit_list$`1`$FF)),
    rep("2", nrow(fit_list$`1`$FF)),
    rep("3", nrow(fit_list$`1`$FF))
  ), 3),
  l = c(
    as.vector(fit_list$`1e-04`$FF),
    as.vector(fit_list$`1`$FF),
    as.vector(fit_list$`1000`$FF)
  ),
  cc = c(
    rep("1e-4", prod(dim(fit_list$`1`$FF))),
    rep("1", prod(dim(fit_list$`1`$FF))),
    rep("1e4", prod(dim(fit_list$`1`$FF)))
  )
)

l_df$cc <- factor(l_df$cc, levels = c("1e-4", "1", "1e4"))

ggplot(l_df, aes(x = log1p(l))) +
  geom_histogram(bins = 40, fill = "blue", color = "white", alpha = 0.7) +
  facet_wrap(~ cc + factr, scales = "free_x", labeller = "label_both") +
  labs(title = "Histograms of factors by c",
       y = "Count") +
  theme_minimal() +
  cowplot::theme_cowplot() +
  xlab("log1p(f)")
```

```{r, fig.height=8}
l_df <- data.frame(
  factr = rep(c(
    rep("1", nrow(fit_list$`1`$LL)),
    rep("2", nrow(fit_list$`1`$LL)),
    rep("3", nrow(fit_list$`1`$LL))
  ), 3),
  l = c(
    as.vector(fit_list$`1e-04`$LL),
    as.vector(fit_list$`1`$LL),
    as.vector(fit_list$`1000`$LL)
  ),
  cc = c(
    rep("1e-4", prod(dim(fit_list$`1`$LL))),
    rep("1", prod(dim(fit_list$`1`$LL))),
    rep("1e4", prod(dim(fit_list$`1`$LL)))
  )
)

l_df$cc <- factor(l_df$cc, levels = c("1e-4", "1", "1e4"))

ggplot(l_df, aes(x = log1p(l))) +
  geom_histogram(bins = 40, fill = "blue", color = "white", alpha = 0.7) +
  facet_wrap(~ cc + factr, scales = "free_x", labeller = "label_both") +
  labs(title = "Histograms of loadings by c",
       y = "Count") +
  theme_minimal() +
  cowplot::theme_cowplot() +
  xlab("log1p(l)")
```


