---
title: Analysis of the "pancreas cytokine" data set (untreated mouse only) using the Poisson log1p NMF model
author: Eric Weine and Peter Carbonetto
output:
  workflowr::wflow_html:
    toc: no
    theme: readable
    highlight: textmate
    lib_dir: site_libs
    self_contained: no
---

Here we analyze the
["pancreas cytokine" data set](pancreas_cytokine.html), focussing on
the scRNA-seq data from untreated mouse only.

```{r knitr-opts, include=FALSE}
knitr::opts_chunk$set(comment = "#",collapse = TRUE,results = "hold",
                      fig.align = "center",dpi = 120)
```

Load packages used to process the data, perform the analyses, and
create the plots.

```{r load-pkgs, message=FALSE}
library(Matrix)
library(NNLM)
library(fastTopics)
library(log1pNMF)
library(ggplot2)
library(cowplot)
```

Set the seed for reproducibility:

```{r set-seed}
set.seed(1)
```

Load the prepared data set:

```{r load-data}
load("~/Downloads/pancreas_cytokine.RData")
```

Here we will analyze the cells from the untreated mouse only:

```{r filter-cells-1}
i       <- which(samples$mouse == "S1")
samples <- samples[i,]
counts  <- counts[i,]
```

Remove three cells that appear to be outliers (one of them appears to
be an acinar cell based on Eric's analysis):

```{r filter-cells-2}
outliers <- c("TTTGTTGTCGTTAGTG-1","TTTGTTGGTAGAGCTG-1","CCCAACTCACTCATAG-1")
i        <- which(!is.element(samples$barcode,outliers))
samples  <- samples[i,]
counts   <- counts[i,]
```

Remove genes that are expressed in fewer than 5 cells:

```{r filter-genes}
j      <- which(colSums(counts > 0) > 4)
genes  <- genes[j,]
counts <- counts[,j]
```

This is the dimension of the data set we will analyze:

```{r dim-counts}
dim(counts)
```

For the Gaussian-based analyses (later), we will need the shifted log
counts:

```{r shifted-log-counts}
a <- 1
s <- rowSums(counts)
s <- s/mean(s)
shifted_log_counts <- log1p(counts/(a*s))
rownames(shifted_log_counts) <- NULL
```

## Topic model (fastTopics)

Fit a topic model to the counts (with $K = 13$ topics):

```{r fit-topic-model, cache=TRUE}
set.seed(1)
tm <- fit_poisson_nmf(counts,k = 13,init.method = "random",method = "em",
                      numiter = 40,verbose = "none",
					  control = list(numiter = 4,nc = 8,extrapolate = FALSE))
tm <- fit_poisson_nmf(counts,fit0 = tm,method = "scd",numiter = 40,
                      control = list(numiter = 4,nc = 8,extrapolate = TRUE),
					  verbose = "none")
```

Structure plots comparing the topics to the clusters (some of which
are inferred from the topics):

```{r structure-plot-topic-model, fig.height=5, fig.width=6.5, results="hide", message=FALSE}
set.seed(1)
celltype_topics <- paste0("k",c(2,3,4,5,7:9,11,12,13))
other_topics <- paste0("k",c(1,6,10))
L <- poisson2multinom(tm)$L
clusters <- as.character(samples$cluster)
clusters[clusters == "islet"]                  <- "beta"
clusters[clusters == "beta" & L[,"k3"] > 0.25] <- "alpha"
clusters[clusters == "beta" & L[,"k8"] > 0.25] <- "delta+epsilon"
clusters[clusters == "beta" & L[,"k9"] > 0.25] <- "gamma"
clusters <- factor(clusters,c("beta","alpha","delta+epsilon","gamma","duct",
                              "endothelial-mesenchymal","macrophage"))
i <- c(sample(which(clusters == "beta"),400),
       which(clusters != "beta"))
p1 <- structure_plot(L[i,],grouping = clusters[i],topics = celltype_topics,
                     gap = 10,n = Inf) +
  labs(fill = "")
p2 <- structure_plot(L[i,],grouping = clusters[i],topics = other_topics,
                     gap = 10,n = Inf) +
  labs(fill = "")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```

## Poisson log1p NMF, with c = 1

Fit a Poisson log1p NMF model to the counts, first by running NMF on
the shifted log counts (this seems to greatly speed up the model
fitting):

```{r fit-log1p-nmf, cache=TRUE}
set.seed(1)
shifted_log_counts_dense <- as.matrix(shifted_log_counts)
nmf <- nnmf(shifted_log_counts_dense,k = 13,loss = "mse",method = "scd",
            max.iter = 100,verbose = 0,n.threads = 8)
rm(shifted_log_counts_dense)
fit_log1p <- fit_poisson_log1p_nmf(counts,K = 13,cc = 1,init_LL = nmf$W,
                                   init_FF = t(nmf$H),loglik = "exact",
                                   control = list(maxiter=100,verbose=FALSE))
```

Compare the factors from the Poisson log1p NMF model to the clusters:

```{r structure-plot-log1p-nmf, fig.height=5, fig.width=6.5, results="hide", message=FALSE}
scale_cols <- function (A, b)
  t(t(A) * b)
celltype_topics <- paste0("k",c(3,5:6,8:12))
other_topics <- paste0("k",c(1,2,4,7,13))
set.seed(1)
L <- fit_log1p$LL
d <- apply(L,2,max)
L <- scale_cols(L,1/d)
colnames(L) <- paste0("k",1:13)
i <- c(sample(which(clusters == "beta"),400),
       which(clusters != "beta"))
i <- c(sample(which(clusters == "beta"),400),
       which(clusters != "beta"))
p1 <- structure_plot(L[i,],grouping = clusters[i],topics = celltype_topics,
                     gap = 10,n = Inf) +
  labs(y = "membership",fill = "")					
p2 <- structure_plot(L[i,],grouping = clusters[i],topics = other_topics,
                     gap = 10,n = Inf) +
  labs(y = "membership",fill = "")
plot_grid(p1,p2,nrow = 2,ncol = 1)
```

Factor 4 is clearly picking up systematic differences among the two
replicates:

```{r boxplot-replicate, fig.height=2, fig.width=2}
pdat <- cbind(samples,L)
ggplot(pdat,aes(x = replicate,y = k4)) +
  geom_boxplot() +
  theme_cowplot()
```

(Topic 6 also picks up this same batch effect, however the correlation
with the topic 6 proportion is not as strong as the factor 4 membership.)
