---
title: "The calibrated simulation: a paper-ready description"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

This page describes **exactly** the simulation that the exact-vs-approx /
sparsity experiment will run. To make drift impossible, every constant below
is read from the experiment's own configuration file
(`inst/paper_figures/experiments/approx_sim_common.R` in the log1pNMF repo),
which is what the cluster workers source — including the calibrated knobs,
which that file in turn reads from `approx_sim_calibration.rds` with a
consistency check. An earlier version of this page developed the method with
a MOCA anchor and a single $c_{\textrm{true}}$; that material is in the git
history.

```{r source-config}
library(ggplot2); library(cowplot)
theme_set(theme_cowplot(font_size = 11))

exp_dir <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
source(file.path(exp_dir, "approx_sim_common.R"), chdir = TRUE)

knitr::kable(data.frame(
  parameter = c("n (samples)", "p (features)", "K (rank, true = fitted)",
                "Dirichlet concentration a", "t degrees of freedom",
                "rate-scale magnitude floor", "c_true grid"),
  value = c(N_ROWS, N_COLS, K_TRUE, A_SIG, T_DF, CAL$rate_floor,
            paste(sapply(C_TRUE_GRID, fmtc), collapse = ", ")),
  role = "fixed by design"),
  caption = "The design constants, read from approx_sim_common.R.")

kb <- KNOBS
kb$c_true <- sapply(kb$c_true, fmtc)
knitr::kable(kb, digits = 4,
  caption = "The calibrated knobs, one row per c_true. pi0, sigma and cap are
  searched; mu is profiled (solved so the expected mean count equals the real
  0.674); floor is the fixed rate-scale floor mapped through each link.")
```

# The description, as the paper would give it

> Counts were simulated from the log1p NMF model itself at each of three
> link constants $c_{\textrm{true}} \in \{10^{-3}, 1, \infty\}$, with
> parameters calibrated so that all three regimes reproduce the count
> characteristics of the pancreas dataset. For each of ten seeds we drew an
> $n \times K$ loading matrix $L$ with rows
> $\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(0.3, \dots, 0.3)$ — the same
> $L$ at every $c_{\textrm{true}}$, so the true loadings and their sparsity
> are identical across regimes — and a $p \times K$ factor matrix $F$ with
> independent entries that are zero with probability $\pi_0$ and otherwise
> $\exp(\mu + \sigma t)$, $t \sim t_4$, truncated to
> $[T_{\textrm{lo}}, T]$. Each feature was conditioned to load on at least
> one factor, and the lower truncation (0.05 expected counts, mapped through
> the link) guarantees it does so at a detectable level; together these are
> the analogue of filtering genes for detectable expression. Counts were
> drawn as $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$ with
> $\lambda_{ij} = c_{\textrm{true}}(e^{\theta_{ij}} - 1)$ for finite
> $c_{\textrm{true}}$ and $\lambda_{ij} = \theta_{ij}$ at
> $c_{\textrm{true}} = \infty$, where $\Theta = LF^{\top}$. The constants
> $(\pi_0, \mu, \sigma, T)$ were calibrated separately at each
> $c_{\textrm{true}}$ against the same two targets: the real mean count,
> matched exactly, and the curve
> $c \mapsto \operatorname{pm}(\log(1 + Y/c))$ — the peak-to-mean ratio of
> the data on the link scale, which by [the sparsity results] governs the
> factor sparsity attainable at each fitting $c$ — matched in least squares
> across $c \in [10^{-4}, 10^{4}]$ after adjusting the real curve to the
> simulation's size. Calibrating every regime to one shared target is what
> makes the datasets comparable across $c_{\textrm{true}}$.

Everything in that paragraph is verified below on the thirty datasets
(ten seeds $\times$ three $c_{\textrm{true}}$) the experiment will actually
use.

# The generative model, precisely

For seed $s \in \{1, \dots, `r N_SEEDS`\}$ and each $c_{\textrm{true}}$, with
knobs $(\pi_0, \mu, \sigma, T, T_{\textrm{lo}})$ from that
$c_{\textrm{true}}$'s row of the table above:

1. $\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(a, \dots, a)$ independently,
   $a = `r A_SIG`$. The draw depends only on the seed, so the same $L$
   underlies all three $c_{\textrm{true}}$ (as do the raw $u_{jk}$ and
   $t_{jk}$ draws behind $F$).
2. $u_{jk} \sim \textrm{Unif}(0,1)$ and $t_{jk} \sim t_{`r T_DF`}$
   independently, and
   $$f_{jk} = \textrm{clamp}\left(\exp(\mu + \sigma t_{jk});\
   T_{\textrm{lo}},\, T\right) \cdot \mathbf{1}\{u_{jk} > \pi_0\},$$
   except that a feature $j$ with $u_{jk} \leq \pi_0$ for every $k$ keeps its
   single largest-$u_{jk}$ factor (so no feature is dead; deterministic given
   $u$).
3. $\Theta = LF^{\top}$;
   $\lambda_{ij} = c_{\textrm{true}}(e^{\theta_{ij}} - 1)$ for finite
   $c_{\textrm{true}}$, and $\lambda_{ij} = \theta_{ij}$ at
   $c_{\textrm{true}} = \infty$ (the plain Poisson NMF limit);
   $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$. In the rare event a row or
   column of $Y$ is entirely zero, the Poisson draw (only) is repeated with
   the next sub-seed; the fitting code rejects all-zero columns, as do real
   pipelines.

There are no size factors ($s_i = 1$; everything is fit with `s = FALSE`).
The two truncation points have distinct jobs. The cap bounds the maximum
count, which otherwise swings by an order of magnitude across seeds under an
unbounded heavy tail. The floor exists because a feature active in a single
factor can otherwise draw a magnitude so small that its column of $Y$ is all
zero with probability near one, which the fitting code rejects — the seed-7
dataset did exactly this before the floor was added. The floor is defined on
the **rate scale** — `r CAL$rate_floor` expected counts — and mapped through
each link ($T_{\textrm{lo}} = \log(1 + 0.05/c_{\textrm{true}})$, i.e.
`r round(log1p(0.05/1e-3), 2)` at $c_{\textrm{true}} = 10^{-3}$ and $0.05$ at
$c_{\textrm{true}} = \infty$), so it imposes the same detectability
requirement in every regime.

# How the calibrated numbers were chosen

The target — **identical for every $c_{\textrm{true}}$, which is the point**
— is the real pancreas count matrix (7,606 cells $\times$ 18,195 genes;
82.2% zeros, mean 0.674). Two properties are matched:

- **The budget curve** $c \mapsto \operatorname{pm}(\log(1+Y/c))$, in least
  squares on the log scale over the experiment's own fitting grid. The curve
  depends only on the count histogram, and it is the quantity our sparsity
  theory identifies as governing what any fit can do at each $c$. Because it
  involves a maximum it is **size-adjusted**: the target uses
  $t^{*} = `r CAL$tstar`$, the median maximum of $np = `r N_ROWS * N_COLS`$
  draws from the real marginal, not the real matrix's own maximum.
- **The mean count, exactly.** The approximation experiments consume the
  absolute scale of the counts, and the curve alone pins it only loosely.
  $\mu$ is therefore profiled — solved by one-dimensional root-finding inside
  every calibration step — rather than searched; Nelder–Mead searches only
  $(\pi_0, \sigma, T)$, with common random numbers, and the final $\mu$ is
  re-solved on 20 seeds.

The calibration code is `calibrate_approx_sim.R`, next to the experiment
scripts; it loops over the $c_{\textrm{true}}$ grid and writes
`approx_sim_calibration.rds`, which `approx_sim_common.R` reads back with a
consistency check.

# Verification on the thirty datasets the experiment will use

Not held-out seeds — these are the (seed, $c_{\textrm{true}}$) datasets
exactly as the workers will regenerate them.

```{r verify, cache=TRUE}
ds <- list()
for (ct in C_TRUE_GRID)
  for (s in seq_len(N_SEEDS))
    ds[[paste(fmtc(ct), s)]] <- sim_dataset(s, ct)

curve_of <- function(Y) sapply(CAL$CS, function(cc) {
  g <- log1p(Y / cc); max(g) / mean(g) })

ratio_tab <- data.frame(`c for fitting` = format(CAL$CS), check.names = FALSE)
for (ct in C_TRUE_GRID) {
  cvs <- sapply(seq_len(N_SEEDS),
                function(s) curve_of(ds[[paste(fmtc(ct), s)]]$Y))
  r <- cvs / CAL$target
  ratio_tab[[paste0("c_true = ", fmtc(ct))]] <-
    sprintf("%.2f  (%.2f–%.2f)", apply(r, 1, median),
            apply(r, 1, min), apply(r, 1, max))
}
knitr::kable(ratio_tab, caption = "Budget-curve ratio to the shared
  size-adjusted pancreas target: median (range) over the ten seeds, per
  c_true. 1.00 = exact match.")

stat_tab <- data.frame(
  what = c("zero fraction", "mean count", "max count"), check.names = FALSE)
for (ct in C_TRUE_GRID) {
  Ys <- lapply(seq_len(N_SEEDS), function(s) ds[[paste(fmtc(ct), s)]]$Y)
  zf <- sapply(Ys, function(Y) mean(Y == 0)); mn <- sapply(Ys, mean)
  mx <- sapply(Ys, max)
  stat_tab[[paste0("c_true = ", fmtc(ct))]] <- c(
    sprintf("%.3f (%.3f–%.3f)", median(zf), min(zf), max(zf)),
    sprintf("%.3f (%.3f–%.3f)", median(mn), min(mn), max(mn)),
    sprintf("%.0f (%d–%d)", median(mx), min(mx), max(mx)))
}
stat_tab$real <- c("0.822", "0.674", paste0(CAL$tstar, " (size-adj.)"))
knitr::kable(stat_tab, caption = "Count characteristics, median (range) over
  the ten seeds.")
```

```{r tail-plot, fig.width=7, fig.height=4}
zp   <- readRDS("data/pancreas_count_histogram.rds")
p_tp <- c(zp$np - sum(zp$h), zp$h) / zp$np
ccdf_real <- rev(cumsum(rev(p_tp[-1])))
d2 <- data.frame(t = seq_along(ccdf_real), p = ccdf_real,
                 what = "real (pancreas)")
for (ct in C_TRUE_GRID) {
  Yall <- unlist(lapply(seq_len(N_SEEDS),
                        function(s) { Y <- ds[[paste(fmtc(ct), s)]]$Y; Y[Y > 0] }))
  cc <- rev(cumsum(rev(tabulate(Yall, nbins = max(Yall))))) /
        (N_SEEDS * N_ROWS * N_COLS)
  d2 <- rbind(d2, data.frame(t = seq_along(cc), p = cc,
                             what = paste0("simulated, c_true = ", fmtc(ct))))
}
d2 <- d2[d2$p > 0 & d2$t <= 3000, ]
ggplot(d2, aes(t, p, colour = what)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c(`real (pancreas)` = "#C2410C",
    `simulated, c_true = 0.001` = "#0E8FA8",
    `simulated, c_true = 1` = "#0E5C6B",
    `simulated, c_true = Inf` = "#B45309"), name = NULL) +
  labs(x = "count value t", y = "P(Y >= t)",
       title = "Tail of the count distribution, per c_true",
       subtitle = "not targeted directly; inherited from the curve match")
```

The controlled side of the design — the truth the fits will be scored
against. The loading rows are identical across $c_{\textrm{true}}$ by
construction, and the table shows it:

```{r truth-side}
hoyer <- function(x) { m <- length(x)
  (sqrt(m) - sum(x)/sqrt(sum(x^2)))/(sqrt(m) - 1) }
tt <- data.frame(
  what = c("mean Hoyer sparsity of loading columns",
           "mean # factors carrying 90% of a sample",
           "fraction of exact zeros in F",
           "mean # factors per feature",
           "mean Hoyer sparsity of factor columns"), check.names = FALSE)
for (ct in C_TRUE_GRID) {
  D <- lapply(seq_len(N_SEEDS), function(s) ds[[paste(fmtc(ct), s)]])
  tt[[paste0("c_true = ", fmtc(ct))]] <- c(
    round(median(sapply(D, function(d) mean(apply(d$L, 2, hoyer)))), 3),
    round(median(sapply(D, function(d) mean(apply(d$L, 1, function(r)
      min(which(cumsum(sort(r, decreasing = TRUE)) >= 0.9)))))), 2),
    round(median(sapply(D, function(d) mean(d$FF == 0))), 3),
    round(median(sapply(D, function(d) mean(rowSums(d$FF > 0)))), 2),
    round(median(sapply(D, function(d) mean(apply(d$FF, 2, hoyer)))), 3))
}
knitr::kable(tt, caption = "Truth structure, median over the ten seeds. The
  loading rows (first two lines) are byte-identical across c_true; the factor
  law differs because its knobs are calibrated per c_true.")
```

# Honest limits of the match at K = 5

The same machinery matched the pancreas curve to within $\pm 16\%$ at
$600 \times 600$, $K = 10$ (git history of this page). At $K = 5$ it cannot
do as well, and the reason is structural: with five factors and
Dirichlet-peaked rows, each entry's rate is nearly a single-factor draw, and
the knobs cannot feed the zero fraction, the mean, and the extreme tail
simultaneously. Two richer variants — a free tail degree of freedom and
two-component magnitudes — fixed one end of the curve only by sacrificing
the other, and were rejected.

The compromise differs by regime, as the tables above show.
$c_{\textrm{true}} = 1$ is the best matched. At
$c_{\textrm{true}} = 10^{-3}$ the exponential link amplifies the magnitudes,
which helps the tail (maxima near 5,000) at some cost in zeros.
$c_{\textrm{true}} = \infty$ is the weakest match: with the identity link the
tail must come entirely from the factor magnitudes themselves, with no
amplification and every loading below one attenuating them, so the maxima
reach only $\sim$1,000 and the zeros run high. This is worth knowing when
reading the results, but it does not undercut the comparison design: all
three regimes are calibrated to the same target, and the fits at every
$c_{\textrm{fit}}$ within a regime see identical data.

# The experiment this feeds

Read from the same configuration file:

```{r protocol}
knitr::kable(data.frame(
  setting = c("datasets", "c_true grid (generation)", "c grid for fitting",
              "methods", "initialization", "convergence tolerance",
              "iteration safety cap", "optimization-time cap",
              "SLURM wall clock", "threads per fit", "tasks"),
  value = c(sprintf("%d seeds per c_true (same L across c_true)", N_SEEDS),
            paste(sapply(C_TRUE_GRID, fmtc), collapse = ", "),
            paste(format(CC_GRID, scientific = TRUE, trim = TRUE), collapse = ", "),
            paste(METHODS, collapse = ", "),
            "rank-1 (each method's own warm-up, seeded)",
            format(TOL, scientific = TRUE),
            format(MAXITER, big.mark = ","),
            "none",
            "24h (backstop only)",
            THREADS,
            sprintf("%d  (one fit per task)", N_TASKS))),
  caption = "The protocol, from approx_sim_common.R and run_approx_sim.sbatch.")
```

Every fit runs until the change in **its own objective** falls below the
tolerance — the approximate methods converge on the approximate objective —
and the comparison number recorded for every fit is the **exact** Poisson
log-likelihood of its fitted rates. The exact fits across the
$c_{\textrm{fit}}$ grid double as the sparsity experiment: fitted $L$ and
$F$ are saved for every task. After the round, `check_approx_sim.R` lists
anything missing or unconverged for resubmission.

# Results

The experiment has now run. All $810$ fits are present and all converged.
Twenty-one needed more than the 250k-iteration cap and were rerun from
scratch with a 1M cap; eighteen of those were Taylor fits at
$c_{\textrm{fit}} \leq 10^{-3}$, a first symptom of the Taylor pathology
quantified below. Metrics were computed on the cluster by
`analyze_approx_sim.R` (one row per fit; Hoyer sparsity and pairwise
Spearman correlations of the fitted $L$ and $F$, plus the stored exact
log-likelihoods), and only that table is read here.

```{r load-results}
res <- readRDS(file.path(exp_dir, "approx_sim_metrics.rds"))
res$ct <- factor(paste0("c_true = ", sapply(res$c_true, fmtc)),
                 levels = paste0("c_true = ", sapply(C_TRUE_GRID, fmtc)))
res$method <- factor(res$method, levels = METHODS)

## per-entry exact-loglik gap of each fit vs the exact method's fit on the
## SAME (seed, c_true, c_fit)
key <- with(res, paste(seed, c_true, cc))
exi <- res$method == "exact"
res$gap_pe <- (res$loglik_exact -
               res$loglik_exact[exi][match(key, key[exi])]) / (N_ROWS * N_COLS)

meth_cols <- c(exact = "#111111", taylor = "#C2410C", chebyshev = "#0E8FA8")

knitr::kable(data.frame(
  method = METHODS,
  `converged / fits` = sapply(METHODS, function(m)
    sprintf("%d / %d", sum(res$converged[res$method == m]),
            sum(res$method == m))),
  `median seconds` = sapply(METHODS, function(m)
    round(median(res$seconds[res$method == m]), 1)),
  `max seconds` = sapply(METHODS, function(m)
    round(max(res$seconds[res$method == m]))),
  check.names = FALSE, row.names = NULL),
  caption = "Convergence and timing. The Taylor maximum is a c_fit <= 1e-3
  fit that needed the extended 1M-iteration cap.")
```

## Approximation accuracy

The comparison number is the **exact** Poisson log-likelihood of the fitted
rates, whichever objective produced them, expressed per matrix entry and
relative to the exact method's fit of the same dataset at the same
$c_{\textrm{fit}}$. Negative means the approximation ended somewhere worse.

```{r approx-gap-tables}
for (m in c("taylor", "chebyshev")) {
  s <- res$method == m
  tb <- tapply(res$gap_pe[s], list(res$ct[s], res$cc[s]), mean)
  print(knitr::kable(signif(tb, 2), caption = paste0(
    "Mean per-entry gap, ", m, " (rows = c_true, cols = c_fit).")))
}
```

```{r approx-gap-fig, fig.width=8.5, fig.height=3.6}
ap <- subset(res, method != "exact")
ap$mag  <- pmax(abs(ap$gap_pe), 1e-12)
ap$sign <- ifelse(ap$gap_pe < 0, "worse than exact", "better than exact")
ggplot(ap, aes(cc, mag, colour = method)) +
  geom_point(aes(shape = sign), size = 1.7, alpha = 0.65,
             position = position_jitter(width = 0.07, height = 0)) +
  stat_summary(fun = median, geom = "line", linewidth = 0.7) +
  geom_hline(yintercept = 0.01, linetype = 3) +
  scale_x_log10(breaks = 10^seq(-4, 4, 2),
                labels = scales::label_log()) +
  scale_y_log10(breaks = 10^seq(-10, 2, 3), labels = scales::label_log()) +
  scale_colour_manual(values = meth_cols[-1], name = NULL) +
  scale_shape_manual(values = c(`worse than exact` = 16,
                                `better than exact` = 2), name = NULL) +
  facet_wrap(~ ct, nrow = 1) +
  labs(x = "c used for fitting", y = "|per-entry gap|  (log scale)",
       title = "Exact log-likelihood given up by each approximation",
       subtitle = "one point per seed; line = median; dotted line = 0.01 per entry")
```

Two regimes are visible. For $c_{\textrm{fit}} \geq 0.1$ **both**
approximations track the exact objective closely — per-entry gaps of
$10^{-2}$ down to $10^{-10}$, i.e. from negligible to identical, and the
scattered "better than exact" triangles (approximation landing in a slightly
better local optimum of the common exact criterion) confirm the remaining
spread is local-optimum noise rather than approximation error, as does the
occasional cell where taylor and chebyshev land the *same* gap. For
$c_{\textrm{fit}} \leq 10^{-2}$ the two methods part company: **Chebyshev
degrades gracefully** (never worse than $\sim 0.03$ per entry anywhere on
the grid) while **Taylor fails outright**, giving up tens to thousands of
log-likelihood units per entry at $c_{\textrm{fit}} \leq 10^{-3}$ — and it
fails slowly, burning hours to 1M iterations on the way (the timing table
above). The Chebyshev coefficients are fit to the relevant interval, while
the Taylor expansion is anchored at a point that small $c$ pushes far from
where the zero-term sum is evaluated.

## Sparsity of the fitted loadings and factors

Mean Hoyer sparsity of the columns of the fitted $L$ and $F$, against the
truth's own sparsity (dashed). The theory says the attainable sparsity is
governed by $c_{\textrm{fit}}$ through the budget
$\operatorname{pm}(\log(1+Y/c))$, increasing in $c$; the design says the
truth was generated with loading sparsity **identical across regimes**
(dashed line at the same height in every panel of the top row).

```{r sparsity-fig, fig.width=8.5, fig.height=5.4}
agg <- aggregate(cbind(hoyer_L, hoyer_F) ~ method + cc + ct, res, mean)
sp  <- rbind(
  data.frame(agg[c("method", "cc", "ct")], what = "loadings L",
             value = agg$hoyer_L),
  data.frame(agg[c("method", "cc", "ct")], what = "factors F",
             value = agg$hoyer_F))
sp$what <- factor(sp$what, levels = c("loadings L", "factors F"))
tru <- unique(res[c("ct", "hoyer_L_true", "hoyer_F_true")])
tru <- aggregate(cbind(hoyer_L_true, hoyer_F_true) ~ ct, tru, mean)
truth_df <- rbind(
  data.frame(ct = tru$ct, what = "loadings L", value = tru$hoyer_L_true),
  data.frame(ct = tru$ct, what = "factors F",  value = tru$hoyer_F_true))
truth_df$what <- factor(truth_df$what, levels = levels(sp$what))
ggplot(sp, aes(cc, value, colour = method)) +
  geom_hline(data = truth_df, aes(yintercept = value), linetype = 2,
             colour = "grey40") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.6) +
  scale_x_log10(breaks = 10^seq(-4, 4, 2), labels = scales::label_log()) +
  scale_colour_manual(values = meth_cols, name = NULL) +
  facet_grid(what ~ ct) +
  labs(x = "c used for fitting", y = "mean Hoyer sparsity of columns",
       title = "Sparsity of the fit is set by the fitting c",
       subtitle = "mean over 10 seeds; dashed = the truth's own sparsity")
```

Three things to read off:

- **Monotonicity.** At every $c_{\textrm{true}}$, for both $L$ and $F$,
  sparsity increases in $c_{\textrm{fit}}$ across the entire eight-decade
  grid — the budget mechanism, on data whose budget curve was calibrated to
  a real dataset's.
- **The fit recovers the truth's sparsity when fit at the true $c$.** The
  fitted curves cross their dashed truth lines at
  $c_{\textrm{fit}} \approx c_{\textrm{true}}$ in every panel: e.g. for
  loadings (truth $0.398$ everywhere) the exact fits give $0.41$ at
  $(10^{-3}, 10^{-3})$, $0.41$ at $(1, 1)$, and $0.40$ at
  $(\infty, 10^{3\text{--}4})$; for factors the crossing is equally tight
  ($|{\Delta}| = 0.004$ at $c_{\textrm{true}} = 1$). Misfit in $c$
  translates directly into over- or under-sparse estimates.
- **The approximations preserve the sparsity story wherever they are
  accurate.** The three method curves are near-coincident for
  $c_{\textrm{fit}} \geq 0.1$; at the small-$c$ end Taylor's curves pull
  away from exact (its fits there are not usable anyway, per the previous
  section), while Chebyshev stays within a few hundredths.

## Rank correlation of the factors

The redundancy measure: mean Spearman correlation over the
$\binom{K}{2} = 10$ pairs of fitted factor columns (the loadings analogue
is shown alongside). The sparsity theory's companion argument predicts that
as columns sparsify, their rankings decouple — pairwise rank correlations
should drift with $c_{\textrm{fit}}$ alongside the sparsity itself.

```{r rankcor-fig, fig.width=8.5, fig.height=5.4}
agg <- aggregate(cbind(rankcor_L, rankcor_F) ~ method + cc + ct, res, mean)
rc  <- rbind(
  data.frame(agg[c("method", "cc", "ct")], what = "loadings L",
             value = agg$rankcor_L),
  data.frame(agg[c("method", "cc", "ct")], what = "factors F",
             value = agg$rankcor_F))
rc$what <- factor(rc$what, levels = c("loadings L", "factors F"))
tru <- aggregate(cbind(rankcor_L_true, rankcor_F_true) ~ ct,
                 unique(res[c("ct", "rankcor_L_true", "rankcor_F_true")]),
                 mean)
truth_df <- rbind(
  data.frame(ct = tru$ct, what = "loadings L", value = tru$rankcor_L_true),
  data.frame(ct = tru$ct, what = "factors F",  value = tru$rankcor_F_true))
truth_df$what <- factor(truth_df$what, levels = levels(rc$what))
ggplot(rc, aes(cc, value, colour = method)) +
  geom_hline(data = truth_df, aes(yintercept = value), linetype = 2,
             colour = "grey40") +
  geom_line(linewidth = 0.7) + geom_point(size = 1.6) +
  scale_x_log10(breaks = 10^seq(-4, 4, 2), labels = scales::label_log()) +
  scale_colour_manual(values = meth_cols, name = NULL) +
  facet_grid(what ~ ct) +
  labs(x = "c used for fitting",
       y = "mean pairwise Spearman correlation",
       title = "Rank correlation between fitted columns",
       subtitle = "mean over 10 seeds; dashed = the truth's own value")
```

For the **loadings** at $c_{\textrm{true}} = 1$ and $\infty$ the effect is
clean: pairwise rank correlation rises toward zero as $c_{\textrm{fit}}$
grows, reaching the truth's value ($-0.174$) near
$c_{\textrm{fit}} \approx c_{\textrm{true}}$, and all three methods lie on
top of each other. At $c_{\textrm{true}} = 10^{-3}$ the exact curve is not
monotone — it peaks around $c_{\textrm{fit}} = 10$ and comes back down where
the fits are heavily over-sparsified — and the methods separate, a first
hint of what the factor panels show clearly. For the **factors** the
movement with $c$ is small and the three methods scatter around each other
even at moderate $c_{\textrm{fit}}$, where their log-likelihoods are nearly
identical: pairwise Spearman between sparse, heavy-tailed columns is
dominated by which local optimum a fit landed in, not by the objective it
optimized, so we read little into the fine structure here. The one robust
feature is that **every fit's factor columns are more rank-correlated than
the truth's** (all curves sit above the dashed lines): the fits never fully
reproduce the decoupling of the generative factors.

## Summary

The simulation does what it was designed to do. On counts calibrated to a
real dataset's budget curve in all three generative regimes: (i) the
sparsity of the fitted loadings and factors is controlled by the fitting
$c$, monotonically and over a wide range; (ii) fitting at the generative
$c$ reproduces the generative sparsity almost exactly, so misfit in $c$ is
visible as over- or under-sparsification; (iii) the Chebyshev approximation
is faithful to the exact objective everywhere on the grid (worst case
$\sim 0.03$ log-likelihood units per entry, typically orders of magnitude
less), while the Taylor approximation is faithful only for
$c_{\textrm{fit}} \gtrsim 0.1$ and fails badly — slowly and to bad optima —
below that.
