---
title: "The calibrated simulation: a paper-ready description"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

This page describes **exactly** the simulation that the exact-vs-approx /
sparsity experiment runs, and reports its results. To make drift impossible, every constant below
is read from the experiment's own configuration file
(`inst/paper_figures/experiments/approx_sim_common.R` in the log1pNMF repo),
which is what the cluster workers source — including the calibrated knobs,
which that file in turn reads from `approx_sim_calibration.rds` with a
consistency check. An earlier version of this page developed the method with
a MOCA anchor and a single $c_{\textrm{true}}$; that material is in the git
history.

```{r source-config}
library(ggplot2); library(cowplot)
theme_set(theme_cowplot(font_size = 11))

exp_dir <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
source(file.path(exp_dir, "approx_sim_common.R"), chdir = TRUE)

knitr::kable(data.frame(
  parameter = c("n (samples)", "p (features)", "K (rank, true = fitted)",
                "Dirichlet concentration a", "t degrees of freedom",
                "rate-scale magnitude floor", "c_true grid"),
  value = c(N_ROWS, N_COLS, K_TRUE, A_SIG, T_DF, CAL$rate_floor,
            paste(sapply(C_TRUE_GRID, fmtc), collapse = ", ")),
  role = "fixed by design"),
  caption = "The design constants, read from approx_sim_common.R.")

kb <- KNOBS
kb$c_true <- sapply(kb$c_true, fmtc)
knitr::kable(kb, digits = 4,
  caption = "The calibrated knobs, one row per c_true. pi0, sigma and cap are
  searched; mu is profiled (solved so the expected mean count equals the real
  0.674); floor is the fixed rate-scale floor mapped through each link.")
```

# The description, as the paper would give it

> Counts were simulated from the log1p NMF model itself at each of three
> link constants $c_{\textrm{true}} \in \{10^{-3}, 1, \infty\}$, with
> parameters calibrated so that all three regimes reproduce the count
> characteristics of the pancreas dataset. For each of thirty seeds we drew an
> $n \times K$ loading matrix $L$ with rows
> $\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(0.3, \dots, 0.3)$ — the same
> $L$ at every $c_{\textrm{true}}$, so the true loadings and their sparsity
> are identical across regimes — and a $p \times K$ factor matrix $F$ with
> independent entries that are zero with probability $\pi_0$ and otherwise
> $\exp(\mu + \sigma t)$, $t \sim t_4$, truncated to
> $[T_{\textrm{lo}}, T]$. Each feature was conditioned to load on at least
> one factor, and the lower truncation (0.05 expected counts, mapped through
> the link) guarantees it does so at a detectable level; together these are
> the analogue of filtering genes for detectable expression. Counts were
> drawn as $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$ with
> $\lambda_{ij} = c_{\textrm{true}}(e^{\theta_{ij}} - 1)$ for finite
> $c_{\textrm{true}}$ and $\lambda_{ij} = \theta_{ij}$ at
> $c_{\textrm{true}} = \infty$, where $\Theta = LF^{\top}$. The constants
> $(\pi_0, \mu, \sigma, T)$ were calibrated separately at each
> $c_{\textrm{true}}$ against the same two targets: the real mean count,
> matched exactly, and the curve
> $c \mapsto \operatorname{pm}(\log(1 + Y/c))$ — the peak-to-mean ratio of
> the data on the link scale, which by [the sparsity results] governs the
> factor sparsity attainable at each fitting $c$ — matched in least squares
> across $c \in [10^{-4}, 10^{4}]$ after adjusting the real curve to the
> simulation's size. Calibrating every regime to one shared target is what
> makes the datasets comparable across $c_{\textrm{true}}$.

Everything in that paragraph is verified below on the ninety datasets
(thirty seeds $\times$ three $c_{\textrm{true}}$) the experiment will actually
use.

# The generative model, precisely

For seed $s \in \{1, \dots, `r N_SEEDS`\}$ and each $c_{\textrm{true}}$, with
knobs $(\pi_0, \mu, \sigma, T, T_{\textrm{lo}})$ from that
$c_{\textrm{true}}$'s row of the table above:

1. $\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(a, \dots, a)$ independently,
   $a = `r A_SIG`$. The draw depends only on the seed, so the same $L$
   underlies all three $c_{\textrm{true}}$ (as do the raw $u_{jk}$ and
   $t_{jk}$ draws behind $F$).
2. $u_{jk} \sim \textrm{Unif}(0,1)$ and $t_{jk} \sim t_{`r T_DF`}$
   independently, and
   $$f_{jk} = \textrm{clamp}\left(\exp(\mu + \sigma t_{jk});\
   T_{\textrm{lo}},\, T\right) \cdot \mathbf{1}\{u_{jk} > \pi_0\},$$
   except that a feature $j$ with $u_{jk} \leq \pi_0$ for every $k$ keeps its
   single largest-$u_{jk}$ factor (so no feature is dead; deterministic given
   $u$).
3. $\Theta = LF^{\top}$;
   $\lambda_{ij} = c_{\textrm{true}}(e^{\theta_{ij}} - 1)$ for finite
   $c_{\textrm{true}}$, and $\lambda_{ij} = \theta_{ij}$ at
   $c_{\textrm{true}} = \infty$ (the plain Poisson NMF limit);
   $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$. In the rare event a row or
   column of $Y$ is entirely zero, the Poisson draw (only) is repeated with
   the next sub-seed; the fitting code rejects all-zero columns, as do real
   pipelines.

There are no size factors ($s_i = 1$; everything is fit with `s = FALSE`).
The two truncation points have distinct jobs. The cap bounds the maximum
count, which otherwise swings by an order of magnitude across seeds under an
unbounded heavy tail. The floor exists because a feature active in a single
factor can otherwise draw a magnitude so small that its column of $Y$ is all
zero with probability near one, which the fitting code rejects — the seed-7
dataset did exactly this before the floor was added. The floor is defined on
the **rate scale** — `r CAL$rate_floor` expected counts — and mapped through
each link ($T_{\textrm{lo}} = \log(1 + 0.05/c_{\textrm{true}})$, i.e.
`r round(log1p(0.05/1e-3), 2)` at $c_{\textrm{true}} = 10^{-3}$ and $0.05$ at
$c_{\textrm{true}} = \infty$), so it imposes the same detectability
requirement in every regime.

# How the calibrated numbers were chosen

The target — **identical for every $c_{\textrm{true}}$, which is the point**
— is the real pancreas count matrix (7,606 cells $\times$ 18,195 genes;
82.2% zeros, mean 0.674). Two properties are matched:

- **The budget curve** $c \mapsto \operatorname{pm}(\log(1+Y/c))$, in least
  squares on the log scale over the experiment's own fitting grid. The curve
  depends only on the count histogram, and it is the quantity our sparsity
  theory identifies as governing what any fit can do at each $c$. Because it
  involves a maximum it is **size-adjusted**: the target uses
  $t^{*} = `r CAL$tstar`$, the median maximum of $np = `r N_ROWS * N_COLS`$
  draws from the real marginal, not the real matrix's own maximum.
- **The mean count, exactly.** The approximation experiments consume the
  absolute scale of the counts, and the curve alone pins it only loosely.
  $\mu$ is therefore profiled — solved by one-dimensional root-finding inside
  every calibration step — rather than searched; Nelder–Mead searches only
  $(\pi_0, \sigma, T)$, with common random numbers, and the final $\mu$ is
  re-solved on 20 seeds.

The calibration code is `calibrate_approx_sim.R`, next to the experiment
scripts; it loops over the $c_{\textrm{true}}$ grid and writes
`approx_sim_calibration.rds`, which `approx_sim_common.R` reads back with a
consistency check.

# Verification on the ninety datasets the experiment uses

Not held-out seeds — these are the (seed, $c_{\textrm{true}}$) datasets
exactly as the workers will regenerate them.

```{r verify, cache=TRUE}
ds <- list()
for (ct in C_TRUE_GRID)
  for (s in seq_len(N_SEEDS))
    ds[[paste(fmtc(ct), s)]] <- sim_dataset(s, ct)

curve_of <- function(Y) sapply(CAL$CS, function(cc) {
  g <- log1p(Y / cc); max(g) / mean(g) })

ratio_tab <- data.frame(`c for fitting` = format(CAL$CS), check.names = FALSE)
for (ct in C_TRUE_GRID) {
  cvs <- sapply(seq_len(N_SEEDS),
                function(s) curve_of(ds[[paste(fmtc(ct), s)]]$Y))
  r <- cvs / CAL$target
  ratio_tab[[paste0("c_true = ", fmtc(ct))]] <-
    sprintf("%.2f  (%.2f–%.2f)", apply(r, 1, median),
            apply(r, 1, min), apply(r, 1, max))
}
knitr::kable(ratio_tab, caption = "Budget-curve ratio to the shared
  size-adjusted pancreas target: median (range) over the thirty seeds, per
  c_true. 1.00 = exact match.")

stat_tab <- data.frame(
  what = c("zero fraction", "mean count", "max count"), check.names = FALSE)
for (ct in C_TRUE_GRID) {
  Ys <- lapply(seq_len(N_SEEDS), function(s) ds[[paste(fmtc(ct), s)]]$Y)
  zf <- sapply(Ys, function(Y) mean(Y == 0)); mn <- sapply(Ys, mean)
  mx <- sapply(Ys, max)
  stat_tab[[paste0("c_true = ", fmtc(ct))]] <- c(
    sprintf("%.3f (%.3f–%.3f)", median(zf), min(zf), max(zf)),
    sprintf("%.3f (%.3f–%.3f)", median(mn), min(mn), max(mn)),
    sprintf("%.0f (%d–%d)", median(mx), min(mx), max(mx)))
}
stat_tab$real <- c("0.822", "0.674", paste0(CAL$tstar, " (size-adj.)"))
knitr::kable(stat_tab, caption = "Count characteristics, median (range) over
  the thirty seeds.")
```

```{r tail-plot, fig.width=7, fig.height=4}
zp   <- readRDS("data/pancreas_count_histogram.rds")
p_tp <- c(zp$np - sum(zp$h), zp$h) / zp$np
ccdf_real <- rev(cumsum(rev(p_tp[-1])))
d2 <- data.frame(t = seq_along(ccdf_real), p = ccdf_real,
                 what = "real (pancreas)")
for (ct in C_TRUE_GRID) {
  Yall <- unlist(lapply(seq_len(N_SEEDS),
                        function(s) { Y <- ds[[paste(fmtc(ct), s)]]$Y; Y[Y > 0] }))
  cc <- rev(cumsum(rev(tabulate(Yall, nbins = max(Yall))))) /
        (N_SEEDS * N_ROWS * N_COLS)
  d2 <- rbind(d2, data.frame(t = seq_along(cc), p = cc,
                             what = paste0("simulated, c_true = ", fmtc(ct))))
}
d2 <- d2[d2$p > 0 & d2$t <= 3000, ]
ggplot(d2, aes(t, p, colour = what)) +
  geom_line(linewidth = 0.9) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c(`real (pancreas)` = "#C2410C",
    `simulated, c_true = 0.001` = "#0E8FA8",
    `simulated, c_true = 1` = "#0E5C6B",
    `simulated, c_true = Inf` = "#B45309"), name = NULL) +
  labs(x = "count value t", y = "P(Y >= t)",
       title = "Tail of the count distribution, per c_true",
       subtitle = "not targeted directly; inherited from the curve match")
```

The controlled side of the design — the truth the fits will be scored
against. The loading rows are identical across $c_{\textrm{true}}$ by
construction, and the table shows it:

```{r truth-side}
hoyer <- function(x) { m <- length(x)
  (sqrt(m) - sum(x)/sqrt(sum(x^2)))/(sqrt(m) - 1) }
tt <- data.frame(
  what = c("mean Hoyer sparsity of loading columns",
           "mean # factors carrying 90% of a sample",
           "fraction of exact zeros in F",
           "mean # factors per feature",
           "mean Hoyer sparsity of factor columns"), check.names = FALSE)
for (ct in C_TRUE_GRID) {
  D <- lapply(seq_len(N_SEEDS), function(s) ds[[paste(fmtc(ct), s)]])
  tt[[paste0("c_true = ", fmtc(ct))]] <- c(
    round(median(sapply(D, function(d) mean(apply(d$L, 2, hoyer)))), 3),
    round(median(sapply(D, function(d) mean(apply(d$L, 1, function(r)
      min(which(cumsum(sort(r, decreasing = TRUE)) >= 0.9)))))), 2),
    round(median(sapply(D, function(d) mean(d$FF == 0))), 3),
    round(median(sapply(D, function(d) mean(rowSums(d$FF > 0)))), 2),
    round(median(sapply(D, function(d) mean(apply(d$FF, 2, hoyer)))), 3))
}
knitr::kable(tt, caption = "Truth structure, median over the thirty seeds. The
  loading rows (first two lines) are byte-identical across c_true; the factor
  law differs because its knobs are calibrated per c_true.")
```

# Honest limits of the match at K = 5

The same machinery matched the pancreas curve to within $\pm 16\%$ at
$600 \times 600$, $K = 10$ (git history of this page). At $K = 5$ it cannot
do as well, and the reason is structural: with five factors and
Dirichlet-peaked rows, each entry's rate is nearly a single-factor draw, and
the knobs cannot feed the zero fraction, the mean, and the extreme tail
simultaneously. Two richer variants — a free tail degree of freedom and
two-component magnitudes — fixed one end of the curve only by sacrificing
the other, and were rejected.

The compromise differs by regime, as the tables above show.
$c_{\textrm{true}} = 1$ is the best matched. At
$c_{\textrm{true}} = 10^{-3}$ the exponential link amplifies the magnitudes,
which helps the tail (maxima near 5,000) at some cost in zeros.
$c_{\textrm{true}} = \infty$ is the weakest match: with the identity link the
tail must come entirely from the factor magnitudes themselves, with no
amplification and every loading below one attenuating them, so the maxima
reach only $\sim$1,000 and the zeros run high. This is worth knowing when
reading the results, but it does not undercut the comparison design: all
three regimes are calibrated to the same target, and the fits at every
$c_{\textrm{fit}}$ within a regime see identical data.

# The experiment this feeds

Read from the same configuration file:

```{r protocol}
knitr::kable(data.frame(
  setting = c("datasets", "c_true grid (generation)", "c grid for fitting",
              "methods", "initialization", "convergence tolerance",
              "iteration safety cap", "optimization-time cap",
              "SLURM wall clock", "threads per fit", "tasks"),
  value = c(sprintf("%d seeds per c_true (same L across c_true)", N_SEEDS),
            paste(sapply(C_TRUE_GRID, fmtc), collapse = ", "),
            paste(format(CC_GRID, scientific = TRUE, trim = TRUE), collapse = ", "),
            paste0(paste(METHODS, collapse = ", "),
                   " (c = Inf: exact only, fastTopics)"),
            "rank-1 (each method's own warm-up, seeded)",
            format(TOL, scientific = TRUE),
            format(MAXITER, big.mark = ","),
            "none",
            "24h (backstop only)",
            THREADS,
            sprintf("%d  (one fit per task)", N_TASKS))),
  caption = "The protocol, from approx_sim_common.R and run_approx_sim.sbatch.")
```

Every fit runs until the change in **its own objective** falls below the
tolerance — the approximate methods converge on the approximate objective —
and the comparison number recorded for every fit is the **exact** Poisson
log-likelihood of its fitted rates. The exact fits across the
$c_{\textrm{fit}}$ grid double as the sparsity experiment: fitted $L$ and
$F$ are saved for every task. After the round, `check_approx_sim.R` lists
anything missing or unconverged for resubmission.

# Results

The full experiment has now run: $2{,}520$ fits — thirty seeds, three
$c_{\textrm{true}}$, the nine-point finite $c_{\textrm{fit}}$ grid under
all three methods, **plus $c_{\textrm{fit}} = \infty$** (plain Poisson NMF,
i.e. the topic model, fit with fastTopics from the same rank-1-padded
initialization; the quadratic zero-term approximations have no
$c = \infty$ analogue, so that column is exact-only). All $2{,}520$
converged. The 810 seed-1–10 fits are the first round's, reused verbatim
(their generator, grid and tags are unchanged); the extension ran at a 1M
iteration cap, which the first round's Taylor pathology motivated — this
time nothing hit it (worst fit: 850k iterations, 17.4h). Metrics were
computed on the cluster by `analyze_approx_sim.R` (one row per fit; Hoyer
sparsity and pairwise Spearman correlations of the fitted $L$ and $F$,
plus the stored exact log-likelihoods), and only that table is read here.

```{r load-results}
res <- readRDS(file.path(exp_dir, "approx_sim_metrics.rds"))
res$ct <- factor(paste0("c_true = ", sapply(res$c_true, fmtc)),
                 levels = paste0("c_true = ", sapply(C_TRUE_GRID, fmtc)))
res$method <- factor(res$method, levels = METHODS)

## per-entry exact-loglik gap of each fit vs the exact method's fit on the
## SAME (seed, c_true, c_fit)
key <- with(res, paste(seed, c_true, cc))
exi <- res$method == "exact"
res$gap_pe <- (res$loglik_exact -
               res$loglik_exact[exi][match(key, key[exi])]) / (N_ROWS * N_COLS)

meth_cols <- c(exact = "#111111", taylor = "#C2410C", chebyshev = "#0E8FA8")

knitr::kable(data.frame(
  method = METHODS,
  `converged / fits` = sapply(METHODS, function(m)
    sprintf("%d / %d", sum(res$converged[res$method == m]),
            sum(res$method == m))),
  `median seconds` = sapply(METHODS, function(m)
    round(median(res$seconds[res$method == m]), 1)),
  `max seconds` = sapply(METHODS, function(m)
    round(max(res$seconds[res$method == m]))),
  check.names = FALSE, row.names = NULL),
  caption = "Convergence and timing. The Taylor maximum is a c_fit = 1e-4
  fit that ran 850k iterations (17.4h) before converging; the exact column's
  maximum is likewise a c_fit = 1e-4 fit.")
```

## Approximation accuracy

The comparison number is the **exact** Poisson log-likelihood of the fitted
rates, whichever objective produced them, expressed per matrix entry and
relative to the exact method's fit of the same dataset at the same
$c_{\textrm{fit}}$. Negative means the approximation ended somewhere worse.

```{r approx-gap-tables}
for (m in c("taylor", "chebyshev")) {
  s <- res$method == m
  tb <- tapply(res$gap_pe[s], list(res$ct[s], res$cc[s]), mean)
  print(knitr::kable(signif(tb, 2), caption = paste0(
    "Mean per-entry gap, ", m, " (rows = c_true, cols = c_fit).")))
}
```

The display matches the paper's approximation-quality figure: the
**per-entry likelihood ratio** $\exp(\bar\ell_{\textrm{approx}} -
\bar\ell_{\textrm{exact}})$, where $\bar\ell$ is the per-entry exact
log-likelihood of the fitted rates — $1$ means the approximation loses
nothing, $0$ means it collapses. Mean over the thirty seeds, error bars a 95%
confidence interval ($t_{9}$-based).

```{r approx-ratio-fig, fig.width=8.5, fig.height=3.4}
ap <- subset(res, method != "exact")
ap$ratio <- exp(ap$gap_pe)
ap$Approximation <- ifelse(ap$method == "taylor", "Taylor", "Chebyshev")
agg <- do.call(data.frame,
  aggregate(ratio ~ Approximation + cc + ct, ap,
            function(x) c(mean = mean(x),
                          half = qt(0.975, length(x) - 1) *
                                 sd(x) / sqrt(length(x)))))
ymax <- max(1.05, agg$ratio.mean + agg$ratio.half)
ggplot(agg, aes(cc, ratio.mean, colour = Approximation)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_line() + geom_point() +
  geom_errorbar(aes(ymin = ratio.mean - ratio.half,
                    ymax = ratio.mean + ratio.half), width = 0.15) +
  scale_x_continuous(transform = "log10",
                     breaks = c(1e-4, 1e-2, 1, 1e2, 1e4),
                     labels = scales::label_log()) +
  coord_cartesian(ylim = c(0, ymax)) +
  facet_wrap(~ ct, nrow = 1) +
  labs(x = "Value of c for Optimization", y = "Likelihood Ratio")
```

Two regimes are visible. For $c_{\textrm{fit}} \geq 0.1$ at
$c_{\textrm{true}} = 1$ and $\infty$, **both** approximations sit on the
dashed line: per-entry gaps of $10^{-3}$ down to $10^{-10}$, negligible to
identical (the tables above). The $c_{\textrm{true}} = 10^{-3}$ panel is
noisier across the whole grid, and thirty seeds make the reason plain:
that regime's optimum landscape is heavy-tailed, and an occasional single
fit — by any method — lands in a substantially better or worse local
optimum of the common exact criterion (the worst Chebyshev fit anywhere is
one seed at $c_{\textrm{fit}} = 1$ losing $1.3$ per entry; the worst
Taylor fit at $c_{\textrm{fit}} \geq 0.1$ loses $27$; other seeds land
*above* the line). That spread is local-optimum variation, not
approximation error, and it is what the wide intervals in the first panel
are showing. For $c_{\textrm{fit}} \leq 10^{-2}$ the two methods part
company systematically: **Chebyshev degrades gracefully** (mean gap at
most $\sim 0.03$ per entry in every cell) while **Taylor collapses to
ratio $0$** at $c_{\textrm{fit}} \leq 10^{-3}$, giving up tens to
thousands of log-likelihood units per entry in the mean. The Chebyshev
coefficients are fit to the relevant interval, while the Taylor expansion
is anchored at a point that small $c$ pushes far from where the zero-term
sum is evaluated.

## Iterations to convergence

Every fit ran to the same tolerance on its own objective, so the iteration
count is comparable within a method and diagnostic across methods.

```{r iterations-fig, fig.width=8.5, fig.height=3.4}
## finite c only: the c = Inf fits use a different optimizer (fastTopics
## SCD; median 53 iterations), so their counts are not comparable
ggplot(subset(res, is.finite(cc)), aes(cc, n_iter, colour = method)) +
  geom_point(size = 1, alpha = 0.3,
             position = position_jitter(width = 0.06, height = 0)) +
  stat_summary(fun = median, geom = "line", linewidth = 0.8) +
  scale_x_log10(breaks = 10^seq(-4, 4, 2), labels = scales::label_log()) +
  scale_y_log10(labels = scales::label_comma()) +
  scale_colour_manual(values = meth_cols, name = NULL) +
  facet_wrap(~ ct, nrow = 1) +
  labs(x = "c used for fitting", y = "iterations to convergence",
       title = "Iterations to convergence",
       subtitle = "one point per seed; line = median over seeds")
```

Iteration counts rise steadily as $c_{\textrm{fit}}$ decreases for every
method — at $c_{\textrm{fit}} = 10^{4}$ the median fit converges in
$\sim 55$ iterations, at $c_{\textrm{fit}} = 10^{-4}$ the exact method
needs $\sim 17{,}000$ — consistent with the increasingly ill-conditioned
small-$c$ regime. The methods only separate where the approximation is
poor: Chebyshev tracks exact throughout, but Taylor detaches below
$c_{\textrm{fit}} = 10^{-2}$, with a *median* of 301k iterations at
$c_{\textrm{fit}} = 10^{-4}$ and a worst case of 850k (17.4 hours) —
under the 1M cap, but only just. Taylor's failure at small $c$ is thus
doubly expensive: it converges to a much worse fit, and takes far longer
to get there. (The $c_{\textrm{fit}} = \infty$ fits are excluded here:
fastTopics' SCD updates are a different algorithm whose median 53
iterations are not comparable to the log1p fits' cycle counts.)

## Sparsity of the fitted loadings and factors

Mean Hoyer sparsity of the columns of the fitted $L$ and $F$, against the
truth's own sparsity (dashed). The theory says the attainable sparsity is
governed by $c_{\textrm{fit}}$ through the budget
$\operatorname{pm}(\log(1+Y/c))$, increasing in $c$; the design says the
truth was generated with loading sparsity **identical across regimes**
(dashed line at the same height in every panel of the top row).

Exact fits only from here on (the previous sections established where the
approximations can be trusted). As in the c-grid figures: thin lines are
the thirty individual seeds, the thick line their mean, the dashed line the
truth's own value, and the circled point marks
$c_{\textrm{fit}} = c_{\textrm{true}}$ — now present in **all three**
panels, since the fitting grid includes $c = \infty$ (the topic model).
Because $\infty$ has no place on a log axis, $c$ is a discrete axis from
here on, as in the c-grid figures.

```{r seed-panels-helper}
## long format + the c-grid panel style, shared by the sparsity and
## rank-correlation figures below. c is DISCRETE (c_fit = Inf is a real
## level of the grid and has no place on a log axis).
cf_of <- function(cc) factor(sapply(cc, fmtc),
                             levels = sapply(CC_GRID, fmtc))
make_long <- function(ex, vL, vF, tL, tF) {
  sp <- rbind(
    data.frame(ex[c("seed", "cc", "ct", "c_true")], what = "loadings L",
               value = ex[[vL]], truth = ex[[tL]]),
    data.frame(ex[c("seed", "cc", "ct", "c_true")], what = "factors F",
               value = ex[[vF]], truth = ex[[tF]]))
  sp$what <- factor(sp$what, levels = c("loadings L", "factors F"))
  sp$cf   <- cf_of(sp$cc)
  sp
}
seed_panels <- function(sp, ylab, title) {
  avg <- aggregate(value ~ what + cf + ct, sp, mean)
  tru <- aggregate(truth ~ what + ct, sp, mean)
  dg  <- unique(sp[(is.finite(sp$c_true) & sp$cc == sp$c_true) |
                   (is.infinite(sp$c_true) & is.infinite(sp$cc)),
                   c("what", "cf", "ct")])
  dg  <- merge(dg, avg)
  ggplot(sp, aes(cf, value)) +
    geom_line(aes(group = seed), colour = "#9BB7BD",
              linewidth = 0.35, alpha = 0.7) +
    geom_hline(data = tru, aes(yintercept = truth), linetype = "dashed",
               colour = "#C2410C", linewidth = 0.5) +
    geom_line(data = avg, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
    geom_point(data = avg, colour = "#0E5C6B", size = 1.6) +
    geom_point(data = dg, shape = 21, size = 3.6, stroke = 1.1,
               fill = NA, colour = "#C2410C") +
    facet_grid(what ~ ct) +
    labs(x = "c used for fitting", y = ylab, title = title,
         subtitle = paste("thin lines: 30 individual seeds;  thick line:",
                          "mean;  dashed: the true value;  circled:",
                          "c_fit = c_true")) +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 10),
          plot.subtitle = element_text(size = 10, colour = "grey35"),
          panel.border = element_rect(colour = "grey80", fill = NA),
          axis.text.x = element_text(size = 7, angle = 45, hjust = 1))
}
```

```{r sparsity-fig, fig.width=8.5, fig.height=5.4}
ex <- subset(res, method == "exact")
seed_panels(make_long(ex, "hoyer_L", "hoyer_F",
                      "hoyer_L_true", "hoyer_F_true"),
            ylab  = "mean Hoyer sparsity of columns",
            title = "Sparsity of the fit is set by the fitting c")
```

Three things to read off:

- **Monotonicity, seed by seed.** At every $c_{\textrm{true}}$, for both $L$
  and $F$, sparsity increases in $c_{\textrm{fit}}$ across the entire grid —
  and the thin lines show this is true of essentially every individual
  seed, not just the average; the seed spread is a small band around the
  mean.
- **The topic model is the continuous endpoint of the grid.** The new
  $c_{\textrm{fit}} = \infty$ column continues every curve smoothly from
  $c_{\textrm{fit}} = 10^{4}$ — plain Poisson NMF behaves as the
  large-$c$ limit of the log1p family, not as a different method.
- **The fit recovers the truth's sparsity when fit at the true $c$.** The
  circled points sit almost exactly on the dashed truth lines in all three
  panels: for loadings (truth $0.396$ everywhere) the exact fits give
  $0.41$ at $(10^{-3}, 10^{-3})$, $0.41$ at $(1, 1)$, and — the case the
  grid could not show before — the topic model on topic-model truth gives
  $0.406$ at $(\infty, \infty)$; for factors the crossings are equally
  tight ($|\Delta| \leq 0.005$ at $c_{\textrm{true}} = 1$ and $\infty$).
  Misfit in $c$ translates directly into over- or under-sparse estimates.

The approximate fits (not shown) trace the same curves wherever the
approximation is accurate; Taylor's small-$c$ fits deviate, but those are
exactly the fits the accuracy section already ruled out.

## Fit quality relative to the fit at the true c

The sparsity panels show what misfit in $c$ does to the *representation*;
the two figures here show what it costs in *fit quality*, using the fit at
$c_{\textrm{fit}} = c_{\textrm{true}}$ on the same dataset as the
reference. Both are per-seed-referenced, so every thin line passes through
the reference value exactly at the circled point.

**Likelihood ratio.** For each exact fit, the per-entry likelihood ratio
against the same seed's fit at the true $c$:
$\exp((\bar\ell(c_{\textrm{fit}}) - \bar\ell(c_{\textrm{true}}))$, where
$\bar\ell$ is the per-entry exact Poisson log-likelihood of the fitted
rates. Values above $1$ mean the misfit $c$ actually achieved a better
exact likelihood than the true one on that dataset.

```{r lr-true-fig, fig.width=8.5, fig.height=3.4}
exq <- subset(res, method == "exact")
ref <- exq[(is.finite(exq$c_true) & exq$cc == exq$c_true) |
           (is.infinite(exq$c_true) & is.infinite(exq$cc)),
           c("seed", "ct", "loglik_exact")]
names(ref)[3] <- "ll_ref"
lr <- merge(exq, ref, by = c("seed", "ct"))
lr$lr_pe <- exp((lr$loglik_exact - lr$ll_ref) / (N_ROWS * N_COLS))
lr$cf <- cf_of(lr$cc)
avg <- aggregate(lr_pe ~ cf + ct, lr, mean)
dg  <- merge(unique(lr[(is.finite(lr$c_true) & lr$cc == lr$c_true) |
                       (is.infinite(lr$c_true) & is.infinite(lr$cc)),
                       c("cf", "ct")]), avg)
ggplot(lr, aes(cf, lr_pe)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "#C2410C",
             linewidth = 0.5) +
  geom_line(aes(group = seed), colour = "#9BB7BD",
            linewidth = 0.35, alpha = 0.7) +
  geom_line(data = avg, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
  geom_point(data = avg, colour = "#0E5C6B", size = 1.6) +
  geom_point(data = dg, shape = 21, size = 3.6, stroke = 1.1,
             fill = NA, colour = "#C2410C") +
  facet_wrap(~ ct, nrow = 1) +
  labs(x = "c used for fitting",
       y = "per-entry likelihood ratio",
       title = "Likelihood ratio relative to the fit at the true c",
       subtitle = paste("thin lines: 30 individual seeds;  thick line:",
                        "mean;  circled: c_fit = c_true (ratio 1 by",
                        "definition)")) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 10, colour = "grey35"),
        panel.border = element_rect(colour = "grey80", fill = NA),
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1))
```

Three readings. First — and this is the figure's headline — the
likelihood surface over $c$ is remarkably **flat** compared to the
representational changes in the sparsity figure: misfitting $c$ by two
decades in either direction costs at most a few percent of per-entry
likelihood, while the fitted sparsity moves across most of its range. The
likelihood alone cannot choose $c$; the representation is what changes.
Second, the $c_{\textrm{true}} = 10^{-3}$ panel again shows the
heavy-tailed local-optimum variation of that regime (individual seeds dip
to $\sim 0.6$ at mid-grid $c_{\textrm{fit}}$ while others sit at $1$).
Third, at $c_{\textrm{true}} = \infty$ the finite-$c$ fits sit slightly
*above* $1$ on average. This is not a convergence-criterion artifact (the
fastTopics reference uses the same stopping rule, an absolute
own-objective change below $10^{-6}$ per iteration, and the median seed's
ratio is $1.000$): the whole mean excess comes from three seeds (5, 11,
20 — ordinary datasets, not the degenerate one) on which fastTopics' SCD
updates converged into a local optimum $\sim 0.3$ nats per entry worse
than what the log1p fits reach at every finite $c$. An optimizer-basin
effect on 3 of 30 datasets, worth knowing about before the topic-model
column anchors any paper claim.

**Relative RMSE on the rates.** For each exact fit, the RMSE of its rates
against the true rates, divided by the same seed's RMSE at
$c_{\textrm{fit}} = c_{\textrm{true}}$ — on the raw rate scale (dominated
by the largest rates) and on the log1p rate scale (weights the low-rate
structure).

```{r rel-rmse-fig, fig.width=8.5, fig.height=5.4}
if (!"rmse_log1p" %in% names(res)) {
  cat("(The RMSE columns are not yet in approx_sim_metrics.rds; rerun",
      "analyze_approx_sim.R on the cluster and re-download the metrics",
      "table to render this figure.)")
} else {
  refm <- exq[(is.finite(exq$c_true) & exq$cc == exq$c_true) |
              (is.infinite(exq$c_true) & is.infinite(exq$cc)),
              c("seed", "ct", "rmse", "rmse_log1p")]
  names(refm)[3:4] <- c("rmse_ref", "rmse_log1p_ref")
  rr <- merge(exq, refm, by = c("seed", "ct"))
  rr <- rbind(
    data.frame(rr[c("seed", "cc", "ct", "c_true")], what = "rate scale",
               value = rr$rmse / rr$rmse_ref),
    data.frame(rr[c("seed", "cc", "ct", "c_true")], what = "log1p rate scale",
               value = rr$rmse_log1p / rr$rmse_log1p_ref))
  rr$what <- factor(rr$what, levels = c("rate scale", "log1p rate scale"))
  rr$cf <- cf_of(rr$cc)
  avg <- aggregate(value ~ what + cf + ct, rr, mean)
  dg  <- merge(unique(rr[(is.finite(rr$c_true) & rr$cc == rr$c_true) |
                         (is.infinite(rr$c_true) & is.infinite(rr$cc)),
                         c("what", "cf", "ct")]), avg)
  print(ggplot(rr, aes(cf, value)) +
    geom_hline(yintercept = 1, linetype = "dashed", colour = "#C2410C",
               linewidth = 0.5) +
    geom_line(aes(group = seed), colour = "#9BB7BD",
              linewidth = 0.35, alpha = 0.7) +
    geom_line(data = avg, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
    geom_point(data = avg, colour = "#0E5C6B", size = 1.6) +
    geom_point(data = dg, shape = 21, size = 3.6, stroke = 1.1,
               fill = NA, colour = "#C2410C") +
    scale_y_log10() +
    facet_grid(what ~ ct) +
    labs(x = "c used for fitting",
         y = "RMSE relative to the fit at the true c",
         title = "Rate-recovery error relative to the fit at the true c",
         subtitle = paste("thin lines: 30 individual seeds;  thick line:",
                          "mean;  circled: c_fit = c_true (ratio 1 by",
                          "definition)")) +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 10),
          plot.subtitle = element_text(size = 10, colour = "grey35"),
          panel.border = element_rect(colour = "grey80", fill = NA),
          axis.text.x = element_text(size = 7, angle = 45, hjust = 1)))
}
```

The striking feature is the **asymmetry**: fitting at a $c$ *below* the
truth costs almost nothing in rate recovery (at $c_{\textrm{true}} = 1$,
even $c_{\textrm{fit}} = 10^{-4}$ leaves the rate-scale RMSE at $\sim 1.05$;
the log1p scale shows a milder $\sim 2\times$ penalty in the low-rate
structure), while fitting *above* the truth degrades recovery steadily —
up to $\sim 4\times$ at $c_{\textrm{true}} = 1$ and, with the usual
heavy-tailed seed scatter, $\sim 5$–$10\times$ in the mean at
$c_{\textrm{true}} = 10^{-3}$. At $c_{\textrm{true}} = \infty$ the
rate-scale curve is flat at $1$ across the entire grid — every $c$
recovers the rates of additive truth about equally well — and the lines
*below* $1$ are the three SCD-trapped seeds from the likelihood-ratio
figure, for which the finite-$c$ fits recover the true rates better than
the topic-model reference itself.

## Fitted rates across c, one typical dataset per regime

The summaries above compress each fit to a number; here we look at the
fitted rates themselves. For one representative dataset per regime (the
seed whose mean count is closest to its regime's median: seed 13 at
$c_{\textrm{true}} = 10^{-3}$, seed 1 at $c_{\textrm{true}} = 1$, seed 10
at $c_{\textrm{true}} = \infty$), we compare the exact fits at
$c_{\textrm{fit}} \in \{10^{-3}, 1, \infty\}$ pairwise: each panel
scatters all $250{,}000$ fitted rates of one fit against another — raw
rates, with the axes on the log1p scale so the labels read in rate units.
Rows are the generative regimes, columns the three pairs; the diagonal is
equality, and the annotation gives the Pearson correlation of
$\log(1 + \lambda)$.

```{r rates-pairs-fig, fig.width=8.5, fig.height=8}
fit_dir <- file.path(exp_dir, "approx_sim_fits")
need <- expand.grid(seed = c(13, 1, 10), cfit = c("0.001", "1", "Inf"))
need$ctrue <- rep(c("0.001", "1", "Inf"), 3)
need$file <- sprintf("approxsim_seed%02d_ctrue%s_cfit%s_exact.rds",
                     need$seed, need$ctrue, need$cfit)
if (!all(file.exists(file.path(fit_dir, need$file)))) {
  cat("(Waiting on the nine per-fit .rds files in",
      "inst/paper_figures/experiments/approx_sim_fits/ -- see the scp",
      "command in the session notes; the figure renders once they are",
      "present.)")
} else {
  rates_of <- function(f, cc) {
    x  <- readRDS(file.path(fit_dir, f))
    th <- tcrossprod(as.matrix(x$LL), as.matrix(x$FF))
    if (is.infinite(cc)) th else cc * expm1(th / max(1, cc))
  }
  scen <- data.frame(ctrue = c("0.001", "1", "Inf"), seed = c(13, 1, 10))
  pairs <- list(c("0.001", "1"), c("0.001", "Inf"), c("1", "Inf"))
  pan <- list(); ann <- list()
  for (i in seq_len(nrow(scen))) for (pr in pairs) {
    fa <- sprintf("approxsim_seed%02d_ctrue%s_cfit%s_exact.rds",
                  scen$seed[i], scen$ctrue[i], pr[1])
    fb <- sprintf("approxsim_seed%02d_ctrue%s_cfit%s_exact.rds",
                  scen$seed[i], scen$ctrue[i], pr[2])
    a <- as.vector(rates_of(fa, as.numeric(pr[1])))
    b <- as.vector(rates_of(fb, as.numeric(pr[2])))
    row_lab <- paste0("truth: c = ", scen$ctrue[i], " (seed ", scen$seed[i], ")")
    col_lab <- paste0("c = ", pr[1], "  vs  c = ", pr[2])
    pan[[length(pan) + 1]] <- data.frame(x = a, y = b,
                                         row = row_lab, col = col_lab)
    ann[[length(ann) + 1]] <- data.frame(
      row = row_lab, col = col_lab,
      lab = sprintf("r = %.3f", cor(log1p(a), log1p(b))))
  }
  pan <- do.call(rbind, pan); ann <- do.call(rbind, ann)
  rl <- paste0("truth: c = ", scen$ctrue, " (seed ", scen$seed, ")")
  pan$row <- factor(pan$row, levels = rl); ann$row <- factor(ann$row, levels = rl)
  print(ggplot(pan, aes(x, y)) +
    geom_bin2d(bins = 70) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed",
                colour = "#C2410C", linewidth = 0.4) +
    geom_text(data = ann, aes(label = lab), x = -Inf, y = Inf,
              hjust = -0.15, vjust = 1.6, size = 3, colour = "grey20") +
    scale_fill_gradient(low = "#DCE8EB", high = "#0E3A44",
                        trans = "log10", name = "entries") +
    scale_x_continuous(trans = "log1p", breaks = c(0, 1, 10, 100, 1000),
                       labels = scales::comma) +
    scale_y_continuous(trans = "log1p", breaks = c(0, 1, 10, 100, 1000),
                       labels = scales::comma) +
    facet_grid(row ~ col) +
    coord_equal() +
    labs(x = "fitted rate (log1p scale), smaller c of the pair",
         y = "fitted rate (log1p scale), larger c of the pair",
         title = "Fitted rates compared across fitting c, entry by entry") +
    theme(strip.background = element_blank(),
          strip.text = element_text(face = "bold", size = 9),
          panel.border = element_rect(colour = "grey80", fill = NA)))
}
```

The high-rate tail is pinned to the diagonal in every panel — above
$\lambda \approx 30$ the fits at different $c$ essentially agree entry by
entry, whichever truth generated the data. All of the disagreement lives
in the low-rate mass, and its shape depends on the regime: on additive
truth (bottom row) even that vanishes ($r \geq 0.99$ everywhere —
matching the flat relative-RMSE row), while on the log-link truths the
panels involving a larger fitting $c$ show a lobe *above* the diagonal at
low rates — the larger-$c$ fit assigns systematically higher rates to
entries the smaller-$c$ fit shrinks toward zero. This is the entry-level
face of the sparsity story: what the budget takes away at small $c$ is
precisely the ability to hold many entries near zero while a few grow
large, and what a large-$c$ fit loses on log-link truth is the aggressive
shrinkage of the low-rate mass.

## Rank correlation of the factors

The redundancy measure: mean Spearman correlation over the
$\binom{K}{2} = 10$ pairs of fitted factor columns (the loadings analogue
is shown alongside). The sparsity theory's companion argument predicts that
as columns sparsify, their rankings decouple — pairwise rank correlations
should drift with $c_{\textrm{fit}}$ alongside the sparsity itself.

```{r rankcor-fig, fig.width=8.5, fig.height=5.4}
seed_panels(make_long(ex, "rankcor_L", "rankcor_F",
                      "rankcor_L_true", "rankcor_F_true"),
            ylab  = "mean pairwise Spearman correlation",
            title = "Rank correlation between fitted columns")
```

(One thin line in the $c_{\textrm{true}} = \infty$ panels shoots upward at
the large-$c$ end: that is the single near-degenerate seed of the
identity-link generator's heavy-tailed law, whose dataset has
$\max(Y) = 30$ and a mean count of $0.096$ — visible for the same reason
in the budget figure below. It is left in deliberately; the mean curves
barely feel it.)

For the **loadings** at $c_{\textrm{true}} = 1$ and $\infty$ the effect is
clean: pairwise rank correlation rises toward zero as $c_{\textrm{fit}}$
grows, reaching the truth's value ($-0.176$) near
$c_{\textrm{fit}} \approx c_{\textrm{true}}$, with a tight seed band. At
$c_{\textrm{true}} = 10^{-3}$ the mean curve is not monotone — it peaks
around $c_{\textrm{fit}} = 10$ and comes back down where the fits are
heavily over-sparsified — and the topic-model column adds a sharp coda:
at $c_{\textrm{fit}} = \infty$ on small-$c$ truth the loading rank
correlation drops to $-0.194$, *below* the truth's value, the one place on
the grid where a fit over-decouples the loadings. For the **factors** the
curves are non-monotone but, at $c_{\textrm{true}} = 1$ and $\infty$,
systematic — the seed bands are tight, and the pairwise rank correlation
is at its most negative, i.e. closest to the truth's decoupled factors, in
a trough near $c_{\textrm{fit}} \approx c_{\textrm{true}}$ (at
$c_{\textrm{fit}} = 1$ and $10$ respectively), rising toward zero or above
on both sides, where the fits are mis-sparsified. At
$c_{\textrm{true}} = 10^{-3}$ the factor curve is flat and the seed spread
dominates. The one robust feature across all regimes is that **every
fit's factor columns are more rank-correlated than the truth's** (the
seed band sits above the dashed lines, coming closest to them in the
topic-model column at $c_{\textrm{true}} = 10^{-3}$): the fits never fully
reproduce the decoupling of the generative factors.

## The sparsity bounds in practice

The manuscript's appendix (*Hoyer Sparsity Results*) proves a chain of
inequalities that constrain how sparse the columns of a fit can be. All of
them are stated for an arbitrary non-negative factorization
$\mathbf{B} = \mathbf{L}\mathbf{F}^{\top}$, so they can be checked directly
on every fitted model here, using each fit's own $\mathbf{B}$. Writing
$\operatorname{pm}(x) = \max(x)/\operatorname{mean}(x)$ for the max-to-mean
ratio, $d(x) = \|x\|_1 / (\sqrt{m}\|x\|_2)$ for the density of an
$m$-vector, $H(x) = \kappa_m(1 - d(x))$ with
$\kappa_m = \sqrt{m}/(\sqrt{m}-1)$ for the Hoyer sparsity, and
$$w_k = \frac{\operatorname{mean}(\mathbf{l}_{\cdot k})
\operatorname{mean}(\mathbf{f}_{\cdot k})}{\operatorname{mean}(\mathbf{B})},
\qquad \textstyle\sum_{k=1}^{K} w_k = 1,$$
the results used below are:

1. **The pm-budget lemma.** For every factor $k$,
   $$\operatorname{pm}(\mathbf{l}_{\cdot k})\,
     \operatorname{pm}(\mathbf{f}_{\cdot k})\, w_k
     \;\leq\; \operatorname{pm}(\mathbf{B}).$$
   The matrix's max-to-mean ratio is a *budget* that the factors' own
   max-to-mean ratios, weighted by $w_k$, must fit inside.
2. **The hyperbola theorem.** Combining (1) with the elementary bound
   $d(x) \geq 1/\sqrt{\operatorname{pm}(x)}$ gives
   $$d(\mathbf{l}_{\cdot k})\, d(\mathbf{f}_{\cdot k})
     \;\geq\; \sqrt{w_k / \operatorname{pm}(\mathbf{B})}:$$
   a loading and its factor cannot both be sparse, and the constraint
   tightens as $\operatorname{pm}(\mathbf{B})$ falls.
3. **The sum-form corollary.** In Hoyer units (here $n = p$, so
   $\kappa_n = \kappa_p = \kappa$),
   $$H(\mathbf{l}_{\cdot k}) + H(\mathbf{f}_{\cdot k})
     \;\leq\; 2\kappa\left(1 -
     \left(w_k/\operatorname{pm}(\mathbf{B})\right)^{1/4}\right):$$
   the pair shares a single sparsity allowance set by
   $\operatorname{pm}(\mathbf{B})$.
4. **The monotone-budget proposition.** For any fixed rates, the budget
   $\operatorname{pm}(\mathbf{B}_c)$ is non-decreasing in $c$ — small $c$
   does not merely fail to produce sparse factors, it makes them
   unavailable.

One caveat established before plotting: the appendix's *averaged* ceiling
(mean Hoyer $\leq \kappa(1 - 1/(K\sqrt{\operatorname{pm}(\mathbf{B})}))$)
is vacuous at this design — with $K = 5$ and budgets that never fall below
$\approx 3$, it sits at $\approx 1$ across the whole grid. The averaging
step costs a factor of $K$; the per-column forms above are the ones with
empirical content, and they are what we examine. All figures in this
section use the exact fits.

**The realized budget.** First, the budget itself. The proposition concerns
fixed rates, whereas each fit re-estimates its rates at its own $c$ — yet
the realized $\operatorname{pm}(\mathbf{B})$ of the fitted models is
monotone in $c_{\textrm{fit}}$ anyway — all the way through the topic
model at $c_{\textrm{fit}} = \infty$, where $\mathbf{B}$ is the fitted
rate matrix itself — spanning three orders of magnitude across the grid
($\sim 3$ to $\sim 5{,}900$), and nearly identical across seeds:

```{r bounds-load}
bounds <- readRDS(file.path(exp_dir, "approx_sim_bounds.rds"))
bounds$ct <- factor(paste0("c_true = ", sapply(bounds$c_true, fmtc)),
                    levels = levels(res$ct))
be <- subset(bounds, method == "exact")
```

```{r budget-fig, fig.width=8.5, fig.height=3.2}
exm <- subset(res, method == "exact")
exm$cf <- cf_of(exm$cc)
avg <- aggregate(pm_B ~ cf + ct, exm, mean)
ggplot(exm, aes(cf, pm_B)) +
  geom_line(aes(group = seed), colour = "#9BB7BD",
            linewidth = 0.35, alpha = 0.7) +
  geom_line(data = avg, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
  geom_point(data = avg, colour = "#0E5C6B", size = 1.6) +
  scale_y_log10(labels = scales::label_log()) +
  facet_wrap(~ ct, nrow = 1) +
  labs(x = "c used for fitting", y = "pm(B) of the fitted model",
       title = "The realized sparsity budget is monotone in the fitting c",
       subtitle = "thin lines: 30 individual seeds;  thick line: mean") +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 10, colour = "grey35"),
        panel.border = element_rect(colour = "grey80", fill = NA),
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1))
```

**The proposition's fixed-rate budget predicts the fitted budget.** The
monotone-budget proposition is a statement about a *fixed* matrix of rates
viewed through different links: for rates $\boldsymbol{\Lambda}$ it
concerns $\operatorname{pm}(\mathbf{B}^{*}_{c})$ with
$b^{*}_{ij}(c) = \log(1 + \lambda_{ij}/c)$. In this simulation the natural
fixed rates are the truth itself, and the quantity
$c \mapsto \operatorname{pm}(\log(1 + \boldsymbol{\Lambda}_{\textrm{true}}/c))$
is computable from the generative model with no fitting at all. Alongside
it we show the same curve computed from the observed counts,
$\operatorname{pm}(\log(1 + \mathbf{Y}/c))$ — the version available in
practice (it is the calibration target of this whole design), and, by the
saturated-case remark in the appendix, *exactly* the fitted budget when
$K \geq \min(n, p)$. Comparing both to the budget the rank-$5$ fits
actually realize:

```{r budget-true-fig, fig.width=8.5, fig.height=5.6}
## at c = Inf the link is the identity: B* = Lambda, and the data budget is
## pm(Y) itself
pm_of <- function(M, cc) {
  g <- if (is.finite(cc)) log1p(M / cc) else M
  max(g) / mean(g)
}
pm_true <- do.call(rbind, lapply(ds, function(d) data.frame(
  seed = d$seed, cc = CC_GRID,
  ct = factor(paste0("c_true = ", fmtc(d$c_true)), levels = levels(res$ct)),
  pm_true = sapply(CC_GRID, pm_of, M = d$lambda),
  pm_data = sapply(CC_GRID, pm_of, M = d$Y))))
cmp <- merge(exm[c("seed", "cc", "ct", "pm_B")],
             pm_true, by = c("seed", "cc", "ct"))
cmp$cf         <- cf_of(cmp$cc)
cmp$ratio      <- cmp$pm_B / cmp$pm_true
cmp$ratio_data <- cmp$pm_B / cmp$pm_data

avg  <- aggregate(cbind(pm_B, pm_true, pm_data) ~ cf + ct, cmp, mean)
p1 <- ggplot(cmp, aes(cf, pm_B)) +
  geom_line(aes(group = seed), colour = "#9BB7BD",
            linewidth = 0.35, alpha = 0.7) +
  geom_line(data = avg, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
  geom_line(data = avg, aes(y = pm_true, group = 1), colour = "#C2410C",
            linetype = "dashed", linewidth = 0.7) +
  geom_line(data = avg, aes(y = pm_data, group = 1), colour = "#B45309",
            linetype = "dotdash", linewidth = 0.7) +
  scale_y_log10(labels = scales::label_log()) +
  facet_wrap(~ ct, nrow = 1) +
  labs(x = NULL, y = "pm(B)",
       title = "Fitted budget vs the true-rate and data budgets",
       subtitle = paste("thin: fitted, per seed;  thick: fitted, mean;",
                        "dashed: pm(log(1 + Λ_true/c));",
                        "dot-dash: pm(log(1 + Y/c))")) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 10, colour = "grey35"),
        panel.border = element_rect(colour = "grey80", fill = NA),
        axis.text.x = element_blank())

ravg <- aggregate(cbind(ratio, ratio_data) ~ cf + ct, cmp, mean)
p2 <- ggplot(cmp, aes(cf, ratio)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey40") +
  geom_line(aes(group = seed), colour = "#9BB7BD",
            linewidth = 0.35, alpha = 0.7) +
  geom_line(data = ravg, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
  geom_point(data = ravg, colour = "#0E5C6B", size = 1.4) +
  geom_line(data = ravg, aes(y = ratio_data, group = 1), colour = "#B45309",
            linetype = "dotdash", linewidth = 0.8) +
  geom_point(data = ravg, aes(y = ratio_data), colour = "#B45309",
             size = 1.4) +
  scale_y_log10(breaks = c(0.125, 0.25, 0.5, 1),
                labels = c("0.125", "0.25", "0.5", "1")) +
  facet_wrap(~ ct, nrow = 1) +
  labs(x = "c used for fitting", y = "fitted / reference budget",
       subtitle = paste("blue: fitted / true-rate budget (thin: seeds);",
                        "amber: fitted / data budget")) +
  theme(strip.background = element_blank(),
        strip.text = element_blank(),
        plot.subtitle = element_text(size = 10, colour = "grey35"),
        panel.border = element_rect(colour = "grey80", fill = NA),
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1))

cowplot::plot_grid(p1, p2, ncol = 1, rel_heights = c(1.35, 1), align = "v")
```

The dashed curve is barely distinguishable from the fitted one: the mean
ratio stays within $0.84$–$1.02$ across the entire grid, including the
$c_{\textrm{fit}} = \infty$ column (bottom row, blue), and the individual
seeds mostly within $0.8$–$1.05$ — the visible exception being the largest
$c_{\textrm{fit}}$ values at finite $c_{\textrm{true}}$, where a rank-$5$
fit can trim the single most extreme entry of $\mathbf{B}$ relative to the
truth and an occasional seed drops to $\approx 0.55$. The mild, systematic
compression elsewhere ($\approx 0.9$) is the same rank-$K$ smoothing in
gentler form, and at $c_{\textrm{true}} = \infty$ with large
$c_{\textrm{fit}}$ the fits slightly *exceed* the true-rate budget
instead.

The **data budget** (dot-dash) tells a complementary story. At
$c_{\textrm{fit}} \gtrsim 1$ it agrees with the other two curves to within
$\sim 15\%$, all the way out to $c_{\textrm{fit}} = \infty$, where it is
simply $\operatorname{pm}(\mathbf{Y})$ and the fitted topic model matches
it closely. At small $c_{\textrm{fit}}$ it splits off upward, exceeding
the fitted budget by a factor of $4$–$8$ (bottom row, amber, dropping to
$0.13$–$0.23$): on the link scale the $\sim 89\%$ zero entries of
$\mathbf{Y}$ stay exactly zero and drag the matrix mean down, while the
true rates — and every entry of a rank-$5$ fit — are strictly positive.
This is the saturated-case remark playing out in reverse: at
$K \geq \min(n,p)$ the fit would reproduce $\mathbf{Y}$ and its budget
exactly, but a low-rank fit smooths the zeros up to small positive values,
and its budget lands on the *true-rate* curve rather than the data curve.
Practically: for sparse count data,
$\operatorname{pm}(\log(1 + \mathbf{Y}/c))$ preserves the shape and
ordering of the budget — it remains monotone and spans the same three
decades, which is all its role as a calibration target requires — but as a
forecast of the budget a low-rank fit will realize at small $c$ it
overstates the level, and the true-rate (equivalently, fitted-rate) version
is the sharper reference. Practically, this closes the loop on the proposition: the
fixed-rate budget curve, computable from rates (or, by the same
calculation, from data) without fitting anything, is an accurate forecast
of the budget the fitted model will actually have — which is what makes
the budget a usable a-priori diagnostic rather than a post-hoc
description.

**How tight are the bounds?** For each fitted factor we compute the slack
of inequality (1), $\operatorname{pm}(\mathbf{B}) /
(\operatorname{pm}(\mathbf{l}_{\cdot k})\operatorname{pm}(\mathbf{f}_{\cdot k})
w_k) \geq 1$, and of inequality (2),
$d(\mathbf{l}_{\cdot k}) d(\mathbf{f}_{\cdot k}) \big/
\sqrt{w_k/\operatorname{pm}(\mathbf{B})} \geq 1$. Slack $1$ means the fit
sits exactly on the constraint.

```{r slack-fig, fig.width=8.5, fig.height=5.4}
be$cf <- cf_of(be$cc)
sl <- rbind(
  data.frame(be[c("seed", "cf", "ct", "k")], bound = "pm budget (1)",
             slack = be$slack_pm),
  data.frame(be[c("seed", "cf", "ct", "k")], bound = "density hyperbola (2)",
             slack = be$slack_hyper))
sl$bound <- factor(sl$bound, levels = c("pm budget (1)",
                                        "density hyperbola (2)"))
med <- aggregate(slack ~ bound + cf + ct, sl, median)
ggplot(sl, aes(cf, slack)) +
  geom_point(size = 0.6, alpha = 0.2, colour = "#0E5C6B",
             position = position_jitter(width = 0.15, height = 0)) +
  geom_line(data = med, aes(group = 1), colour = "#0E5C6B", linewidth = 1) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "#C2410C") +
  scale_y_log10() +
  facet_grid(bound ~ ct) +
  labs(x = "c used for fitting", y = "slack of the bound  (1 = active)",
       title = "Slack of the per-factor sparsity bounds",
       subtitle = paste("one point per (seed, factor); line: median;",
                        "dashed: the bound itself")) +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 10, colour = "grey35"),
        panel.border = element_rect(colour = "grey80", fill = NA),
        axis.text.x = element_text(size = 7, angle = 45, hjust = 1))
```

The two rows say different things. The **pm-budget lemma is close to
equality wherever the budget is scarce**: for $c_{\textrm{fit}} \leq 0.1$
the median slack is $1.09$–$1.15$ at the finite $c_{\textrm{true}}$
regimes (minimum $1.000$ — some factors sit exactly on the constraint),
growing to $2$–$6$ only at the large-$c$ end — including the topic-model
column — where the budget is plentiful.
Small $c$ genuinely exhausts the budget. The **density hyperbola inherits
extra slack** from the step $d(x) \geq 1/\sqrt{\operatorname{pm}(x)}$,
which is loose for the dense columns that small $c$ forces; its slack
floor is $\approx 1.3$–$2$ across the grid. So in Hoyer units the bounds
are valid but conservative envelopes at this $K$, while in pm units the
budget accounting is nearly sharp.

**The sum-form allowance.** Finally, inequality (3) read as intended: each
factor's observed $H(\mathbf{l}_{\cdot k}) + H(\mathbf{f}_{\cdot k})$
against its own allowance
$2\kappa(1 - (w_k/\operatorname{pm}(\mathbf{B}))^{1/4})$.

```{r allowance-fig, fig.width=8.5, fig.height=3.6}
kn <- sqrt(N_ROWS) / (sqrt(N_ROWS) - 1)
be$allow <- 2 * kn * (1 - (be$w / be$pm_B)^0.25)
be$Hsum  <- be$H_L + be$H_F
be$cf    <- factor(be$cc, levels = sort(unique(be$cc)),
                   labels = sapply(sort(unique(be$cc)), fmtc))
ggplot(be, aes(allow, Hsum, colour = cf)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed",
              colour = "grey40") +
  geom_point(size = 1, alpha = 0.6) +
  scale_colour_viridis_d(name = "c_fit", option = "mako", end = 0.9,
                         direction = -1) +
  facet_wrap(~ ct, nrow = 1) +
  coord_equal(xlim = c(0, 2.1), ylim = c(0, 2.1)) +
  labs(x = "allowance  2κ(1 − (w/pm(B))^{1/4})",
       y = "observed  H(l) + H(f)",
       title = "Observed sparsity vs the sum-form allowance, per factor",
       subtitle = "dashed: allowance used in full; points cannot lie above it") +
  theme(strip.background = element_blank(),
        strip.text = element_text(face = "bold", size = 10),
        plot.subtitle = element_text(size = 10, colour = "grey35"),
        panel.border = element_rect(colour = "grey80", fill = NA))
```

Every point lies below the diagonal, as it must. Two practical readings.
First, the allowance and the observed sparsity move together: small
$c_{\textrm{fit}}$ (light points) sits at small allowance and small
observed sum, large $c_{\textrm{fit}}$ (dark) at large allowance and large
observed sum — the budget is the axis along which $c$ operates, exactly as
the theory says. Second, the *fraction* of the allowance actually used
grows with $c_{\textrm{fit}}$: the median factor uses $39$–$65\%$ of its
allowance at $c_{\textrm{fit}} = 10^{-4}$ and $73$–$89\%$ at
$c_{\textrm{fit}} \geq 10^{2}$, topic model included (highest in the
$c_{\textrm{true}} = 10^{-3}$ regime, whose truth is sparsest). In other words: at small $c$ the fits sit
against the *pm form* of the constraint while the Hoyer conversion leaves
headroom, and at large $c$ the fits approach the Hoyer allowance itself.
The bounds are therefore practically relevant in both regimes, but through
different faces of the same inequality chain.

## Summary

The simulation does what it was designed to do. On counts calibrated to a
real dataset's budget curve in all three generative regimes, with thirty
seeds and the topic model anchoring the large-$c$ end of the grid: (i) the
sparsity of the fitted loadings and factors is controlled by the fitting
$c$, monotonically and over a wide range, with plain Poisson NMF behaving
as the continuous $c \to \infty$ endpoint of the log1p family; (ii)
fitting at the generative $c$ reproduces the generative sparsity almost
exactly — now including the topic model on topic-model truth — so misfit
in $c$ is visible as over- or under-sparsification; (iii) the appendix
bounds are live constraints, not formalities — the realized budget
$\operatorname{pm}(\mathbf{B})$ is monotone in $c_{\textrm{fit}}$ over
three orders of magnitude, the pm-budget lemma is within $9$–$15\%$ of
equality wherever the budget is scarce, and the fits use up to $\sim 89\%$
of the sum-form Hoyer allowance at the large-$c$ end; and (iv) the
Chebyshev approximation is faithful to the exact objective everywhere on
the grid (mean gap at most $\sim 0.03$ log-likelihood units per entry in
every cell, typically orders of magnitude less; isolated single-fit
outliers in the $c_{\textrm{true}} = 10^{-3}$ regime are local-optimum
variation shared with the exact fits), while the Taylor approximation is
faithful only for $c_{\textrm{fit}} \gtrsim 0.1$ and fails badly — slowly
and to bad optima — below that.
