---
title: "A calibrated simulation: realistic counts with controlled factors"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r pkgs}
library(ggplot2); library(cowplot)
theme_set(theme_cowplot(font_size = 11))
```

# What this simulation is for

One generator, used everywhere a simulation appears: the sparsity and
identifiability experiments across values of $c$, and the assessment of the
approximation accuracy. The design has three requirements.

1. **The true loadings have fixed, controlled sparsity.** Whatever else varies,
   the rows of $L$ are drawn the same way, so differences in *fitted* sparsity
   across $c$ can never be blamed on the truth changing.
2. **"Realistic" is a checked property, not an adjective.** The simulated
   counts are calibrated to a real dataset against a specific criterion,
   stated below, and the match is verified on held-out seeds.
3. **The whole recipe fits in five lines**, so it can be described completely
   in a paper paragraph.

# The generative recipe

Fix the dimensions $n$, $p$, the rank $K$, a link constant $c_{\textrm{true}}$,
and the Dirichlet concentration $a = 0.3$. The calibration below supplies four
numbers: a zero probability $\pi_0$, a location $\mu$, a scale $\sigma$, and a
cap $T$. Then:

1. **Loadings.** Each row of $L$ is drawn
   $\;\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(a, \dots, a)$.
   With $a = 0.3$ each sample places most of its weight on a few factors —
   membership is soft but concentrated, and its sparsity is set once by $a$.
2. **Factors.** Each entry of $F$ is independently
   $$f_{jk} = \begin{cases} 0 & \textrm{with probability } \pi_0 \\
   \min\{\exp(\mu + \sigma t_{jk}),\; T\} & \textrm{otherwise}, \end{cases}
   \qquad t_{jk} \sim t_4 .$$
   A point mass at zero, a heavy-tailed body, and a hard ceiling.
3. **Linear predictor.** $\Theta = L F^{\top}$.
4. **Rates.** $\lambda_{ij} = c_{\textrm{true}}\left(e^{\theta_{ij}} - 1\right)$
   — the inverse of the log1p link.
5. **Counts.** $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$.

Each ingredient of step 2 has one job. The point mass makes most rates exactly
zero, which is where the zeros of real count matrices come from. The log-$t$
body gives the counts a heavy tail, which lognormal magnitudes cannot produce
(we tried: the maximum count comes out several times too small). The cap
exists because the *maximum* of unbounded heavy-tailed draws varies wildly from
seed to seed — with a cap, the extreme counts land in the same place every
time, as they should for a matrix of fixed size.

# What "realistic" means, exactly

The property of the data that our theory identifies as consequential is the
**budget curve**

$$ c \;\mapsto\; \operatorname{pm}\!\left(\log(1 + Y/c)\right),
\qquad \operatorname{pm}(x) = \frac{\max(x)}{\operatorname{mean}(x)} , $$

the peak-to-mean ratio of the data seen through the link, as a function of the
$c$ used to fit. By the results in
[Why the sparsity of a log1p fit is decided by Y and c](hoyer_sparsity_c.html)
(and the corresponding appendix of the paper), this curve caps the Hoyer
sparsity that *any* rank-$K$ fit can attain at each $c$ — and on both our
simulations and the paper's real-data fits, the cap is essentially attained.
So if a simulation gets this curve right, it reproduces the phenomenon the
sparsity experiments study; if it gets it wrong, no other resemblance helps.

Two facts make the curve a practical calibration target.

- **It is structure-blind.** $\operatorname{pm}$ sees only the multiset of
  count values, so the curve is a functional of the count *histogram* — it can
  be matched without matching any of the real data's latent structure.
- **It must be size-adjusted.** The curve involves $\max(Y)$, and a
  $26{,}000 \times 1.3$M matrix reaches deeper into its tail than any
  simulation-sized matrix can. The honest target for a simulation with $np$
  entries is the curve of $np$ draws from the real marginal, whose median
  maximum $t^{*}$ solves $F(t)^{np} = \tfrac12$.

The four calibrated numbers map onto the curve's features: $\pi_0$ sets the
small-$c$ floor (the zero fraction), $\mu$ the overall level, $\sigma$ the
transition, and $T$ the large-$c$ ceiling.

# The target: MOCA

We calibrate against the MOCA counts (26,183 genes $\times$ 1,331,984 cells)
— the same dataset used in the paper's approximation-accuracy experiments,
so a single real dataset anchors every simulation in the paper. The histogram
was extracted once with `tabulate()`; it is a 1KB file.

```{r target}
z  <- readRDS("data/moca_count_histogram.rds")   # h = counts of value t>=1; np
p_t <- c(z$np - sum(z$h), z$h) / z$np            # P(Y = t), t = 0, 1, ...
tv  <- 0:(length(p_t) - 1)

N_S <- 600L; P_S <- 600L; NP_S <- N_S * P_S      # simulation dimensions
K <- 10L; A_SIG <- 0.3; C_TRUE <- 1

tstar <- tv[min(which(cumsum(p_t)^NP_S >= 0.5))] # median max of NP_S draws
CS <- c(1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3, Inf)  # the fitting grid
target <- sapply(CS, function(cc) {
  g <- function(t) if (is.infinite(cc)) t else log1p(t/cc)
  g(tstar) / sum(p_t * g(tv))
})

knitr::kable(data.frame(
  `real zero fraction` = round(p_t[1], 4),
  `real mean` = round(sum(p_t * tv), 4),
  `real max` = max(tv),
  `size-adjusted max t*` = tstar, check.names = FALSE))
knitr::kable(t(setNames(round(target, 1), paste0("c=", format(CS)))),
             caption = "The size-adjusted target budget curve.")
```

# Calibration

Nelder–Mead over the four knobs. The loss is the squared log-mismatch of the
curves, averaged over three fixed seeds (fixing seeds inside the loss makes it
a deterministic function, which is all Nelder–Mead asks; averaging keeps the
optimizer from exploiting one seed's luck).

```{r calibrate, cache=TRUE}
sim_Y <- function(par, seed, n = N_S, p = P_S) {
  PI0 <- plogis(par[1]); mu <- par[2]; sig <- exp(par[3]); CAP <- exp(par[4])
  set.seed(seed)
  L  <- matrix(rgamma(n*K, A_SIG), n, K); L <- L / rowSums(L)
  Fm <- matrix(pmin(exp(mu + sig * rt(p*K, df = 4)), CAP) *
               (runif(p*K) > PI0), p, K)
  matrix(rpois(n*p, C_TRUE * expm1(tcrossprod(L, Fm))), n, p)
}
curve_of <- function(Y) sapply(CS, function(cc) {
  g <- if (is.infinite(cc)) as.numeric(Y) else log1p(Y/cc)
  max(g) / mean(g) })
loss <- function(par) mean(sapply(1:3, function(s) {
  cv <- curve_of(sim_Y(par, s))
  if (any(!is.finite(cv)) || any(cv <= 0)) return(1e6)
  sum((log(cv) - log(target))^2) }))

opt <- optim(c(qlogis(0.9), -1.3, log(0.6), log(4.3)), loss,
             method = "Nelder-Mead", control = list(maxit = 500, reltol = 1e-6))
stopifnot(opt$convergence == 0)
knobs <- c(pi0 = plogis(opt$par[1]), mu = opt$par[2],
           sigma = exp(opt$par[3]), cap = exp(opt$par[4]))
knitr::kable(t(round(knobs, 3)), caption = "The four calibrated numbers.")
saveRDS(list(par = opt$par, knobs = knobs, target = target, CS = CS,
             tstar = tstar, n = N_S, p = P_S, K = K, a_sig = A_SIG,
             c_true = C_TRUE, df = 4),
        "output/sim_calibration.rds")
```

# Does it hold on seeds the optimizer never saw?

Everything below uses seeds 101–105, disjoint from the three calibration
seeds.

```{r acceptance, cache=TRUE}
fresh <- sapply(101:105, function(s) curve_of(sim_Y(opt$par, s)))
knitr::kable(data.frame(
  c = format(CS), target = round(target, 1),
  `achieved (median of 5)` = round(apply(fresh, 1, median), 1),
  `ratio` = round(apply(fresh, 1, median)/target, 2),
  `ratio range` = paste0(round(apply(fresh/target, 1, min), 2), " – ",
                         round(apply(fresh/target, 1, max), 2)),
  check.names = FALSE), caption = "Budget curve on held-out seeds.")

Y <- sim_Y(opt$par, 101)
knitr::kable(data.frame(
  what = c("zero fraction", "mean count", "max count"),
  simulated = c(round(mean(Y == 0), 3), round(mean(Y), 3), max(Y)),
  real = c(round(p_t[1], 3), round(sum(p_t*tv), 3),
           paste0(tstar, " (size-adjusted)"))),
  caption = "Byproducts, none of which were targeted directly.")
```

The curve lands within about $\pm 25\%$ of the target across six decades of
$c$, and the zero fraction, mean, and maximum come along for free — which is
the point of matching a sufficient statistic rather than a list of summary
numbers.

For contrast, the same curves for the *previous* simulation design (the
300$\times$300, rank-3 grid used in
[the earlier experiments](c_grid_sim.html), which calibrated the zero fraction
to 0.5 and capped counts near 20):

```{r overlay, fig.width=7, fig.height=4}
old <- c(1.55, 1.87, 2.7, 5.6, 13.0, 18.3, 19.3, 19.4)   # pm of old-sim fits
lab <- ifelse(is.infinite(CS), "Inf", format(CS, scientific = TRUE))
d <- rbind(
  data.frame(c = CS, pm = target, what = "target (MOCA, size-adjusted)"),
  data.frame(c = CS, pm = apply(fresh, 1, median), what = "new simulation"),
  data.frame(c = CS, pm = old, what = "old simulation"))
d$x <- log10(ifelse(is.infinite(d$c), 1e4, d$c))
ggplot(d, aes(x, pm, colour = what)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_log10() +
  scale_x_continuous(breaks = log10(c(1e-3, 1e-1, 10, 1e3, 1e4)),
                     labels = c("1e-3", "0.1", "10", "1000", "Inf")) +
  scale_colour_manual(values = c("target (MOCA, size-adjusted)" = "#C2410C",
                                 "new simulation" = "#0E5C6B",
                                 "old simulation" = "grey55"), name = NULL) +
  labs(x = "c used for fitting", y = "pm(log1p(Y/c))",
       title = "The budget curve: target, new design, old design",
       subtitle = "the old design was 40–100x below the real curve everywhere")
```

The old design was not wrong for its original purpose, but its budget curve
sat two orders of magnitude below the real data's: it probed the regime where
the sparsity ceiling collapses toward zero, which real data — with its 97%
zeros and heavy tail — never enters.

```{r tail, fig.width=6.5, fig.height=4}
ccdf_sim <- rev(cumsum(rev(tabulate(Y[Y > 0], nbins = max(Y))))) / length(Y)
ccdf_real <- rev(cumsum(rev(p_t[-1])))
tt <- 1:min(length(ccdf_sim), length(ccdf_real), 200)
d2 <- rbind(data.frame(t = tt, p = ccdf_sim[tt],  what = "simulated"),
            data.frame(t = tt, p = ccdf_real[tt], what = "real (MOCA)"))
d2 <- d2[d2$p > 0, ]
ggplot(d2, aes(t, p, colour = what)) +
  geom_line(linewidth = 1) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c(`real (MOCA)` = "#C2410C",
                                 simulated = "#0E5C6B"), name = NULL) +
  labs(x = "count value t", y = "P(Y >= t)",
       title = "Tail of the count distribution",
       subtitle = "not targeted by the calibration; inherited from the curve")
```

# Calibrating to the pancreas data instead

The dataset at the center of the paper's applications is the pancreas data
(7,606 cells $\times$ 18,195 genes), so it is the natural primary anchor. Its
count histogram was extracted from the raw counts
(`pancreas_cytokine_lsa.Rdata`) with one `tabulate()` call.

This calibration also serves the approximation-accuracy experiments, and
those consume something the budget curve only pins loosely: the *absolute
scale* of the counts (the approximation error depends on where the linear
predictors actually live). So here the mean count is not added to the loss —
it is **profiled out**: inside every loss evaluation, the location $\mu$ is
solved by one-dimensional root-finding so that the expected mean count equals
the real one exactly, and the optimizer searches only the remaining three
knobs $(\pi_0, \sigma, T)$ against the curve. This answers the cleaner
question — the best curve match *among* generators with the correct mean —
rather than blending two errors with an arbitrary weight. (The additive
alternative, a weighted mean term in the loss, was tried and traded away the
tail of the curve; profiling keeps both.)

```{r pancreas-target}
zp   <- readRDS("data/pancreas_count_histogram.rds")
p_tp <- c(zp$np - sum(zp$h), zp$h) / zp$np
tvp  <- 0:(length(p_tp) - 1)
tstar_p <- tvp[min(which(cumsum(p_tp)^NP_S >= 0.5))]
target_p <- sapply(CS, function(cc) {
  g <- function(t) if (is.infinite(cc)) t else log1p(t/cc)
  g(tstar_p) / sum(p_tp * g(tvp))
})
knitr::kable(data.frame(
  dataset = c("pancreas", "MOCA"),
  `zero fraction` = c(round(p_tp[1], 3), round(p_t[1], 3)),
  `mean count` = c(round(sum(p_tp*tvp), 3), round(sum(p_t*tv), 3)),
  `max count` = c(max(tvp), max(tv)),
  `size-adjusted max t*` = c(tstar_p, tstar),
  `budget floor (c = 0.001)` = round(c(target_p[1], target[1]), 1),
  `budget ceiling (c = Inf)` = round(c(target_p[8], target[8]), 0),
  check.names = FALSE),
  caption = "The two anchors. Pancreas has fewer zeros (lower floor) and a far
  heavier tail (higher ceiling): its budget curve spans three decades where
  MOCA's spans one and a half.")
```

The comparison explains, through the theory, why the pancreas panels of the
paper's sparsity figure show the most dramatic trend across $c$: its budget
curve moves the most.

```{r pancreas-calibrate, cache=TRUE}
M_REAL_P <- sum(p_tp * tvp)

## fixed randomness per seed (common random numbers): loadings, t-draws,
## zero indicators. The inner root needs no Poisson draws, so the expected
## mean count given these parts is deterministic and monotone in mu.
make_parts <- function(seeds) lapply(seeds, function(s) {
  set.seed(s)
  L <- matrix(rgamma(N_S*K, A_SIG), N_S, K); L <- L/rowSums(L)
  list(L = L, tmat = matrix(rt(P_S*K, df = 4), P_S, K),
       u = matrix(runif(P_S*K), P_S, K))
})
parts3 <- make_parts(1:3)
mean_rate <- function(mu, sig, CAP, PI0, parts)
  mean(sapply(parts, function(pp) {
    Fm <- pmin(exp(mu + sig*pp$tmat), CAP) * (pp$u > PI0)
    mean(C_TRUE * expm1(tcrossprod(pp$L, Fm))) }))
solve_mu <- function(sig, CAP, PI0, parts)
  uniroot(function(m) mean_rate(m, sig, CAP, PI0, parts) - M_REAL_P,
          c(-9, 3), tol = 1e-4)$root

Y_from <- function(pp, mu, sig, CAP, PI0, pseed) {
  Fm <- pmin(exp(mu + sig*pp$tmat), CAP) * (pp$u > PI0)
  set.seed(pseed)
  matrix(rpois(NP_S, C_TRUE*expm1(tcrossprod(pp$L, Fm))), N_S, P_S)
}
loss_p <- function(par) {
  PI0 <- plogis(par[1]); sig <- exp(par[2]); CAP <- exp(par[3])
  mu <- tryCatch(solve_mu(sig, CAP, PI0, parts3), error = function(e) NA)
  if (is.na(mu)) return(1e6)
  mean(sapply(1:3, function(s) {
    cv <- curve_of(Y_from(parts3[[s]], mu, sig, CAP, PI0, s + 5000))
    if (any(!is.finite(cv)) || any(cv <= 0)) return(1e6)
    sum((log(cv) - log(target_p))^2) }))
}
opt_p <- optim(c(qlogis(0.75), log(0.6), log(10)), loss_p,
               method = "Nelder-Mead", control = list(maxit = 400, reltol = 1e-6))
stopifnot(opt_p$convergence == 0)

## the final mu is re-solved on 20 seeds: the root is cheap (no Poisson
## draws), and three seeds are not enough to center the mean.
PI0_p <- plogis(opt_p$par[1]); sig_p <- exp(opt_p$par[2]); cap_p <- exp(opt_p$par[3])
mu_p  <- solve_mu(sig_p, cap_p, PI0_p, make_parts(1:20))
knobs_p <- c(pi0 = PI0_p, mu = mu_p, sigma = sig_p, cap = cap_p)
knitr::kable(rbind(pancreas = round(knobs_p, 3), MOCA = round(knobs, 3)),
             caption = "Calibrated knobs under the two anchors. For pancreas,
             mu is profiled (solved to match the real mean count exactly),
             not searched.")
saveRDS(list(knobs = knobs_p, target = target_p, CS = CS,
             tstar = tstar_p, mean_target = M_REAL_P, n = N_S, p = P_S,
             K = K, a_sig = A_SIG, c_true = C_TRUE, df = 4,
             method = "curve-matched with profiled mean"),
        "output/sim_calibration_pancreas.rds")
```

```{r pancreas-acceptance, cache=TRUE}
sim_pancreas <- function(seed) {
  set.seed(seed)
  L <- matrix(rgamma(N_S*K, A_SIG), N_S, K); L <- L/rowSums(L)
  Fm <- matrix(pmin(exp(mu_p + sig_p*rt(P_S*K, 4)), cap_p) *
               (runif(P_S*K) > PI0_p), P_S, K)
  matrix(rpois(NP_S, C_TRUE*expm1(tcrossprod(L, Fm))), N_S, P_S)
}
fresh_l <- lapply(101:110, sim_pancreas)
fresh_p <- sapply(fresh_l, curve_of)
knitr::kable(data.frame(
  c = format(CS), target = round(target_p, 1),
  `achieved (median of 10)` = round(apply(fresh_p, 1, median), 1),
  ratio = round(apply(fresh_p, 1, median)/target_p, 2),
  `ratio range` = paste0(round(apply(fresh_p/target_p, 1, min), 2), " – ",
                         round(apply(fresh_p/target_p, 1, max), 2)),
  check.names = FALSE),
  caption = "Pancreas-calibrated budget curve on held-out seeds.")

mns <- sapply(fresh_l, mean)
knitr::kable(data.frame(
  what = c("zero fraction (median)", "mean count (median)",
           "mean count (range)", "max count (median)"),
  simulated = c(round(median(sapply(fresh_l, function(Y) mean(Y == 0))), 3),
                round(median(mns), 3),
                paste0(round(min(mns), 2), " – ", round(max(mns), 2)),
                median(sapply(fresh_l, max))),
  real = c(round(p_tp[1], 3), round(M_REAL_P, 3), "",
           paste0(tstar_p, " (size-adjusted)"))),
  caption = "Byproducts under the pancreas calibration, 10 held-out seeds.")
```

The profiled mean lands centered on the target (the seed-to-seed range is
genuine: with a tail this heavy, the few largest rates carry several percent
of the mean, so the mean of any one dataset wobbles), and the curve keeps the
same quality as the curve-only fit rather than trading its tail away.

The knobs move in interpretable directions relative to MOCA: less zero mass,
a higher location, and a doubled cap, matching a dataset with more signal per
entry and a heavier tail. The interesting difference is $t^{*}$: the pancreas
tail is heavy enough that even a 600$\times$600 subsample is expected to
contain a count near $10^{4}$, so the simulated rates must reach that scale —
which is exactly what the raised cap provides. Since the pancreas data is the
paper's main application — and the mean is now pinned, which is what the
approximation experiments need — **this is the calibration for all
simulation experiments**; the MOCA calibration remains as a secondary anchor.

# What one dataset looks like

```{r truth, cache=TRUE}
set.seed(101)
L  <- matrix(rgamma(N_S*K, A_SIG), N_S, K); L <- L / rowSums(L)
hoyer <- function(x) { m <- length(x)
  (sqrt(m) - sum(x)/sqrt(sum(x^2)))/(sqrt(m) - 1) }
knitr::kable(data.frame(
  `mean Hoyer sparsity of true loading columns` = round(mean(apply(L, 2, hoyer)), 3),
  `mean # factors carrying 90% of a sample` = round(mean(apply(L, 1, function(r)
     min(which(cumsum(sort(r, decreasing = TRUE)) >= 0.9)))), 1),
  check.names = FALSE),
  caption = "The controlled side: loading structure, identical for every c and seed.")
```

# How this feeds the paper's experiments

- **Sparsity and identifiability across $c$** — the grid experiments rerun
  with this generator under the **pancreas** calibration
  (`output/sim_calibration_pancreas.rds`), since the pancreas data is the
  paper's main application. If several $c_{\textrm{true}}$ values are kept,
  the calibration is rerun per $c_{\textrm{true}}$ with the *same* target
  curve, which is the principled version of "the datasets are comparable
  across $c_{\textrm{true}}$".
- **Approximation accuracy** — the approximation's error depends on where the
  counts and linear predictors actually live, which is why the pancreas
  calibration profiles the mean to match exactly rather than only matching
  the (scale-loose) curve. The same
  `output/sim_calibration_pancreas.rds` therefore serves here too, giving one
  generator for every simulation in the paper. MOCA
  (`output/sim_calibration.rds`) remains as a secondary anchor, and
  recalibrating for a different $n$, $p$ takes seconds ($t^*$ depends on
  $np$).

# What is matched, and what is not

Matched, and verified on held-out seeds: the budget curve across six decades
of $c$; the zero fraction; the mean; the maximum; the count tail.

Not matched, by design: the real data's shared gene-abundance profile (our
factor columns are exchangeable, real genes are not), size-factor
heterogeneity across cells ($s_i = 1$ here), and gene-level mean structure.
Consequently the factor-*correlation* phenomenology of the real fits — which
is driven by the shared abundance profile — will not appear in this
simulation, and no claim about it should rest on these data. If that ever
becomes the target, the right tool is a parametric bootstrap of a real fit,
not a redesign of this generator.
