---
title: "The calibrated simulation: a paper-ready description"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

This page describes **exactly** the simulation that the exact-vs-approx
experiment will run. To make drift impossible, every constant below is read
from the experiment's own configuration file
(`inst/paper_figures/experiments/approx_sim_common.R` in the log1pNMF repo),
which is what the cluster workers source. An earlier version of this page
developed the method with a MOCA anchor; that material is in the git history.

```{r source-config}
library(ggplot2); library(cowplot)
theme_set(theme_cowplot(font_size = 11))

exp_dir <- path.expand("~/Documents/log1pNMF/inst/paper_figures/experiments")
source(file.path(exp_dir, "approx_sim_common.R"))
cal <- readRDS(file.path(exp_dir, "approx_sim_calibration.rds"))
stopifnot(all.equal(unname(cal$knobs),
                    c(PI0, MU, SIGMA, CAP), tolerance = 1e-4))

knitr::kable(data.frame(
  parameter = c("n (samples)", "p (features)", "K (rank, true = fitted)",
                "Dirichlet concentration a", "zero probability pi0",
                "location mu", "scale sigma", "cap T", "floor T_lo",
                "t degrees of freedom", "c used to generate"),
  value = c(N_ROWS, N_COLS, K_TRUE, A_SIG, round(PI0, 4), round(MU, 4),
            round(SIGMA, 4), round(CAP, 4), FLOOR, T_DF, C_TRUE),
  role = c("fixed by design", "fixed by design", "fixed by design",
           "fixed by design (controls loading sparsity)",
           "calibrated", "calibrated (profiled)", "calibrated", "calibrated",
           "fixed by design (see below)",
           "fixed by design", "fixed by design")),
  caption = "Every number the generator uses, read from approx_sim_common.R.")
```

# The description, as the paper would give it

> Counts were simulated from the log1p NMF model itself, with parameters
> calibrated so that the simulated data reproduce the count characteristics
> of the pancreas dataset. For each of ten seeds we drew an
> $n \times K$ loading matrix $L$ with rows
> $\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(0.3, \dots, 0.3)$, so that
> every sample concentrates on a small number of factors, and a
> $p \times K$ factor matrix $F$ with independent entries that are zero with
> probability $\pi_0$ and otherwise
> $\exp(\mu + \sigma t)$, $t \sim t_4$, truncated to $[T_{\textrm{lo}}, T]$ —
> a point mass at zero and a heavy-tailed body, truncated on both sides.
> Each feature was conditioned to load on at least one factor, and the lower
> truncation guarantees it does so at a detectable level; together these are
> the analogue of filtering genes for detectable expression. Counts were then drawn as
> $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$ with
> $\lambda_{ij} = e^{\theta_{ij}} - 1$ and $\Theta = LF^{\top}$, i.e. from
> the model at $c = 1$. The four constants
> $(\pi_0, \mu, \sigma, T)$ were chosen by matching two properties of the
> real data: the mean count, matched exactly, and the curve
> $c \mapsto \operatorname{pm}(\log(1 + Y/c))$ — the peak-to-mean ratio of
> the data on the link scale, which by [the sparsity results] governs the
> factor sparsity attainable at each $c$ — matched in least squares across
> $c \in [10^{-4}, 10^{4}]$, after adjusting the real curve to the
> simulation's size. The resulting datasets are 88–89% zeros with a mean
> count of about 0.6 and maximum counts in the low thousands, against 82%,
> 0.67, and (size-adjusted) $\sim$8,800 for the real data.

Everything in that paragraph is verified below on the ten seeds the
experiment will actually use.

# The generative model, precisely

For seed $s \in \{1, \dots, `r N_SEEDS`\}$:

1. $\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(a, \dots, a)$ independently,
   $a = `r A_SIG`$, giving $L \in \mathbb{R}_{\geq 0}^{`r N_ROWS` \times `r K_TRUE`}$.
2. $u_{jk} \sim \textrm{Unif}(0,1)$ and $t_{jk} \sim t_{`r T_DF`}$
   independently, and
   $$f_{jk} = \textrm{clamp}\left(\exp(\mu + \sigma t_{jk});\; T_{\textrm{lo}},\, T\right) \cdot
   \mathbf{1}\{u_{jk} > \pi_0\},$$
   except that a feature $j$ with $u_{jk} \leq \pi_0$ for every $k$ keeps its
   single largest-$u_{jk}$ factor (so no feature is dead; deterministic given
   $u$).
3. $\Theta = LF^{\top}$, $\;\lambda_{ij} = c_{\textrm{true}}(e^{\theta_{ij}} - 1)$
   with $c_{\textrm{true}} = `r C_TRUE`$, $\;y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$.
   In the rare event a row or column of $Y$ is entirely zero, the Poisson
   draw (only) is repeated with the next sub-seed; the fitting code rejects
   all-zero columns, as do real pipelines.

There are no size factors ($s_i = 1$; everything is fit with `s = FALSE`).
The two truncation points have distinct jobs. The cap bounds the rates at
$e^{T} - 1 \approx `r round(expm1(CAP))`$, which is what makes the maximum
count land in the same place on every seed — unbounded heavy tails made
$\max(Y)$ swing by an order of magnitude across seeds when we tried them.
The floor $T_{\textrm{lo}} = `r FLOOR`$ exists because a feature active in a
single factor can otherwise draw a magnitude so small that its column of $Y$
is all zero with probability near one, which the fitting code rejects — the
seed-7 dataset did exactly this before the floor was added. At the floor, a
single-factor feature's column has expected total count
$\approx T_{\textrm{lo}} \cdot n/K = `r round(FLOOR * N_ROWS / K_TRUE, 1)`$.
The floor moves about 4% of the nonzero entries; the knobs above were
recalibrated with it in place.

# How the four calibrated numbers were chosen

The target is the real pancreas count matrix (7,606 cells $\times$ 18,195
genes; 82.2% zeros, mean 0.674). Two properties are matched:

- **The budget curve** $c \mapsto \operatorname{pm}(\log(1+Y/c))$, in least
  squares on the log scale over the experiment's own $c$ grid. This curve
  depends only on the count histogram, and it is the quantity our sparsity
  theory identifies as governing what any fit can do at each $c$. Because it
  involves a maximum it must be **size-adjusted**: the target uses
  $t^{*} = `r cal$tstar`$, the median maximum of $np = `r N_ROWS * N_COLS`$
  draws from the real marginal, not the real matrix's own maximum
  (`r max(0:(length(readRDS("data/pancreas_count_histogram.rds")$h)))`).
- **The mean count, exactly.** The approximation experiments consume the
  absolute scale of the counts, and the curve alone pins it only loosely. So
  $\mu$ is not searched: inside every calibration step it is solved by
  one-dimensional root-finding so the expected mean count equals the real
  0.674 (profiling). Nelder–Mead searches only $(\pi_0, \sigma, T)$ against
  the curve, with common random numbers across evaluations; the final $\mu$
  is re-solved on 20 seeds.

The calibration code is `calibrate_approx_sim.R`, next to the experiment
scripts; it writes `approx_sim_calibration.rds`, from which the knobs in the
table above were checked.

# Verification on the ten seeds the experiment will use

Not held-out seeds — these are datasets 1–10, exactly as the workers will
regenerate them.

```{r verify, cache=TRUE}
ds <- lapply(seq_len(N_SEEDS), sim_dataset)
curve_of <- function(Y) sapply(cal$CS, function(cc) {
  g <- log1p(Y / cc); max(g) / mean(g) })
cvs <- sapply(ds, function(d) curve_of(d$Y))

knitr::kable(data.frame(
  c = format(cal$CS), `target pm` = round(cal$target, 1),
  `simulated pm (median of 10)` = round(apply(cvs, 1, median), 1),
  ratio = round(apply(cvs, 1, median) / cal$target, 2),
  `ratio range` = paste0(round(apply(cvs / cal$target, 1, min), 2), " – ",
                         round(apply(cvs / cal$target, 1, max), 2)),
  check.names = FALSE),
  caption = "Budget curve of the ten experiment datasets vs the size-adjusted
  pancreas target.")

zf <- sapply(ds, function(d) mean(d$Y == 0))
mn <- sapply(ds, function(d) mean(d$Y))
mx <- sapply(ds, function(d) max(d$Y))
knitr::kable(data.frame(
  what = c("zero fraction", "mean count", "max count"),
  `simulated, median (range)` = c(
    sprintf("%.3f  (%.3f – %.3f)", median(zf), min(zf), max(zf)),
    sprintf("%.3f  (%.3f – %.3f)", median(mn), min(mn), max(mn)),
    sprintf("%.0f  (%d – %d)", median(mx), min(mx), max(mx))),
  real = c("0.822", "0.674", paste0(cal$tstar, " (size-adjusted)")),
  check.names = FALSE),
  caption = "Count characteristics of the ten datasets.")
```

```{r tail-plot, fig.width=6.5, fig.height=4}
zp   <- readRDS("data/pancreas_count_histogram.rds")
p_tp <- c(zp$np - sum(zp$h), zp$h) / zp$np
ccdf_real <- rev(cumsum(rev(p_tp[-1])))
Yall <- unlist(lapply(ds, function(d) d$Y[d$Y > 0]))
ccdf_sim <- rev(cumsum(rev(tabulate(Yall, nbins = max(Yall))))) / (N_SEEDS * N_ROWS * N_COLS)
tt <- 1:min(length(ccdf_real), length(ccdf_sim), 3000)
d2 <- rbind(data.frame(t = tt, p = ccdf_sim[tt],  what = "simulated (10 datasets pooled)"),
            data.frame(t = tt, p = ccdf_real[tt], what = "real (pancreas)"))
d2 <- d2[d2$p > 0, ]
ggplot(d2, aes(t, p, colour = what)) +
  geom_line(linewidth = 1) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c(`real (pancreas)` = "#C2410C",
                                 `simulated (10 datasets pooled)` = "#0E5C6B"),
                      name = NULL) +
  labs(x = "count value t", y = "P(Y >= t)",
       title = "Tail of the count distribution",
       subtitle = "not targeted directly; inherited from the curve match")
```

The controlled side of the design — the truth the fits will be scored
against:

```{r truth-side}
hoyer <- function(x) { m <- length(x)
  (sqrt(m) - sum(x)/sqrt(sum(x^2)))/(sqrt(m) - 1) }
knitr::kable(data.frame(
  what = c("mean Hoyer sparsity of loading columns",
           "mean # factors carrying 90% of a sample",
           "fraction of exact zeros in F",
           "mean # factors per feature",
           "mean Hoyer sparsity of factor columns"),
  `median over the 10 seeds` = c(
    round(median(sapply(ds, function(d) mean(apply(d$L, 2, hoyer)))), 3),
    round(median(sapply(ds, function(d) mean(apply(d$L, 1, function(r)
      min(which(cumsum(sort(r, decreasing = TRUE)) >= 0.9)))))), 2),
    round(median(sapply(ds, function(d) mean(d$FF == 0))), 3),
    round(median(sapply(ds, function(d) mean(rowSums(d$FF > 0)))), 2),
    round(median(sapply(ds, function(d) mean(apply(d$FF, 2, hoyer)))), 3)),
  check.names = FALSE),
  caption = "Truth structure: identical in distribution across all seeds and
  all c; the loading sparsity is set once by a = 0.3.")
```

# Honest limits of the match at K = 5

The same calibration machinery matched the pancreas curve to within
$\pm 16\%$ at $600 \times 600$, $K = 10$ (git history of this page). At
$K = 5$ it cannot do as well, and the reason is structural: with five
factors and Dirichlet-peaked rows, each entry's rate is nearly a
single-factor draw, and the knobs cannot feed the zero fraction, the mean,
and the extreme tail simultaneously. Two richer variants — a free tail
degree of freedom, and two-component magnitudes — fixed one end of the curve
only by sacrificing the other, and were rejected. The realized compromise,
visible in the tables above: zeros ~0.89 vs 0.822 (so the budget curve runs
~1.4–1.5$\times$ high at the small-$c$ end), and a maximum around 2,000 vs
the size-adjusted 8,820 (so the curve runs ~0.3–0.8$\times$ at the large-$c$
end). The mean is centered by construction. If the tail mismatch ever
matters, the honest fix is $K = 10$, not more knobs.

# The experiment this feeds

Read from the same configuration file:

```{r protocol}
knitr::kable(data.frame(
  setting = c("datasets", "c grid for fitting", "methods",
              "initialization", "convergence tolerance",
              "iteration safety cap", "optimization-time cap",
              "SLURM wall clock", "threads per fit", "tasks"),
  value = c(sprintf("%d seeds, generated once each at c = %s", N_SEEDS, C_TRUE),
            paste(format(CC_GRID, scientific = TRUE, trim = TRUE), collapse = ", "),
            paste(METHODS, collapse = ", "),
            "rank-1 (each method's own warm-up, seeded)",
            format(TOL, scientific = TRUE),
            format(MAXITER, big.mark = ","),
            "none",
            "24h (backstop only)",
            THREADS,
            sprintf("%d  (one fit per task)", N_TASKS))),
  caption = "The exact-vs-approx protocol, from approx_sim_common.R and
  run_approx_sim.sbatch.")
```

Every fit runs until the change in **its own objective** falls below the
tolerance — the approximate methods converge on the approximate objective —
and the comparison number recorded for every fit is the **exact** Poisson
log-likelihood of its fitted rates. Each task writes one .rds; after the
round, `check_approx_sim.R` lists anything missing or unconverged for
resubmission.
