---
title: "A calibrated simulation: realistic counts with controlled factors"
output:
  workflowr::wflow_html:
    code_folding: hide
    toc: true
    toc_float: true
date: "2026-08-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dpi = 110)
```

```{r pkgs}
library(ggplot2); library(cowplot)
theme_set(theme_cowplot(font_size = 11))
```

# What this simulation is for

One generator, used everywhere a simulation appears: the sparsity and
identifiability experiments across values of $c$, and the assessment of the
approximation accuracy. The design has three requirements.

1. **The true loadings have fixed, controlled sparsity.** Whatever else varies,
   the rows of $L$ are drawn the same way, so differences in *fitted* sparsity
   across $c$ can never be blamed on the truth changing.
2. **"Realistic" is a checked property, not an adjective.** The simulated
   counts are calibrated to a real dataset against a specific criterion,
   stated below, and the match is verified on held-out seeds.
3. **The whole recipe fits in five lines**, so it can be described completely
   in a paper paragraph.

# The generative recipe

Fix the dimensions $n$, $p$, the rank $K$, a link constant $c_{\textrm{true}}$,
and the Dirichlet concentration $a = 0.3$. The calibration below supplies four
numbers: a zero probability $\pi_0$, a location $\mu$, a scale $\sigma$, and a
cap $T$. Then:

1. **Loadings.** Each row of $L$ is drawn
   $\;\mathbf{l}_{i\cdot} \sim \textrm{Dirichlet}(a, \dots, a)$.
   With $a = 0.3$ each sample places most of its weight on a few factors —
   membership is soft but concentrated, and its sparsity is set once by $a$.
2. **Factors.** Each entry of $F$ is independently
   $$f_{jk} = \begin{cases} 0 & \textrm{with probability } \pi_0 \\
   \min\{\exp(\mu + \sigma t_{jk}),\; T\} & \textrm{otherwise}, \end{cases}
   \qquad t_{jk} \sim t_4 .$$
   A point mass at zero, a heavy-tailed body, and a hard ceiling.
3. **Linear predictor.** $\Theta = L F^{\top}$.
4. **Rates.** $\lambda_{ij} = c_{\textrm{true}}\left(e^{\theta_{ij}} - 1\right)$
   — the inverse of the log1p link.
5. **Counts.** $y_{ij} \sim \textrm{Poisson}(\lambda_{ij})$.

Each ingredient of step 2 has one job. The point mass makes most rates exactly
zero, which is where the zeros of real count matrices come from. The log-$t$
body gives the counts a heavy tail, which lognormal magnitudes cannot produce
(we tried: the maximum count comes out several times too small). The cap
exists because the *maximum* of unbounded heavy-tailed draws varies wildly from
seed to seed — with a cap, the extreme counts land in the same place every
time, as they should for a matrix of fixed size.

# What "realistic" means, exactly

The property of the data that our theory identifies as consequential is the
**budget curve**

$$ c \;\mapsto\; \operatorname{pm}\!\left(\log(1 + Y/c)\right),
\qquad \operatorname{pm}(x) = \frac{\max(x)}{\operatorname{mean}(x)} , $$

the peak-to-mean ratio of the data seen through the link, as a function of the
$c$ used to fit. By the results in
[Why the sparsity of a log1p fit is decided by Y and c](hoyer_sparsity_c.html)
(and the corresponding appendix of the paper), this curve caps the Hoyer
sparsity that *any* rank-$K$ fit can attain at each $c$ — and on both our
simulations and the paper's real-data fits, the cap is essentially attained.
So if a simulation gets this curve right, it reproduces the phenomenon the
sparsity experiments study; if it gets it wrong, no other resemblance helps.

Two facts make the curve a practical calibration target.

- **It is structure-blind.** $\operatorname{pm}$ sees only the multiset of
  count values, so the curve is a functional of the count *histogram* — it can
  be matched without matching any of the real data's latent structure.
- **It must be size-adjusted.** The curve involves $\max(Y)$, and a
  $26{,}000 \times 1.3$M matrix reaches deeper into its tail than any
  simulation-sized matrix can. The honest target for a simulation with $np$
  entries is the curve of $np$ draws from the real marginal, whose median
  maximum $t^{*}$ solves $F(t)^{np} = \tfrac12$.

The four calibrated numbers map onto the curve's features: $\pi_0$ sets the
small-$c$ floor (the zero fraction), $\mu$ the overall level, $\sigma$ the
transition, and $T$ the large-$c$ ceiling.

# The target: MOCA

We calibrate against the MOCA counts (26,183 genes $\times$ 1,331,984 cells)
— the same dataset used in the paper's approximation-accuracy experiments,
so a single real dataset anchors every simulation in the paper. The histogram
was extracted once with `tabulate()`; it is a 1KB file.

```{r target}
z  <- readRDS("data/moca_count_histogram.rds")   # h = counts of value t>=1; np
p_t <- c(z$np - sum(z$h), z$h) / z$np            # P(Y = t), t = 0, 1, ...
tv  <- 0:(length(p_t) - 1)

N_S <- 600L; P_S <- 600L; NP_S <- N_S * P_S      # simulation dimensions
K <- 10L; A_SIG <- 0.3; C_TRUE <- 1

tstar <- tv[min(which(cumsum(p_t)^NP_S >= 0.5))] # median max of NP_S draws
CS <- c(1e-3, 1e-2, 1e-1, 1, 10, 100, 1e3, Inf)  # the fitting grid
target <- sapply(CS, function(cc) {
  g <- function(t) if (is.infinite(cc)) t else log1p(t/cc)
  g(tstar) / sum(p_t * g(tv))
})

knitr::kable(data.frame(
  `real zero fraction` = round(p_t[1], 4),
  `real mean` = round(sum(p_t * tv), 4),
  `real max` = max(tv),
  `size-adjusted max t*` = tstar, check.names = FALSE))
knitr::kable(t(setNames(round(target, 1), paste0("c=", format(CS)))),
             caption = "The size-adjusted target budget curve.")
```

# Calibration

Nelder–Mead over the four knobs. The loss is the squared log-mismatch of the
curves, averaged over three fixed seeds (fixing seeds inside the loss makes it
a deterministic function, which is all Nelder–Mead asks; averaging keeps the
optimizer from exploiting one seed's luck).

```{r calibrate, cache=TRUE}
sim_Y <- function(par, seed, n = N_S, p = P_S) {
  PI0 <- plogis(par[1]); mu <- par[2]; sig <- exp(par[3]); CAP <- exp(par[4])
  set.seed(seed)
  L  <- matrix(rgamma(n*K, A_SIG), n, K); L <- L / rowSums(L)
  Fm <- matrix(pmin(exp(mu + sig * rt(p*K, df = 4)), CAP) *
               (runif(p*K) > PI0), p, K)
  matrix(rpois(n*p, C_TRUE * expm1(tcrossprod(L, Fm))), n, p)
}
curve_of <- function(Y) sapply(CS, function(cc) {
  g <- if (is.infinite(cc)) as.numeric(Y) else log1p(Y/cc)
  max(g) / mean(g) })
loss <- function(par) mean(sapply(1:3, function(s) {
  cv <- curve_of(sim_Y(par, s))
  if (any(!is.finite(cv)) || any(cv <= 0)) return(1e6)
  sum((log(cv) - log(target))^2) }))

opt <- optim(c(qlogis(0.9), -1.3, log(0.6), log(4.3)), loss,
             method = "Nelder-Mead", control = list(maxit = 500, reltol = 1e-6))
stopifnot(opt$convergence == 0)
knobs <- c(pi0 = plogis(opt$par[1]), mu = opt$par[2],
           sigma = exp(opt$par[3]), cap = exp(opt$par[4]))
knitr::kable(t(round(knobs, 3)), caption = "The four calibrated numbers.")
saveRDS(list(par = opt$par, knobs = knobs, target = target, CS = CS,
             tstar = tstar, n = N_S, p = P_S, K = K, a_sig = A_SIG,
             c_true = C_TRUE, df = 4),
        "output/sim_calibration.rds")
```

# Does it hold on seeds the optimizer never saw?

Everything below uses seeds 101–105, disjoint from the three calibration
seeds.

```{r acceptance, cache=TRUE}
fresh <- sapply(101:105, function(s) curve_of(sim_Y(opt$par, s)))
knitr::kable(data.frame(
  c = format(CS), target = round(target, 1),
  `achieved (median of 5)` = round(apply(fresh, 1, median), 1),
  `ratio` = round(apply(fresh, 1, median)/target, 2),
  `ratio range` = paste0(round(apply(fresh/target, 1, min), 2), " – ",
                         round(apply(fresh/target, 1, max), 2)),
  check.names = FALSE), caption = "Budget curve on held-out seeds.")

Y <- sim_Y(opt$par, 101)
knitr::kable(data.frame(
  what = c("zero fraction", "mean count", "max count"),
  simulated = c(round(mean(Y == 0), 3), round(mean(Y), 3), max(Y)),
  real = c(round(p_t[1], 3), round(sum(p_t*tv), 3),
           paste0(tstar, " (size-adjusted)"))),
  caption = "Byproducts, none of which were targeted directly.")
```

The curve lands within about $\pm 25\%$ of the target across six decades of
$c$, and the zero fraction, mean, and maximum come along for free — which is
the point of matching a sufficient statistic rather than a list of summary
numbers.

For contrast, the same curves for the *previous* simulation design (the
300$\times$300, rank-3 grid used in
[the earlier experiments](c_grid_sim.html), which calibrated the zero fraction
to 0.5 and capped counts near 20):

```{r overlay, fig.width=7, fig.height=4}
old <- c(1.55, 1.87, 2.7, 5.6, 13.0, 18.3, 19.3, 19.4)   # pm of old-sim fits
lab <- ifelse(is.infinite(CS), "Inf", format(CS, scientific = TRUE))
d <- rbind(
  data.frame(c = CS, pm = target, what = "target (MOCA, size-adjusted)"),
  data.frame(c = CS, pm = apply(fresh, 1, median), what = "new simulation"),
  data.frame(c = CS, pm = old, what = "old simulation"))
d$x <- log10(ifelse(is.infinite(d$c), 1e4, d$c))
ggplot(d, aes(x, pm, colour = what)) +
  geom_line(linewidth = 1) + geom_point(size = 2) +
  scale_y_log10() +
  scale_x_continuous(breaks = log10(c(1e-3, 1e-1, 10, 1e3, 1e4)),
                     labels = c("1e-3", "0.1", "10", "1000", "Inf")) +
  scale_colour_manual(values = c("target (MOCA, size-adjusted)" = "#C2410C",
                                 "new simulation" = "#0E5C6B",
                                 "old simulation" = "grey55"), name = NULL) +
  labs(x = "c used for fitting", y = "pm(log1p(Y/c))",
       title = "The budget curve: target, new design, old design",
       subtitle = "the old design was 40–100x below the real curve everywhere")
```

The old design was not wrong for its original purpose, but its budget curve
sat two orders of magnitude below the real data's: it probed the regime where
the sparsity ceiling collapses toward zero, which real data — with its 97%
zeros and heavy tail — never enters.

```{r tail, fig.width=6.5, fig.height=4}
ccdf_sim <- rev(cumsum(rev(tabulate(Y[Y > 0], nbins = max(Y))))) / length(Y)
ccdf_real <- rev(cumsum(rev(p_t[-1])))
tt <- 1:min(length(ccdf_sim), length(ccdf_real), 200)
d2 <- rbind(data.frame(t = tt, p = ccdf_sim[tt],  what = "simulated"),
            data.frame(t = tt, p = ccdf_real[tt], what = "real (MOCA)"))
d2 <- d2[d2$p > 0, ]
ggplot(d2, aes(t, p, colour = what)) +
  geom_line(linewidth = 1) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = c(`real (MOCA)` = "#C2410C",
                                 simulated = "#0E5C6B"), name = NULL) +
  labs(x = "count value t", y = "P(Y >= t)",
       title = "Tail of the count distribution",
       subtitle = "not targeted by the calibration; inherited from the curve")
```

# Calibrating to the pancreas data instead

MOCA anchors the approximation-accuracy experiments, but the dataset at the
center of the paper's applications is the pancreas data (7,606 cells $\times$
18,195 genes). The same machinery calibrates to it — only the target changes.
Its count histogram was extracted from the raw counts
(`pancreas_cytokine_lsa.Rdata`) with one `tabulate()` call.

```{r pancreas-target}
zp   <- readRDS("data/pancreas_count_histogram.rds")
p_tp <- c(zp$np - sum(zp$h), zp$h) / zp$np
tvp  <- 0:(length(p_tp) - 1)
tstar_p <- tvp[min(which(cumsum(p_tp)^NP_S >= 0.5))]
target_p <- sapply(CS, function(cc) {
  g <- function(t) if (is.infinite(cc)) t else log1p(t/cc)
  g(tstar_p) / sum(p_tp * g(tvp))
})
knitr::kable(data.frame(
  dataset = c("pancreas", "MOCA"),
  `zero fraction` = c(round(p_tp[1], 3), round(p_t[1], 3)),
  `mean count` = c(round(sum(p_tp*tvp), 3), round(sum(p_t*tv), 3)),
  `max count` = c(max(tvp), max(tv)),
  `size-adjusted max t*` = c(tstar_p, tstar),
  `budget floor (c = 0.001)` = round(c(target_p[1], target[1]), 1),
  `budget ceiling (c = Inf)` = round(c(target_p[8], target[8]), 0),
  check.names = FALSE),
  caption = "The two anchors. Pancreas has fewer zeros (lower floor) and a far
  heavier tail (higher ceiling): its budget curve spans three decades where
  MOCA's spans one and a half.")
```

The comparison explains, through the theory, why the pancreas panels of the
paper's sparsity figure show the most dramatic trend across $c$: its budget
curve moves the most.

```{r pancreas-calibrate, cache=TRUE}
loss_p <- function(par) mean(sapply(1:3, function(s) {
  cv <- curve_of(sim_Y(par, s))
  if (any(!is.finite(cv)) || any(cv <= 0)) return(1e6)
  sum((log(cv) - log(target_p))^2) }))
opt_p <- optim(c(qlogis(0.8), -0.7, log(0.8), log(7)), loss_p,
               method = "Nelder-Mead", control = list(maxit = 600, reltol = 1e-6))
stopifnot(opt_p$convergence == 0)
knobs_p <- c(pi0 = plogis(opt_p$par[1]), mu = opt_p$par[2],
             sigma = exp(opt_p$par[3]), cap = exp(opt_p$par[4]))
knitr::kable(rbind(pancreas = round(knobs_p, 3), MOCA = round(knobs, 3)),
             caption = "Calibrated knobs under the two anchors.")
saveRDS(list(par = opt_p$par, knobs = knobs_p, target = target_p, CS = CS,
             tstar = tstar_p, n = N_S, p = P_S, K = K, a_sig = A_SIG,
             c_true = C_TRUE, df = 4),
        "output/sim_calibration_pancreas.rds")
```

```{r pancreas-acceptance, cache=TRUE}
fresh_p <- sapply(101:105, function(s) curve_of(sim_Y(opt_p$par, s)))
knitr::kable(data.frame(
  c = format(CS), target = round(target_p, 1),
  `achieved (median of 5)` = round(apply(fresh_p, 1, median), 1),
  ratio = round(apply(fresh_p, 1, median)/target_p, 2),
  `ratio range` = paste0(round(apply(fresh_p/target_p, 1, min), 2), " – ",
                         round(apply(fresh_p/target_p, 1, max), 2)),
  check.names = FALSE),
  caption = "Pancreas-calibrated budget curve on held-out seeds.")

Yp <- sim_Y(opt_p$par, 101)
knitr::kable(data.frame(
  what = c("zero fraction", "mean count", "max count"),
  simulated = c(round(mean(Yp == 0), 3), round(mean(Yp), 3), max(Yp)),
  real = c(round(p_tp[1], 3), round(sum(p_tp*tvp), 3),
           paste0(tstar_p, " (size-adjusted)"))),
  caption = "Byproducts under the pancreas calibration.")
```

The knobs move in interpretable directions: less zero mass ($\pi_0$ 0.75
versus 0.90), a higher location, and a higher cap, matching a dataset with
more signal per entry and a heavier tail. The interesting difference from
MOCA is $t^{*}$: the pancreas tail is heavy enough that even a
600$\times$600 subsample is expected to contain a count near $10^{4}$, so the
simulated rates must reach that scale — which is exactly what the raised cap
provides. Since the pancreas data is the paper's main application, **this is
the calibration the simulation experiments should use**; the MOCA calibration
remains available for anchoring the approximation-accuracy work.

# What one dataset looks like

```{r truth, cache=TRUE}
set.seed(101)
L  <- matrix(rgamma(N_S*K, A_SIG), N_S, K); L <- L / rowSums(L)
hoyer <- function(x) { m <- length(x)
  (sqrt(m) - sum(x)/sqrt(sum(x^2)))/(sqrt(m) - 1) }
knitr::kable(data.frame(
  `mean Hoyer sparsity of true loading columns` = round(mean(apply(L, 2, hoyer)), 3),
  `mean # factors carrying 90% of a sample` = round(mean(apply(L, 1, function(r)
     min(which(cumsum(sort(r, decreasing = TRUE)) >= 0.9)))), 1),
  check.names = FALSE),
  caption = "The controlled side: loading structure, identical for every c and seed.")
```

# How this feeds the paper's experiments

- **Sparsity and identifiability across $c$** — the grid experiments rerun
  with this generator under the **pancreas** calibration
  (`output/sim_calibration_pancreas.rds`), since the pancreas data is the
  paper's main application. If several $c_{\textrm{true}}$ values are kept,
  the calibration is rerun per $c_{\textrm{true}}$ with the *same* target
  curve, which is the principled version of "the datasets are comparable
  across $c_{\textrm{true}}$".
- **Approximation accuracy** — the approximation's error depends on where the
  counts and linear predictors actually live, so the realism criterion that
  matters is exactly the one calibrated here. Either anchor works; MOCA
  (`output/sim_calibration.rds`) is the natural one there because it is the
  dataset used in the paper's timing and accuracy experiments. Recalibrating
  for a different $n$, $p$ takes seconds ($t^*$ depends on $np$).

# What is matched, and what is not

Matched, and verified on held-out seeds: the budget curve across six decades
of $c$; the zero fraction; the mean; the maximum; the count tail.

Not matched, by design: the real data's shared gene-abundance profile (our
factor columns are exchangeable, real genes are not), size-factor
heterogeneity across cells ($s_i = 1$ here), and gene-level mean structure.
Consequently the factor-*correlation* phenomenology of the real fits — which
is driven by the shared abundance profile — will not appear in this
simulation, and no claim about it should rest on these data. If that ever
becomes the target, the right tool is a parametric bootstrap of a real fit,
not a redesign of this generator.
